### 2019-07-12 applications/pass_through/doc/pass_through.tex

\documentclass[preprint, 12pt, authoryear]{elsarticle}
\pdfoutput=1
\usepackage{etoolbox}
%\renewcommand*{\today}{August 29, 2018}
\patchcmd{\MaketitleBox}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}%
\patchcmd{\pprintMaketitle}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}%

\usepackage{eurosym} 
\usepackage{gensymb}
\usepackage{makecell}
\usepackage{subcaption}
\usepackage{url}
\usepackage{breakurl}
\def\UrlBreaks{\do\/\do-}

\usepackage[version=4]{mhchem} 
\usepackage{amsmath} 
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}

\usepackage{multibib}
\newcites{sec}{Bibliography}

%\bibliographystyle{elsarticle-harv}

\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{}%
 \let\@evenfoot\@oddfoot}
\makeatother

\begin{document}
\begin{frontmatter}
\title{District heating systems under high \ce{CO2} emission prices: the role of the pass-through from emission cost to electricity prices}

\author[1,2]{Sebastian Wehrle\corref{cor1}}
\ead{sebastian.wehrle@boku.ac.at}

\author[1]{Johannes Schmidt}
\ead{johannes.schmidt@boku.ac.at}

\author[1]{Olga Turkovska}
\ead{olga.turkovska@boku.ac.at}

\cortext[cor1]{Corresponding author}

\address[1]{Institute for Sustainable Economic Development, University of Natural Resources and Life Sciences, Feistmantelstrasse 4, 1180 Vienna, Austria}

\address[2]{Wiener Netze GmbH, Erdbergstrasse 236, 1110 Vienna, Austria}

\begin{abstract}
In spite of recent increases, carbon prices in the European emission trading system are still below levels considered to be consistent with the goals of the Paris Agreement \citep[e.g.][]{Parry2018, OECD2018}.
The cost of \ce{CO2} emissions affects, inter alia, integrated district heating systems that use a variety of emission intense heat generation technologies. 
We study the effect of high carbon prices on district heating systems, operating within the current and a prospective future German electricity system.
%In addition, we examine how district heating system fare under a prospective further expansion of renewable power generation and an intended coal phase out.
Therefore, we (i) develop a simplified analytical framework to analyse optimal decisions of a district heating operator, and (ii) investigate the market-wide effects of increasing emission prices, in particular the pass-through from emission costs to electricity prices.
Using MEDEA, a numerical model of the power market in the German bidding area, we estimate a pass-through from \ce{CO2} emission prices to power prices between $0.82$ and $0.64$ as of 2017, depending on the absolute emission price level.
This suffices to improve the profitability of  natural gas-fired district heating systems of at least moderate efficiency under rising carbon prices. 
However, the emission cost pass-through is set to decline markedly in prospective electricity systems, rendering this a short-term opportunity.
\end{abstract}

\begin{keyword}
electricity markets \sep pass-through \sep carbon pricing \sep district heating
\JEL L94 \sep Q41 \sep C61
\end{keyword}
\end{frontmatter}
\newpage

\section{Introduction} \label{intro}
Combined heat and power (CHP) units produced $124$ TWh of electricity in 2017, accounting for $20.0\%$ of total German electricity generation \citep{AGEB2018a}. 
The CHP heat is used in industrial processes or for space heating through heat grids. In Germany, such district heating grids are found in virtually all major urban areas. 
In consequence, Germany is by far the largest consumer of district heat in Europe, with sales of $121.4$ TWh ($8.3\%$ of final heating demand) in 2017 \citep{AGEB2018b}. 
However, not all sold district heat is generated in natural gas or coal-fired CHP plants, as typical district heating systems rely on a broader portfolio of technologies. 
Many systems include waste incineration units and most use boilers to supply heat in times of peak load. 
Recently, district heating system operators also seek to include low-emission heat sources such as geothermal, solar thermal, biomass, or waste heat from commercial and industrial activities.

District heating systems are typically vertically integrated, with municipalities or federal states either holding majority stakes in district heating systems (e.g. in Cologne, Stuttgart, Frankfurt am Main, D\"usseldorf, Dortmund) or fully owning district heating systems in their district (e.g. Munich).\footnote{In Hamburg, the federal state currently holds a minority stake in the district heating system and is set to regain full control in 2019, while Berlin seeks to regain control over the district heating system, which was privatized in 2001. }
Against this backdrop, Germany aims for $120$ TWh of electricity generation from CHP units by 2025 \citep{KWKG2016}, while expanding renewable power generation to $55-60 \%$ of electricity consumption by 2035 and phasing out nuclear power generation by 2022. 
In addition, the government-installed “Kohlekommission” (“coal commission”) recently proposed to end power generation from coal and lignite by 2038 \citep{KommissionWSB2019}. 
Ambitions for tighter pricing of carbon dioxide emissions are complementing the planned changes in the power generation fleet.
With the backloading of EU emission allowances (EUA) and the recent start of the market stability reserve, EUA prices rose from an average of $5.89$ \euro{}/t\ce{CO2} in 2017 to more than $20$~\euro{}/t\ce{CO2} in early 2019 \citep{EEX2019}.
Several national and international institutions consider carbon prices above this levels necessary to limit carbon dioxide emissions to levels consistent with the Paris Agreement \citep{UNFCCC2015} in the longer term \citep[e.g.][]{Parry2018, OECD2018}. 
In the following, we study the effect of high carbon prices on district heating systems, both within the current and a prospective future electricity system. 
%Our aim in doing so is two-fold. 
%First, we want to address the question how district heating systems and their owners (which are mostly political actors) fare under high emission prices. 
%Second, we want to shed a light on the potential future role of combined heat and power plants, which are currently considered as ‘climate friendly’ due to their comparatively high total fuel efficiency.

The related literature on electricity generation has shown that rents for electricity producers with low \ce{CO2} emission intensity can increase substantially, when carbon  prices increase \citep{Sijm2006, Jouvet2013}. 
This is due to the fact that changes in carbon prices have a stronger effect on the generation cost of more emission intense generators. 
In response to a cost increase, the marginal power plant (i.e. the most expensive still dispatched power plant that sets the market price) will be increasingly often emission intense. And when an emission intense power plant is marginal, it will pass its higher cost of generation through to the electricity price.
%Similarly, we show in section \ref{dh_model} that rents of a given district heating system can increase in response to rising emission price, depending on the system's technical characteristics and the \emph{emission cost pass-through to electricity prices}.

\citet{Weyl2013} develop a theoretical framework for this so-called \emph{pass-through} effect, generalizing results obtained by \citet{Jenkin1872} for perfectly competitive markets to more general market structures. 
They find the cost pass-through to depend on market competitiveness as well as on the price elasticities of demand and supply.

In market-based electricity systems, the price elasticity of supply results from marginal cost-based power plant dispatch. 
The effect of carbon prices on marginal cost depends on specific features of the considered power plant, such as conversion efficiency and used fuels.
In consequence, the effect of emission prices can not be generalized to any electricity system, but will be specific to the power generation fleet under consideration.

In line with this, a stream of the literature uses econometric methods to identify empirical emission cost pass-through levels for various European power systems \citep[e.g.][]{Sijm2006, Honkatukia2006, Zachmann2008, Fabra2014, Hintermann2016}. 
Econometric analysis has the advantage of not requiring specific assumptions regarding the price elasticities of demand and supply or market competitiveness.
However, econometric estimation is necessarily based on historical observations, which limits the analysis to the range of observed and close-to-observed emission prices in current electricity systems. Moreover, it requires very detailed data of high quality to avoid potential endogeneity issues in the estimation \citep[c.f.][]{Fabra2014, Hintermann2016}. Yet, such data is not available for the German bidding zone.

Another strand of the literature \citep{Sijm2006, Chen2008, Lise2010} investigates emission cost pass-through to electricity prices with the numerical model COMPETES, which simulates electricity generation and transmission under imperfect competition in several EU member states.
%(The exact geographic coverage depends on the model version used.) 
The model uses twelve distinct periods with different demand levels to represent all 8760 hours of the year \citep{Sijm2005}. 
This coarse time resolution is insufficient to capture the variations in intermittent electricity generation even at the country level \citep{Ernst1999} and should be treated cautiously, depending on the research question \citep{Pfenninger2017}.

%To overcome these issues, we use the hourly power system model \emph{MEDEA}\footnote{See appendix \ref{Apndx2} for a description} to analyse emission cost pass-through given current and projected future market penetration of renewable electricity generators. 
In this study we aim at exploring the effect of high carbon prices on district heating systems, both within the current and a potential future electricity system with high shares of electricity generation from intermittent sources. First, we address the question how district heating systems and their owners, i.e. mostly the public, fare under high emission prices.
Second, we shed some light on the potential future role of combined heat and power plants, which are currently considered as `climate friendly' due to their comparatively high total fuel efficiency. 
To do this, we analyse emission cost pass-through, given current and projected future market penetration of renewable electricity generators, with the power system model MEDEA.%\footnote{See appendix \ref{Apndx2} for a description}

In section \ref{dh_model}, we develop a simple, stylized analytical framework to analyse decisions of a district heating operator. 
%Next, we introduce the power system model \emph{MEDEA} and lay out the used data in section \ref{sec_datref}. 
Next, we examine the drivers of \ce{CO2} cost pass-through in current and projected future electricity systems with high shares of intermittent renewable electricity generation. Subsequent sections \ref{sec_discuss} and \ref{sec_conclude} discuss findings and conclude.

\section{A stylized model of district heating operations} \label{dh_model}
\subsection{Cost minimization in district heating systems}
District heat is typically supplied by a broad portfolio of heat generation technologies.
Some of these technologies, for example heat boilers, are generating heat only, while others, such as combined heat and power (CHP) plants, couple the generation of heat and electricity.
A cost-minimizing operator will dispatch heat generation units according to their marginal cost, with the lowest cost unit being dispatched first.
For co-generation units, marginal cost is affected by the prevailing prices for electricity, fuels and \ce{CO2} emissions.
To analyze the effect of emission costs on the decisions of district heating suppliers, a simplified portfolio consisting of co-generation units $K$ and natural gas boilers $B$ is considered.
The cost function $C$ of the CHP operator is then given by
\begin{align}\label{1}
C(q_{f}^{B}, q_{f}^{K}) = p_{f} q_{f}^{B} + p_{e} em_{f} q_{f}^{B} + p_{f} q_{f}^{K} + p_{e} em_{f} q_{f}^{K} - p_{el} \eta_{el} q_{f}^{K}
\end{align}
where $p$ denotes prices of fuel $f$ (e.g.~coal or natural gas), $e$ emissions of \ce{CO2}, $el$ stands for electricity, and $q$ denotes quantities of fuel used in boiler $B$ or CHP units $K$.
Plant efficiency, measured as MWh of output per MWh of input, is denoted by $\eta$. $em$ represents the fuel emission factor in tons of \ce{CO2} emitted per MWh of fuel used.
At each point in time, heat generation from the portfolio has to match heat demand from the system, such that $q_{f}^{B} \eta_{th}^{B} + q_{f}^{K} \eta_{th}^{K} = d_{th}$.
The corresponding cost-minimization problem can be represented by the Langrangian
\begin{align}\label{2}
\begin{split}
\mathcal{L}(q_{f}^{B},q_{f}^{K},\lambda)= p_{f} q_{f}^{B} + p_{e} em_{f} q_{f}^{B} + p_{f} q_{f}^{K} + p_{e} em_{f} q_{f}^{K} - p_{el} \eta_{el} q_{f}^{K} \\ - \lambda ( q_{f}^{B} \eta_{th}^{B} + q_{f}^{K} \eta_{th}^{K} - d_{th})
\end{split}
\end{align}
The first-order conditions of the problem in (\ref{2}) allow to express shadow prices (cost $c$) of heat generated by a heat boiler and a CHP-plant, respectively.
\begin{align}\label{3_4}
c_{th}^{B} = \frac{1}{\eta_{th}^{B}} (p_{f} + p_{e} em_{f}) \\
c_{th}^{K} = \frac{1}{\eta_{th}^{K}} (p_{f} + p_{e} em_{f} - \eta_{el}^{K} p_{el})
\end{align}
As long as sufficient capacities are available (e.g.~in summer, when space heating demand is low), a cost minimizing operator will dispatch the lowest cost unit first.
CHP units will have an absolute cost advantage over boilers $(c_{th}^{K} \leq c_{th}^{B})$, if the 'mark-up' of the electricity price over fuel and \ce{CO2} emission cost is sufficiently high.
\begin{align}\label{5}
\frac{p_{el}}{(p_{f} + p_{e} em_{f})} \geq \frac{\eta_{th}^{B} - \eta_{th}^{K}}{\eta_{th}^{B} \eta_{el}^{K}}
\end{align}
Boilers will be dispatched when electricity prices are low (below the threshold in (\ref{5})) or when CHP units are capacity constrained (which is typically the case in winter when space heating demand peaks).

\subsection{Effect of rising \ce{CO2} prices on district heating systems} \label{CO2DHeffect}
Upon the introduction of measures to increase the price $p_{e}$ the \ce{CO2} emission cost of CHP units and boilers will increase.
Moreover, any measure that induces a general increase in carbon prices will also affect generation costs of \ce{CO2} emitting power generators. In turn, they will pass some of the cost increase on to electricity prices $p_{el}$.
Thus, the electricity price depends, amongst others, on the emission price, i.e. $p_{el}(\cdot, p_{e})$.
Rising electricity prices generate additional revenues for CHP units from electricity sales.
If these additional revenues outweigh the additional generation cost, district heating companies benefit from rising \ce{CO2} prices.
In other words, the total cost of district heat generation $c^{tot} = c_{th}^{K} q_{th}^{K} + c_{th}^{B} q_{th}^{B}$ need to decline in emission prices.
This will be the case if
\begin{align}\label{6}
\frac{\partial c^{tot}}{\partial p_{e}} = \frac{\partial c_{th}^{K}}{\partial p_{e}} q_{th}^{K} + c_{th}^{K}
\frac{\partial q_{th}^K}{\partial p_e} + \frac{\partial c_{th}^{B}}{\partial p_{e}} q_{th}^{B} +c_{th}^{B}
\frac{\partial q_{th}^{B}}{\partial p_{e}} < 0
\end{align}
To simplify this expression, we make use of the fact that (exogenous) heat demand $d_{th}$ must be supplied entirely from the district heating system, such that any reduction in heat generation from one source must lead to a corresponding increase in heat generation of another source, i.e.
$q_{th}^{K} + q_{th}^{B} = d_{th} \Rightarrow \partial q^{B} /\partial p_{e} = - \partial q^{K} / \partial p_{e}$.
Based on this, we rewrite the 'declining total cost-condition' in (\ref{6}) as
\begin{align}\label{7}
\frac{\partial c_{th}^{K}}{\partial p_{e}} q^{K} + \frac{\partial c_{th}^{B}}{\partial p_{e}} q^{B} < (c_{th}^{B} - c_{th}^{K})\frac{\partial q^{K}}{\partial p_{e}}
\end{align}
The inequality in (\ref{7}) relates produced quantities from boilers and CHPs to these units' heat cost.
To simplify (\ref{7}) further, we use the identity $q^{B} = q^{T} - q^{K}$, with $q^{T}$ being total heat production from emission intense generators. 
Then, the inequality in (\ref{7}) can be rearranged to yield the lower threshold on the pass-through for which a district heating system with a given share of CHP heat generation and given electrical efficiency reduces its total cost.
%Using the identity $q^{B} = q^{T} - q^{K}$, with $q^{T}$ being total heat production from emission intense generators, equation (\ref{7}) can be rearranged to yield the lower threshold on the pass-through from emission cost to electricity prices for which a district heating system with a given share of CHP heat generation and given electrical efficiency reduces its total cost.

In general, total costs of heat generation are declining in emission prices, if
\begin{align}\label{8}
\frac{\partial p_{el}}{\partial p_e} > \underbrace{\left( \frac{q^T - q^K}{q^K} \frac{\eta_{th}^{K}}{\eta_{th}^{B} \eta_{el}^{K}} + \frac{\eta_{th}^{B}}{\eta_{th}^{B} \eta_{el}^{K}} \right) em_{f}}_\text{A} - \underbrace{ \frac{(c_{th}^{B} - c_{th}^{K})}{p_{el}} \varepsilon_{s^{K}}^{p_{el}} \frac{\eta_{th}^{K}}{\eta_{el}^{K}} \frac{\partial p_{el}}{\partial p_{e}}}_\text{B}
\end{align}
where $\varepsilon_{s^{K}}^{p_{el}} = (\partial q^{K} / \partial p_{el}) (p_{el} / q^{K}) \geq 0$ is the electricity price elasticity of heat supply from CHP units.
\begin{figure}[t]
%\flushright
\centering
\includegraphics[scale=0.35]{Figure1_dhpt.png}
\caption{Minimally required pass-through for total cost reduction. Dot indicates the lowest pass-through level compatible with cost reduction for a district heating system with electrical efficiency of $0.5$ and a CHP heat generation share of $0.8$.}\label{Fig1}
\end{figure}

Part $\text{A}$ of the right-hand side of (\ref{8}) is strictly positive for district heating portfolios, while the sign of $\text{B}$ depends on the difference in heat cost between boilers and CHP units.
$\text{B}$ will be zero either if CHP units are capacity constrained (i.e. $\varepsilon_{s^{K}}^{p_{el}} = 0$) or if the heat cost of boilers and CHP units is equal.
If (i) CHP units have a cost advantage (cf. equation (\ref{5})) the whole of $\text{B}$ will be positive, resulting in a lower threshold for the minimally required pass-through from \ce{CO2} emission cost to electricity prices.
On the other hand, if (ii) electricity prices are so low that boilers have a cost advantage, heat will be preferentially generated in boilers, up to the available boiler capacity and heat generation from CHP units is at the minimal level.
In consequence, heat generation does not change with the electricity price, i.e. $\varepsilon_{s^K}^{p_{el}} = 0$.
This holds true until the electricity price reaches a level at which $c_{th}^{B} = c_{th}^{K}$ and we are back in case~(i).
Hence, we can restrict further analysis to the most adverse case (highest level of pass-through required for total cost reduction) in which B equals zero and (\ref{8}) simplifies to
\begin{align}\label{9}
\frac{\partial p_{el}}{\partial p_{e}} > \left( \frac{q^{T} - q^{K}}{q^{K}} \frac{\eta_{th}^{K}}{\eta_{th}^{B} \eta_{el}^{K}} + \frac{\eta_{th}^{B}}{\eta_{th}^{B} \eta_{el}^{K}} \right) em_{f}
\end{align}

Figure~\ref{Fig1} depicts the minimal total cost reducing pass-through for given shares of CHP heat generation in total emission intense heat generation and for given electrical efficiency of CHP units. 
As is to be expected, a high CHP share and high electrical efficiency both lead to lower required pass-through levels.

As the required pass-through depends on technical characteristics of district heating systems, Table~\ref{Tab1} presents exemplary requirements on the minimal pass-through that guarantees total cost reduction.
In addition, Table~\ref{Tab1} also indicates the minimal electricity mark-up over fuel and emission cost that allows CHP plants to enjoy a cost advantage over heat boilers.
For district heating systems with the least efficient power generators, the minimal pass-through compatible with reduction of total cost is more than twice as high ($0.759$) as for district heating systems with the most efficient power generators ($0.352$).
More typical district heating systems (electrical efficiency $0.4$) are able to reduce their cost with rising emission prices, provided that at least 55.6\% of the increase in \ce{CO2} emission prices is passed on to electricity prices.

In the following, we use the power system model \emph{MEDEA} to investigate the effect of emission prices on electricity prices, i.e. to estimate the pass-through $\partial p_{el} / \partial p_{e}$. A detailed description of the model is provided in \ref{Apndx2}.
%Power plant dispatch and corresponding power prices are derived for \ce{CO2} prices of $5.89$ \euro{}/t\ce{CO2} (which is equal to the average EUA price in 2017) and all $5$ \euro{}/t\ce{CO2} increments up to $80.89$ \euro{}/t\ce{CO2}, given the current inventory of power plants.
%The results are used to approximate $\partial p_{el} / \partial p_{e} \approxeq \Delta p_{el} / \Delta p_{e}$, the pass-through from emission prices to electricity prices.

\begin{table}[t]
\caption{Effects of technical parameters on heat generation units}
\smallskip
\centering
\begin{tabular}{c c c}
\hline
\makecell{Electrical \\efficiency} & \makecell{Min. mark-up \\(cf. eq. (\ref{5}))} &
\makecell{Min. pass-through \\ (cf. eq. (\ref{9}))} \\ [1.5 ex]
\hline  \hline
$0.3$ & $0.481$ & $0.759$ \\ [0.5 ex] \hline
$0.4$ & $0.389$ & $0.556$ \\ [0.5 ex] \hline
$0.5$ & $0.333$ & $0.433$ \\ [0.5 ex] \hline
$0.6$ & $0.296$ & $0.352$ \\ [1.0 ex]
\hline \hline
\end{tabular}
\label{Tab1}
\end{table}

\section{Results} \label{sec_results}
To approximate pass-through rates from \ce{CO2} emission costs to electricity prices, we determine dispatch of installed power plant capacities in $16$ emission price scenarios $s = 0,1,\ldots,15$, both for the current and a projected future electricity system.
In either setting, we start from actually observed daily EU emission allowance prices (annual average in 2017: $5.89$ \euro{}/t\ce{CO2}) and increase the price by $5$ \euro{}/t\ce{CO2} in each subsequent scenario up to an annual average price of $80.89$ \euro{}/t\ce{CO2}, leaving everything else unchanged.
The marginals on the electricity supply and demand balance equation (see equation (\ref{12}) in~\ref{Apndx2}) of our power system model reflect the endogenously determined hourly electricity prices.
We use the electricity base price $p_{el}$ (i.e. the annual average of the hourly spot price) to approximate the pass-through from emission costs to electricity prices by
\begin{align}\label{10}
\frac{\Delta p_{el}}{\Delta p_{e}} = \frac{p_{el, s} - p_{el, s-1}}{p_{e, s} - p_{e, s-1}}
\end{align}

\subsection{Effects of increasing emission prices in the current electricity system}
Our pass-through estimates for the electricity system as of 2017 are presented in Figure \ref{Fig3}, once for the current system allowing for power generation from all installed capacities and once for the current system with restricted operable capacities, e.g. due to maintenance. 
The possibility to substitute electricity generation from emission intense generators with electricity generation from low-emitting units affects pass-through estimates considerably.
If substitution is impossible, emission cost pass-through will be constant. 
With limited potential for substitution, cost pass-through can increase as emission prices rise. This happens when emission intense generators, e.g. lignite-fired plants, are inframarginal at low emission prices. As emission prices increase, marginal cost of these units rise strongly. At some emission price, one of these units will be the price setting, most expensive still dispatched unit. As emission prices increase further, this unit would be priced out of the market if sufficient capacity for replacing this unit would be available. 
However, in the absence of substitution potential, emission cost pass-through will remain high.

In the following, we will confine our attention to the more interesting case in which generation capacities can be substituted up to installed capacities. 
In this case, emission cost pass-through to electricity prices depends on the absolute level of the emission price, and is declining from $0.82$ at emission prices between $5$ and $10$ \euro{}/t\ce{CO2} to $0.64$ at emission prices between $50$ and $55$~\euro{}/t\ce{CO2}.

\begin{figure}[t]
\centering
\includegraphics[scale=0.75]{Figure3a_current_capacity.png}
\caption{Estimated pass-through from emission prices to electricity prices (2017)}\label{Fig3}
\end{figure}

The co-movement of \ce{CO2} emission prices and the pass-through can be explained by (i) a 'fuel switch' triggered by rising emissions prices, (ii) a 'CHP generation switch' and (iii) an increase in the overall fuel efficiency.

The 'fuel switch' (i) in response to increasing emission costs is illustrated in Figure~\ref{Fig4a}.
At low \ce{CO2} prices, lignite and coal fired power plants dominate the generation mix, with approximately $440$ TWh of lignite and $277$ TWh of hard coal  being burned over the course of the year.
Price-setting marginal power plants are predominantly fueled by coal and natural gas.
With emission prices rising above $20$~\euro{}/t\ce{CO2}, generation cost of lignite-fired plants increase disproportionately strong, as lignite is the fuel with the highest specific \ce{CO2} emissions. 
In effect, lignite-fired generation becomes more expensive relative to generation from coal and natural gas.
While some lignite-fired units are driven out of the market, the remaining units are increasingly often becoming the price setting, marginal (i.e. most expensive still dispatched) units.

\begin{figure}[t]
\centering
\begin{subfigure}[t]{0.475\textwidth}
\centering
\includegraphics[scale=0.75]{Figure4a_fuel_burn.png}
\caption{Fuel burn for power generation} \label{Fig4a}
\end{subfigure}
%~
\begin{subfigure}[t]{0.475\textwidth}
\centering
\includegraphics[scale=0.75]{Figure4b_emissions.png}
\caption{\ce{CO2} emissions from power generation}\label{Fig4b}
\end{subfigure}
\caption{Results from emission price scenarios}\label{Fig4}
\end{figure}

At emission prices at or above $20$~\euro{}/t\ce{CO2}, lignite loses most of its cost advantage relative to coal and natural gas and the use of lignite for power generation begins to decline markedly.
Under these conditions, lignite-fired units are frequently the most expensive dispatched units.
In consequence, any further increase in \ce{CO2} emission prices leads to a particularly strong effect on electricity prices.
However, this effect vanishes as more and more lignite-fired generators become uncompetitive and cease generation.
With \ce{CO2} emission prices reaching $40$~\euro{}/t\ce{CO2}, the decline in lignite use slows down.
Further emission reductions are caused by the substitution of lignite and coal fired plants with natural gas fired generators.
However, even for emission prices above $70$~\euro{}/t\ce{CO2} coal and lignite account for roughly one quarter of total thermal power generation as the power plant stock is fixed in our analysis and cannot adjust to higher \ce{CO2} emission costs.

The 'CHP generation switch' (ii) occurs in parallel to the fuel switch.
In response to rising emission prices, the cost of power generation from emission intense lignite increases particularly, leading to a marked decline of power generation (but a slight increase in heat generation) from these units. This is compensated by additional power generation from natural gas and hard coal. While natural gas-fired units expand their fuel use to raise their output of electricity and heat, hard coal-fired co-generation units shift their output from heat towards power. 
In total, less lignite is burned ($-27\%$as emission prices rise from $5.89$~\euro{}/t\ce{CO2} to $15.89$~\euro{}/t\ce{CO2}), but with higher total efficiency (total generation declines only by $-15\%$ over the same price range). 
Conversely, burning of hard coal and natural gas increases by $2.6$\% and $124.6$\%, respectively.

In addition, we observe (iii) an increase in overall fuel efficiency, as visible in Figure~\ref{Fig4a}.
Total fuel combustion for power generation declines from $1219$ TWh at low emission prices to $1106$ TWh at the highest \ce{CO2} emission prices that we investigated.
Taken together, these effects induce relatively steep marginal \ce{CO2} emission reductions for emission prices up to $20$~\euro{}/t\ce{CO2} (see Figure~\ref{Fig4b}).
Above this price level marginal \ce{CO2} emission reductions gradually level off as the substitution potential for lignite-fired generators vanishes. 
At emission prices above $40$~\euro{}/t\ce{CO2} most of the (short-run) substitution potential is exploited and marginal \ce{CO2}~emission reductions recede considerably.

\subsection{Robustness of estimates}\label{robustn}
To assess the robustness of our pass-through estimates with respect to model assumptions, we conduct sensitivity runs, in which we vary the parameter(s) of interest, holding everything else constant.
Figure~\ref{Fig5} summarizes all resulting estimates for the pass-through from \ce{CO2} emission costs to electricity prices.
\begin{figure}[t]
\centering
\includegraphics[scale=0.75]{Figure3c_current_sensitivity.png}
\caption{Sensitivity of pass-through estimates}\label{Fig5}
\end{figure}
\subsubsection*{International electricity exchange}

An increase of $25\%$ in net imports (quantities imported less quantities exported) leads to an average increase in the emission cost pass-through by $0.01$ compared to our baseline estimate.
Germany was a net exporter of electricity in 2017, i.e. over the course of the year electricity generation was higher than it would have been in the absence of international electricity exchange. 
Ceteris paribus, this marginal generation has to be sourced from thermal generators. In effect, the emission cost pass-through increases.

\subsubsection*{Relative price of coal}
We find that a $25\%$ increase in (relative) coal prices has a considerable impact on the emission cost pass-through at low emission prices. The higher price of coal relative to natural gas induces a $26$ TWh shift in power generation from coal to natural gas, as compared to our baseline. In effect, marginal plants are more frequently natural gas-fired, leading to initially lower pass-through levels. As the specific \ce{CO2} emissions of coal are about two-thirds higher than those of natural gas, rising emission prices lead to a disproportionate increase in the marginal cost of coal-fired generators. In effect, such generators are increasingly often the marginal (most expensive still dispatched) unit, inducing a rising pass-though level. At high emission prices, emission cost dominates fuel cost and the effect on the pass-through recedes.

\subsubsection*{Ancillary Services and system flexibility} 
Lowering the capacity required to be operational at any point in time for the provision of ancillary services by $50\%$ increases our estimated pass-through on average by $0.19$.
The minimum capacity requirement for the provision of ancillary services is effective only if the residual electricity demand falls below the minimal capacity level.
In this case, supply and demand must be balanced either through curtailment of renewable electricity generation (which would raise residual demand) or through increased pumping by pumped hydro storages (which has the additional benefit that ancillary services can also be provided through pumping).
In consequence, the marginal, price-setting unit is a renewable generator and the cost of \ce{CO2} emissions is not passed through at that time.
With less capacity required to be operational for the provision of ancillary services, such incidents occur less frequently and thermal, \ce{CO2} emission intense generators are setting the price more often. This results in a higher emission cost pass-through.\subsubsection*{Rigidity of heat supply}
To examine the effect of fully substitutability of heat generation sources, we introduce a lower bound on the hourly generation of heat from lignite and coal that is proportional to the fuels’ share in total installed CHP heat generation capacity.
For emission prices up to $15$~\euro{}/t\ce{CO2} this has little effect on our pass-through estimates, as emission intense generators are anyway dispatched. 
At higher emission prices, the constraint gets binding, requiring emission intense generators to be dispatched. In effect, less emission intense generators are determining the price and emission cost pass-through declines by up to $0.07$.

\subsection{Effect of \ce{CO2} prices in a future electricity system}
The power system undergoes rapid changes, as renewable electricity generation is added to the system while generation from fossil power plants retreats. In Germany, the Renewable Energy Act~\citep{EEG2017} governs the further expansion of renewable electricity generation. In addition, the government-installed, so-called “Coal Commission” recently proposed to limit power generation from coal and lignite and eventually end it by 2038 \citep{KommissionWSB2019}.
In line with these plans, we assume that by 2030 electricity generation from solar PV and wind turbines will increase by $70\%$ and $100\%$, respectively. Over the same time period, coal and lignite fired generation capacities are assumed to be reduced to $35\%$ and $45\%$ of capacities installed in 2017 while nuclear generation is already phased-out. 
In addition, we simulate a scenario for 2040, which assumes an end to power generation from coal and lignite, while generation from wind energy increases by more than 140\% and solar photovoltaics by 130\% as compared to our base year 2017. 
For all scenarios, we assume sufficient additional natural gas-fired generation capacities so as to guarantee resource adequacy.
\begin{figure}[t]
\centering
\includegraphics[scale=0.75]{Figure3d_future_sensitivity.png}
\caption{Pass-through estimates for future scenarios}\label{Fig3c}
\end{figure}

Given our assumptions for 2030, we estimate the carbon price pass-through at close to 0.28, with minor variation in response to the carbon price level. Due to the reduction of power generation capacities, substitution potential in response to rising carbon prices is firmly limited. Moreover, during most hours of the year the marginal unit is a low-emitting generator, reducing the carbon cost pass-through by between $0.57$ and $0.37$ compared to our baseline estimation for the year 2017. Increasing power system flexibility, modeled as halving the must-run capacity, has the potential to increase the pass-through level by $0.06$ on average. 
By 2040, in the absence of power generation from coal or lignite, emission cost pass-through declines to $0.11$ for all considered carbon prices as power generation is dominated by intermittent renewables, while low-emitting natural gas-fired generators are marginal infrequently.
Under these conditions, halving must-run capacity is not sufficient to raise the pass-through. Eliminating any must-run requirement leads to an average increase of $0.04$ in the emission cost pass-through.

\section{Discussion} \label{sec_discuss}
Our estimates of emission cost pass-through in the current German electricity system are well in line with findings from econometric \citep{Hintermann2016} and model-based \citep{Lise2010} assessments of carbon-price pass-through in the literature. This is remarkable, as in contrast to some previous studies, our analysis features high (hourly) temporal  resolution to realistically incorporate power generation from intermittent renewable sources and endogenizes the price elasticity of electricity supply.
Yet, our results are valid only for the specific electricity systems that we studied and rest on several simplifying assumptions, most notably perfect competition on electricity wholesale markets and perfectly price-inelastic electricity demand.

\begin{table}[h!]
\caption{Cost pass-through estimates for Germany based on COMPETES}
\smallskip
\centering
\begin{tabular}{l c c c c c}
\hline
\makecell{Market\\structure} & \makecell{Price elasticity\\of demand} & \makecell{Change in\\ \ce{CO2} cost} & \makecell{Chen et al.\\(2008)} & \makecell{Lise et al.\\(2010)} \\ \hline \hline
\makecell{perfect\\competition} & $0$ & $10$ & $0.92$ & -- \\ \hline
\makecell{perfect\\competition} & $0$ & $20$ & $1.18$ & $0.87$ \\ \hline
\makecell{perfect\\competition} & $0.2$ & $20$ & $0.8$ & $0.75$ \\ \hline
\makecell{oligopolistic} & $0.2$ & $20$ & $0.69$\footnote{Assuming EDF excercises market power in France} & $0.76$  \\ \hline

\makecell{perfect\\competition} & $0$ & $40$ & -- & $0.89$ \\ \hline
\makecell{perfect\\competition} & $0.2$ & $40$ & -- & $0.77$ \\ \hline
\makecell{oligopolistic} & $0.2$ & $40$ & -- & $0.65$ \\ \hline
\end{tabular}
\label{competes_results}
\end{table}

As \citet{Chen2008} and \citet{Lise2010} demonstrate, both assumptions can have significant effect on the estimated pass-through. 
For further illustration, Table \ref{competes_results} summarizes pass-through estimates from these publications employing the numerical model COMPETES to asses emission cost pass-through in imperfectly competitive European electricity markets.\footnote{Please note that reported pass-through estimates are not directly comparable to ours, as (marginal) emission cost pass-through is approximated for a change in emission cost from $0$~\euro{}/t\ce{CO2} to the level indicated as 'Change in \ce{CO2} cost'.}

Assuming the price elasticity of demand is $0.2$ instead of $0$ reduces the pass-through by approximately $0.12$. Likewise, going from perfect competition to oligopolistic competition can reduce pass-through by up to $0.12$ \citep{Lise2010}.
While the magnitude of the effect is potentially significant, one also needs to take into account how realistic these deviations from our assumptions are, when assessing our results.

The German \citet{MPK2017}, which monitors the electricity market and advises the government on competition policy and regulation, finds no significant evidence of excessive market power on the German power market. 
% This is also acknowledged by \citet{XX}, who state “”.
Moreover, the short‐run price elasticity of electricity demand is empirically found to be very low \citep{Lijesen2007}. 
This suggests, that the estimation bias caused by assuming perfectly competitive markets and inelastic demand is likely small.
Likewise, moderate changes in net imports should not alter pass-through substantially, as our analysis in section \ref{robustn} suggests very limited sensitivity of carbon price pass-through to net electricity imports.
However, we only changed the magnitude of net imports but not the pattern of (net) imports over time. Hence, our analysis might not reveal the full impact of changes in the pan-European power plant dispatch on the emission cost pass-through caused by rising \ce{CO2} emission prices.
More frequent electricity imports from low-carbon sources (e.g. French nuclear power plants) could crowd out some of the \ce{CO2} emission intense production with high marginal emission cost.
This would reduce pass-through rates.
If, on the other hand, exports would increase in response to rising \ce{CO2} prices, pass-through rates could turn out to be higher than estimated.
Similarly, changes in the relative prices of coal and natural gas have little effect on carbon price pass-through at high levels of \ce{CO2} prices. Yet, at lower carbon prices, shifts in relative coal prices affect the pattern of power plant substitution, compressing or stretching the pass-through curvature, without systematically reducing cost pass-through. 
Further potential bias stems from the assumption of perfect substitutability of coal and natural gas in district heat generation. Requiring more emission intense heat generation than dispatched by our model would lower pass-through estimates , as the inflexibility works much like must-run constraints.

\section{Conclusions} \label{sec_conclude}
We have shown that emission cost pass-through to electricity prices is an important determinant of the current and even more so the future profitability of district heating systems. Moreover, emission cost pass-through with heterogeneous suppliers, such as in the electricity system, is affected by a wide variety of factors, including the emission price level itself.

Given our conservative estimation of technically required pass-through levels (see section \ref{CO2DHeffect}) and the small and likely upward-skewed bias in our estimates of system-wide pass-through, we currently see operators of district heating systems of moderate or better electrical efficiency ($\eta_{el}^{K} \geq 0.375$) in a position to benefit from any \ce{CO2} price increase up to the highest considered level (provided that boilers generate less than $20\%$ of total fossil heat generation)\footnote{As boilers are typically used for peak generation, this should hold for many, if not most DHC systems.}.
However, the long-run transition towards a decarbonized, highly renewable electricity system is a substantial challenge for DHSO profitability, in particular if the renewables expansion is accompanied by rising \ce{CO2} emission prices.
By 2030, even the most efficient CHP plants will no longer benefit from high emission prices, as the still remaining coal and lignite fired power plants will be inframarginal throughout most of the year. Increasing power system flexibility, which we modeled as a reduction in must-run capacity, could raise emission cost pass-through somewhat. Yet, at pass-through levels below $0.3$ profitability will still deteriorate even for the most efficient natural gas-fueled CHP plants. 
With the complete phase-out of coal, that the ‘coal commission’ dates to the late 2030s, emission prices will have little effect on electricity prices. In the face of high emission prices, DHSOs will then either have to pass on increasing costs to final heat prices or seek for other, low-emission heat generation options.

From the perspective of an owner of a natural gas-fired district heating system, an immediate increase in emission prices could be attractive, as the resulting 'windfall profits' could be invested in low-emission heat generation technologies. 
This would relieve such an DHSO from the pressure of high carbon prices that will intensify over time. 
However, this window of opportunity for DHSOs is closing at the speed of the coal exit.

Policy makers need to reconsider whether conventional district heating systems, which source their heat largely as a by-product from fossil power generation, are indeed a promising component of future energy systems. The current German target of $120$ TWh of heat generated by co-generation in 2025 \citep{KWKG2016} is likely to lock DHSOs into a generation structure built on fossil fuels that undermines DHSO profits, raises heat costs and falls short of potential carbon emission reductions.
\section*{Acknowledgements}
We gratefully acknowledge support from the European Research Council (“reFUEL" ERC-2017-STG 758149).

\newpage
\bibliographystyle{elsarticle-harv}
\bibliography{passtru_main}

\newpage
\appendix
\nocitesec{*}
\section{Description of the power system Model \emph{MEDEA}} \label{sec_datref} \label{Apndx2}
\subsection{Overview}
We use a simple, stylized and parsimonious model of the German and Austrian power system to estimate the pass-through from emission prices to electricity prices in the common bidding area.
Our power system model determines the cost-minimizing hourly dispatch of thermal and hydro storage power plants that is required to meet price-inelastic (residual) demand for electricity and district heat.
In total, $552$ thermal power plants and $62$ hydro storage plants are grouped in $34$ technology clusters. We differentiate the clusters by generation technologies (e.g. steam turbine, combustion turbine, combined cycle, etc.) and by fuels (uranium, lignite, hard coal, natural gas, mineral oil, biomass, water).
Thermal power plants burn fuels to generate electricity and are constrained in their operation by installed capacities for heat and power generation.
Heat generation is possible in heat boilers (aggregate capacity $30$ GWth, efficiency $0.9$) or in combined heat and power (CHP) plants, which must respect the limits of their feasible operation region.\footnote{The feasible operation region specifies all viable combinations of heat and power generation along with the required fuel use.}
Heat generation and consumption are aggregated, i.e. heat generators are not serving specific district heating systems.
Electricity generation from intermittent renewable sources (wind, solar, run-of-river) is given exogenously but can be curtailed.
Electricity can also be stored in reservoir and pumped hydro storages.
Generation from hydro storage plants is constrained by turbine capacity and energy contained in reservoirs of limited size.
Reservoirs are filled by inflows or by pumping (pumped storages only).
To better capture operational differences, we model daily, weekly and seasonal reservoir and pumped storage plants separately.
Electricity exchange with countries outside Austria and Germany is held fixed at hourly quantities realized in 2017.
To ensure a stable and secure operation of the electricity system, power plants must provide ancillary services such as frequency control or voltage support.
We assume that this requires generators with an installed capacity of at least $21$ GW\footnote{This is equal to 12.5\% of peak load plus 7.5\% of installed solar and wind power capacity and is broadly in line with findings by \citet{Hirth2015} and \citet{Nicolosi2012}.} to be operational (either generating or pumping in case of pumped hydro storages) at any point in time.
For a mathematical description see \ref{Apndx2}. The model is implemented in GAMS and was solved by Gurobi on an Intel Xeon Gold $6144$ with $264$ GB RAM.
%The model code can be found at \url{https://github.com/sebwehrle/medea}.

\subsection{Mathematical Description}
\subsubsection{List of Sets}
\begin{tabular}{c l}
$f \in F$ & fuels \\
$g \in G$ & generators \\
$CHP \in G$ & combined heat and power generators \\
$PSP \in G$ & hydro storage generators \\
$l \in L$ & limits to the feasible operation region of CHP plants \\
$p \in P$ & products generated \\
$t \in T$ & time periods (hours)
\end{tabular}

\subsubsection{List of Parameters}
\begin{tabular}{c l}
$\underline{a}$ & \makecell[l]{minimum generation level required for provision of \\ancillary services [MW]} \\
$\overline{C}_{g,p}$ & \makecell[l]{maximal generation of product $p$ by generator in \\cluster $g$ [MW]} \\
$\overline{STOR}_{g}$ & \makecell[l]{reservor storage capacity of hydro storage plant $g$ [MWh]} \\
$D_{t,p}$ & demand for product $p$ at time $t$ [MW] \\
$N_{g}$ & number of plants in cluster $g$ \\
$om_g$  & \makecell[l]{variable operation and maintenance cost of \\generator $g$ [\euro/MWh]} \\
$ORF_{g,l,f}$ & \makecell[l]{use of fuel $f$ at limit $l$ of the operating region of a \\CHP unit in cluster $g$ [MWh]} \\
$ORP_{g,l,p}$ & \makecell[l]{generation of product $p$ at limit $l$ of the operating \\region of a CHP unit in cluster $g$ [MWh]} \\
$P_{t,eua}$ & price of \ce{CO2} emissions at time $t$ [\euro/t~\ce{CO2}e] \\
$P_{t,f}$ & price of fuel $f$ at time $t$ [\euro/MWh] \\
$Q_{t,pv}$ & solar energy generated at time $t$ [MW] \\
$Q_{t,ror}$ & energy generated by run-of-river plants at time $t$ [MW]\\
$Q_{t,we}$ & wind energy generated at time $t$ [MW]\\
$Q_{t,nip}$ & net imports at time $t$ [MWh] \\
$Q_{t,res}$ & inflows into hydro reservoirs at time $t$ [MWh/h]\\
$em_{f}$ & emission factor of fuel $f$ [t \ce{CO2}/MWh fuel used]\\
$\eta_{g,f,p}$ & \makecell[l]{efficiency of generator $g$ using fuel $f$ to generate product $p$}\\

\end{tabular}

\subsubsection{List of Variables}
\begin{tabular}{c l}
$ppsp_{t,g}$ & \makecell[l]{energy pumped into pumped storage reservoirs at \\time $t$ [MWh]} \\
$qf_{t,g,f}$ & quantity of fuel $f$ used by generator $g$ at time $t$ [MWh]\\
$qp_{t,g,p}$ & energy generated by generator $g$ at time $t$ [MWh]\\
$qpsp_{t,g}$ & power generated by hydro storage plant $g$ at time $t$ [MWh]\\
$sconv_{t,g,l}$ & convexity variable $g$\\
$qstor_{t,g}$ & \makecell[l]{quantity of energy stored in hydro storage plant $g$ at \\time $t$ [MWh]}\\
$qns_{t,p}$ & non-served load of product $p$ at time $t$ [MWh]\\
$qct_{t}$ & quantity of energy curtailed at time $t$ [MWh]\\
\end{tabular}

\subsubsection{Mathematical model}
Our power system model uses a linear programming formulation of the economic dispatch problem for thermal units within the Austro-German bidding zone. Operation of pumped storage plants is also formulated as a linear problem.
The model's objective is to minimize total system cost, the sum of fuel, emission and operation and maintenance cost along with the cost associated with curtailment of renewable energies and loss of load.
\begin{align}
\min \left(\sum_{t,g,f}\left(\left(P_{t,f} + em_{f} P_{t,eua}+om_{g}\right) qf_{t,g,f} + qns_{t,p} M + qct_{t} N \right) \right)
\end{align}
In each hour the market has to clear, such that electricity supply from thermal and net generation from hydro storage plants plus power generation from non-dispatchable sources (wind energy, photovoltaics, and run-of-river hydro plants) equals electricity demand less net imports of electricity.
\begin{align}\label{12}
\begin{split}
D_{t,pwr} - Q_{t,nip} = \sum_{g} \left(q p_{t,g,pwr} \right) + \sum_{g\in PSP} \left(qpsp_{t,g} - ppsp_{t,g} \right) \\
+ Q_{t,we} + Q_{t,pv} + Q_{t,ror}, \forall t
\end{split}
\end{align}
In linear (economic dispatch) models, the marginals ('shadow prices') on equation (\ref{12}) can be interpreted as power prices in an energy-only market. We use these marginals to derive the pass-through from emission prices to power prices.
As we also consider co-generation of heat and power in our model, we introduce the heat balance equation (\ref{13}). Heat supply from CHP units and heat boilers must be adequate to meet district heating demand $D_{t,ht}$.
\begin{align}\label{13}
D_{t,ht} \leq \sum_{g \in CHP} \left( qp_{t,g,ht} \right) + qns_{t,p}, \forall t
\end{align}
Hourly generation of power and heat is constrained by installed capacity. 
\begin{align}
qp_{t,g,p} \leq \bar{C}_{g,p}, \forall t
\end{align}
Coproduction of heat and power in CHP-plants is governed by
\begin{align}
\sum_{l} sconv_{t,g,l} = 1, \forall g \in CHP \\
\sum_{l} sconv_{t,g,l} ORP_{g,l,p} = qp_{t,g,p}, \forall g \in CHP \\
\sum_{l} sconv_{t,g,l} ORF_{g,l,f} \leq qfuel_{t,g,f}
\end{align}
Power production by power-only generators is modelled as a linear function of plant efficiency
\begin{align}
qp_{t,g,pwr} \leq \sum_{f} \eta_{g,f,pwr} qf_{t,g,f}, \forall t, g \notin CHP  
\end{align}
Ancillary services must be provided by operating thermal generation units or active hydro storage plants (regardless of whether they are pumping or turbining).
\begin{align}
\underline{a} \leq qp_{t,g,pwr} + qpsp_{h,g} + qstor_{t,g}, \forall t
\end{align}
Operation of pumped storage plants is subject to the equations
\begin{align}
qpsp_{h,g} \eta_{g} \leq \overline{C}_{g}, \forall t, g \in PSP \\
ppsp_{t,g} \leq \overline{C}_{g}, \forall t, g \in PSP \\
qstor_{t,g} - qstor_{t-1, g} = ppsp_{h,g} \eta_{g} - qpsp_{h,g} \\
qstor_{t,g} \leq \overline{STOR}_{g}, \forall t, g \in PSP
\end{align}

\subsection{Data}
Information regarding the power plant stock (generation capacities, technology, locations) in Germany are based on data provided by the Open Power System Data (OPSD) project \citep{OPSD2018a}.
Data on Austrian power plants was collected through own research and includes information from the regulatory body~\citep{EControl2003}, sector associations~\citep{OeEn2017}, operating companies, and water registers of federal authorities~\citep{LandVorarlberg2018, LandTirol2018}.
Hourly electricity generation from intermittent sources (solar PV, wind) and load in Austria and Germany is also sourced from \citet{OPSD2018b}. 
Further time series on international commercial electricity exchanges, the aggregate filling rate of hydro reservoirs and storage plants, and the actual electricity generation and consumption of hydro power plants (including run-of-river plants) are obtained from ENTSO-E's transparency platform~\citep{ENTSOE2018}.
We approximate inflows of water to reservoirs in Austria by combining downsampled data on weekly water reservoir levels with hourly electricity generation and pumping from hydro reservoirs and pumped storage plants.\footnote{In Germany, 98\% of the installed hydro storage capacity is pumped hydro power so that we abstract from inflows to German hydro storage plants.}
Hourly district heating demand is estimated based on synthetic load profiles for natural gas demand~\citep{Almbauer2009}.
These load profile make use of daily average temperatures from MERRA-2 satellite data~\citep{GMAO2015} and are scaled to total final consumption of district heat in Germany and Austria.\footnote{As final data for the consumption of district heat in 2017 was not published at the time of writing, we scale 2016 district heat consumption by the relative change in heating degree days from 2016 to 2017~\citep{AGEB2018a, OeStat2018}.}
Realized prices of hard coal, natural gas and EU emission allowances for the year 2017 are taken from the \citet{EEX2019}.
Prices for mineral oil are approximated on the basis of prices for Brent crude oil as published by the~\citet{EIA2018}.
As there are no market prices for nuclear fuel and lignite, we estimate lignite cost at $5.50$~\euro{}/MWh (including mining, but excluding emission cost) and $3.50$~\euro{}/MWh for nuclear fuel.
Descriptive Statistics for all used time series are displayed in Table~\ref{Tab2}.

\begin{table}[t!]
\caption{Descriptive statistics}
\smallskip
\centering
\begin{tabular}{c c c c c c c}
\hline
 & Unit & Min & Max & Mean & Std. Dev & Source \\ \hline \hline
Wind & $GW$ & $0.279$ & $41.72$ & $12.49$ & $9.03$ & OPSD \\ \hline
Solar PV & $GW$ & $0$ & $28.33$ & $4.23$ & $6.49$ & OPSD \\ \hline
Run-of-river & $GW$ & $2.08$ & $7.13$ & $4.78$ & $1.13$ & ENTSO-E \\ \hline
\makecell{Reservoir \\ inflows} & $GW$ & $0$ & $2.14$ & $0.79$ & $0.59$ & \makecell{own\\ calculations} \\ \hline
\makecell{Electricity \\consumption} & $GW$ & $38.52$ & $96.77$ & $69.09$ & $12.59$ & ENTSO-E \\ \hline
\makecell{District heat \\ consumption} & $GW$ & $1.54$ & $42.04$ & $13.49$ & $9.60$ & \makecell{own \\calculations} \\ \hline
\makecell{Net imports \\ (commercial \\exchange)} & $GW$ & $-13.74$ & $7.07$ & $-4.87$ & $2.93$ & ENTSO-E\\ \hline
\makecell{Coal price \\(API2)} & \euro{}$/MWh_{th}$ & $8.32$ & $11.36$ & $9.65$ & $0.72$ & eex \\ \hline
\makecell{Natural gas \\price (NCG)} & \euro{}$/MWh_{th}$ & $15.91$ & $19.02$ & $17.27$ & $0.87$ & eex\\ \hline
\makecell{Mineral oil \\price} & \euro{}$/MWh_{th}$ & $23.18$ & $32.93$ & $28.25$ & $2.57$ & EIA\\ \hline
\makecell{EU Emission \\Allowance price} & \euro{}$/MWh_{th}$ & $4.41$ & $8.21$ & $5.89$ & $1.11$ & eex\\
\hline \hline
\end{tabular}
\label{Tab2}
\end{table}

\subsection{Model calibration and goodness-of-fit}
\begin{figure}[t]
\centering
\includegraphics[scale=0.75]{Figure2_calibration.png}
\caption{Actual and model derived fuel burn (2017)}\label{Fig2}
\end{figure}

We use power plant efficiencies and availabilities to calibrate our power system model on actual data for quantities of fuel burnt for power generation~\citep{AGEB2018, OeStat2018} and greenhouse gas emissions \citep{UBA_DE2018a, UBA_DE2018b} from 2017.
\footnote{For Austria, the most recent data on GHG emissions and fuel burn for power generation available at the time of writing relates to 2016. As Austria’s electricity consumption is only around one tenth of Germany’s and roughly two-third of power generation in Austria stems from hydro power plants, the induced error should, however, be small.}

As visible in figure \ref{Fig2}, our model is able to replicate fuel burn and emissions fairly well, although the use of coal is somewhat overstated ($+5.9$ TWh), while consumption of natural gas and oil are underestimated by $4.0$ TWh and $5.7$ TWh. Overall, modelled \ce{CO2} emissions from power generation of $316.4$ million tonnes are falling short of near time estimates by around $2.4\%$ or $7.9$ million tonnes.
Comparing the model-derived hourly shadow prices of electricity to actual day-ahead prices at the European Energy Exchange in 2017, we observe a correlation of $0.80$ and a root mean squared error (RMSE) of $10.91$.

\bibliographystylesec{elsarticle-harv}
\bibliographysec{passtru_appendix}

\end{document}

### 2019-08-01 doc/mathematical_description.tex
add mathematical description

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

\usepackage{natbib}
\usepackage{makecell}
\usepackage{eurosym} 
\usepackage[version=4]{mhchem}
%\usepackage{courier} 
\usepackage[inner=2.5cm,outer=2.5cm]{geometry}

%\usepackage{gensymb}
%\usepackage{subcaption}
%\usepackage{url}
%\usepackage{breakurl}
%\def\UrlBreaks{\do\/\do-}
%
%\usepackage{multibib}
%\newcites{sec}{Bibliography} 
\begin{document}
\title{Mathematical description of the power system model \emph{medea}}
\section{Overview}
\emph{medea} is a simple, stylized and parsimonious power system model.
It simulates investment in intermittent and conventional electricity and heat generation technologies as well as in cross-border electricity transmission capacities.
At the same time, the model determines the system-cost minimizing hourly dispatch of electricity and heat generators to meet price-inelastic demand.
Model results include hourly energy generation by technology and the associated fuel use and CO2 emissions, investment in and decommissioning of conventional and renewable generators and energy storages, hourly cross-border flows of electricity and potentially required transmission capacity expansion, as well as producer and consumer surplus.

A detailed description of the model is provided in the following. Section \ref{sets} gives an overview of the sets and set elements used in \emph{medea}. Sections \ref{parameters} and \ref{variables} introduce the model's parameters and variables, while section \ref{mathmodel} gives a detailed description of the model's mathematical formulation.

% decisions covered - intermittent generation given, thermal and storage dispatch, investment in generation and storage technologies, interconnector capacity
% intermittent technologies: wind onshore \& offshore, pv, run-of-river
% dispatchable technologies: $36$ technologies for dispatchable power and heat generation, including co-generation units for the combined generation of power and heat (for industry or district heating).

%\emph{medea} numerically determines a partial-equilibrium in interconnected electricity markets as would be implemented by a social planner seeking to minimize the total system cost. 
%In particular, this implies that social planners seeking to minimize system cost of individual electricity markets might have an incentive to deviate from the overall cost minimum.
%The model assumes perfectly price inelastic demand for electricity and heat. 
%Within each country we assume sufficient transmission capacities to be in place ('copperplate'). 
%Moreover, heat demand can be met by any heat generator operating in the same country.

%It covers the hourly dispatch of thermal and hydro storage plants in seven European countries, while taking electricity generation from intermittent sources such as wind or solar energy as given. (Not true because curtailment is possible)
%In addition, the model can determine optimal investment in intermittent and dispatchable generation technologies, storage technologies, and international transmission capacities.

% It can be used to simulate short-term dispatch decisions, but also currently covers

% determines the system cost-minimizing hourly dispatch of thermal and hydro storage power plants required to meet price-inelastic (residual) demand for electricity and district heat.

% Thermal power plants and hydro storage plants are grouped in $34$ technology clusters. We differentiate the clusters by generation technologies (e.g. steam turbine, combustion turbine, combined cycle, etc.) and by fuels (uranium, lignite, hard coal, natural gas, mineral oil, biomass, water).

% Thermal power plants burn fuels to generate electricity and are constrained in their operation by installed capacities for heat and power generation.

% Heat generation is possible in heat boilers (aggregate capacity $30$ GWth, efficiency $0.9$) or in combined heat and power (CHP) plants, which must respect the limits of their feasible operation region.\footnote{The feasible operation region specifies all viable combinations of heat and power generation along with the required fuel use.}

% Heat generation and consumption are aggregated, i.e. heat generators are not serving specific district heating systems.

% Electricity generation from intermittent renewable sources (wind, solar, run-of-river) is given exogenously but can be curtailed.

% Electricity can also be stored in reservoir and pumped hydro storages.

%Generation from hydro storage plants is constrained by turbine capacity and energy contained in reservoirs of limited size.
%Reservoirs are filled by inflows or by pumping (pumped storages only).
%To better capture operational differences, we model daily, weekly and seasonal reservoir and pumped storage plants separately.
%Electricity exchange with countries outside Austria and Germany is held fixed at hourly quantities realized in 2017.
%To ensure a stable and secure operation of the electricity system, power plants must provide ancillary services such as frequency control or voltage support.
%We assume that this requires generators with an installed capacity of at least $21$ GW \footnote{This is equal to 12.5\% of peak load plus 7.5\% of installed solar and wind power capacity and is broadly in line with findings by \citet{Hirth2015} and \citet{Nicolosi2012}.} to be operational (either generating or pumping in case of pumped hydro storages) at any point in time.

%For a mathematical description see \ref{Apndx2}. The model is implemented in GAMS and was solved by Gurobi on an Intel Xeon Gold $6144$ with $264$ GB RAM.
%The model code can be found at \url{https://github.com/sebwehrle/medea}.

\section{Sets} \label{sets}
\begin{tabular}{l l l l}
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{elements} \\
\hline \hline
market zones & $z \in Z$ & \texttt{r} & \texttt{AT, DE} \\ \hline

\makecell[l]{time periods \\(hours)} & $t \in T$ & \texttt{t} & \texttt{t1, t2, \ldots, t8760} \\ \hline

\makecell[l]{power generating \\technologies} & $i \in I$ & \texttt{tec} &

\makecell[l]{ \texttt{nuc, lig\_stm, lig\_stm\_chp,}\\
	\texttt{lig\_boa, lig\_boa\_chp,} \\
	\texttt{coal\_sub, coal\_sub\_chp,} \\
	\texttt{coal\_sc, coal\_sc\_chp,}\\
	\texttt{coal\_usc, coal\_usc\_chp,} \\
	\texttt{coal\_igcc,}\\
	\texttt{ng\_stm, ng\_stm\_chp,} \\
	\texttt{ng\_ctb\_lo, ng\_ctb\_lo\_chp,}\\ 
	\texttt{ng\_ctb\_hi, ng\_ctb\_hi\_chp,}\\
	\texttt{ng\_cc\_lo, ng\_cc\_lo\_chp,}\\
	\texttt{ng\_cc\_hi, ng\_cc\_hi\_chp,}\\
	\texttt{ng\_mtr, ng\_mtr\_chp,} \\
	\texttt{ng\_boiler\_chp,} \\
	\texttt{oil\_stm, oil\_stm\_chp,} \\ 
	\texttt{oil\_ctb, oil\_ctb\_chp,} \\
	\texttt{oil\_cc, oil\_cc\_chp,} \\
	\texttt{bio, bio\_chp,}
	\texttt{heatpump\_pth} 
} \\ \hline

CHP technologies & $j \in I$ & \texttt{tec\_chp} & \makecell[l]{
	\texttt{lig\_stm\_chp, lig\_boa\_chp,}\\
	\texttt{coal\_sub\_chp, coal\_sc\_chp,}\\
	\texttt{coal\_usc\_chp, ng\_stm\_chp,}\\ 
	\texttt{ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp,}\\
	\texttt{ng\_cc\_lo\_chp, ng\_cc\_hi\_chp,}\\
	\texttt{ng\_mtr\_chp, ng\_boiler\_chp,}\\
	\texttt{oil\_stm\_chp, oil\_ctb\_chp,} \\ 
	\texttt{oil\_cc\_chp, bio\_chp} } \\ \hline

\makecell[l]{power to heat \\ technologies} & $h \in I$ & \texttt{tec\_pth} & \texttt{heatpump\_pth}  \\ \hline

storage technologies & $k \in K$ & \texttt{tec\_strg} & \makecell[l]{
	\texttt{res\_day, res\_week, res\_season,} \\ 
	\texttt{psp\_day, psp\_week, psp\_season,} \\
	\texttt{battery}
	} \\ \hline

\makecell[l]{intermittent \\generators} & $n \in N$ & \texttt{tec\_itm} & \texttt{wind\_on, wind\_off, pv, ror}\\ \hline

fuels & $f \in F$ & \texttt{f} & \makecell[l]{\texttt{nuclear, lignite, coal,} \\ \texttt{gas, oil, biomass, power}} \\ \hline

\makecell[l]{feasible operation \\region limits}& $l \in L$ & \texttt{l} & \texttt{l1, l2, l3, l4} \\ \hline

energy products & $p \in P$ & \texttt{prd} & \texttt{power, heat} \\ \hline \hline
\end{tabular}

\section{Parameters} \label{parameters}
\begin{tabular}{l l l l}
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\
\hline \hline
\makecell[l]{minimal conventional \\generation} & $\mu_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(r)} & GW \\ \hline
energy demand & $D_{z,t,p}$ & \texttt{CONSUMPTION(r,t,prd)} & GW \\ \hline
power plant efficiency & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & $MW / MW_{th}$\\ \hline
fuel emission intensity & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $t \ce{CO2}/MWh_{th}$\\ \hline
\makecell[l]{feasible operating \\region} & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & MWh ? \\ \hline
\makecell[l]{feasible operating \\region} & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & MWh ? \\ \hline
\makecell[l]{intermittent \\generation profile} & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(r,t,tec\_itm)} & \% \\ \hline
%\makecell[l]{exports to not \\modelled regions} & $X$ & \texttt{FLOW\_EXPORTS(r,t)} & GW \\ \hline
%\makecell[l]{imports from not \\modelled regions} & $X$ & \texttt{FLOW\_IMPORTS(r,t)} & GW \\ \hline
\makecell[l]{installed capacity \\of intermittent \\generators} & $r^{0}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(r,tec\_itm)} & GW \\ \hline
\makecell[l]{installed capacity \\of thermal generators} & $g^{0}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(r,tec)} & GW \\ \hline
\makecell[l]{capital cost of \\intermittent generators \\(specific, annuity)} & $\kappa_{z,n}$ & \texttt{INVESTCOST\_ITM(r,tec\_itm)} & $\frac{mn EUR}{GW}$ \\ \hline
\makecell[l]{capital cost of \\thermal generators \\(specific, annuity)} & $\kappa_{z,i}$ & \texttt{INVESTCOST\_THERMAL(r,tec)} & $\frac{mn EUR}{GW}$ \\ \hline
\makecell[l]{capital cost of \\storages - power \\(specific, annuity)} & $\kappa^{P}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'cost\_power')} & $\frac{mn EUR}{GW}$ \\ \hline
\makecell[l]{capital cost of \\storages - energy \\(specific, annuity)} & $\kappa^{E}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'cost\_energy')} & $\frac{mn EUR}{GW}$ \\ \hline
net transfer capacity & $\tau_{z,zz}$ & \texttt{NTC(r,rr)} & GW \\ \hline
\makecell[l]{count of power plants \\of same technology} & $|i_{z}|$ & \texttt{NUM(r,tec)} & . \\ \hline
fixed O\&M cost & $\kappa^o_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & EUR \\ \hline
variable O\&M cost & $\kappa^o_{i}$ & \texttt{OM\_VAR\_COST(tec)} & EUR / MWh \\ \hline
% day-ahead price & $p_{el}$ & \texttt{PRICE\_DA(t,r)} & EUR / MWh \\ \hline
\ce{CO2} price & $p^{e}_{t,z}$ & \texttt{PRICE\_EUA(t,r)} & EUR / t CO2 \\ \hline
fuel price & $p_{t,z,f}$ & \texttt{PRICE\_FUEL(t,r,f)} & EUR / MWh \\ \hline
reservoir inflows & $\rho_{z,t,k}$ & \texttt{RESERVOIR\_INFLOWS(r,t,tec\_strg)} & MW \\ \hline
max power out & $\bar{s}^{out}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'power\_out')} & GW \\ \hline
max power in & $\bar{s}^{in}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'power\_in')} & GW \\ \hline
max energy stored & $\bar{v}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'energy\_max')} & . \\ \hline
efficiency power out & $\eta^{out}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'efficiency\_out')} & . \\ \hline
efficiency power in & $\eta^{in}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'efficiency\_in')} & . \\ \hline
\hline
\end{tabular}

\section{Variables} \label{variables}
\begin{tabular}{l l l l}
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\
\hline \hline
%cost variables
system cost & $C_{z}$ & \texttt{cost(r)} & EUR \\ \hline
emission cost & $C^{e}_{z,t,i}$ & \texttt{cost\_emission(r,t,tec)} & EUR \\ \hline
fuel cost & $C^{f}_{z,t,i}$ & \texttt{cost\_fuel(r,t,tec)} & EUR \\ \hline
total o\&m cost & $C^{om}_{z,i}$ & \texttt{cost\_om(r,tec)} & EUR \\ \hline
\makecell[l]{capital cost of \\generators} & $C_{z}^{inv,i}$ & \texttt{cost\_invgen(r)} & EUR \\ \hline
\makecell[l]{capital cost of \\storages} & $C_{z}^{inv,k}$ & \texttt{cost\_invstrg(r)} & EUR \\ \hline
\makecell[l]{capital cost of \\interconnectors} & $C_{z}^{inv,ic}$ & \texttt{cost\_gridexpansion(r)} & EUR \\ \hline
% investment and decommissioning variables
\makecell[l]{added capacity of \\intermittents} & $r^{+}_{z,n}$ & \texttt{invest\_res(r,tec\_itm)} & GW \\ \hline
\makecell[l]{added capacity of \\conventionals} & $g^{+}_{z,i}$ & \texttt{invest\_thermal(r,tec)} & GW \\ \hline
\makecell[l]{added storage \\capacity (power)} & $v^{+}_{z,k}$ & \texttt{invest\_storage\_power(r,tec\_strg)} & EUR \\ \hline
\makecell[l]{added in storage \\capacity (energy)} & $s^{+}_{z,k}$ & \texttt{invest\_storage\_energy(r,tec\_strg)} & EUR \\ \hline
\makecell[l]{added transmission \\capacity} & $x^{+}_{z,zz}$ & \texttt{invest\_ntc(r,rr)} & EUR \\ \hline
\makecell[l]{decommissioned \\capacity of \\conventionals} & $g^{-}_{z,i}$ & \texttt{decommission(r,tec)} & GW \\ \hline
% dispatch variables
\makecell[l]{energy generated by \\conventionals} & $g_{z,t,i,p}$ & \texttt{q\_gen(r,t,tec,prd)} & GW \\ \hline
\makecell[l]{electricity generated \\by intermittents} & $r_{z,t,n}$ & \texttt{q\_itm(r,t,tec\_itm)} & GW \\ \hline
\makecell[l]{operating region \\weight} & $w_{z,t,i,l}$ & \texttt{cc\_weights(r,t,tec,l)} & . \\ \hline
\makecell[l]{fuel burn for energy \\generation} & $b_{z,t,i,f}$ & \texttt{q\_fueluse(r,t,tec,f)} & GW \\ \hline
energy stored in & $s^{in}_{z,t,k}$ & \texttt{q\_store\_in(r,t,tec\_strg)} & GW \\ \hline
energy stored out & $s^{out}_{z,t,k}$ & \texttt{q\_store\_out(r,t,tec\_strg)} & GW \\ \hline
\makecell[l]{storage energy \\content} & $v_{z,t,k}$ & \texttt{storage\_level(r,t,tec\_strg)} & GWh \\ \hline
\makecell[l]{electricity net \\export} & $x_{z,zz,t}$ & \texttt{flow(r,rr,t)} & GW \\ \hline
curtailed energy & $\Omega^{-}_{z,t,p}$ & \texttt{q\_curtail(r,t,prd)} & GW \\ \hline
non-served energy & $\Omega^{+}_{z,t,p}$ & \texttt{q\_nonserved(r,t,prd)} & GW \\ \hline
% other variables
\ce{CO2} emissions & $e_{z}$ & \texttt{emissions(r)} & t \ce{CO2} \\ \hline

\end{tabular}

\newpage
\section{Mathematical description} \label{mathmodel}
Objective function: total system cost, minimization
\begin{align}
\min C = \sum_{z} (C_{z})
\end{align}
System cost consist of fuel cost $C^{f}_{z,t,i}$, emission cost $C^{e}_{z,t,i}$, operation and maintenance cost $C^{om}_{z,i}$, and capital costs of investment in generation, storage and transmission equipment plus the cost of non-served load that accrues when demand is not met, i.e. when there is a black out.
\begin{align}
\begin{split}
C_{z} = \sum_{t,i} & C^{f}_{z,t,i} + \sum_{t,i} C^{e}_{z,t,i} + \sum_{i} C^{om}_{z,i} + C^{inv,i}_{z} + \\
& C^{inv,n}_{z} + C^{inv,k}_{z} + C^{inv,ic}_{z} + \sum_{t,p} C^{nse}_{z,t,p} \qquad \qquad \forall z
\end{split}
\end{align}

Our power system model uses a linear programming formulation of the economic dispatch problem for thermal units within the Austro-German bidding zone. Operation of pumped storage plants is also formulated as a linear problem.
The model's objective is to minimize total system cost, the sum of fuel, emission and operation and maintenance cost along with the cost associated with curtailment of renewable energies and loss of load.
\begin{align}
\min \left(\sum_{t,g,f}\left(\left(P_{t,f} + em_{f} P_{t,eua}+om_{g}\right) qf_{t,g,f} + qns_{t,p} M + qct_{t} N \right) \right)
\end{align}
In each hour the market has to clear, such that electricity supply from thermal and net generation from hydro storage plants plus power generation from non-dispatchable sources (wind energy, photovoltaics, and run-of-river hydro plants) equals electricity demand less net imports of electricity.
\begin{align}\label{12}
\begin{split}
D_{t,pwr} - Q_{t,nip} = \sum_{g} \left(q p_{t,g,pwr} \right) + \sum_{g\in PSP} \left(qpsp_{t,g} - ppsp_{t,g} \right) \\
+ Q_{t,we} + Q_{t,pv} + Q_{t,ror}, \forall t
\end{split}
\end{align}
In linear (economic dispatch) models, the marginals ('shadow prices') on equation (\ref{12}) can be interpreted as power prices in an energy-only market. We use these marginals to derive the pass-through from emission prices to power prices.
As we also consider co-generation of heat and power in our model, we introduce the heat balance equation (\ref{13}). Heat supply from CHP units and heat boilers must be adequate to meet district heating demand $D_{t,ht}$.
\begin{align}\label{13}
D_{t,ht} \leq \sum_{g \in CHP} \left( qp_{t,g,ht} \right) + qns_{t,p}, \forall t
\end{align}
Hourly generation of power and heat is constrained by installed capacity. 
\begin{align}
qp_{t,g,p} \leq \bar{C}_{g,p}, \forall t
\end{align}
Coproduction of heat and power in CHP-plants is governed by
\begin{align}
\sum_{l} sconv_{t,g,l} = 1, \forall g \in CHP \\
\sum_{l} sconv_{t,g,l} ORP_{g,l,p} = qp_{t,g,p}, \forall g \in CHP \\
\sum_{l} sconv_{t,g,l} ORF_{g,l,f} \leq qfuel_{t,g,f}
\end{align}
Power production by power-only generators is modelled as a linear function of plant efficiency
\begin{align}
qp_{t,g,pwr} \leq \sum_{f} \eta_{g,f,pwr} qf_{t,g,f}, \forall t, g \notin CHP  
\end{align}
Ancillary services must be provided by operating thermal generation units or active hydro storage plants (regardless of whether they are pumping or turbining).
\begin{align}
\underline{a} \leq qp_{t,g,pwr} + qpsp_{h,g} + qstor_{t,g}, \forall t
\end{align}
Operation of pumped storage plants is subject to the equations
\begin{align}
qpsp_{h,g} \eta_{g} \leq \overline{C}_{g}, \forall t, g \in PSP \\
ppsp_{t,g} \leq \overline{C}_{g}, \forall t, g \in PSP \\
qstor_{t,g} - qstor_{t-1, g} = ppsp_{h,g} \eta_{g} - qpsp_{h,g} \\
qstor_{t,g} \leq \overline{STOR}_{g}, \forall t, g \in PSP
\end{align}

\section{Data}
Information regarding the power plant stock (generation capacities, technology, locations) in Germany are based on data provided by the Open Power System Data (OPSD) project \citep{OPSD2018a}.
Data on Austrian power plants was collected through own research and includes information from the regulatory body~\citep{EControl2003}, sector associations~\citep{OeEn2017}, operating companies, and water registers of federal authorities~\citep{LandVorarlberg2018, LandTirol2018}.
Hourly electricity generation from intermittent sources (solar PV, wind) and load in Austria and Germany is also sourced from \citet{OPSD2018b}. 
Further time series on international commercial electricity exchanges, the aggregate filling rate of hydro reservoirs and storage plants, and the actual electricity generation and consumption of hydro power plants (including run-of-river plants) are obtained from ENTSO-E's transparency platform~\citep{ENTSOE2018}.
We approximate inflows of water to reservoirs in Austria by combining downsampled data on weekly water reservoir levels with hourly electricity generation and pumping from hydro reservoirs and pumped storage plants.\footnote{In Germany, 98\% of the installed hydro storage capacity is pumped hydro power so that we abstract from inflows to German hydro storage plants.}
Hourly district heating demand is estimated based on synthetic load profiles for natural gas demand~\citep{Almbauer2009}.
These load profile make use of daily average temperatures from MERRA-2 satellite data~\citep{GMAO2015} and are scaled to total final consumption of district heat in Germany and Austria.\footnote{As final data for the consumption of district heat in 2017 was not published at the time of writing, we scale 2016 district heat consumption by the relative change in heating degree days from 2016 to 2017~\citep{AGEB2018a, OeStat2018}.}
Realized prices of hard coal, natural gas and EU emission allowances for the year 2017 are taken from the \citet{EEX2019}.
Prices for mineral oil are approximated on the basis of prices for Brent crude oil as published by the~\citet{EIA2018}.
As there are no market prices for nuclear fuel and lignite, we estimate lignite cost at $5.50$~\euro{}/MWh (including mining, but excluding emission cost) and $3.50$~\euro{}/MWh for nuclear fuel.
Descriptive Statistics for all used time series are displayed in Table~\ref{Tab2}.

\begin{table}[t!]
\caption{Descriptive statistics}
\smallskip
\centering
\begin{tabular}{c c c c c c c}
\hline
 & Unit & Min & Max & Mean & Std. Dev & Source \\ \hline \hline
Wind & $GW$ & $0.279$ & $41.72$ & $12.49$ & $9.03$ & OPSD \\ \hline
Solar PV & $GW$ & $0$ & $28.33$ & $4.23$ & $6.49$ & OPSD \\ \hline
Run-of-river & $GW$ & $2.08$ & $7.13$ & $4.78$ & $1.13$ & ENTSO-E \\ \hline
\makecell{Reservoir \\ inflows} & $GW$ & $0$ & $2.14$ & $0.79$ & $0.59$ & \makecell{own\\ calculations} \\ \hline
\makecell{Electricity \\consumption} & $GW$ & $38.52$ & $96.77$ & $69.09$ & $12.59$ & ENTSO-E \\ \hline
\makecell{District heat \\ consumption} & $GW$ & $1.54$ & $42.04$ & $13.49$ & $9.60$ & \makecell{own \\calculations} \\ \hline
\makecell{Net imports \\ (commercial \\exchange)} & $GW$ & $-13.74$ & $7.07$ & $-4.87$ & $2.93$ & ENTSO-E\\ \hline
\makecell{Coal price \\(API2)} & \euro{}$/MWh_{th}$ & $8.32$ & $11.36$ & $9.65$ & $0.72$ & eex \\ \hline
\makecell{Natural gas \\price (NCG)} & \euro{}$/MWh_{th}$ & $15.91$ & $19.02$ & $17.27$ & $0.87$ & eex\\ \hline
\makecell{Mineral oil \\price} & \euro{}$/MWh_{th}$ & $23.18$ & $32.93$ & $28.25$ & $2.57$ & EIA\\ \hline
\makecell{EU Emission \\Allowance price} & \euro{}$/MWh_{th}$ & $4.41$ & $8.21$ & $5.89$ & $1.11$ & eex\\
\hline \hline
\end{tabular}
\label{Tab2}
\end{table}

%\section{Model calibration and goodness-of-fit}
%\begin{figure}[t]
%\centering
%\includegraphics[scale=0.75]{Figure2_calibration.png}
%\caption{Actual and model derived fuel burn (2017)}\label{Fig2}
%\end{figure}
%
%We use power plant efficiencies and availabilities to calibrate our power system model on actual data for quantities of fuel burnt for power generation~\citep{AGEB2018, OeStat2018} and greenhouse gas emissions \citep{UBA_DE2018a, UBA_DE2018b} from 2017.
%\footnote{For Austria, the most recent data on GHG emissions and fuel burn for power generation available at the time of writing relates to 2016. As Austria’s electricity consumption is only around one tenth of Germany’s and roughly two-third of power generation in Austria stems from hydro power plants, the induced error should, however, be small.} 
%
%As visible in figure \ref{Fig2}, our model is able to replicate fuel burn and emissions fairly well, although the use of coal is somewhat overstated ($+5.9$ TWh), while consumption of natural gas and oil are underestimated by $4.0$ TWh and $5.7$ TWh. Overall, modelled \ce{CO2} emissions from power generation of $316.4$ million tonnes are falling short of near time estimates by around $2.4\%$ or $7.9$ million tonnes.
%Comparing the model-derived hourly shadow prices of electricity to actual day-ahead prices at the European Energy Exchange in 2017, we observe a correlation of $0.80$ and a root mean squared error (RMSE) of $10.91$.
\end{document}

### 2019-08-01 doc/mathematical_description.tex

@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsmath}
+\usepackage[fleqn]{amsmath}
 \usepackage{amsfonts}
 \usepackage{amssymb}
+\DeclareMathOperator*{\argmax}{arg\,max}
 
@@
 
-%\usepackage{gensymb}
-%\usepackage{subcaption}
-%\usepackage{url}
-%\usepackage{breakurl}
-%\def\UrlBreaks{\do\/\do-}
-%
-%\usepackage{multibib}
-%\newcites{sec}{Bibliography} 
 \begin{document}
@@
 
-% decisions covered - intermittent generation given, thermal and storage dispatch, investment in generation and storage technologies, interconnector capacity
-% intermittent technologies: wind onshore \& offshore, pv, run-of-river
-% dispatchable technologies: $36$ technologies for dispatchable power and heat generation, including co-generation units for the combined generation of power and heat (for industry or district heating).
-
-%\emph{medea} numerically determines a partial-equilibrium in interconnected electricity markets as would be implemented by a social planner seeking to minimize the total system cost. 
-%In particular, this implies that social planners seeking to minimize system cost of individual electricity markets might have an incentive to deviate from the overall cost minimum.
-%The model assumes perfectly price inelastic demand for electricity and heat. 
-%Within each country we assume sufficient transmission capacities to be in place ('copperplate'). 
-%Moreover, heat demand can be met by any heat generator operating in the same country.
-
-%It covers the hourly dispatch of thermal and hydro storage plants in seven European countries, while taking electricity generation from intermittent sources such as wind or solar energy as given.
-(Not true because curtailment is possible)
-%In addition, the model can determine optimal investment in intermittent and dispatchable generation technologies, storage technologies, and international transmission capacities.
-
-% It can be used to simulate short-term dispatch decisions, but also currently covers
-
-% determines the system cost-minimizing hourly dispatch of thermal and hydro storage power plants required to meet price-inelastic (residual) demand for electricity and district heat.
-
-% Thermal power plants and hydro storage plants are grouped in $34$ technology clusters.
-We differentiate the clusters by generation technologies (e.g.
-steam turbine, combustion turbine, combined cycle, etc.) and by fuels (uranium, lignite, hard coal, natural gas, mineral oil, biomass, water).
-
-% Thermal power plants burn fuels to generate electricity and are constrained in their operation by installed capacities for heat and power generation.
-
-% Heat generation is possible in heat boilers (aggregate capacity $30$ GWth, efficiency $0.9$) or in combined heat and power (CHP) plants, which must respect the limits of their feasible operation region.\footnote{The feasible operation region specifies all viable combinations of heat and power generation along with the required fuel use.}
-
-% Heat generation and consumption are aggregated, i.e. heat generators are not serving specific district heating systems.
-
-% Electricity generation from intermittent renewable sources (wind, solar, run-of-river) is given exogenously but can be curtailed.
-
-% Electricity can also be stored in reservoir and pumped hydro storages.
-
-%Generation from hydro storage plants is constrained by turbine capacity and energy contained in reservoirs of limited size.
-%Reservoirs are filled by inflows or by pumping (pumped storages only).
-%To better capture operational differences, we model daily, weekly and seasonal reservoir and pumped storage plants separately.
-%Electricity exchange with countries outside Austria and Germany is held fixed at hourly quantities realized in 2017.
-%To ensure a stable and secure operation of the electricity system, power plants must provide ancillary services such as frequency control or voltage support.
-%We assume that this requires generators with an installed capacity of at least $21$ GW \footnote{This is equal to 12.5\% of peak load plus 7.5\% of installed solar and wind power capacity and is broadly in line with findings by \citet{Hirth2015} and \citet{Nicolosi2012}.} to be operational (either generating or pumping in case of pumped hydro storages) at any point in time.
-
-%For a mathematical description see \ref{Apndx2}. The model is implemented in GAMS and was solved by Gurobi on an Intel Xeon Gold $6144$ with $264$ GB RAM.
-%The model code can be found at \url{https://github.com/sebwehrle/medea}.
-
 \section{Sets} \label{sets}
@@
 
-\makecell[l]{power generating \\technologies} & $i \in I$ & \texttt{tec} &
+\makecell[l]{power \\generation \\technologies} & $i \in I$ & \texttt{tec} &
 
@@
 
-CHP technologies & $j \in I$ & \texttt{tec\_chp} & \makecell[l]{
+\makecell[l]{CHP \\technologies} & $j \in J \subset I$ & \texttt{tec\_chp} & \makecell[l]{
 	\texttt{lig\_stm\_chp, lig\_boa\_chp,}\\
@@
 
-\makecell[l]{power to heat \\ technologies} & $h \in I$ & \texttt{tec\_pth} & \texttt{heatpump\_pth}  \\ \hline
+\makecell[l]{power to heat \\ technologies} & $h \in H \subset I$ & \texttt{tec\_pth} & \texttt{heatpump\_pth}  \\ \hline
 
-storage technologies & $k \in K$ & \texttt{tec\_strg} & \makecell[l]{
+\makecell[l]{storage \\technologies} & $k \in K$ & \texttt{tec\_strg} & \makecell[l]{
 	\texttt{res\_day, res\_week, res\_season,} \\ 
@@
 
-energy products & $p \in P$ & \texttt{prd} & \texttt{power, heat} \\ \hline \hline
+energy products & $m \in M=\{\text{el}, \text{ht}\}$ & \texttt{prd} & \texttt{power, heat} \\ \hline \hline
 \end{tabular}
@@
 \hline \hline
-\makecell[l]{minimal conventional \\generation} & $\mu_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(r)} & GW \\ \hline
+\makecell[l]{minimal conventional \\generation} & $\sigma_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(r)} & GW \\ \hline
 energy demand & $D_{z,t,p}$ & \texttt{CONSUMPTION(r,t,prd)} & GW \\ \hline
@@
 %\makecell[l]{imports from not \\modelled regions} & $X$ & \texttt{FLOW\_IMPORTS(r,t)} & GW \\ \hline
-\makecell[l]{installed capacity \\of intermittent \\generators} & $r^{0}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(r,tec\_itm)} & GW \\ \hline
-\makecell[l]{installed capacity \\of thermal generators} & $g^{0}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(r,tec)} & GW \\ \hline
+\makecell[l]{installed capacity \\of intermittent \\generators} & $\bar{r}^{0}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(r,tec\_itm)} & GW \\ \hline
+\makecell[l]{installed capacity \\of thermal generators} & $\bar{g}^{0}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(r,tec)} & GW \\ \hline
 \makecell[l]{capital cost of \\intermittent generators \\(specific, annuity)} & $\kappa_{z,n}$ & \texttt{INVESTCOST\_ITM(r,tec\_itm)} & $\frac{mn EUR}{GW}$ \\ \hline
@@
 \makecell[l]{capital cost of \\storages - energy \\(specific, annuity)} & $\kappa^{E}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'cost\_energy')} & $\frac{mn EUR}{GW}$ \\ \hline
-net transfer capacity & $\tau_{z,zz}$ & \texttt{NTC(r,rr)} & GW \\ \hline
-\makecell[l]{count of power plants \\of same technology} & $|i_{z}|$ & \texttt{NUM(r,tec)} & . \\ \hline
-fixed O\&M cost & $\kappa^o_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & EUR \\ \hline
-variable O\&M cost & $\kappa^o_{i}$ & \texttt{OM\_VAR\_COST(tec)} & EUR / MWh \\ \hline
+\makecell[l]{installed available \\transfer capacity} & $\bar{x}^{0}_{z,zz}$ & \texttt{ATC(r,rr)} & GW \\ \hline
+\makecell[l]{count 100 MW slices \\of same technology} & $|i_{z}|$ & \texttt{NUM(r,tec)} & . \\ \hline
+quasi-fixed O\&M cost & $o^{q}_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & EUR \\ \hline
+variable O\&M cost & $o^{v}_{i}$ & \texttt{OM\_VAR\_COST(tec)} & EUR / MWh \\ \hline
 % day-ahead price & $p_{el}$ & \texttt{PRICE\_DA(t,r)} & EUR / MWh \\ \hline
-\ce{CO2} price & $p^{e}_{t,z}$ & \texttt{PRICE\_EUA(t,r)} & EUR / t CO2 \\ \hline
-fuel price & $p_{t,z,f}$ & \texttt{PRICE\_FUEL(t,r,f)} & EUR / MWh \\ \hline
+\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_EUA(t,r)} & EUR / t CO2 \\ \hline
+fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,r,f)} & EUR / MWh \\ \hline
 reservoir inflows & $\rho_{z,t,k}$ & \texttt{RESERVOIR\_INFLOWS(r,t,tec\_strg)} & MW \\ \hline
@@
 max power in & $\bar{s}^{in}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'power\_in')} & GW \\ \hline
-max energy stored & $\bar{v}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'energy\_max')} & . \\ \hline
+max energy stored & $\bar{v}^{0}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'energy\_max')} & . \\ \hline
 efficiency power out & $\eta^{out}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'efficiency\_out')} & . \\ \hline
 efficiency power in & $\eta^{in}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'efficiency\_in')} & . \\ \hline
+value of lost load & $\mu$ & \texttt{not set yet} & EUR \\ \hline
 \hline
@@
 %cost variables
-system cost & $C_{z}$ & \texttt{cost(r)} & EUR \\ \hline
+zonal system cost & $C_{z}$ & \texttt{cost(r)} & EUR \\ \hline
 emission cost & $C^{e}_{z,t,i}$ & \texttt{cost\_emission(r,t,tec)} & EUR \\ \hline
@@
 % investment and decommissioning variables
-\makecell[l]{added capacity of \\intermittents} & $r^{+}_{z,n}$ & \texttt{invest\_res(r,tec\_itm)} & GW \\ \hline
-\makecell[l]{added capacity of \\conventionals} & $g^{+}_{z,i}$ & \texttt{invest\_thermal(r,tec)} & GW \\ \hline
-\makecell[l]{added storage \\capacity (power)} & $v^{+}_{z,k}$ & \texttt{invest\_storage\_power(r,tec\_strg)} & EUR \\ \hline
-\makecell[l]{added in storage \\capacity (energy)} & $s^{+}_{z,k}$ & \texttt{invest\_storage\_energy(r,tec\_strg)} & EUR \\ \hline
-\makecell[l]{added transmission \\capacity} & $x^{+}_{z,zz}$ & \texttt{invest\_ntc(r,rr)} & EUR \\ \hline
-\makecell[l]{decommissioned \\capacity of \\conventionals} & $g^{-}_{z,i}$ & \texttt{decommission(r,tec)} & GW \\ \hline
+\makecell[l]{added capacity of \\intermittents} & $\bar{r}^{+}_{z,n}$ & \texttt{invest\_res(r,tec\_itm)} & GW \\ \hline
+\makecell[l]{added capacity of \\conventionals} & $\bar{g}^{+}_{z,i}$ & \texttt{invest\_thermal(r,tec)} & GW \\ \hline
+\makecell[l]{added storage \\capacity (power)} & $\bar{s}^{+}_{z,k}$ & \texttt{invest\_storage\_power(r,tec\_strg)} & EUR \\ \hline
+\makecell[l]{added storage \\capacity (energy)} & $\bar{v}^{+}_{z,k}$ & \texttt{invest\_storage\_energy(r,tec\_strg)} & EUR \\ \hline
+\makecell[l]{added transmission \\capacity} & $\bar{x}^{+}_{z,zz}$ & \texttt{invest\_atc(r,rr)} & EUR \\ \hline
+\makecell[l]{decommissioned \\capacity of \\conventionals} & $\bar{g}^{-}_{z,i}$ & \texttt{decommission(r,tec)} & GW \\ \hline
 % dispatch variables
@@
 \makecell[l]{electricity net \\export} & $x_{z,zz,t}$ & \texttt{flow(r,rr,t)} & GW \\ \hline
-curtailed energy & $\Omega^{-}_{z,t,p}$ & \texttt{q\_curtail(r,t,prd)} & GW \\ \hline
-non-served energy & $\Omega^{+}_{z,t,p}$ & \texttt{q\_nonserved(r,t,prd)} & GW \\ \hline
+curtailed energy & $\Omega^{+}_{z,t}$ & \texttt{q\_curtail(r,t)} & GW \\ \hline
+non-served energy & $\Omega^{-}_{z,t,p}$ & \texttt{q\_nonserved(r,t,prd)} & GW \\ \hline
 % other variables
@@
 \section{Mathematical description} \label{mathmodel}
-Objective function: total system cost, minimization
+\paragraph{Model objective}
+\emph{medea} minimizes total system cost, i.e. the total cost of generating electricity and heat from technologies and capacities adequate to meet demand.
 \begin{align}
@@
 \end{align}
-System cost consist of fuel cost $C^{f}_{z,t,i}$, emission cost $C^{e}_{z,t,i}$, operation and maintenance cost $C^{om}_{z,i}$, and capital costs of investment in generation, storage and transmission equipment plus the cost of non-served load that accrues when demand is not met, i.e.
-when there is a black out.
+System costs consist of fuel cost $C^{f}_{z,t,i}$, emission cost $C^{e}_{z,t,i}$, operation and maintenance cost $C^{om}_{z,i}$, capital costs of investment in generation, storage and transmission equipment, and the cost of non-served load that accrues when demand is not met, e.g.
+when there is a power outage.
 \begin{align}
@@
 C_{z} = \sum_{t,i} & C^{f}_{z,t,i} + \sum_{t,i} C^{e}_{z,t,i} + \sum_{i} C^{om}_{z,i} + C^{inv,i}_{z} + \\
-& C^{inv,n}_{z} + C^{inv,k}_{z} + C^{inv,ic}_{z} + \sum_{t,p} C^{nse}_{z,t,p} \qquad \qquad \forall z
+& C^{inv,n}_{z} + C^{inv,k}_{z} + C^{inv,ic}_{z} + C^{nse}_{z} \qquad \qquad \forall z
 \end{split}
 \end{align}
+Upper-case $C$ represent total cost, while $\kappa$ denotes specific, annualized capital cost of technology investment. Prices for fuels and \ce{CO2} are denoted by uppercase $P$.
 
-Our power system model uses a linear programming formulation of the economic dispatch problem for thermal units within the Austro-German bidding zone.
-Operation of pumped storage plants is also formulated as a linear problem.
-The model's objective is to minimize total system cost, the sum of fuel, emission and operation and maintenance cost along with the cost associated with curtailment of renewable energies and loss of load.
+The components of total cost are calculated according to equations \ref{fuel_cost} to \ref{lost_load_cost}.
 \begin{align}
-\min \left(\sum_{t,g,f}\left(\left(P_{t,f} + em_{f} P_{t,eua}+om_{g}\right) qf_{t,g,f} + qns_{t,p} M + qct_{t} N \right) \right)
-\end{align}
-In each hour the market has to clear, such that electricity supply from thermal and net generation from hydro storage plants plus power generation from non-dispatchable sources (wind energy, photovoltaics, and run-of-river hydro plants) equals electricity demand less net imports of electricity.
-\begin{align}\label{12}
-\begin{split}
-D_{t,pwr} - Q_{t,nip} = \sum_{g} \left(q p_{t,g,pwr} \right) + \sum_{g\in PSP} \left(qpsp_{t,g} - ppsp_{t,g} \right) \\
-+ Q_{t,we} + Q_{t,pv} + Q_{t,ror}, \forall t
-\end{split}
-\end{align}
-In linear (economic dispatch) models, the marginals ('shadow prices') on equation (\ref{12}) can be interpreted as power prices in an energy-only market.
-We use these marginals to derive the pass-through from emission prices to power prices.
-As we also consider co-generation of heat and power in our model, we introduce the heat balance equation (\ref{13}).
-Heat supply from CHP units and heat boilers must be adequate to meet district heating demand $D_{t,ht}$.
-\begin{align}\label{13}
-D_{t,ht} \leq \sum_{g \in CHP} \left( qp_{t,g,ht} \right) + qns_{t,p}, \forall t
-\end{align}
-Hourly generation of power and heat is constrained by installed capacity. 
-\begin{align}
-qp_{t,g,p} \leq \bar{C}_{g,p}, \forall t
-\end{align}
-Coproduction of heat and power in CHP-plants is governed by
-\begin{align}
-\sum_{l} sconv_{t,g,l} = 1, \forall g \in CHP \\
-\sum_{l} sconv_{t,g,l} ORP_{g,l,p} = qp_{t,g,p}, \forall g \in CHP \\
-\sum_{l} sconv_{t,g,l} ORF_{g,l,f} \leq qfuel_{t,g,f}
-\end{align}
-Power production by power-only generators is modelled as a linear function of plant efficiency
-\begin{align}
-qp_{t,g,pwr} \leq \sum_{f} \eta_{g,f,pwr} qf_{t,g,f}, \forall t, g \notin CHP  
-\end{align}
-Ancillary services must be provided by operating thermal generation units or active hydro storage plants (regardless of whether they are pumping or turbining).
-\begin{align}
-\underline{a} \leq qp_{t,g,pwr} + qpsp_{h,g} + qstor_{t,g}, \forall t
-\end{align}
-Operation of pumped storage plants is subject to the equations
-\begin{align}
-qpsp_{h,g} \eta_{g} \leq \overline{C}_{g}, \forall t, g \in PSP \\
-ppsp_{t,g} \leq \overline{C}_{g}, \forall t, g \in PSP \\
-qstor_{t,g} - qstor_{t-1, g} = ppsp_{h,g} \eta_{g} - qpsp_{h,g} \\
-qstor_{t,g} \leq \overline{STOR}_{g}, \forall t, g \in PSP
+&C^{f}_{z,t,i}& =&\ \sum_{f} \left( P_{t,z,f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i \label{fuel_cost} \\
+&C^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: \varepsilon_{f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i\\
+&C^{om}_{z,i}& =&\ \left(|i_{z}| - \bar{g}^{-}_{z,i} + \bar{g}^{+}_{z,i}\right) o^{q}_{i} + \sum_{t,m} \left(o^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
+&C^{inv,i}_{z}& =&\ \kappa_{z,i} \: \bar{g}^{+}_{z,i} \qquad \qquad &\forall z\\
+&C^{inv,n}_{z}& =&\ \kappa_{z,n} \: \bar{r}^{+}_{z,n} \qquad \qquad &\forall z\\
+&C^{inv,k}_{z}& =&\ \kappa^{P}_{z,k} \: \bar{s}^{+}_{z,k} + \kappa^{E}_{z,k} \: v^{+}_{z,k} \qquad \qquad &\forall z\\
+&C^{inv,ic}_{z}& =&\ \kappa^{ic}_{z} \: \bar{x}^{+}_{z,zz} \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
+&C^{nse}_{z}& =&\ \mu \sum_{t,m} \Omega^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
 \end{align}
 
-\section{Data}
-Information regarding the power plant stock (generation capacities, technology, locations) in Germany are based on data provided by the Open Power System Data (OPSD) project \citep{OPSD2018a}.
-Data on Austrian power plants was collected through own research and includes information from the regulatory body~\citep{EControl2003}, sector associations~\citep{OeEn2017}, operating companies, and water registers of federal authorities~\citep{LandVorarlberg2018, LandTirol2018}.
-Hourly electricity generation from intermittent sources (solar PV, wind) and load in Austria and Germany is also sourced from \citet{OPSD2018b}. 
-Further time series on international commercial electricity exchanges, the aggregate filling rate of hydro reservoirs and storage plants, and the actual electricity generation and consumption of hydro power plants (including run-of-river plants) are obtained from ENTSO-E's transparency platform~\citep{ENTSOE2018}.
-We approximate inflows of water to reservoirs in Austria by combining downsampled data on weekly water reservoir levels with hourly electricity generation and pumping from hydro reservoirs and pumped storage plants.\footnote{In Germany, 98\% of the installed hydro storage capacity is pumped hydro power so that we abstract from inflows to German hydro storage plants.}
-Hourly district heating demand is estimated based on synthetic load profiles for natural gas demand~\citep{Almbauer2009}.
-These load profile make use of daily average temperatures from MERRA-2 satellite data~\citep{GMAO2015} and are scaled to total final consumption of district heat in Germany and Austria.\footnote{As final data for the consumption of district heat in 2017 was not published at the time of writing, we scale 2016 district heat consumption by the relative change in heating degree days from 2016 to 2017~\citep{AGEB2018a, OeStat2018}.}
-Realized prices of hard coal, natural gas and EU emission allowances for the year 2017 are taken from the \citet{EEX2019}.
-Prices for mineral oil are approximated on the basis of prices for Brent crude oil as published by the~\citet{EIA2018}.
-As there are no market prices for nuclear fuel and lignite, we estimate lignite cost at $5.50$~\euro{}/MWh (including mining, but excluding emission cost) and $3.50$~\euro{}/MWh for nuclear fuel.
-Descriptive Statistics for all used time series are displayed in Table~\ref{Tab2}.
+\paragraph{Market clearing}
+In each hour, the markets for electricity and heat have to clear.
+Equation \ref{market_clearing_el} ensures that the total supply from conventional and intermittent sources, and storages equals total electricity demand plus net exports, electricity stored and used for heat generation.
+Likewise, equation \ref{market_clearing_ht} clears the heat market by equating heat generation to heat demand.
 
-\begin{table}[t!]
-\caption{Descriptive statistics}
-\smallskip
-\centering
-\begin{tabular}{c c c c c c c}
-\hline
- & Unit & Min & Max & Mean & Std. Dev & Source \\ \hline \hline
-Wind & $GW$ & $0.279$ & $41.72$ & $12.49$ & $9.03$ & OPSD \\ \hline
-Solar PV & $GW$ & $0$ & $28.33$ & $4.23$ & $6.49$ & OPSD \\ \hline
-Run-of-river & $GW$ & $2.08$ & $7.13$ & $4.78$ & $1.13$ & ENTSO-E \\ \hline
-\makecell{Reservoir \\ inflows} & $GW$ & $0$ & $2.14$ & $0.79$ & $0.59$ & \makecell{own\\ calculations} \\ \hline
-\makecell{Electricity \\consumption} & $GW$ & $38.52$ & $96.77$ & $69.09$ & $12.59$ & ENTSO-E \\ \hline
-\makecell{District heat \\ consumption} & $GW$ & $1.54$ & $42.04$ & $13.49$ & $9.60$ & \makecell{own \\calculations} \\ \hline
-\makecell{Net imports \\ (commercial \\exchange)} & $GW$ & $-13.74$ & $7.07$ & $-4.87$ & $2.93$ & ENTSO-E\\ \hline
-\makecell{Coal price \\(API2)} & \euro{}$/MWh_{th}$ & $8.32$ & $11.36$ & $9.65$ & $0.72$ & eex \\ \hline
-\makecell{Natural gas \\price (NCG)} & \euro{}$/MWh_{th}$ & $15.91$ & $19.02$ & $17.27$ & $0.87$ & eex\\ \hline
-\makecell{Mineral oil \\price} & \euro{}$/MWh_{th}$ & $23.18$ & $32.93$ & $28.25$ & $2.57$ & EIA\\ \hline
-\makecell{EU Emission \\Allowance price} & \euro{}$/MWh_{th}$ & $4.41$ & $8.21$ & $5.89$ & $1.11$ & eex\\
-\hline \hline
-\end{tabular}
-\label{Tab2}
-\end{table}
+\begin{align}
+\sum_{i} g_{z,t,i,\text{el}} + \sum_{k} s^{out}_{z,t,k} + \sum_{n} r_{z,t,n} =
+D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} 
+- \Omega^{-}_{z,t,\text{el}} + \Omega^{+}_{z,t} \qquad \forall z,t \label{market_clearing_el} \\
+\sum_{i} g_{z,t,i,\text{ht}} = D_{z,t,\text{ht}} - \Omega^{-}_{z,t,\text{ht}} \qquad \forall z,t \label{market_clearing_ht}
+\end{align}
 
-%\section{Model calibration and goodness-of-fit}
-%\begin{figure}[t]
-%\centering
-%\includegraphics[scale=0.75]{Figure2_calibration.png}
-%\caption{Actual and model derived fuel burn (2017)}\label{Fig2}
-%\end{figure}
-%
-%We use power plant efficiencies and availabilities to calibrate our power system model on actual data for quantities of fuel burnt for power generation~\citep{AGEB2018, OeStat2018} and greenhouse gas emissions \citep{UBA_DE2018a, UBA_DE2018b} from 2017.
-%\footnote{For Austria, the most recent data on GHG emissions and fuel burn for power generation available at the time of writing relates to 2016.
-As Austria’s electricity consumption is only around one tenth of Germany’s and roughly two-third of power generation in Austria stems from hydro power plants, the induced error should, however, be small.}
-%
-%As visible in figure \ref{Fig2}, our model is able to replicate fuel burn and emissions fairly well, although the use of coal is somewhat overstated ($+5.9$ TWh), while consumption of natural gas and oil are underestimated by $4.0$ TWh and $5.7$ TWh.
-Overall, modelled \ce{CO2} emissions from power generation of $316.4$ million tonnes are falling short of near time estimates by around $2.4\%$ or $7.9$ million tonnes.
-%Comparing the model-derived hourly shadow prices of electricity to actual day-ahead prices at the European Energy Exchange in 2017, we observe a correlation of $0.80$ and a root mean squared error (RMSE) of $10.91$.
+In its current formulation, the model represents energy-only electricity and heat markets without capacity payments.
+Thus, the marginals of the market clearing equations ($\partial C / \partial D_{z,t,m}$) can be interpreted as the zonal prices for electricity and heat, respectively.
+
+\paragraph{Thermal electricity generation}
+Energy generation is constrained by available installed capacity, which can be adjusted through investment and decommissioning.
+\begin{align}
+g_{z,t,i,m} \leq \argmax_{l}(\psi_{i,l,m}) \left( |i_{z}| + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \right) \qquad \qquad \forall z,t,i,m
+\end{align}
+Generator efficiency $\eta$ determines the amount of fuel that needs to be burnt in order to generate a given amount of electricity.
+\begin{align}
+g_{z,t,i,\text{el}} = \sum_{f} \eta_{i,\text{el},f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \notin J
+\end{align}
+
+\paragraph{Electric heat generation}
+Heat pumps and electric boilers use electricity to generate heat.
+The amount of electricity required to generate a given amount of heat depends on the unit's efficiency $\eta$.
+For heat pumps, $\eta$ is assumed to be $3$.
+\begin{align}
+g_{z,t,i,\text{ht}} = \sum_{f} \eta_{i,\text{ht},f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \in H
+\end{align}
+
+\paragraph{Thermal co-generation}
+Co-generation units jointly generate heat and electricity.
+All feasible combinations of heat and electricity generation along with the corresponding fuel requirement are reflected in so-called `feasible operating region'.
+The elements $l \in L$ span up a three-dimensional, convex feasible operating region for each co-generation technology. 
+The weights $w$ form a convex combination of the corners $l$ scaled to the available installed capacity of each co-generation technology.
+Heat and electricity output along with the corresponding fuel requirement is then set according to the chosen weights.
+\begin{align}
+\sum_{l} w_{z,t,i,l} = |i_{z}| + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \qquad \qquad \forall z,t,i \in J \\
+g_{z,t,i,m} \leq \sum_{l} w_{z,t,i,l} \: \psi_{i,l,m} \qquad \qquad \forall z,t,i \in J, m \\
+b_{z,t,i,f} \geq \sum_{l} w_{z,t,i,l} \: \chi_{i,l,f} \qquad \qquad \forall z,t,i \in J, f
+\end{align}
+
+\paragraph{Intermittent electricity generation}
+Electricity generation from intermittent sources wind (on-shore and off-shore), solar irradiation, and river runoff follows generation profiles $\phi_{z,t,n} \in [0,1]$ and is scaled according to corresponding installed and added capacity.
+\begin{align}
+r_{z,t,n} = \phi_{z,t,n} \: \left( \bar{r}^{0}_{z,n} + \bar{r}^{+}_{z,n} \right) \qquad \qquad \forall z,t,n
+\end{align}
+
+\paragraph{Electricity storages}
+Charging and discharging of storages is constrained by the storages' power capacity $\bar{s}^{in}_{z,k}$ and $\bar{s}^{out}_{z,k}$, respectively.
+Similarly, the total energy that can be stored is constrained by the storage technology's energy capacity $\bar{v}_{z,k}$.
+\begin{align}
+s^{out}_{z,t,k} \leq \bar{s}^{out}_{z,k} + \bar{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
+s^{in}_{z,t,k} \leq \bar{s}^{in}_{z,k} + \bar{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
+v_{z,t,k} \leq \bar{v}^{0}_{z,k} + \bar{v}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
+\end{align}
+Storage operation is subject to a storage balance, such that the current energy content must be equal to the previous period's energy content plus all energy flowing into the storage less all energy flowing out of the storage.
+\begin{align}
+v_{z,t,k} = v_{z,t-1,k} + \rho_{z,t,k} + \eta^{in}_{z,k} \: s^{in}_{z,t,k} - (\eta^{out}_{z,k})^{-1} \: s^{out}_{z,t,k} \qquad \qquad \forall z,t,k
+\end{align}
+Since the model can add storage power capacity and energy capacity independently, we require a storage to hold at least as much energy as it could store in or out in one hour.
+\begin{align}
+\bar{v}^{+}_{z,k} \geq \bar{s}^{+}_{z,k} \qquad \qquad \forall z,k
+\end{align}
+
+\paragraph{Electricity exchange}
+Implicitly, \emph{medea} assumes that there are no transmission constraints within market zones. 
+However, electricity exchange between market zones is subject to several constraints.
+
+First, exchange between market zones is constrained by available transfer capacities.
+Transfer capacities can be expanded at constant, specific investment cost (see equation (\ref{transmission_expansion_cost})), ruling out economies of scale in transmission investment.
+\begin{align}
+x_{z,zz,t} \leq \bar{x}^{0}_{z,zz} + \bar{x}^{+}_{z,zz} \qquad \qquad \forall z, zz, t \\
+x_{z,zz,t} \geq - \left( \bar{x}^{0}_{z,zz} + \bar{x}^{+}_{z,zz} \right) \qquad \qquad \forall z, zz, t
+\end{align}
+By definition, electricity net exports $x_{z,zz,t}$ from $z$ to $zz$ must equal electricity net imports of $zz$ from $z$.
+\begin{align}
+x_{z,zz,t} = -x_{zz,z,t} \qquad \qquad \forall z, zz, t
+\end{align}
+Added transmission capacities can be used in either direction.
+\begin{align}
+\bar{x}^{+}_{z,zz} = \bar{x}^{+}_{zz,z} \qquad \qquad \forall z, zz
+\end{align}
+Finally, electricity cannot flow between zones where there is no transmission infrastructure in place (including intra-zonal flows).
+\begin{align}
+x_{z,zz,t} = 0 \qquad \qquad \forall z \notin NTC, zz \notin NTC, t \\
+x_{zz,z,t} = 0 \qquad \qquad \forall z \notin NTC, zz \notin NTC, t
+\end{align}
+
+\paragraph{Decommissioning of thermal units}
+Any conventional unit can be decommission to save the quasi-fixed cost associated with keeping the plant potentially active.
+\begin{align}
+\bar{g}^{-}_{z,i} \leq |i_{z}| + \bar{g}^{+}_{z,i} \qquad \qquad \forall z,i
+\end{align}
+
+\paragraph{Ancillary services}
+Power systems require various system services for secure and reliable operation, such as balancing services or voltage support.
+Many of these system services are mostly or only supplied by operational generators.
+Thus, we approximate system service provision by a requirement on the minimal amount of spinning reserves operational at each hour.
+We assume that conventional (thermal) power plants as well as hydro power plants (even when in pumping mode) can supply ancillary services.
+\begin{align}
+g_{z,t,i,\text{el}} + r_{z,t,\text{ror}} + s^{out}_{z,t,k} + s^{in}_{z,t,k} \geq \sigma_{z} \qquad \qquad \forall z,t,i,k
+\end{align}
+
+\paragraph{Curtailment}
+Electricity generated from intermittent sources can be curtailed (disposed) without any further cost (apart from opportunity cost). 
+\begin{align}
+\Omega^{+}_{z,t} \leq \sum_{n} r_{z,t,n} \qquad \qquad \forall z, t
+\end{align}
+
 \end{document}

### 2019-08-01 doc/mathematical_description.tex
improved naming consistency

@@
 
+\title{Description of the power system model \emph{medea}}
+%\title{Mathematical description of the power system model \emph{medea}}
+\date{\today}
+\author{Sebastian Wehrle\thanks{Over recent years, \emph{medea} has grown into its current state thanks to inspiration and contributions from many friends and colleagues. 
+In particular, I'd like to thank Johannes Schmidt, who contributed significant parts of the model code, for getting \emph{medea} started and for keeping us running; Dieter Mayr and Stefan Höltinger for early contributions; Peter Regener for inspiration and endurance with all things programming.
+Current work on \emph{medea} is funded through grateful support from the European Research Council ("reFUEL" ERC-2017-STG 758149)} \\University of Natural Resources and Life Sciences, Vienna}
+
 \begin{document}
-\title{Mathematical description of the power system model \emph{medea}}
+\maketitle
 \section{Overview}
-\emph{medea} is a simple, stylized and parsimonious power system model.
+\emph{medea} is a simple, stylized and parsimonious model of interconnected power and heating systems in Western and Central Europe.
 It simulates investment in intermittent and conventional electricity and heat generation technologies as well as in cross-border electricity transmission capacities.
@@
 
+\newpage
+
 \section{Sets} \label{sets}
@@
 \hline \hline
-market zones & $z \in Z$ & \texttt{r} & \texttt{AT, DE} \\ \hline
+market zones & $z \in Z$ & \texttt{z} & \texttt{AT, DE} \\ \hline
 
@@
 
-energy products & $m \in M=\{\text{el}, \text{ht}\}$ & \texttt{prd} & \texttt{power, heat} \\ \hline \hline
+energy products & $m \in M$ & \texttt{prd} & \texttt{el, ht} \\ \hline \hline
 \end{tabular}
@@
 \begin{tabular}{l l l l}
-\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\
+\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\
 \hline \hline
-\makecell[l]{minimal conventional \\generation} & $\sigma_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(r)} & GW \\ \hline
-energy demand & $D_{z,t,p}$ & \texttt{CONSUMPTION(r,t,prd)} & GW \\ \hline
-power plant efficiency & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & $MW / MW_{th}$\\ \hline
-fuel emission intensity & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $t \ce{CO2}/MWh_{th}$\\ \hline
-\makecell[l]{feasible operating \\region} & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & MWh ? \\ \hline
-\makecell[l]{feasible operating \\region} & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & MWh ? \\ \hline
-\makecell[l]{intermittent \\generation profile} & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(r,t,tec\_itm)} & \% \\ \hline
-%\makecell[l]{exports to not \\modelled regions} & $X$ & \texttt{FLOW\_EXPORTS(r,t)} & GW \\ \hline
-%\makecell[l]{imports from not \\modelled regions} & $X$ & \texttt{FLOW\_IMPORTS(r,t)} & GW \\ \hline
-\makecell[l]{installed capacity \\of intermittent \\generators} & $\bar{r}^{0}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(r,tec\_itm)} & GW \\ \hline
-\makecell[l]{installed capacity \\of thermal generators} & $\bar{g}^{0}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(r,tec)} & GW \\ \hline
-\makecell[l]{capital cost of \\intermittent generators \\(specific, annuity)} & $\kappa_{z,n}$ & \texttt{INVESTCOST\_ITM(r,tec\_itm)} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{capital cost of \\thermal generators \\(specific, annuity)} & $\kappa_{z,i}$ & \texttt{INVESTCOST\_THERMAL(r,tec)} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{capital cost of \\storages - power \\(specific, annuity)} & $\kappa^{P}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'cost\_power')} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{capital cost of \\storages - energy \\(specific, annuity)} & $\kappa^{E}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'cost\_energy')} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{installed available \\transfer capacity} & $\bar{x}^{0}_{z,zz}$ & \texttt{ATC(r,rr)} & GW \\ \hline
-\makecell[l]{count 100 MW slices \\of same technology} & $|i_{z}|$ & \texttt{NUM(r,tec)} & . \\ \hline
+\makecell[l]{minimal \\conventional \\generation} & $\sigma_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(z)} & GW \\ \hline
+energy demand & $D_{z,t,p}$ & \texttt{CONSUMPTION(z,t,prd)} & GW \\ \hline
+\makecell[l]{power plant \\efficiency} & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & $MW / MW_{th}$\\ \hline
+\makecell[l]{fuel emission \\intensity} & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $t \ce{CO2}/MWh_{th}$\\ \hline
+\makecell[l]{inputs of \\feasible operating \\region} & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & MWh ? \\ \hline
+\makecell[l]{output tuples of \\feasible operating \\region} & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & MWh ? \\ \hline
+\makecell[l]{intermittent \\generation profile} & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(z,t,tec\_itm)} & \% \\ \hline
+%\makecell[l]{exports to not \\modelled regions} & $X$ & \texttt{FLOW\_EXPORTS(z,t)} & GW \\ \hline
+%\makecell[l]{imports from not \\modelled regions} & $X$ & \texttt{FLOW\_IMPORTS(z,t)} & GW \\ \hline
+\makecell[l]{installed capacity \\of intermittent \\generators} & $\bar{r}^{0}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(z,tec\_itm)} & GW \\ \hline
+\makecell[l]{installed capacity \\of thermal generators} & $\bar{g}^{0}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(z,tec)} & GW \\ \hline
+\makecell[l]{capital cost of \\intermittent generators \\(specific, annuity)} & $\kappa_{z,n}$ & \texttt{INVESTCOST\_ITM(z,tec\_itm)} & $\frac{mn EUR}{GW}$ \\ \hline
+\makecell[l]{capital cost of \\thermal generators \\(specific, annuity)} & $\kappa_{z,i}$ & \texttt{INVESTCOST\_THERMAL(z,tec)} & $\frac{mn EUR}{GW}$ \\ \hline
+\makecell[l]{capital cost of \\storages - power \\(specific, annuity)} & $\kappa^{P}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_power')}} & $\frac{mn EUR}{GW}$ \\ \hline
+\makecell[l]{capital cost of \\storages - energy \\(specific, annuity)} & $\kappa^{E}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_energy')}} & $\frac{mn EUR}{GW}$ \\ \hline
+\makecell[l]{installed available \\transfer capacity} & $\bar{x}^{0}_{z,zz}$ & \texttt{ATC(z,zz)} & GW \\ \hline
+\makecell[l]{count 100 MW slices \\of same technology} & $|i_{z}|$ & \texttt{NUM(z,tec)} & . \\ \hline
 quasi-fixed O\&M cost & $o^{q}_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & EUR \\ \hline
 variable O\&M cost & $o^{v}_{i}$ & \texttt{OM\_VAR\_COST(tec)} & EUR / MWh \\ \hline
-% day-ahead price & $p_{el}$ & \texttt{PRICE\_DA(t,r)} & EUR / MWh \\ \hline
-\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_EUA(t,r)} & EUR / t CO2 \\ \hline
-fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,r,f)} & EUR / MWh \\ \hline
-reservoir inflows & $\rho_{z,t,k}$ & \texttt{RESERVOIR\_INFLOWS(r,t,tec\_strg)} & MW \\ \hline
-max power out & $\bar{s}^{out}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'power\_out')} & GW \\ \hline
-max power in & $\bar{s}^{in}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'power\_in')} & GW \\ \hline
-max energy stored & $\bar{v}^{0}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'energy\_max')} & . \\ \hline
-efficiency power out & $\eta^{out}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'efficiency\_out')} & . \\ \hline
-efficiency power in & $\eta^{in}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'efficiency\_in')} & . \\ \hline
-value of lost load & $\mu$ & \texttt{not set yet} & EUR \\ \hline
+\hline
+\end{tabular}
+
+\begin{tabular}{l l l l}
+\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\
+\hline \hline
+% day-ahead price & $p_{el}$ & \texttt{PRICE\_DA(t,z)} & EUR / MWh \\ \hline
+\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_EUA(t,z)} & EUR / t CO2 \\ \hline
+fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & EUR / MWh \\ \hline
+reservoir inflows & $\rho_{z,t,k}$ & \makecell[l]{\texttt{RESERVOIR\_INFLOWS} \\ \texttt{(z,t,tec\_strg)}} & MW \\ \hline
+max power out & $\bar{s}^{out}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'power\_out')}} & GW \\ \hline
+max power in & $\bar{s}^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'power\_in')}} & GW \\ \hline
+max energy stored & $\bar{v}^{0}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'energy\_max')}} & . \\ \hline
+efficiency power out & $\eta^{out}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_out')}} & . \\ \hline
+efficiency power in & $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_in')}} & . \\ \hline
+value of lost load & $\mu$ & \texttt{VALUE\_NSE(z)} & EUR per MWh \\ \hline
 \hline
@@
 zonal system cost & $C_{z}$ & \texttt{cost(r)} & EUR \\ \hline
-emission cost & $C^{e}_{z,t,i}$ & \texttt{cost\_emission(r,t,tec)} & EUR \\ \hline
-fuel cost & $C^{f}_{z,t,i}$ & \texttt{cost\_fuel(r,t,tec)} & EUR \\ \hline
-total o\&m cost & $C^{om}_{z,i}$ & \texttt{cost\_om(r,tec)} & EUR \\ \hline
-\makecell[l]{capital cost of \\generators} & $C_{z}^{inv,i}$ & \texttt{cost\_invgen(r)} & EUR \\ \hline
-\makecell[l]{capital cost of \\storages} & $C_{z}^{inv,k}$ & \texttt{cost\_invstrg(r)} & EUR \\ \hline
-\makecell[l]{capital cost of \\interconnectors} & $C_{z}^{inv,ic}$ & \texttt{cost\_gridexpansion(r)} & EUR \\ \hline
+emission cost & $C^{e}_{z,t,i}$ & \texttt{cost\_emission(z,t,tec)} & EUR \\ \hline
+fuel cost & $C^{f}_{z,t,i}$ & \texttt{cost\_fuel(z,t,tec)} & EUR \\ \hline
+total o\&m cost & $C^{om}_{z,i}$ & \texttt{cost\_om(z,tec)} & EUR \\ \hline
+\makecell[l]{capital cost of \\generators} & $C_{z}^{inv,i}$ & \texttt{cost\_invgen(z)} & EUR \\ \hline
+\makecell[l]{capital cost of \\storages} & $C_{z}^{inv,k}$ & \texttt{cost\_invstrg(z)} & EUR \\ \hline
+\makecell[l]{capital cost of \\interconnectors} & $C_{z}^{inv,ic}$ & \texttt{cost\_gridexpansion(z)} & EUR \\ \hline
 % investment and decommissioning variables
-\makecell[l]{added capacity of \\intermittents} & $\bar{r}^{+}_{z,n}$ & \texttt{invest\_res(r,tec\_itm)} & GW \\ \hline
-\makecell[l]{added capacity of \\conventionals} & $\bar{g}^{+}_{z,i}$ & \texttt{invest\_thermal(r,tec)} & GW \\ \hline
-\makecell[l]{added storage \\capacity (power)} & $\bar{s}^{+}_{z,k}$ & \texttt{invest\_storage\_power(r,tec\_strg)} & EUR \\ \hline
-\makecell[l]{added storage \\capacity (energy)} & $\bar{v}^{+}_{z,k}$ & \texttt{invest\_storage\_energy(r,tec\_strg)} & EUR \\ \hline
-\makecell[l]{added transmission \\capacity} & $\bar{x}^{+}_{z,zz}$ & \texttt{invest\_atc(r,rr)} & EUR \\ \hline
-\makecell[l]{decommissioned \\capacity of \\conventionals} & $\bar{g}^{-}_{z,i}$ & \texttt{decommission(r,tec)} & GW \\ \hline
+\makecell[l]{added capacity of \\intermittents} & $\bar{r}^{+}_{z,n}$ & \texttt{invest\_res(z,tec\_itm)} & GW \\ \hline
+\makecell[l]{added capacity of \\conventionals} & $\bar{g}^{+}_{z,i}$ & \texttt{invest\_thermal(z,tec)} & GW \\ \hline
+\makecell[l]{added storage \\capacity (power)} & $\bar{s}^{+}_{z,k}$ & \texttt{invest\_storage\_power(z,tec\_strg)} & EUR \\ \hline
+\makecell[l]{added storage \\capacity (energy)} & $\bar{v}^{+}_{z,k}$ & \texttt{invest\_storage\_energy(z,tec\_strg)} & EUR \\ \hline
+\makecell[l]{added transmission \\capacity} & $\bar{x}^{+}_{z,zz}$ & \texttt{invest\_atc(z,zz)} & EUR \\ \hline
+\makecell[l]{decommissioned \\capacity of \\conventionals} & $\bar{g}^{-}_{z,i}$ & \texttt{decommission(z,tec)} & GW \\ \hline
 % dispatch variables
-\makecell[l]{energy generated by \\conventionals} & $g_{z,t,i,p}$ & \texttt{q\_gen(r,t,tec,prd)} & GW \\ \hline
-\makecell[l]{electricity generated \\by intermittents} & $r_{z,t,n}$ & \texttt{q\_itm(r,t,tec\_itm)} & GW \\ \hline
-\makecell[l]{operating region \\weight} & $w_{z,t,i,l}$ & \texttt{cc\_weights(r,t,tec,l)} & . \\ \hline
-\makecell[l]{fuel burn for energy \\generation} & $b_{z,t,i,f}$ & \texttt{q\_fueluse(r,t,tec,f)} & GW \\ \hline
-energy stored in & $s^{in}_{z,t,k}$ & \texttt{q\_store\_in(r,t,tec\_strg)} & GW \\ \hline
-energy stored out & $s^{out}_{z,t,k}$ & \texttt{q\_store\_out(r,t,tec\_strg)} & GW \\ \hline
-\makecell[l]{storage energy \\content} & $v_{z,t,k}$ & \texttt{storage\_level(r,t,tec\_strg)} & GWh \\ \hline
-\makecell[l]{electricity net \\export} & $x_{z,zz,t}$ & \texttt{flow(r,rr,t)} & GW \\ \hline
-curtailed energy & $\Omega^{+}_{z,t}$ & \texttt{q\_curtail(r,t)} & GW \\ \hline
-non-served energy & $\Omega^{-}_{z,t,p}$ & \texttt{q\_nonserved(r,t,prd)} & GW \\ \hline
+\makecell[l]{energy generated by \\conventionals} & $g_{z,t,i,p}$ & \texttt{q\_gen(z,t,tec,prd)} & GW \\ \hline
+\makecell[l]{electricity generated \\by intermittents} & $r_{z,t,n}$ & \texttt{q\_itm(z,t,tec\_itm)} & GW \\ \hline
+\makecell[l]{operating region \\weight} & $w_{z,t,i,l}$ & \texttt{cc\_weights(z,t,tec,l)} & . \\ \hline
+\makecell[l]{fuel burn for energy \\generation} & $b_{z,t,i,f}$ & \texttt{q\_fueluse(z,t,tec,f)} & GW \\ \hline
+energy stored in & $s^{in}_{z,t,k}$ & \texttt{q\_store\_in(z,t,tec\_strg)} & GW \\ \hline
+energy stored out & $s^{out}_{z,t,k}$ & \texttt{q\_store\_out(z,t,tec\_strg)} & GW \\ \hline
+\makecell[l]{storage energy \\content} & $v_{z,t,k}$ & \texttt{storage\_level(z,t,tec\_strg)} & GWh \\ \hline
+\makecell[l]{electricity net \\export} & $x_{z,zz,t}$ & \texttt{flow(z,zz,t)} & GW \\ \hline
+curtailed energy & $\Omega^{+}_{z,t}$ & \texttt{q\_curtail(z,t)} & GW \\ \hline
+non-served energy & $\Omega^{-}_{z,t,p}$ & \texttt{q\_nonserved(z,t,prd)} & GW \\ \hline
 % other variables
-\ce{CO2} emissions & $e_{z}$ & \texttt{emissions(r)} & t \ce{CO2} \\ \hline
-
+\ce{CO2} emissions & $e_{z}$ & \texttt{emissions(z)} & t \ce{CO2} \\ \hline
+\hline
 \end{tabular}
@@
 \section{Mathematical description} \label{mathmodel}
+
 \paragraph{Model objective}
-\emph{medea} minimizes total system cost, i.e. the total cost of generating electricity and heat from technologies and capacities adequate to meet demand.
+\emph{medea} minimizes total system cost $C$, i.e.
+the total cost of generating electricity and heat from technologies and capacities adequate to meet demand, over a large number  of decision variables, essentially representing investment and dispatch decisions in each market zone $z$ of the modelled energy systems.
 \begin{align}
@@
 \end{align}
-System costs consist of fuel cost $C^{f}_{z,t,i}$, emission cost $C^{e}_{z,t,i}$, operation and maintenance cost $C^{om}_{z,i}$, capital costs of investment in generation, storage and transmission equipment, and the cost of non-served load that accrues when demand is not met, e.g.
+Zonal system costs $C_{z}$ are the sum of fuel cost $C^{f}_{z,t,i}$, emission cost $C^{e}_{z,t,i}$, operation and maintenance cost $C^{om}_{z,i}$, capital costs of investment in conventional and intermittent generation ($C^{inv,i}_{z}$, $C^{inv,n}_{z}$), storage ($C^{inv,k}_{z}$) and transmission ($C^{inv,ic}_{z}$) equipment, and the cost of non-served load ($C^{nse}_{z}$) that accrues when demand is not met, e.g.
 when there is a power outage.
 \begin{align}
-\begin{split}
-C_{z} = \sum_{t,i} & C^{f}_{z,t,i} + \sum_{t,i} C^{e}_{z,t,i} + \sum_{i} C^{om}_{z,i} + C^{inv,i}_{z} + \\
-& C^{inv,n}_{z} + C^{inv,k}_{z} + C^{inv,ic}_{z} + C^{nse}_{z} \qquad \qquad \forall z
-\end{split}
+%\begin{split}
+C_{z} = \sum_{t,i}  C^{f}_{z,t,i} + \sum_{t,i} C^{e}_{z,t,i} + \sum_{i} C^{om}_{z,i} + C^{inv,i}_{z} + 
+ C^{inv,n}_{z} + C^{inv,k}_{z} + C^{inv,ic}_{z} + C^{nse}_{z} \qquad \qquad \forall z
+%\end{split}
 \end{align}
+The components of zonal system costs are calculated as given in equations (\ref{fuel_cost}) to (\ref{lost_load_cost}).
 Upper-case $C$ represent total cost, while $\kappa$ denotes specific, annualized capital cost of technology investment. Prices for fuels and \ce{CO2} are denoted by uppercase $P$.
-
-The components of total cost are calculated according to equations \ref{fuel_cost} to \ref{lost_load_cost}.
 \begin{align}
@@
 \end{align}
+In line with \cite{citation}, we estimate the specific cost of non-served load $\mu$ at 12500 EUR per MWh.
 
@@
 In each hour, the markets for electricity and heat have to clear.
-Equation \ref{market_clearing_el} ensures that the total supply from conventional and intermittent sources, and storages equals total electricity demand plus net exports, electricity stored and used for heat generation.
-Likewise, equation \ref{market_clearing_ht} clears the heat market by equating heat generation to heat demand.
-
+Equation (\ref{market_clearing_el}) ensures that the total supply from conventional and intermittent sources, and storages equals total electricity demand plus net exports, electricity stored and used for heat generation.
+Likewise, equation (\ref{market_clearing_ht}) clears the heat market by equating heat generation to heat demand.
 \begin{align}
@@
 \end{align}
-
-In its current formulation, the model represents energy-only electricity and heat markets without capacity payments.
-Thus, the marginals of the market clearing equations ($\partial C / \partial D_{z,t,m}$) can be interpreted as the zonal prices for electricity and heat, respectively.
+\emph{medea} can be thought of as representing energy-only electricity and heat markets without capacity payments.
+Then, the marginals of the market clearing equations (\ref{market_clearing_el}) and (\ref{market_clearing_ht}), $\partial C / \partial D_{z,t,m}$, can be interpreted as the zonal prices for electricity and heat, respectively.
 
@@
 The amount of electricity required to generate a given amount of heat depends on the unit's efficiency $\eta$.
-For heat pumps, $\eta$ is assumed to be $3$.
+For heat pumps, the annual COP $\eta$ is assumed to be $3$.
 \begin{align}
@@
 Co-generation units jointly generate heat and electricity.
-All feasible combinations of heat and electricity generation along with the corresponding fuel requirement are reflected in so-called `feasible operating region'.
+All feasible combinations of heat and electricity generation along with the corresponding fuel requirement are reflected in so-called `feasible operating regions'.
 The elements $l \in L$ span up a three-dimensional, convex feasible operating region for each co-generation technology. 
-The weights $w$ form a convex combination of the corners $l$ scaled to the available installed capacity of each co-generation technology.
+The weights $w$ form a convex combination of the corners $l$, which are scaled to the available installed capacity of each co-generation technology.
 Heat and electricity output along with the corresponding fuel requirement is then set according to the chosen weights.
@@
 \end{align}
-Since the model can add storage power capacity and energy capacity independently, we require a storage to hold at least as much energy as it could store in or out in one hour.
+Since the model can add storage power capacity and energy capacity independently, we require a storage to hold at least as much energy as it could store in (or out) in one hour.
 \begin{align}
@@
 First, exchange between market zones is constrained by available transfer capacities.
-Transfer capacities can be expanded at constant, specific investment cost (see equation (\ref{transmission_expansion_cost})), ruling out economies of scale in transmission investment.
+Transfer capacities can be expanded at constant, specific investment cost (see equation (\ref{transmission_expansion_cost})).
+This rules out economies of scale in transmission investment that might arise in interconnected, meshed grids.
 \begin{align}
@@
 \paragraph{Decommissioning of thermal units}
-Any conventional unit can be decommission to save the quasi-fixed cost associated with keeping the plant potentially active.
+Keeping a plant available for dispatch gives rise to quasi-fixed operation and maintenance costs. 
+Such cost can be avoided by decommissioning an energy generator. This is modelled as a reduction in generation capacity, which cannot exceed installed capacity.
 \begin{align}
@@
 \paragraph{Ancillary services}
-Power systems require various system services for secure and reliable operation, such as balancing services or voltage support.
-Many of these system services are mostly or only supplied by operational generators.
-Thus, we approximate system service provision by a requirement on the minimal amount of spinning reserves operational at each hour.
-We assume that conventional (thermal) power plants as well as hydro power plants (even when in pumping mode) can supply ancillary services.
+Power systems require various system services for secure and reliable operation, such as balancing services or voltage support through the provision of reactive power.
+Such system services can only be supplied by operational generators.
+Thus, we approximate system service provision by a requirement on the minimal amount of spinning reserves operating at each hour.
+We assume that ancillary services are supplied by conventional (thermal) power plants, hydro power plants, and storages.
 \begin{align}
@@
 \paragraph{Curtailment}
-Electricity generated from intermittent sources can be curtailed (disposed) without any further cost (apart from opportunity cost). 
+Electricity generated from intermittent sources can be curtailed (disposed of) without any further cost (apart from implicit opportunity cost). 
 \begin{align}

### 2019-08-01 applications/pass_through/results_pass_through.py

@@ -39,7 +39,7 @@ for scenario in capacity_scenario_name:
         result_dict = {key: gdx2df(db_output, key, value[0], value[1])
                        for key, value in read_variables.items()}
 
-        df = pd.DataFrame(columns=cfg.regions)
+        df = pd.DataFrame(columns=cfg.zones)
         for key, value in result_dict.items():
             if value.index.any() == 'Value':
                 value.index = [key]

### 2019-08-01 applications/pass_through/run_pass_through.py

@@ -7,12 +7,13 @@ use medea to estimate emission cost pass-through to electricity prices in the cu
 import os
 import subprocess
 from shutil import copyfile
+
 import pandas as pd
 from gams import *
 
 import config as cfg
-from medea.gams_io import reset_parameter, gdx2df, df2gdx
 from applications.pass_through.settings_pass_through import *
+from medea.gams_io import reset_parameter, gdx2df, df2gdx
 
 # --------------------------------------------------------------------------- #
 # %% initialize GAMS, GAMS workspace and load model data
@@ -78,7 +79,7 @@ EUA_SCENARIO = df2gdx(db_input, pd.DataFrame(data=[0]), 'EUA_SCENARIO', 'par', 0
 for cap_scenario in range(0, 3, 1):   # capacity_scenarios:
     # conventional capacities
     df_num_mod = df_num
-    for reg in cfg.regions:
+    for reg in cfg.zones:
         for gen in df_num.loc[reg].index:
             df_num_mod.loc[idx[reg, gen]] = (df_num.loc[idx[reg, gen]] *
                                              capacity_scenarios[reg][gen.split('_')[0]][cap_scenario]).round(0)
@@ -87,7 +88,7 @@ for cap_scenario in range(0, 3, 1):   # capacity_scenarios:
 
     # renewable capacities
     df_capitm_mod = df_capitm
-    for reg in cfg.regions:
+    for reg in cfg.zones:
         for itm in itm_dict:
             if itm != 'ror':
                 df_capitm_mod.loc[idx[reg, itm], :] = capacity_scenarios[reg][itm][cap_scenario]

### 2019-08-01 applications/wind_vs_pv/results_wind_vs_pv.py

@@ -42,7 +42,7 @@ for peua in eua_range:
     result_dict = {key: gdx2df(db_output, key, value[0], value[1])
                    for key, value in read_variables.items()}
 
-    df = pd.DataFrame(columns=cfg.regions)
+    df = pd.DataFrame(columns=cfg.zones)
     for key, value in result_dict.items():
         if value.index.any() == 'Value':
             value.index = [key]
@@ -68,7 +68,7 @@ for peua in eua_range:
             result_dict = {key: gdx2df(db_output, key, value[0], value[1])
                            for key, value in read_variables.items()}
 
-            df = pd.DataFrame(columns=cfg.regions)
+            df = pd.DataFrame(columns=cfg.zones)
             for key, value in result_dict.items():
                 if value.index.any() == 'Value':
                     value.index = [key]

### 2019-08-01 applications/wind_vs_pv/run_wind_vs_pv.py

@@ -64,7 +64,7 @@ reset_parameter(db_input, 'FEASIBLE_INPUT', df_fuelreq_mod)
 
 # conventional capacities
 df_num_mod = df_num
-for reg in cfg.regions:
+for reg in cfg.zones:
     for gen in df_num.loc[reg].index:
         df_num_mod.loc[idx[reg, gen]] = (df_num.loc[idx[reg, gen]] *
                                          scenario_2030[reg][gen.split('_')[0]][0]).round(0)
@@ -72,7 +72,7 @@ reset_parameter(db_input, 'NUM', df_num_mod)
 
 # intermittent capacities
 df_capitm_mod = df_capitm
-for reg in cfg.regions:
+for reg in cfg.zones:
     for itm in itm_dict:
         if itm != 'ror':
             df_capitm_mod.loc[idx[reg, itm], :] = scenario_2030[reg][itm][0]

### 2019-08-01 config.py

@@ -16,7 +16,7 @@ iter_num = 1
 # set iteration offset, i.e. number of hours cut off at each iteration-end to avoid last-round effects
 iter_offset = 72
 # covered regions - currently available: 'AT', 'DE'
-regions = ['AT', 'DE']
+zones = ['AT', 'DE']
 # long-run (invest) vs short-run (no invest) model version
 long_run = True
 

### 2019-08-01 legacy/scenario_results.py

@@ -152,7 +152,7 @@ reset_parameter(db_input, 'FEASIBLE_INPUT', df_fuelreq_mod)
 
 # available capacity / number of plants
 df_num_mod = df_num.copy()
-for reg in cfg.regions:
+for reg in cfg.zones:
     for fu in capcy[reg].keys():
         df_num_mod[df_num_mod.index.get_level_values(0).str.contains(reg) &
                (df_num_mod.index.get_level_values(1).str.contains(fu))] = \

### 2019-08-01 legacy/solve.py

@@ -230,7 +230,7 @@ def solve(output_folder, output_name, scenario_iteration, gen_target):
         hgbf_list = []
         fubf_list = []
         emission_list = []
-        for reg in cfg.regions:
+        for reg in cfg.zones:
             # power generation by fuel
             df_pgbf = result_dict['q_gen'].xs('Power', level=2, axis=1).xs(reg, level=0, axis=1).groupby(by=df_fuelmap['f'], axis=1).sum()
             df_pgbf.columns = pd.MultiIndex.from_product([[reg], df_pgbf.columns])
@@ -250,7 +250,7 @@ def solve(output_folder, output_name, scenario_iteration, gen_target):
         htgen_by_fuel = pd.concat(hgbf_list, axis=1).reindex(cols, level=1, axis=1)
         burn_by_fuel = pd.concat(fubf_list, axis=1).reindex(cols, level=1, axis=1)
 
-        for reg in cfg.regions:
+        for reg in cfg.zones:
             df_emis = df_emf['Value'] * burn_by_fuel.xs(reg, axis=1)
             df_emis.columns = pd.MultiIndex.from_product([[reg], df_emis.columns])
             emission_list.append(df_emis)

### 2019-08-01 medea/data/compile_timeseries.py

@@ -60,13 +60,13 @@ ts_medea['AT-pv-profile'] = ts_medea['AT-pv-generation'] / ts_medea['AT-pv-capac
 ts_load_ht = pd.read_csv(os.path.join(cfg.folder, 'data', 'processed', 'heat_hourly_consumption.csv'),
                          index_col=[0], header=[0, 1])
 ts_load_ht.index = pd.DatetimeIndex(ts_load_ht.index).tz_localize('utc')
-for reg in cfg.regions:
+for reg in cfg.zones:
     ts_medea[f'{reg}-heat-load'] = ts_load_ht.loc[:, reg].sum(axis=1)
 
 # read hydro data
 ts_hydro_ror = pd.read_csv(os.path.join(cfg.folder, 'data', 'processed', 'generation_hydro.csv'), index_col=[0])
 ts_hydro_ror.index = pd.DatetimeIndex(ts_hydro_ror.index).tz_localize('utc')
-for reg in cfg.regions:
+for reg in cfg.zones:
     ts_medea[f'{reg}-ror-capacity'] = itm_capacities[(reg, 'ror')]
     ts_medea[f'{reg}-ror-capacity'] = ts_medea[f'{reg}-ror-capacity'].interpolate()
     ts_medea[f'{reg}-ror-profile'] = ts_hydro_ror[f'ror_{reg}'] / 1000 / ts_medea[f'{reg}-ror-capacity']
@@ -74,7 +74,7 @@ for reg in cfg.regions:
 # commercial flows
 ts_flows = pd.read_csv(os.path.join(cfg.folder, 'data', 'processed', 'commercial_flows.csv'), index_col=[0])
 ts_flows.index = pd.DatetimeIndex(ts_flows.index).tz_localize('utc')
-for reg in cfg.regions:
+for reg in cfg.zones:
     ts_medea[f'{reg}-imports-flow'] = ts_flows.loc[:, f'imp_{reg}'] / 1000
     ts_medea[f'{reg}-exports-flow'] = ts_flows.loc[:, f'exp_{reg}'] / 1000
 
@@ -84,17 +84,17 @@ df_hydro_fill.index = pd.DatetimeIndex(df_hydro_fill.index).tz_localize('utc')
 ts_hydro_fill = pd.DataFrame(
     index=pd.date_range(pd.datetime(df_hydro_fill.head(1).index.year[0], 1, 1, 0, 0),
                         pd.datetime(df_hydro_fill.tail(1).index.year[0], 12, 31, 23, 0),
[…]

### 2019-08-01 medea/instantiate.py

@@ -1,15 +1,18 @@
 import logging
 import os
+import sys
 
 import pandas as pd
 from gams import *
 
+sys.path.append(["D:\\git_repos\\medea", "D:\\git_repos\\medea\\medea"])
+
 import config as cfg
 from medea.gams_io import df2gdx
-from medea.prepare import df_fuel, df_lim, df_prd, df_props, df_regions, df_tec_itm, df_tec_strg, \
+from medea.prepare import df_fuel, df_lim, df_prd, df_props, df_zones, df_tec_itm, df_tec_strg, \
     df_time, data_technology, data_atc, tec_props, df_efficiency, df_emission_intensity, df_itm_invest, df_itm_cap, \
     df_ancil, lim_invest_itm, lim_invest_thermal, lim_invest_atc, lim_invest_storage, df_feasops, storage_clusters, \
-    ts_regional, ts_price, ts_inflows
+    ts_zonal, ts_price, ts_inflows
 
 # TODO: Add energy stored in hydro reservoirs - STORAGE_LEVEL
 
@@ -27,7 +30,7 @@ f_set = df2gdx(db, df_fuel, 'f', 'set', [])
 l_set = df2gdx(db, df_lim, 'l', 'set', [])
 prd_set = df2gdx(db, df_prd, 'prd', 'set', [])
 prop_set = df2gdx(db, df_props, 'props', 'set', [])
-r_set = df2gdx(db, df_regions, 'r', 'set', [])
+z_set = df2gdx(db, df_zones, 'z', 'set', [])
 tec_set = df2gdx(db, data_technology['medea_type'], 'tec', 'set', [])
 tec_chp_set = df2gdx(db, pd.DataFrame.from_dict({x: True for x in [y for y in data_technology.index if 'chp' in y]},
                                                 orient='index'), 'tec_chp', 'set', [])
@@ -42,51 +45,52 @@ logging.info('medea sets instantiated')
 # --------------------------------------------------------------------------- #
 # %% instantiate static PARAMETERS
 # --------------------------------------------------------------------------- #
[…]

### 2019-08-01 medea/prepare.py

@@ -21,10 +21,10 @@ fuel_set = {'Nuclear': 10, 'Lignite': 20, 'Coal': 30, 'Gas': 40, 'Oil': 50, 'Hyd
             'Biomass': 70, 'Solar': 80, 'Wind': 90, 'Power': 100, 'Heat': 110}
 df_fuel = pd.DataFrame(fuel_set.values(), fuel_set.keys(), ['Value'])
 df_lim = pd.DataFrame(data=True, index=[f'l{x}' for x in range(1, 6)], columns=['Value'])
-df_prd = pd.DataFrame(data=True, index=['power', 'heat'], columns=['Value'])
+df_prd = pd.DataFrame(data=True, index=['el', 'ht'], columns=['Value'])
 df_props = pd.DataFrame(data=True, index=['power_out', 'power_in', 'energy_max', 'efficiency_out', 'efficiency_in',
                                           'cost_power', 'cost_energy'], columns=['Value'])
-df_regions = pd.DataFrame(data=True, index=cfg.regions, columns=['Value'])
+df_zones = pd.DataFrame(data=True, index=cfg.zones, columns=['Value'])
 df_tec_itm = pd.DataFrame(data=True, index=['pv', 'ror', 'wind_on', 'wind_off'], columns=['Value'])
 df_tec_strg = pd.DataFrame(data=True, index=['psp_day', 'psp_week', 'psp_season', 'res_day', 'res_week', 'res_season',
                                              'battery'], columns=['Value'])
@@ -53,8 +53,8 @@ efficiency_electric = {'nuc': 0.34, 'lig_stm': 0.33, 'lig_stm_chp': 0.33, 'lig_b
 df_efficiency = pd.DataFrame.from_dict(efficiency_electric, orient='index', columns=['l1'])
 dict_name2fuel = {'nuc': 'Nuclear', 'lig': 'Lignite', 'coal': 'Coal', 'ng': 'Gas', 'oil': 'Oil', 'bio': 'Biomass',
                   'heatpump': 'Power'}
-df_efficiency['product'] = 'power'
-df_efficiency.loc[df_efficiency.index.str.contains('pth'), 'product'] = 'heat'
+df_efficiency['product'] = 'el'
+df_efficiency.loc[df_efficiency.index.str.contains('pth'), 'product'] = 'ht'
 df_efficiency['fuel'] = df_efficiency.index.to_series().str.split('_').str.get(0).replace(dict_name2fuel)
 df_efficiency.set_index(['product', 'fuel'], append=True, inplace=True)
[…]

### 2019-08-01 medea/results_template.py

@@ -1,12 +1,14 @@
 import os
-import numpy as np
-import pandas as pd
 
+import pandas as pd
+# import project settings, USER MUST REPLACE {project_name} BY CORRESPONDING PROJECT NAME
+from applications.
 from gams import *
+
 import config as cfg
 from medea.gams_io import gdx2df
-# import project settings, USER MUST REPLACE {project_name} BY CORRESPONDING PROJECT NAME
-from applications.{project_name}.settings_{project_name} import *
+
+{project_name}.settings_{project_name} import *
 
 
 # -------------------------------------------------------------------------------------------------------------------- #
@@ -37,7 +39,7 @@ for price_co2 in range_co2price:
     # write data to disk
 
 
-    df_xxx = pd.DataFrame(columns=cfg.regions)
+    df_xxx = pd.DataFrame(columns=cfg.zones)
     for key, df in dict_results.items():
         if df.index.any() == 'Value':
             df.index = [key]

### 2019-08-01 doc/mathematical_description.tex
improved model description

@@
 %\usepackage{courier} 
-\usepackage[inner=2.5cm,outer=2.5cm]{geometry}
+\usepackage[inner=2.5cm,outer=2.5cm, bottom=3cm]{geometry}
 
@@
 \author{Sebastian Wehrle\thanks{Over recent years, \emph{medea} has grown into its current state thanks to inspiration and contributions from many friends and colleagues. 
-In particular, I'd like to thank Johannes Schmidt, who contributed significant parts of the model code, for getting \emph{medea} started and for keeping us running; Dieter Mayr and Stefan Höltinger for early contributions; Peter Regener for inspiration and endurance with all things programming.
+In particular, I'd like to thank Johannes Schmidt, who contributed substantial parts of the initial model code, for getting \emph{medea} started and for keeping us running; Dieter Mayr and Stefan Höltinger for early contributions; Peter Regener for inspiration and endurance with all things programming.
 Current work on \emph{medea} is funded through grateful support from the European Research Council ("reFUEL" ERC-2017-STG 758149)} \\University of Natural Resources and Life Sciences, Vienna}

### 2019-08-16 doc/mathematical_description.tex
updated mathematical model description * replaces capacity slices with actual capacities * updates calculation of grid expansion cost

@@
 \usepackage{makecell}
-\usepackage{eurosym} 
+\usepackage[gen]{eurosym} 
 \usepackage[version=4]{mhchem}
@@
 energy demand & $D_{z,t,p}$ & \texttt{CONSUMPTION(z,t,prd)} & GW \\ \hline
-\makecell[l]{power plant \\efficiency} & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & $MW / MW_{th}$\\ \hline
-\makecell[l]{fuel emission \\intensity} & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $t \ce{CO2}/MWh_{th}$\\ \hline
-\makecell[l]{inputs of \\feasible operating \\region} & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & MWh ? \\ \hline
-\makecell[l]{output tuples of \\feasible operating \\region} & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & MWh ? \\ \hline
-\makecell[l]{intermittent \\generation profile} & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(z,t,tec\_itm)} & \% \\ \hline
+\makecell[l]{power plant \\efficiency} & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & MWh / MWh\\ \hline
+\makecell[l]{fuel emission \\intensity} & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $\text{t}_{\ce{CO2}}$ / MWh\\ \hline
+\makecell[l]{inputs of \\feasible operating \\region} & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & $[0,1]$ \\ \hline
+\makecell[l]{output tuples of \\feasible operating \\region} & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & $[0,1]$ \\ \hline
+\makecell[l]{intermittent \\generation profile} & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(z,t,tec\_itm)} & $[0,1]$ \\ \hline
 %\makecell[l]{exports to not \\modelled regions} & $X$ & \texttt{FLOW\_EXPORTS(z,t)} & GW \\ \hline
@@
 \makecell[l]{installed capacity \\of thermal generators} & $\bar{g}^{0}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(z,tec)} & GW \\ \hline
-\makecell[l]{capital cost of \\intermittent generators \\(specific, annuity)} & $\kappa_{z,n}$ & \texttt{INVESTCOST\_ITM(z,tec\_itm)} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{capital cost of \\thermal generators \\(specific, annuity)} & $\kappa_{z,i}$ & \texttt{INVESTCOST\_THERMAL(z,tec)} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{capital cost of \\storages - power \\(specific, annuity)} & $\kappa^{P}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_power')}} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{capital cost of \\storages - energy \\(specific, annuity)} & $\kappa^{E}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_energy')}} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{installed available \\transfer capacity} & $\bar{x}^{0}_{z,zz}$ & \texttt{ATC(z,zz)} & GW \\ \hline
-\makecell[l]{count 100 MW slices \\of same technology} & $|i_{z}|$ & \texttt{NUM(z,tec)} & . \\ \hline
-quasi-fixed O\&M cost & $o^{q}_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & EUR \\ \hline
-variable O\&M cost & $o^{v}_{i}$ & \texttt{OM\_VAR\_COST(tec)} & EUR / MWh \\ \hline
+\makecell[l]{capital cost of \\intermittent generators \\(specific, annuity)} & $\kappa_{z,n}$ & \texttt{INVESTCOST\_ITM(z,tec\_itm)} & k\EUR / GW \\ \hline
+\makecell[l]{capital cost of \\thermal generators \\(specific, annuity)} & $\kappa_{z,i}$ & \texttt{INVESTCOST\_THERMAL(z,tec)} & k\EUR / GW \\ \hline
+\makecell[l]{capital cost of \\storages - power \\(specific, annuity)} & $\kappa^{P}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_power')}} & k\EUR / GW \\ \hline
+\makecell[l]{capital cost of \\storages - energy \\(specific, annuity)} & $\kappa^{E}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_energy')}} & k\EUR / GW \\ \hline
+\makecell[l]{installed available \\transmission capacity} & $\bar{x}^{0}_{z,zz}$ & \texttt{ATC(z,zz)} & GW \\ \hline
+\makecell[l]{capital cost of \\transmission capacity} & $\kappa^{ic}_{z}$ & \texttt{INVESTCOST\_ATC} & k\EUR / GW \\ \hline
+\makecell[l]{distance between \\countries} & $\delta_{z,zz}$ & \texttt{KM(z,zz)} & km \\ \hline
+%\makecell[l]{count 100 MW slices \\of same technology} & $|i_{z}|$ & \texttt{NUM(z,tec)} & . \\ \hline
+quasi-fixed O\&M cost & $o^{q}_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & k\EUR \\ \hline
+variable O\&M cost & $o^{v}_{i}$ & \texttt{OM\_VAR\_COST(tec)} & \EUR / MWh \\ \hline
 \hline
@@
 % day-ahead price & $p_{el}$ & \texttt{PRICE\_DA(t,z)} & EUR / MWh \\ \hline
-\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_EUA(t,z)} & EUR / t CO2 \\ \hline
-fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & EUR / MWh \\ \hline
+\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_EUA(t,z)} & \EUR / $\text{t}_{\ce{CO2}}$ \\ \hline
+fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & \EUR / MWh \\ \hline
 reservoir inflows & $\rho_{z,t,k}$ & \makecell[l]{\texttt{RESERVOIR\_INFLOWS} \\ \texttt{(z,t,tec\_strg)}} & MW \\ \hline
@@
 efficiency power in & $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_in')}} & . \\ \hline
-value of lost load & $\mu$ & \texttt{VALUE\_NSE(z)} & EUR per MWh \\ \hline
+value of lost load & $\mu$ & \texttt{VALUE\_NSE(z)} & \EUR / MWh \\ \hline
 \hline
@@
 %cost variables
-zonal system cost & $C_{z}$ & \texttt{cost(r)} & EUR \\ \hline
-emission cost & $C^{e}_{z,t,i}$ & \texttt{cost\_emission(z,t,tec)} & EUR \\ \hline
-fuel cost & $C^{f}_{z,t,i}$ & \texttt{cost\_fuel(z,t,tec)} & EUR \\ \hline
-total o\&m cost & $C^{om}_{z,i}$ & \texttt{cost\_om(z,tec)} & EUR \\ \hline
-\makecell[l]{capital cost of \\generators} & $C_{z}^{inv,i}$ & \texttt{cost\_invgen(z)} & EUR \\ \hline
-\makecell[l]{capital cost of \\storages} & $C_{z}^{inv,k}$ & \texttt{cost\_invstrg(z)} & EUR \\ \hline
-\makecell[l]{capital cost of \\interconnectors} & $C_{z}^{inv,ic}$ & \texttt{cost\_gridexpansion(z)} & EUR \\ \hline
+zonal system cost & $C_{z}$ & \texttt{cost(r)} & k\EUR \\ \hline
+emission cost & $C^{e}_{z,t,i}$ & \texttt{cost\_emission(z,t,tec)} & k\EUR \\ \hline
+fuel cost & $C^{f}_{z,t,i}$ & \texttt{cost\_fuel(z,t,tec)} & k\EUR \\ \hline
+total o\&m cost & $C^{om}_{z,i}$ & \texttt{cost\_om(z,tec)} & k\EUR \\ \hline
+\makecell[l]{capital cost of \\generators} & $C_{z}^{inv,i}$ & \texttt{cost\_invgen(z)} & k\EUR \\ \hline
+\makecell[l]{capital cost of \\storages} & $C_{z}^{inv,k}$ & \texttt{cost\_invstrg(z)} & k\EUR \\ \hline
+\makecell[l]{capital cost of \\interconnectors} & $C_{z}^{inv,ic}$ & \texttt{cost\_gridexpansion(z)} & k\EUR \\ \hline
 % investment and decommissioning variables
@@
 \makecell[l]{added capacity of \\conventionals} & $\bar{g}^{+}_{z,i}$ & \texttt{invest\_thermal(z,tec)} & GW \\ \hline
-\makecell[l]{added storage \\capacity (power)} & $\bar{s}^{+}_{z,k}$ & \texttt{invest\_storage\_power(z,tec\_strg)} & EUR \\ \hline
-\makecell[l]{added storage \\capacity (energy)} & $\bar{v}^{+}_{z,k}$ & \texttt{invest\_storage\_energy(z,tec\_strg)} & EUR \\ \hline
-\makecell[l]{added transmission \\capacity} & $\bar{x}^{+}_{z,zz}$ & \texttt{invest\_atc(z,zz)} & EUR \\ \hline
+\makecell[l]{added storage \\capacity (power)} & $\bar{s}^{+}_{z,k}$ & \texttt{invest\_storage\_power(z,tec\_strg)} & GW \\ \hline
+\makecell[l]{added storage \\capacity (energy)} & $\bar{v}^{+}_{z,k}$ & \texttt{invest\_storage\_energy(z,tec\_strg)} & GWh \\ \hline
+\makecell[l]{added transmission \\capacity} & $\bar{x}^{+}_{z,zz}$ & \texttt{invest\_atc(z,zz)} & GW \\ \hline
 \makecell[l]{decommissioned \\capacity of \\conventionals} & $\bar{g}^{-}_{z,i}$ & \texttt{decommission(z,tec)} & GW \\ \hline
@@
 &C^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: \varepsilon_{f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i\\
-&C^{om}_{z,i}& =&\ \left(|i_{z}| - \bar{g}^{-}_{z,i} + \bar{g}^{+}_{z,i}\right) o^{q}_{i} + \sum_{t,m} \left(o^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
+&C^{om}_{z,i}& =&\ \left(\bar{g}^{0}_{z,i} - \bar{g}^{-}_{z,i} + \bar{g}^{+}_{z,i}\right) o^{q}_{i} + \sum_{t,m} \left(o^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
 &C^{inv,i}_{z}& =&\ \kappa_{z,i} \: \bar{g}^{+}_{z,i} \qquad \qquad &\forall z\\
@@
 &C^{inv,k}_{z}& =&\ \kappa^{P}_{z,k} \: \bar{s}^{+}_{z,k} + \kappa^{E}_{z,k} \: v^{+}_{z,k} \qquad \qquad &\forall z\\
-&C^{inv,ic}_{z}& =&\ \kappa^{ic}_{z} \: \bar{x}^{+}_{z,zz} \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
+&C^{inv,ic}_{z}& =&\ \kappa^{ic}_{z} \: \delta_{z,zz} \: \bar{x}^{+}_{z,zz} \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
 &C^{nse}_{z}& =&\ \mu \sum_{t,m} \Omega^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
@@
 \begin{align}
-g_{z,t,i,m} \leq \argmax_{l}(\psi_{i,l,m}) \left( |i_{z}| + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \right) \qquad \qquad \forall z,t,i,m
+g_{z,t,i,m} \leq \argmax_{l}(\psi_{i,l,m}) \left( \bar{g}^{0}_{z,i} + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \right) \qquad \qquad \forall z,t,i,m
 \end{align}
@@
 \begin{align}
-\sum_{l} w_{z,t,i,l} = |i_{z}| + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \qquad \qquad \forall z,t,i \in J \\
+\sum_{l} w_{z,t,i,l} = \bar{g}^{0}_{z,i} + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \qquad \qquad \forall z,t,i \in J \\
 g_{z,t,i,m} \leq \sum_{l} w_{z,t,i,l} \: \psi_{i,l,m} \qquad \qquad \forall z,t,i \in J, m \\
@@
 \begin{align}
-\bar{g}^{-}_{z,i} \leq |i_{z}| + \bar{g}^{+}_{z,i} \qquad \qquad \forall z,i
+\bar{g}^{-}_{z,i} \leq \bar{g}^{0}_{z,i} + \bar{g}^{+}_{z,i} \qquad \qquad \forall z,i
 \end{align}

### 2019-08-19 doc/mathematical_description.tex
fixed co-generation bug in main model; added biomass to wind_vs_pv analysis

@@
 &C^{inv,k}_{z}& =&\ \kappa^{P}_{z,k} \: \bar{s}^{+}_{z,k} + \kappa^{E}_{z,k} \: v^{+}_{z,k} \qquad \qquad &\forall z\\
-&C^{inv,ic}_{z}& =&\ \kappa^{ic}_{z} \: \delta_{z,zz} \: \bar{x}^{+}_{z,zz} \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
+&C^{inv,ic}_{z}& =&\ \sum_{zz} (\kappa^{ic}_{z} \: \delta_{z,zz} \: \bar{x}^{+}_{z,zz}) / \: 2 \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
 &C^{nse}_{z}& =&\ \mu \sum_{t,m} \Omega^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}

### 2019-09-05 doc/mathematical_description.tex
improved consistency of mathematical description

--- 2019-08-01 doc/mathematical_description.tex
@@
 \usepackage[inner=2.5cm,outer=2.5cm, bottom=3cm]{geometry}
+\usepackage{ctable}
+\usepackage{longtable}
 
 \title{Description of the power system model \emph{medea}}
-%\title{Mathematical description of the power system model \emph{medea}}
 \date{\today}
@@
 \section{Sets} \label{sets}
-\begin{tabular}{l l l l}
+Sets are denoted by upper-case latin letters, while set elements are denoted by lower-case latin letters.
+
+\begin{longtable}{p{3.6cm} c c p{6cm}}
+%\caption{Sets} \\ 
+\hline
 \textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{elements} \\
 \hline \hline
+
 market zones & $z \in Z$ & \texttt{z} & \texttt{AT, DE} \\ \hline
 
-\makecell[l]{time periods \\(hours)} & $t \in T$ & \texttt{t} & \texttt{t1, t2, \ldots, t8760} \\ \hline
+time periods (hours) & $t \in T$ & \texttt{t} & \texttt{t1, t2, \ldots, t8760} \\ \hline
 
-\makecell[l]{power \\generation \\technologies} & $i \in I$ & \texttt{tec} &
+power generation technologies & $i \in I$ & \texttt{tec} & 
+\texttt{nuc, lig\_stm, lig\_stm\_chp, lig\_boa, lig\_boa\_chp,coal\_sub, coal\_sub\_chp, coal\_sc, coal\_sc\_chp, coal\_usc, coal\_usc\_chp, coal\_igcc, ng\_stm, 
+ng\_stm\_chp, ng\_ctb\_lo, ng\_ctb\_lo\_chp, ng\_ctb\_hi, ng\_ctb\_hi\_chp, ng\_cc\_lo, ng\_cc\_lo\_chp, ng\_cc\_hi, ng\_cc\_hi\_chp, ng\_mtr, ng\_mtr\_chp, 
+ng\_boiler\_chp, oil\_stm, oil\_stm\_chp, oil\_ctb, oil\_ctb\_chp, oil\_cc, oil\_cc\_chp, bio, bio\_chp, heatpump\_pth} \\ \hline
 
-\makecell[l]{ \texttt{nuc, lig\_stm, lig\_stm\_chp,}\\
-	\texttt{lig\_boa, lig\_boa\_chp,} \\
-	\texttt{coal\_sub, coal\_sub\_chp,} \\
-	\texttt{coal\_sc, coal\_sc\_chp,}\\
-	\texttt{coal\_usc, coal\_usc\_chp,} \\
-	\texttt{coal\_igcc,}\\
-	\texttt{ng\_stm, ng\_stm\_chp,} \\
-	\texttt{ng\_ctb\_lo, ng\_ctb\_lo\_chp,}\\ 
-	\texttt{ng\_ctb\_hi, ng\_ctb\_hi\_chp,}\\
-	\texttt{ng\_cc\_lo, ng\_cc\_lo\_chp,}\\
-	\texttt{ng\_cc\_hi, ng\_cc\_hi\_chp,}\\
-	\texttt{ng\_mtr, ng\_mtr\_chp,} \\
-	\texttt{ng\_boiler\_chp,} \\
-	\texttt{oil\_stm, oil\_stm\_chp,} \\ 
-	\texttt{oil\_ctb, oil\_ctb\_chp,} \\
-	\texttt{oil\_cc, oil\_cc\_chp,} \\
-	\texttt{bio, bio\_chp,}
-	\texttt{heatpump\_pth} 
-} \\ \hline
+CHP technologies & $j \in J \subset I$ & \texttt{tec\_chp} & \texttt{lig\_stm\_chp, lig\_boa\_chp, coal\_sub\_chp, coal\_sc\_chp, coal\_usc\_chp, ng\_stm\_chp, 
+ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp, ng\_cc\_lo\_chp, ng\_cc\_hi\_chp, ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm\_chp, oil\_ctb\_chp, oil\_cc\_chp, bio\_chp} \\ \hline
 
-\makecell[l]{CHP \\technologies} & $j \in J \subset I$ & \texttt{tec\_chp} & \makecell[l]{
-	\texttt{lig\_stm\_chp, lig\_boa\_chp,}\\
-	\texttt{coal\_sub\_chp, coal\_sc\_chp,}\\
-	\texttt{coal\_usc\_chp, ng\_stm\_chp,}\\ 
-	\texttt{ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp,}\\
-	\texttt{ng\_cc\_lo\_chp, ng\_cc\_hi\_chp,}\\
-	\texttt{ng\_mtr\_chp, ng\_boiler\_chp,}\\
-	\texttt{oil\_stm\_chp, oil\_ctb\_chp,} \\ 
-	\texttt{oil\_cc\_chp, bio\_chp} } \\ \hline
+power to heat technologies & $h \in H \subset I$ & \texttt{tec\_pth} & \texttt{heatpump\_pth}  \\ \hline
 
-\makecell[l]{power to heat \\ technologies} & $h \in H \subset I$ & \texttt{tec\_pth} & \texttt{heatpump\_pth}  \\ \hline
+storage technologies & $k \in K$ & \texttt{tec\_strg} & \texttt{res\_day, res\_week, res\_season, psp\_day, psp\_week, psp\_season, battery} \\ \hline
 
-\makecell[l]{storage \\technologies} & $k \in K$ & \texttt{tec\_strg} & \makecell[l]{
-	\texttt{res\_day, res\_week, res\_season,} \\ 
-	\texttt{psp\_day, psp\_week, psp\_season,} \\
-	\texttt{battery}
-	} \\ \hline
+intermittent generators & $n \in N$ & \texttt{tec\_itm} & \texttt{wind\_on, wind\_off, pv, ror}\\ \hline
 
-\makecell[l]{intermittent \\generators} & $n \in N$ & \texttt{tec\_itm} & \texttt{wind\_on, wind\_off, pv, ror}\\ \hline
+fuels & $f \in F$ & \texttt{f} & \texttt{nuclear, lignite, coal, gas, oil, biomass, power} \\ \hline
 
-fuels & $f \in F$ & \texttt{f} & \makecell[l]{\texttt{nuclear, lignite, coal,} \\ \texttt{gas, oil, biomass, power}} \\ \hline
-
-\makecell[l]{feasible operation \\region limits}& $l \in L$ & \texttt{l} & \texttt{l1, l2, l3, l4} \\ \hline
+feasible operation region limits & $l \in L$ & \texttt{l} & \texttt{l1, l2, l3, l4} \\ \hline
 
 energy products & $m \in M$ & \texttt{prd} & \texttt{el, ht} \\ \hline \hline
-\end{tabular}
+\end{longtable}
 
 \section{Parameters} \label{parameters}
-\begin{tabular}{l l l l}
-\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\
-\hline \hline
-\makecell[l]{minimal \\conventional \\generation} & $\sigma_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(z)} & GW \\ \hline
+Parameters are denoted either by lower-case greek letters or by upper-case latin letters.
+
+\begin{longtable}{p{4.1cm} c l c}
+\hline
+	\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\ \hline \hline
+% LOWER-CASE GREEK LETTERS
+	distance between countries & $\delta_{z,zz}$ & \texttt{KM(z,zz)} & km \\ \hline
+	fuel emission intensity & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $\text{t}_{\ce{CO2}}$ / MWh \\ \hline
+	power plant efficiency & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & MWh / MWh \\ \hline
+	efficiency power out & $\eta^{out}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_out')}} &  \\ \hline
+	efficiency power in & $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_in')}} &  \\ \hline	
+	value of lost load & $\mu_{z}$ & \texttt{VALUE\_NSE(z)} & \EUR / MWh \\ \hline
+	reservoir inflows & $\rho_{z,t,k}$ & \makecell[l]{\texttt{RESERVOIR\_INFLOWS} \\ \texttt{(z,t,tec\_strg)}} & MW \\ \hline	
+	minimal conventional generation & $\sigma_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(z)} & GW \\ \hline
+	intermittent generation profile & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(z,t,tec\_itm)} & $[0,1]$ \\ \hline
+	inputs of feasible operating region & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & $[0,1]$ \\ \hline
+	output tuples of feasible operating region & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & $[0,1]$ \\ \hline
+% UPPER-CASE LATIN LETTERS	
+	capital cost of intermittent generators (specific, annuity) & $C^{r}_{z,n}$ & \texttt{INVESTCOST\_ITM(z,tec\_itm)} & k\EUR / GW \\ \hline
+	capital cost of thermal generators (specific, annuity) & $C^{g}_{z,i}$ & \texttt{INVESTCOST\_THERMAL(z,tec)} & k\EUR / GW \\ \hline
+	capital cost of storages - power (specific, annuity) & $C^{s}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_power')}} & k\EUR / GW \\ \hline
+	capital cost of storages - energy (specific, annuity) & $C^{v}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_energy')}} & k\EUR / GW \\ \hline
+	capital cost of transmission capacity & $C^{x}_{z}$ & \texttt{INVESTCOST\_ATC} & k\EUR / GW \\ \hline
 	energy demand & $D_{z,t,p}$ & \texttt{CONSUMPTION(z,t,prd)} & GW \\ \hline
-\makecell[l]{power plant \\efficiency} & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & $MW / MW_{th}$\\ \hline
-\makecell[l]{fuel emission \\intensity} & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $t \ce{CO2}/MWh_{th}$\\ \hline
-\makecell[l]{inputs of \\feasible operating \\region} & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & MWh ? \\ \hline
-\makecell[l]{output tuples of \\feasible operating \\region} & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & MWh ? \\ \hline
-\makecell[l]{intermittent \\generation profile} & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(z,t,tec\_itm)} & \% \\ \hline
-%\makecell[l]{exports to not \\modelled regions} & $X$ & \texttt{FLOW\_EXPORTS(z,t)} & GW \\ \hline
-%\makecell[l]{imports from not \\modelled regions} & $X$ & \texttt{FLOW\_IMPORTS(z,t)} & GW \\ \hline
-\makecell[l]{installed capacity \\of intermittent \\generators} & $\bar{r}^{0}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(z,tec\_itm)} & GW \\ \hline
-\makecell[l]{installed capacity \\of thermal generators} & $\bar{g}^{0}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(z,tec)} & GW \\ \hline
-\makecell[l]{capital cost of \\intermittent generators \\(specific, annuity)} & $\kappa_{z,n}$ & \texttt{INVESTCOST\_ITM(z,tec\_itm)} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{capital cost of \\thermal generators \\(specific, annuity)} & $\kappa_{z,i}$ & \texttt{INVESTCOST\_THERMAL(z,tec)} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{capital cost of \\storages - power \\(specific, annuity)} & $\kappa^{P}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_power')}} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{capital cost of \\storages - energy \\(specific, annuity)} & $\kappa^{E}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_energy')}} & $\frac{mn EUR}{GW}$ \\ \hline
-\makecell[l]{installed available \\transfer capacity} & $\bar{x}^{0}_{z,zz}$ & \texttt{ATC(z,zz)} & GW \\ \hline
-\makecell[l]{count 100 MW slices \\of same technology} & $|i_{z}|$ & \texttt{NUM(z,tec)} & . \\ \hline
-quasi-fixed O\&M cost & $o^{q}_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & EUR \\ \hline
-variable O\&M cost & $o^{v}_{i}$ & \texttt{OM\_VAR\_COST(tec)} & EUR / MWh \\ \hline
+	initial capacity of thermal generators & $\bar{G}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(z,tec)} & GW \\ \hline
+	quasi-fixed O\&M cost & $O^{q}_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & k\EUR / GW \\ \hline
+	variable O\&M cost & $O^{v}_{i}$ & \texttt{OM\_VAR\_COST(tec)} & \EUR / MWh \\ \hline
+	\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_CO2(t,z)} & \EUR / $\text{t}_{\ce{CO2}}$ \\ \hline
+	fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & \EUR / MWh \\ \hline
+	initial capacity of intermittent generators & $\bar{R}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(z,tec\_itm)} & GW \\ \hline
+\pagebreak
+%\caption[Parameters]{(continued)}\\ 
 \hline
-\end{tabular}
+	% UPPER-CASE LATIN LETTERS (CONTINUED)
+	\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\ \hline \hline
+	max power out & $\bar{S}^{out}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'power\_out')}} & GW \\ \hline
+	max power in & $\bar{S}^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'power\_in')}} & GW \\ \hline
+	max energy stored & $\bar{V}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'energy\_max')}} &  \\ \hline
+	installed available transmission capacity & $\bar{X}_{z,zz}$ & \texttt{ATC(z,zz)} & GW \\ \hline \hline	
+\end{longtable}
 
-\begin{tabular}{l l l l}
-\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\
-\hline \hline
-% day-ahead price & $p_{el}$ & \texttt{PRICE\_DA(t,z)} & EUR / MWh \\ \hline
-\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_EUA(t,z)} & EUR / t CO2 \\ \hline
-fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & EUR / MWh \\ \hline
-reservoir inflows & $\rho_{z,t,k}$ & \makecell[l]{\texttt{RESERVOIR\_INFLOWS} \\ \texttt{(z,t,tec\_strg)}} & MW \\ \hline
-max power out & $\bar{s}^{out}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'power\_out')}} & GW \\ \hline
-max power in & $\bar{s}^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'power\_in')}} & GW \\ \hline
-max energy stored & $\bar{v}^{0}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'energy\_max')}} & . \\ \hline
-efficiency power out & $\eta^{out}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_out')}} & . \\ \hline
-efficiency power in & $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_in')}} & . \\ \hline
-value of lost load & $\mu$ & \texttt{VALUE\_NSE(z)} & EUR per MWh \\ \hline
-\hline
-\end{tabular}
+\newpage
 
 \section{Variables} \label{variables}
-\begin{tabular}{l l l l}
-\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\
-\hline \hline
-%cost variables
-zonal system cost & $C_{z}$ & \texttt{cost(r)} & EUR \\ \hline
-emission cost & $C^{e}_{z,t,i}$ & \texttt{cost\_emission(z,t,tec)} & EUR \\ \hline
-fuel cost & $C^{f}_{z,t,i}$ & \texttt{cost\_fuel(z,t,tec)} & EUR \\ \hline
-total o\&m cost & $C^{om}_{z,i}$ & \texttt{cost\_om(z,tec)} & EUR \\ \hline
-\makecell[l]{capital cost of \\generators} & $C_{z}^{inv,i}$ & \texttt{cost\_invgen(z)} & EUR \\ \hline
-\makecell[l]{capital cost of \\storages} & $C_{z}^{inv,k}$ & \texttt{cost\_invstrg(z)} & EUR \\ \hline
-\makecell[l]{capital cost of \\interconnectors} & $C_{z}^{inv,ic}$ & \texttt{cost\_gridexpansion(z)} & EUR \\ \hline
-% investment and decommissioning variables
-\makecell[l]{added capacity of \\intermittents} & $\bar{r}^{+}_{z,n}$ & \texttt{invest\_res(z,tec\_itm)} & GW \\ \hline
-\makecell[l]{added capacity of \\conventionals} & $\bar{g}^{+}_{z,i}$ & \texttt{invest\_thermal(z,tec)} & GW \\ \hline
-\makecell[l]{added storage \\capacity (power)} & $\bar{s}^{+}_{z,k}$ & \texttt{invest\_storage\_power(z,tec\_strg)} & EUR \\ \hline
-\makecell[l]{added storage \\capacity (energy)} & $\bar{v}^{+}_{z,k}$ & \texttt{invest\_storage\_energy(z,tec\_strg)} & EUR \\ \hline
-\makecell[l]{added transmission \\capacity} & $\bar{x}^{+}_{z,zz}$ & \texttt{invest\_atc(z,zz)} & EUR \\ \hline
-\makecell[l]{decommissioned \\capacity of \\conventionals} & $\bar{g}^{-}_{z,i}$ & \texttt{decommission(z,tec)} & GW \\ \hline
-% dispatch variables
-\makecell[l]{energy generated by \\conventionals} & $g_{z,t,i,p}$ & \texttt{q\_gen(z,t,tec,prd)} & GW \\ \hline
-\makecell[l]{electricity generated \\by intermittents} & $r_{z,t,n}$ & \texttt{q\_itm(z,t,tec\_itm)} & GW \\ \hline
-\makecell[l]{operating region \\weight} & $w_{z,t,i,l}$ & \texttt{cc\_weights(z,t,tec,l)} & . \\ \hline
-\makecell[l]{fuel burn for energy \\generation} & $b_{z,t,i,f}$ & \texttt{q\_fueluse(z,t,tec,f)} & GW \\ \hline
+Variables are denoted by lower-case latin letters.
+
+\begin{longtable}{p{4.1cm} c l c}
+%\caption{Variables}\label{variables} \\ 
+\hline
+\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\ \hline \hline
+fuel burn for energy generation & $b_{z,t,i,f}$ & \texttt{q\_fueluse(z,t,tec,f)} & GW \\ \hline
+zonal system cost & $c_{z}$ & \texttt{cost(r)} & k\EUR \\ \hline
+fuel cost & $c^{b}_{z,t,i}$ & \texttt{cost\_fuel(z,t,tec)} & k\EUR \\ \hline
+emission cost & $c^{e}_{z,t,i}$ & \texttt{cost\_emission(z,t,tec)} & k\EUR \\ \hline
+total o\&m cost & $c^{om}_{z,i}$ & \texttt{cost\_om(z,tec)} & k\EUR \\ \hline
+capital cost of generators & $c^{g}_{z}$ & \texttt{cost\_invgen(z)} & k\EUR \\ \hline
+total cost of non-served load & $c^{q}_{z}$ & \texttt{cost\_nse(z)} & k\EUR \\ \hline
+capital cost of intermittent generators & $c^{r}_{z}$ & \texttt{cost\_invitm(z)} & k\EUR \\ \hline
+capital cost of storages & $c^{s,v}_{z}$ & \texttt{cost\_invstrg(z)} & k\EUR \\ \hline
+capital cost of interconnectors & $c^{x}_{z}$ & \texttt{cost\_gridexpansion(z)} & k\EUR \\ \hline
+\ce{CO2} emissions & $e_{z}$ & \texttt{emissions(z)} & t \ce{CO2} \\ \hline
+added capacity of conventionals & $\bar{g}^{+}_{z,i}$ & \texttt{invest\_thermal(z,tec)} & GW \\ \hline
+decommissioned capacity of conventionals & $\bar{g}^{-}_{z,i}$ & \texttt{decommission(z,tec)} & GW \\ \hline
+energy generated by conventionals & $g_{z,t,i,p}$ & \texttt{q\_gen(z,t,tec,prd)} & GW \\ \hline
+curtailed energy & $q^{+}_{z,t}$ & \texttt{q\_curtail(z,t)} & GW \\ \hline
+non-served energy & $q^{-}_{z,t,p}$ & \texttt{q\_nonserved(z,t,prd)} & GW \\ \hline
+added capacity of intermittents & $\bar{r}^{+}_{z,n}$ & \texttt{invest\_res(z,tec\_itm)} & GW \\ \hline
+electricity generated by intermittents & $r_{z,t,n}$ & \texttt{q\_itm(z,t,tec\_itm)} & GW \\ \hline
+added storage capacity (power) & $\bar{s}^{+}_{z,k}$ & \texttt{invest\_storage\_power(z,tec\_strg)} & GW \\ \hline
 energy stored in & $s^{in}_{z,t,k}$ & \texttt{q\_store\_in(z,t,tec\_strg)} & GW \\ \hline
 energy stored out & $s^{out}_{z,t,k}$ & \texttt{q\_store\_out(z,t,tec\_strg)} & GW \\ \hline
-\makecell[l]{storage energy \\content} & $v_{z,t,k}$ & \texttt{storage\_level(z,t,tec\_strg)} & GWh \\ \hline
-\makecell[l]{electricity net \\export} & $x_{z,zz,t}$ & \texttt{flow(z,zz,t)} & GW \\ \hline
-curtailed energy & $\Omega^{+}_{z,t}$ & \texttt{q\_curtail(z,t)} & GW \\ \hline
-non-served energy & $\Omega^{-}_{z,t,p}$ & \texttt{q\_nonserved(z,t,prd)} & GW \\ \hline
-% other variables
-\ce{CO2} emissions & $e_{z}$ & \texttt{emissions(z)} & t \ce{CO2} \\ \hline
-\hline
-\end{tabular}
+added storage capacity (energy) & $\bar{v}^{+}_{z,k}$ & \texttt{invest\_storage\_energy(z,tec\_strg)} & GWh \\ \hline
+storage energy content & $v_{z,t,k}$ & \texttt{storage\_level(z,t,tec\_strg)} & GWh \\ \hline
+operating region weight & $w_{z,t,i,l}$ & \texttt{cc\_weights(z,t,tec,l)} & . \\ \hline
+added transmission capacity & $\bar{x}^{+}_{z,zz}$ & \texttt{invest\_atc(z,zz)} & GW \\ \hline
+electricity net export & $x_{z,zz,t}$ & \texttt{flow(z,zz,t)} & GW \\ \hline \hline
+\end{longtable}
+
+\newpage
+
+\section{Naming system}
+\ctable[
+	maxwidth = 160mm,
+	pos = h]
+	{X p{2.0cm} p{2.0cm} p{2.0cm} p{2.0cm} p{2.0cm}}
+	{
+	\tnote[$\dagger$]{parameter} 
+	\tnote[$\ddagger$]{variable}
+	}
+	{\FL 
+	 &  initial capacity\tmark[$\dagger$] & added capacity\tmark[$\ddagger$] & \makecell[l]{decom-\\missioned \\capacity\tmark[$\ddagger$]} & \makecell[l]{specific \\investment \\cost\tmark[$\dagger$]} & dispatch\tmark[$\ddagger$] \ML
+	thermal units & $\bar{G}_{z,i}$ & $\bar{g}^{+}_{z,i}$ & $\bar{g}^{-}_{z,i}$ & $C^{g}_{z,i}$ & $g_{z,t,i,p}$ \ML
+	\makecell[l]{intermittent \\units} &  $\bar{R}_{z,n}$ & $\bar{r}^{+}_{z,n}$ & $\bar{r}^{-}_{z,n}$ & $C^{r}_{z,n}$ & $r_{z,t,n}$ \ML
+	\makecell[l]{storages \\(power)} & $\bar{S}_{z,k}$ & $\bar{s}^{+}_{z,k}$ & $\bar{s}^{-}_{z,k}$ & $C^{s}_{z,k}$ & $s_{z,t,k}$ \ML
+	\makecell[l]{storages \\(energy)} & $\bar{V}_{z,k}$ & $\bar{v}^{+}_{z,k}$ & $\bar{v}^{-}_{z,t,k}$ & $C^{v}_{z,k}$ & na \ML
+	transmission &  $\bar{X}_{z,zz}$ & $\bar{x}^{+}_{z,zz}$ & $\bar{x}^{-}_{z,zz}$ & $C^{x}_{z,zz}$ & $x_{z,zz,t}$ \LL
+}
 
@@
 \paragraph{Model objective}
-\emph{medea} minimizes total system cost $C$, i.e.
+\emph{medea} minimizes total system cost $c$, i.e.
 the total cost of generating electricity and heat from technologies and capacities adequate to meet demand, over a large number  of decision variables, essentially representing investment and dispatch decisions in each market zone $z$ of the modelled energy systems.
 \begin{align}
-\min C = \sum_{z} (C_{z})
+\min c = \sum_{z} (c_{z})
 \end{align}
-Zonal system costs $C_{z}$ are the sum of fuel cost $C^{f}_{z,t,i}$, emission cost $C^{e}_{z,t,i}$, operation and maintenance cost $C^{om}_{z,i}$, capital costs of investment in conventional and intermittent generation ($C^{inv,i}_{z}$, $C^{inv,n}_{z}$), storage ($C^{inv,k}_{z}$) and transmission ($C^{inv,ic}_{z}$) equipment, and the cost of non-served load ($C^{nse}_{z}$) that accrues when demand is not met, e.g.
+Zonal system costs $c_{z}$ are the sum of fuel cost $c^{b}_{z,t,i}$, emission cost $c^{e}_{z,t,i}$, operation and maintenance cost $c^{om}_{z,i}$, capital costs of investment in conventional and intermittent generation ($c^{g}_{z}$, $c^{r}_{z}$), storage ($c^{s,v}_{z}$) and transmission ($c^{x}_{z}$) equipment, and the cost of non-served load ($c^{q}_{z}$) that accrues when demand is not met, e.g.
 when there is a power outage.
@@
 %\begin{split}
-C_{z} = \sum_{t,i}  C^{f}_{z,t,i} + \sum_{t,i} C^{e}_{z,t,i} + \sum_{i} C^{om}_{z,i} + C^{inv,i}_{z} + 
- C^{inv,n}_{z} + C^{inv,k}_{z} + C^{inv,ic}_{z} + C^{nse}_{z} \qquad \qquad \forall z
+c_{z} = \sum_{t,i}  c^{b}_{z,t,i} + \sum_{t,i} c^{e}_{z,t,i} + \sum_{i} c^{om}_{z,i} + c^{g}_{z} + 
+ c^{r}_{z} + c^{s,v}_{z} + c^{x}_{z} + c^{q}_{z} \qquad \qquad \forall z
 %\end{split}
@@
 The components of zonal system costs are calculated as given in equations (\ref{fuel_cost}) to (\ref{lost_load_cost}).
-Upper-case $C$ represent total cost, while $\kappa$ denotes specific, annualized capital cost of technology investment. Prices for fuels and \ce{CO2} are denoted by uppercase $P$.
+Lower-case $c$ represent total cost, while upper-case $C$ denotes specific, annualized capital cost of technology investment. Prices for fuels and \ce{CO2} are denoted by $P$.
 \begin{align}
-&C^{f}_{z,t,i}& =&\ \sum_{f} \left( P_{t,z,f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i \label{fuel_cost} \\
-&C^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: \varepsilon_{f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i\\
-&C^{om}_{z,i}& =&\ \left(|i_{z}| - \bar{g}^{-}_{z,i} + \bar{g}^{+}_{z,i}\right) o^{q}_{i} + \sum_{t,m} \left(o^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
-&C^{inv,i}_{z}& =&\ \kappa_{z,i} \: \bar{g}^{+}_{z,i} \qquad \qquad &\forall z\\
-&C^{inv,n}_{z}& =&\ \kappa_{z,n} \: \bar{r}^{+}_{z,n} \qquad \qquad &\forall z\\
-&C^{inv,k}_{z}& =&\ \kappa^{P}_{z,k} \: \bar{s}^{+}_{z,k} + \kappa^{E}_{z,k} \: v^{+}_{z,k} \qquad \qquad &\forall z\\
-&C^{inv,ic}_{z}& =&\ \kappa^{ic}_{z} \: \bar{x}^{+}_{z,zz} \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
-&C^{nse}_{z}& =&\ \mu \sum_{t,m} \Omega^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
+&c^{b}_{z,t,i}& =&\ \sum_{f} \left( P_{t,z,f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i \label{fuel_cost} \\
+&c^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: \varepsilon_{f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i\\
+&c^{om}_{z,i}& =&\ O^{q}_{i} \left(\bar{G}_{z,i} - \bar{g}^{-}_{z,i} + \bar{g}^{+}_{z,i}\right) + \sum_{t,m} \left(O^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
+&c^{g}_{z}& =&\ C^{g}_{z,i} \: \bar{g}^{+}_{z,i} \qquad \qquad &\forall z\\
+&c^{r}_{z}& =&\ C^{r}_{z,n} \: \bar{r}^{+}_{z,n} \qquad \qquad &\forall z\\
+&c^{s,v}_{z}& =&\ C^{s}_{z,k} \: \bar{s}^{+}_{z,k} + C^{v}_{z,k} \: v^{+}_{z,k} \qquad \qquad &\forall z\\
+&c^{x}_{z}& =&\ \frac{1}{2} \: \sum_{zz} (C^{x}_{z} \: \delta_{z,zz} \: \bar{x}^{+}_{z,zz}) \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
+&c^{q}_{z}& =&\ \mu \sum_{t,m} q^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
 \end{align}
-In line with \cite{citation}, we estimate the specific cost of non-served load $\mu$ at 12500 EUR per MWh.
 
@@
 D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} 
-- \Omega^{-}_{z,t,\text{el}} + \Omega^{+}_{z,t} \qquad \forall z,t \label{market_clearing_el} \\
-\sum_{i} g_{z,t,i,\text{ht}} = D_{z,t,\text{ht}} - \Omega^{-}_{z,t,\text{ht}} \qquad \forall z,t \label{market_clearing_ht}
+- q^{-}_{z,t,\text{el}} + q^{+}_{z,t} \qquad \forall z,t \label{market_clearing_el} \\
+\sum_{i} g_{z,t,i,\text{ht}} = D_{z,t,\text{ht}} - q^{-}_{z,t,\text{ht}} \qquad \forall z,t \label{market_clearing_ht}
 \end{align}
@@
 \begin{align}
-g_{z,t,i,m} \leq \argmax_{l}(\psi_{i,l,m}) \left( |i_{z}| + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \right) \qquad \qquad \forall z,t,i,m
+g_{z,t,i,m} \leq \bar{G}_{z,i} + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \qquad \qquad \forall z,t,i,m
 \end{align}
@@
 \begin{align}
-\sum_{l} w_{z,t,i,l} = |i_{z}| + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \qquad \qquad \forall z,t,i \in J \\
-g_{z,t,i,m} \leq \sum_{l} w_{z,t,i,l} \: \psi_{i,l,m} \qquad \qquad \forall z,t,i \in J, m \\
-b_{z,t,i,f} \geq \sum_{l} w_{z,t,i,l} \: \chi_{i,l,f} \qquad \qquad \forall z,t,i \in J, f
+\sum_{l} w_{z,t,i,l} = \bar{G}_{z,i} + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \qquad \qquad \forall z,t,i \in J \\
+g_{z,t,i,m} \leq \sum_{l} \psi_{i,l,m} \: w_{z,t,i,l} \qquad \qquad \forall z,t,i \in J, m \\
+b_{z,t,i,f} \geq \sum_{l} \chi_{i,l,f} \: w_{z,t,i,l} \qquad \qquad \forall z,t,i \in J, f
 \end{align}
@@
 \begin{align}
-r_{z,t,n} = \phi_{z,t,n} \: \left( \bar{r}^{0}_{z,n} + \bar{r}^{+}_{z,n} \right) \qquad \qquad \forall z,t,n
+r_{z,t,n} = \phi_{z,t,n} \: \left( \bar{R}_{z,n} + \bar{r}^{+}_{z,n} \right) \qquad \qquad \forall z,t,n
 \end{align}
@@
 \begin{align}
-s^{out}_{z,t,k} \leq \bar{s}^{out}_{z,k} + \bar{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
-s^{in}_{z,t,k} \leq \bar{s}^{in}_{z,k} + \bar{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
-v_{z,t,k} \leq \bar{v}^{0}_{z,k} + \bar{v}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
+s^{out}_{z,t,k} \leq \bar{S}^{out}_{z,k} + \bar{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
+s^{in}_{z,t,k} \leq \bar{S}^{in}_{z,k} + \bar{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
+v_{z,t,k} \leq \bar{V}_{z,k} + \bar{v}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
 \end{align}
@@
 \begin{align}
-v_{z,t,k} = v_{z,t-1,k} + \rho_{z,t,k} + \eta^{in}_{z,k} \: s^{in}_{z,t,k} - (\eta^{out}_{z,k})^{-1} \: s^{out}_{z,t,k} \qquad \qquad \forall z,t,k
+v_{z,t,k} = \rho_{z,t,k} + \eta^{in}_{z,k} \: s^{in}_{z,t,k} - (\eta^{out}_{z,k})^{-1} \: s^{out}_{z,t,k} + v_{z,t-1,k} \qquad \qquad \forall z,t,k
 \end{align}
@@
 \begin{align}
-x_{z,zz,t} \leq \bar{x}^{0}_{z,zz} + \bar{x}^{+}_{z,zz} \qquad \qquad \forall z, zz, t \\
-x_{z,zz,t} \geq - \left( \bar{x}^{0}_{z,zz} + \bar{x}^{+}_{z,zz} \right) \qquad \qquad \forall z, zz, t
+x_{z,zz,t} \leq \bar{X}_{z,zz} + \bar{x}^{+}_{z,zz} \qquad \qquad \forall z, zz, t \\
+x_{z,zz,t} \geq - \left( \bar{X}_{z,zz} + \bar{x}^{+}_{z,zz} \right) \qquad \qquad \forall z, zz, t
 \end{align}
@@
 \begin{align}
-\bar{g}^{-}_{z,i} \leq |i_{z}| + \bar{g}^{+}_{z,i} \qquad \qquad \forall z,i
+\bar{g}^{-}_{z,i} \leq \bar{G}_{z,i} + \bar{g}^{+}_{z,i} \qquad \qquad \forall z,i
 \end{align}
@@
 \begin{align}
-\Omega^{+}_{z,t} \leq \sum_{n} r_{z,t,n} \qquad \qquad \forall z, t
+q^{+}_{z,t} \leq \sum_{n} r_{z,t,n} \qquad \qquad \forall z, t
 \end{align}

### 2019-09-06 doc/mathematical_description.tex
further consistency improvements in math documentation

@@
 \usepackage[utf8]{inputenc}
-\usepackage[fleqn]{amsmath}
+\usepackage{amsmath}
 \usepackage{amsfonts}
@@
 \hline \hline
-
-market zones & $z \in Z$ & \texttt{z} & \texttt{AT, DE} \\ \hline
-
-time periods (hours) & $t \in T$ & \texttt{t} & \texttt{t1, t2, \ldots, t8760} \\ \hline
-
-power generation technologies & $i \in I$ & \texttt{tec} & 
+fuels & $f \in F$ & \texttt{f} & \texttt{nuclear, lignite, coal, gas, oil, biomass, power} \\ \hline
+power generation technologies & $i \in I$ & \texttt{i} & 
 \texttt{nuc, lig\_stm, lig\_stm\_chp, lig\_boa, lig\_boa\_chp,coal\_sub, coal\_sub\_chp, coal\_sc, coal\_sc\_chp, coal\_usc, coal\_usc\_chp, coal\_igcc, ng\_stm, 
@@
 ng\_boiler\_chp, oil\_stm, oil\_stm\_chp, oil\_ctb, oil\_ctb\_chp, oil\_cc, oil\_cc\_chp, bio, bio\_chp, heatpump\_pth} \\ \hline
-
-CHP technologies & $j \in J \subset I$ & \texttt{tec\_chp} & \texttt{lig\_stm\_chp, lig\_boa\_chp, coal\_sub\_chp, coal\_sc\_chp, coal\_usc\_chp, ng\_stm\_chp, 
+power to heat technologies & $h \in H \subset I$ & \texttt{h(i)} & \texttt{heatpump\_pth}  \\ \hline
+CHP technologies & $j \in J \subset I$ & \texttt{j(i)} & \texttt{lig\_stm\_chp, lig\_boa\_chp, coal\_sub\_chp, coal\_sc\_chp, coal\_usc\_chp, ng\_stm\_chp, 
 ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp, ng\_cc\_lo\_chp, ng\_cc\_hi\_chp, ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm\_chp, oil\_ctb\_chp, oil\_cc\_chp, bio\_chp} \\ \hline
-
-power to heat technologies & $h \in H \subset I$ & \texttt{tec\_pth} & \texttt{heatpump\_pth}  \\ \hline
-
-storage technologies & $k \in K$ & \texttt{tec\_strg} & \texttt{res\_day, res\_week, res\_season, psp\_day, psp\_week, psp\_season, battery} \\ \hline
-
-intermittent generators & $n \in N$ & \texttt{tec\_itm} & \texttt{wind\_on, wind\_off, pv, ror}\\ \hline
-
-fuels & $f \in F$ & \texttt{f} & \texttt{nuclear, lignite, coal, gas, oil, biomass, power} \\ \hline
-
+storage technologies & $k \in K$ & \texttt{k} & \texttt{res\_day, res\_week, res\_season, psp\_day, psp\_week, psp\_season, battery} \\ \hline
 feasible operation region limits & $l \in L$ & \texttt{l} & \texttt{l1, l2, l3, l4} \\ \hline
-
-energy products & $m \in M$ & \texttt{prd} & \texttt{el, ht} \\ \hline \hline
+energy products & $m \in M$ & \texttt{m} & \texttt{el, ht} \\ \hline
+intermittent generators & $n \in N$ & \texttt{n} & \texttt{wind\_on, wind\_off, pv, ror}\\ \hline
+time periods (hours) & $t \in T$ & \texttt{t} & \texttt{t1, t2, \ldots, t8760} \\ \hline
+market zones & $z \in Z$ & \texttt{z} & \texttt{AT, DE} \\ \hline \hline
 \end{longtable}
@@
 % LOWER-CASE GREEK LETTERS
-	distance between countries & $\delta_{z,zz}$ & \texttt{KM(z,zz)} & km \\ \hline
-	fuel emission intensity & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $\text{t}_{\ce{CO2}}$ / MWh \\ \hline
-	power plant efficiency & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & MWh / MWh \\ \hline
-	efficiency power out & $\eta^{out}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_out')}} &  \\ \hline
-	efficiency power in & $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_in')}} &  \\ \hline	
+	distance between countries & $\delta_{z,zz}$ & \texttt{DISTANCE(z,zz)} & km \\ \hline
+	fuel emission intensity & $\varepsilon_{f}$ & \texttt{CO2\_INTENSITY(f)} & $\text{t}_{\ce{CO2}}$ / MWh \\ \hline
+	power plant efficiency & $\eta_{i,m,f}$ & \texttt{EFFICIENCY\_G(i,m,f)} & MWh / MWh \\ \hline
+	efficiency power out & $\eta^{out}_{z,k}$ & \makecell[l]{\texttt{EFFICIENCY\_S\_OUT(k)}} &  \\ \hline
+	efficiency power in & $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{EFFICIENCY\_S\_IN(k)}} &  \\ \hline	
+	scaling factor for peak load & $\lambda_{z}$ & \texttt{LAMBDA(z)} &  \\ \hline
 	value of lost load & $\mu_{z}$ & \texttt{VALUE\_NSE(z)} & \EUR / MWh \\ \hline
-	reservoir inflows & $\rho_{z,t,k}$ & \makecell[l]{\texttt{RESERVOIR\_INFLOWS} \\ \texttt{(z,t,tec\_strg)}} & MW \\ \hline	
-	minimal conventional generation & $\sigma_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(z)} & GW \\ \hline
-	intermittent generation profile & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(z,t,tec\_itm)} & $[0,1]$ \\ \hline
-	inputs of feasible operating region & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & $[0,1]$ \\ \hline
-	output tuples of feasible operating region & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & $[0,1]$ \\ \hline
+	reservoir inflows & $\rho_{z,t,k}$ & \makecell[l]{\texttt{INFLOWS(z,t,k)}} & MW \\ \hline	
+	scaling factor for peak intermittent generation & $\sigma_{z}$ & \texttt{SIGMA(z)} &  \\ \hline
+	intermittent generation profile & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(z,t,n)} & $[0,1]$ \\ \hline
+	peak intermittent generation profile & $\widehat{\phi}_{z,n}$ & \texttt{PEAK\_PROFILE(z,n)} & $[0,1]$ \\ \hline	
+	inputs of feasible operating region & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(i,l,f)} & $[0,1]$ \\ \hline
+	output tuples of feasible operating region & $\psi_{i,l,m}$ & \texttt{FEASIBLE\_OUTPUT(i,l,m)} & $[0,1]$ \\ \hline
 % UPPER-CASE LATIN LETTERS	
-	capital cost of intermittent generators (specific, annuity) & $C^{r}_{z,n}$ & \texttt{INVESTCOST\_ITM(z,tec\_itm)} & k\EUR / GW \\ \hline
-	capital cost of thermal generators (specific, annuity) & $C^{g}_{z,i}$ & \texttt{INVESTCOST\_THERMAL(z,tec)} & k\EUR / GW \\ \hline
-	capital cost of storages - power (specific, annuity) & $C^{s}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_power')}} & k\EUR / GW \\ \hline
-	capital cost of storages - energy (specific, annuity) & $C^{v}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_energy')}} & k\EUR / GW \\ \hline
-	capital cost of transmission capacity & $C^{x}_{z}$ & \texttt{INVESTCOST\_ATC} & k\EUR / GW \\ \hline
-	energy demand & $D_{z,t,p}$ & \texttt{CONSUMPTION(z,t,prd)} & GW \\ \hline
-	initial capacity of thermal generators & $\bar{G}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(z,tec)} & GW \\ \hline
-	quasi-fixed O\&M cost & $O^{q}_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & k\EUR / GW \\ \hline
-	variable O\&M cost & $O^{v}_{i}$ & \texttt{OM\_VAR\_COST(tec)} & \EUR / MWh \\ \hline
-	\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_CO2(t,z)} & \EUR / $\text{t}_{\ce{CO2}}$ \\ \hline
-	fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & \EUR / MWh \\ \hline
-	initial capacity of intermittent generators & $\bar{R}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(z,tec\_itm)} & GW \\ \hline
+	capital cost of intermittent generators (specific, annuity) & $C^{r}_{z,n}$ & \texttt{CAPITALCOST\_R(z,n)} & k\EUR / GW \\ \hline
+	capital cost of thermal generators (specific, annuity) & $C^{g}_{z,i}$ & \texttt{CAPITALCOST\_G(z,i)} & k\EUR / GW \\ \hline
+	capital cost of storages - power (specific, annuity) & $C^{s}_{z,k}$ & \makecell[l]{\texttt{CAPITALCOST\_S(z,k)}} & k\EUR / GW \\ \hline
+	capital cost of storages - energy (specific, annuity) & $C^{v}_{z,k}$ & \makecell[l]{\texttt{CAPITALCOST\_V(z,k)}} & k\EUR / GW \\ \hline
+	capital cost of transmission capacity & $C^{x}$ & \texttt{CAPITALCOST\_X} & k\EUR / GW \\ \hline
+	energy demand & $D_{z,t,m}$ & \texttt{DEMAND(z,t,m)} & GW \\ \hline
+	peak demand & $\widehat{D}_{z,m}$ & \texttt{PEAK\_LOAD(z,m)} & GW \\ \hline
+	initial capacity of dispatchable generators & $\widetilde{G}_{z,i}$ & \texttt{INITIAL\_CAP\_G(z,tec)} & GW \\ \hline
 \pagebreak
@@
 	\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\ \hline \hline
-	max power out & $\bar{S}^{out}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'power\_out')}} & GW \\ \hline
-	max power in & $\bar{S}^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'power\_in')}} & GW \\ \hline
-	max energy stored & $\bar{V}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'energy\_max')}} &  \\ \hline
-	installed available transmission capacity & $\bar{X}_{z,zz}$ & \texttt{ATC(z,zz)} & GW \\ \hline \hline	
+	quasi-fixed O\&M cost & $O^{q}_{i}$ & \texttt{OM\_COST\_QFIX(i)} & k\EUR / GW \\ \hline
+	variable O\&M cost & $O^{v}_{i}$ & \texttt{OM\_COST\_VAR(i)} & \EUR / MWh \\ \hline
+	\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_CO2(t,z)} & \EUR / $\text{t}_{\ce{CO2}}$ \\ \hline
+	fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & \EUR / MWh \\ \hline
+	initial capacity of intermittent generators & $\widetilde{R}_{z,n}$ & \texttt{INITIAL\_CAP\_R(z,n)} & GW \\ \hline
+	max power out & $\widetilde{S}^{out}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_S\_OUT(z,k)}} & GW \\ \hline
+	max power in & $\widetilde{S}^{in}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_S\_IN(z,k)}} & GW \\ \hline
+	max energy stored & $\widetilde{V}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_V(z,k)}} &  \\ \hline
+	installed available transmission capacity & $\widetilde{X}_{z,zz}$ & \texttt{INITIAL\_CAP\_X(z,zz)} & GW \\ \hline \hline	
 \end{longtable}
@@
 \textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\ \hline \hline
-fuel burn for energy generation & $b_{z,t,i,f}$ & \texttt{q\_fueluse(z,t,tec,f)} & GW \\ \hline
-zonal system cost & $c_{z}$ & \texttt{cost(r)} & k\EUR \\ \hline
-fuel cost & $c^{b}_{z,t,i}$ & \texttt{cost\_fuel(z,t,tec)} & k\EUR \\ \hline
-emission cost & $c^{e}_{z,t,i}$ & \texttt{cost\_emission(z,t,tec)} & k\EUR \\ \hline
-total o\&m cost & $c^{om}_{z,i}$ & \texttt{cost\_om(z,tec)} & k\EUR \\ \hline
-capital cost of generators & $c^{g}_{z}$ & \texttt{cost\_invgen(z)} & k\EUR \\ \hline
+fuel burn for energy generation & $b_{z,t,i,f}$ & \texttt{b(z,t,i,f)} & GW \\ \hline
+total system cost & $c$ & \texttt{cost\_system} & k\EUR \\ \hline
+zonal system cost & $c_{z}$ & \texttt{cost\_zonal(z)} & k\EUR \\ \hline
+fuel cost & $c^{b}_{z,t,i}$ & \texttt{cost\_fuel(z,t,i)} & k\EUR \\ \hline
+emission cost & $c^{e}_{z,t,i}$ & \texttt{cost\_co2(z,t,i)} & k\EUR \\ \hline
+total o\&m cost & $c^{om}_{z,i}$ & \texttt{cost\_om(z,i)} & k\EUR \\ \hline
+capital cost of generators & $c^{g}_{z}$ & \texttt{cost\_invest\_g(z)} & k\EUR \\ \hline
 total cost of non-served load & $c^{q}_{z}$ & \texttt{cost\_nse(z)} & k\EUR \\ \hline
-capital cost of intermittent generators & $c^{r}_{z}$ & \texttt{cost\_invitm(z)} & k\EUR \\ \hline
-capital cost of storages & $c^{s,v}_{z}$ & \texttt{cost\_invstrg(z)} & k\EUR \\ \hline
-capital cost of interconnectors & $c^{x}_{z}$ & \texttt{cost\_gridexpansion(z)} & k\EUR \\ \hline
-\ce{CO2} emissions & $e_{z}$ & \texttt{emissions(z)} & t \ce{CO2} \\ \hline
-added capacity of conventionals & $\bar{g}^{+}_{z,i}$ & \texttt{invest\_thermal(z,tec)} & GW \\ \hline
-decommissioned capacity of conventionals & $\bar{g}^{-}_{z,i}$ & \texttt{decommission(z,tec)} & GW \\ \hline
-energy generated by conventionals & $g_{z,t,i,p}$ & \texttt{q\_gen(z,t,tec,prd)} & GW \\ \hline
+capital cost of intermittent generators & $c^{r}_{z}$ & \texttt{cost\_invest\_r(z)} & k\EUR \\ \hline
+capital cost of storages & $c^{s,v}_{z}$ & \texttt{cost\_invest\_sv(z)} & k\EUR \\ \hline
+capital cost of interconnectors & $c^{x}_{z}$ & \texttt{cost\_invest\_x(z)} & k\EUR \\ \hline
+\ce{CO2} emissions & $e_{z}$ & \texttt{emission\_co2(z)} & t \ce{CO2} \\ \hline
+added capacity of dispatchables & $\widetilde{g}^{+}_{z,i}$ & \texttt{add\_g(z,i)} & GW \\ \hline
+decommissioned capacity of dispatchables & $\widetilde{g}^{-}_{z,i}$ & \texttt{deco\_g(z,i)} & GW \\ \hline
+energy generated by conventionals & $g_{z,t,i,m}$ & \texttt{g(z,t,i,m)} & GW \\ \hline
 curtailed energy & $q^{+}_{z,t}$ & \texttt{q\_curtail(z,t)} & GW \\ \hline
-non-served energy & $q^{-}_{z,t,p}$ & \texttt{q\_nonserved(z,t,prd)} & GW \\ \hline
-added capacity of intermittents & $\bar{r}^{+}_{z,n}$ & \texttt{invest\_res(z,tec\_itm)} & GW \\ \hline
-electricity generated by intermittents & $r_{z,t,n}$ & \texttt{q\_itm(z,t,tec\_itm)} & GW \\ \hline
-added storage capacity (power) & $\bar{s}^{+}_{z,k}$ & \texttt{invest\_storage\_power(z,tec\_strg)} & GW \\ \hline
-energy stored in & $s^{in}_{z,t,k}$ & \texttt{q\_store\_in(z,t,tec\_strg)} & GW \\ \hline
-energy stored out & $s^{out}_{z,t,k}$ & \texttt{q\_store\_out(z,t,tec\_strg)} & GW \\ \hline
-added storage capacity (energy) & $\bar{v}^{+}_{z,k}$ & \texttt{invest\_storage\_energy(z,tec\_strg)} & GWh \\ \hline
-storage energy content & $v_{z,t,k}$ & \texttt{storage\_level(z,t,tec\_strg)} & GWh \\ \hline
-operating region weight & $w_{z,t,i,l}$ & \texttt{cc\_weights(z,t,tec,l)} & . \\ \hline
-added transmission capacity & $\bar{x}^{+}_{z,zz}$ & \texttt{invest\_atc(z,zz)} & GW \\ \hline
-electricity net export & $x_{z,zz,t}$ & \texttt{flow(z,zz,t)} & GW \\ \hline \hline
+non-served energy & $q^{-}_{z,t,m}$ & \texttt{q\_nse(z,t,m)} & GW \\ \hline
+added capacity of intermittents & $\widetilde{r}^{+}_{z,n}$ & \texttt{add\_r(z,n)} & GW \\ \hline
+electricity generated by intermittents & $r_{z,t,n}$ & \texttt{r(z,t,n)} & GW \\ \hline
+added storage capacity (power) & $\widetilde{s}^{+}_{z,k}$ & \texttt{add\_s(z,k)} & GW \\ \hline
+energy stored in & $s^{in}_{z,t,k}$ & \texttt{s\_in(z,t,k)} & GW \\ \hline
+energy stored out & $s^{out}_{z,t,k}$ & \texttt{s\_out(z,t,k)} & GW \\ \hline
+added storage capacity (energy) & $\widetilde{v}^{+}_{z,k}$ & \texttt{add\_v(z,k)} & GWh \\ \hline
+storage energy content & $v_{z,t,k}$ & \texttt{v(z,t,k)} & GWh \\ \hline
+operating region weight & $w_{z,t,i,l}$ & \texttt{w(z,t,i,l)} & \\ \hline
+added transmission capacity & $\widetilde{x}^{+}_{z,zz}$ & \texttt{add\_x(z,zz)} & GW \\ \hline
+electricity net export & $x_{z,zz,t}$ & \texttt{x(z,zz,t)} & GW \\ \hline \hline
 \end{longtable}
@@
 	 &  initial capacity\tmark[$\dagger$] & added capacity\tmark[$\ddagger$] & \makecell[l]{decom-\\missioned \\capacity\tmark[$\ddagger$]} & \makecell[l]{specific \\investment \\cost\tmark[$\dagger$]} & dispatch\tmark[$\ddagger$] \ML
-	thermal units & $\bar{G}_{z,i}$ & $\bar{g}^{+}_{z,i}$ & $\bar{g}^{-}_{z,i}$ & $C^{g}_{z,i}$ & $g_{z,t,i,p}$ \ML
-	\makecell[l]{intermittent \\units} &  $\bar{R}_{z,n}$ & $\bar{r}^{+}_{z,n}$ & $\bar{r}^{-}_{z,n}$ & $C^{r}_{z,n}$ & $r_{z,t,n}$ \ML
-	\makecell[l]{storages \\(power)} & $\bar{S}_{z,k}$ & $\bar{s}^{+}_{z,k}$ & $\bar{s}^{-}_{z,k}$ & $C^{s}_{z,k}$ & $s_{z,t,k}$ \ML
-	\makecell[l]{storages \\(energy)} & $\bar{V}_{z,k}$ & $\bar{v}^{+}_{z,k}$ & $\bar{v}^{-}_{z,t,k}$ & $C^{v}_{z,k}$ & na \ML
-	transmission &  $\bar{X}_{z,zz}$ & $\bar{x}^{+}_{z,zz}$ & $\bar{x}^{-}_{z,zz}$ & $C^{x}_{z,zz}$ & $x_{z,zz,t}$ \LL
+	thermal units & $\widetilde{G}_{z,i}$ & $\widetilde{g}^{+}_{z,i}$ & $\widetilde{g}^{-}_{z,i}$ & $C^{g}_{z,i}$ & $g_{z,t,i,m}$ \ML
+	\makecell[l]{intermittent \\units} &  $\widetilde{R}_{z,n}$ & $\widetilde{r}^{+}_{z,n}$ & $\widetilde{r}^{-}_{z,n}$ & $C^{r}_{z,n}$ & $r_{z,t,n}$ \ML
+	\makecell[l]{storages \\(power)} & $\widetilde{S}_{z,k}$ & $\widetilde{s}^{+}_{z,k}$ & $\widetilde{s}^{-}_{z,k}$ & $C^{s}_{z,k}$ & $s_{z,t,k}$ \ML
+	\makecell[l]{storages \\(energy)} & $\widetilde{V}_{z,k}$ & $\widetilde{v}^{+}_{z,k}$ & $\widetilde{v}^{-}_{z,t,k}$ & $C^{v}_{z,k}$ & na \ML
+	transmission &  $\widetilde{X}_{z,zz}$ & $\widetilde{x}^{+}_{z,zz}$ & $\widetilde{x}^{-}_{z,zz}$ & $C^{x}_{z,zz}$ & $x_{z,zz,t}$ \LL
 }
@@
 &c^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: \varepsilon_{f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i\\
-&c^{om}_{z,i}& =&\ O^{q}_{i} \left(\bar{G}_{z,i} - \bar{g}^{-}_{z,i} + \bar{g}^{+}_{z,i}\right) + \sum_{t,m} \left(O^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
-&c^{g}_{z}& =&\ C^{g}_{z,i} \: \bar{g}^{+}_{z,i} \qquad \qquad &\forall z\\
-&c^{r}_{z}& =&\ C^{r}_{z,n} \: \bar{r}^{+}_{z,n} \qquad \qquad &\forall z\\
-&c^{s,v}_{z}& =&\ C^{s}_{z,k} \: \bar{s}^{+}_{z,k} + C^{v}_{z,k} \: v^{+}_{z,k} \qquad \qquad &\forall z\\
-&c^{x}_{z}& =&\ \frac{1}{2} \: \sum_{zz} (C^{x}_{z} \: \delta_{z,zz} \: \bar{x}^{+}_{z,zz}) \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
+&c^{om}_{z,i}& =&\ O^{q}_{i} \left(\widetilde{G}_{z,i} - \widetilde{g}^{-}_{z,i} + \widetilde{g}^{+}_{z,i}\right) + \sum_{t,m} \left(O^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
+&c^{g}_{z}& =&\ \sum_{i} \left( C^{g}_{z,i} \: \widetilde{g}^{+}_{z,i} \right) \qquad \qquad &\forall z\\
+&c^{r}_{z}& =&\ \sum_{n} \left( C^{r}_{z,n} \: \widetilde{r}^{+}_{z,n} \right) \qquad \qquad &\forall z\\
+&c^{s,v}_{z}& =&\ \sum_{k} \left( C^{s}_{z,k} \: \widetilde{s}^{+}_{z,k} + C^{v}_{z,k} \: v^{+}_{z,k} \right) \qquad \qquad &\forall z\\
+&c^{x}_{z}& =&\ \frac{1}{2} \: \sum_{zz} (C^{x} \: \delta_{z,zz} \: \widetilde{x}^{+}_{z,zz}) \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
 &c^{q}_{z}& =&\ \mu \sum_{t,m} q^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
@@
 \begin{align}
-\sum_{i} g_{z,t,i,\text{el}} + \sum_{k} s^{out}_{z,t,k} + \sum_{n} r_{z,t,n} =
-D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} 
-- q^{-}_{z,t,\text{el}} + q^{+}_{z,t} \qquad \forall z,t \label{market_clearing_el} \\
+\begin{split}
+\sum_{i} g_{z,t,i,\text{el}} + \sum_{k} s^{out}_{z,t,k} + \sum_{n} r_{z,t,n} &=  \\ D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + & \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} - q^{-}_{z,t,\text{el}} + q^{+}_{z,t} \qquad \forall z,t
+\end{split}
+\label{market_clearing_el}
+\end{align} 
+\begin{align}
 \sum_{i} g_{z,t,i,\text{ht}} = D_{z,t,\text{ht}} - q^{-}_{z,t,\text{ht}} \qquad \forall z,t \label{market_clearing_ht}
@@
 
-\paragraph{Thermal electricity generation}
+\paragraph{Energy generation}
 Energy generation is constrained by available installed capacity, which can be adjusted through investment and decommissioning.
 \begin{align}
-g_{z,t,i,m} \leq \bar{G}_{z,i} + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \qquad \qquad \forall z,t,i,m
+g_{z,t,i,m} \leq \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} - \widetilde{g}^{-}_{z,i} \qquad \qquad \forall z,t,i,m
 \end{align}
-Generator efficiency $\eta$ determines the amount of fuel that needs to be burnt in order to generate a given amount of electricity.
+Generator efficiency $\eta$ determines the amount of fuel that needs to be spent in order to generate a given amount of energy.
 \begin{align}
-g_{z,t,i,\text{el}} = \sum_{f} \eta_{i,\text{el},f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \notin J
-\end{align}
-
-\paragraph{Electric heat generation}
-Heat pumps and electric boilers use electricity to generate heat.
-The amount of electricity required to generate a given amount of heat depends on the unit's efficiency $\eta$.
-For heat pumps, the annual COP $\eta$ is assumed to be $3$.
-\begin{align}
-g_{z,t,i,\text{ht}} = \sum_{f} \eta_{i,\text{ht},f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \in H
+g_{z,t,i,m} = \sum_{f} \eta_{i,m,f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \notin J
 \end{align}
@@
 \begin{align}
-\sum_{l} w_{z,t,i,l} = \bar{G}_{z,i} + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \qquad \qquad \forall z,t,i \in J \\
+\sum_{l} w_{z,t,i,l} = \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} - \widetilde{g}^{-}_{z,i} \qquad \qquad \forall z,t,i \in J \\
 g_{z,t,i,m} \leq \sum_{l} \psi_{i,l,m} \: w_{z,t,i,l} \qquad \qquad \forall z,t,i \in J, m \\
@@
 \begin{align}
-r_{z,t,n} = \phi_{z,t,n} \: \left( \bar{R}_{z,n} + \bar{r}^{+}_{z,n} \right) \qquad \qquad \forall z,t,n
+r_{z,t,n} = \phi_{z,t,n} \: \left( \widetilde{R}_{z,n} + \widetilde{r}^{+}_{z,n} \right) \qquad \qquad \forall z,t,n
 \end{align}
@@
 \paragraph{Electricity storages}
-Charging and discharging of storages is constrained by the storages' power capacity $\bar{s}^{in}_{z,k}$ and $\bar{s}^{out}_{z,k}$, respectively.
-Similarly, the total energy that can be stored is constrained by the storage technology's energy capacity $\bar{v}_{z,k}$.
+Charging and discharging of storages is constrained by the storages' power capacity $\widetilde{s}^{in}_{z,k}$ and $\widetilde{s}^{out}_{z,k}$, respectively.
+Similarly, the total energy that can be stored is constrained by the storage technology's energy capacity $\widetilde{v}_{z,k}$.
 \begin{align}
-s^{out}_{z,t,k} \leq \bar{S}^{out}_{z,k} + \bar{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
-s^{in}_{z,t,k} \leq \bar{S}^{in}_{z,k} + \bar{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
-v_{z,t,k} \leq \bar{V}_{z,k} + \bar{v}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
+s^{out}_{z,t,k} \leq \widetilde{S}^{out}_{z,k} + \widetilde{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
+s^{in}_{z,t,k} \leq \widetilde{S}^{in}_{z,k} + \widetilde{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
+v_{z,t,k} \leq \widetilde{V}_{z,k} + \widetilde{v}^{+}_{z,k} \qquad \qquad \forall z,t,k
 \end{align}
@@
 \begin{align}
-\bar{v}^{+}_{z,k} \geq \bar{s}^{+}_{z,k} \qquad \qquad \forall z,k
+\widetilde{v}^{+}_{z,k} \geq \widetilde{s}^{+}_{z,k} \qquad \qquad \forall z,k
 \end{align}
@@
 \begin{align}
-x_{z,zz,t} \leq \bar{X}_{z,zz} + \bar{x}^{+}_{z,zz} \qquad \qquad \forall z, zz, t \\
-x_{z,zz,t} \geq - \left( \bar{X}_{z,zz} + \bar{x}^{+}_{z,zz} \right) \qquad \qquad \forall z, zz, t
+x_{z,zz,t} \leq \widetilde{X}_{z,zz} + \widetilde{x}^{+}_{z,zz} \qquad \qquad \forall z, zz, t \\
+x_{z,zz,t} \geq - \left( \widetilde{X}_{z,zz} + \widetilde{x}^{+}_{z,zz} \right) \qquad \qquad \forall z, zz, t
 \end{align}
@@
 \begin{align}
-\bar{x}^{+}_{z,zz} = \bar{x}^{+}_{zz,z} \qquad \qquad \forall z, zz
+\widetilde{x}^{+}_{z,zz} = \widetilde{x}^{+}_{zz,z} \qquad \qquad \forall z, zz
 \end{align}
@@
 \begin{align}
-x_{z,zz,t} = 0 \qquad \qquad \forall z \notin NTC, zz \notin NTC, t \\
-x_{zz,z,t} = 0 \qquad \qquad \forall z \notin NTC, zz \notin NTC, t
+x_{z,zz,t} = 0 \qquad \qquad \forall z, zz, t: \widetilde{X}_{z,zz} = 0 \\
+x_{zz,z,t} = 0 \qquad \qquad \forall z, zz, t: \widetilde{X}_{z,zz} = 0
 \end{align}
@@
 \begin{align}
-\bar{g}^{-}_{z,i} \leq \bar{G}_{z,i} + \bar{g}^{+}_{z,i} \qquad \qquad \forall z,i
+\widetilde{g}^{-}_{z,i} \leq \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} \qquad \qquad \forall z,i
 \end{align}
@@
 We assume that ancillary services are supplied by conventional (thermal) power plants, hydro power plants, and storages.
+The requirement for spinning reserves is proportional to electricity peak load $\widehat{D}_{z,\text{el}} = \max_{t} D_{z,t,\text{el}}$ and peak generation from wind and solar resources, where $\widehat{\phi}_{z,n} = \max_{t} \phi_{z,t,n}$.
 \begin{align}
-g_{z,t,i,\text{el}} + r_{z,t,\text{ror}} + s^{out}_{z,t,k} + s^{in}_{z,t,k} \geq \sigma_{z} \qquad \qquad \forall z,t,i,k
+\sum_{i}\left( g_{z,t,i,\text{el}} \right) + r_{z,t,\text{ror}} + \sum_{k} \left( s^{out}_{z,t,k} + s^{in}_{z,t,k} \right) \geq \lambda_{z} \widehat{D}_{z,\text{el}} + \sigma_{z} \sum_{n \setminus \{ \text{ror}\}} \widehat{\phi}_{z,n} (\widetilde{R}_{z,n} + \widetilde{r}^{+}_{z,n}) \qquad \forall z,t
 \end{align}

### 2019-09-09 medea/instantiate_gdx.py
consistent naming across GAMS and gdx instantiation

@@ -37,26 +37,26 @@ logging.info('medea sets instantiated')
 # --------------------------------------------------------------------------- #
 # %% instantiate static PARAMETERS
 # --------------------------------------------------------------------------- #
-CAPITALCOST_G = df2gdx(db, static_data['i']['annuity'].round(4),
+CAPITALCOST_G = df2gdx(db, static_data['tec']['annuity'].round(4),
                             'CAPITALCOST_G', 'par', [i_set], '[kEUR per GW]')
 CAPITALCOST_R = df2gdx(db, static_data['CAPCOST_R'].stack().reorder_levels([1, 0]).round(4), 'CAPITALCOST_R', 'par',
                        [z_set, n_set], '[kEUR per GW]')
-CAPITALCOST_S = df2gdx(db, plant_data['storage_clusters']['cost_power'].reorder_levels((1, 0)), 'par',
-                           [z_set, k_set], '[GW]')
-CAPITALCOST_V = df2gdx(db, plant_data['storage_clusters']['cost_energy'].reorder_levels((1, 0)), 'par',
-                           [z_set, k_set], '[GW]')
+CAPITALCOST_S = df2gdx(db, plant_data['storage_clusters']['cost_power'].reorder_levels((1, 0)), 'CAPITALCOST_S', 'par',
+                       [z_set, k_set], '[GW]')
+CAPITALCOST_V = df2gdx(db, plant_data['storage_clusters']['cost_energy'].reorder_levels((1, 0)), 'CAPITALCOST_V', 'par',
+                       [z_set, k_set], '[GW]')
 CAPITALCOST_X = df2gdx(db, static_data['CAPCOST_X'], 'CAPITALCOST_X', 'par', [z_set], 'kEUR per GW')
 CO2_INTENSITY = df2gdx(db, dict_instantiate['CO2_INTENSITY'], 'CO2_INTENSITY', 'par', [f_set],
                        '[kt CO2 per GWh fuel input]')
 DEMAND = df2gdx(db, ts_data['zonal'].loc[:, idx[:, :, 'load']].stack((0, 1)).reorder_levels((1, 0, 2)).round(4),
                 'DEMAND', 'par', [z_set, t_set, m_set])
[…]

### 2019-09-09 medea/prepare_data.py

@@ -60,7 +60,7 @@ dict_sets = {
         'Heat': [110]
     },
     'l': {f'l{x}': [True] for x in range(1, 5)},
-    'prd': {
+    'm': {
         'el': True,
         'ht': True
     },
@@ -140,7 +140,9 @@ dict_static = {
     'VALUE_NSE': {
         'AT': [12500],
         'DE': [12500]
-    }
+    },
+    'LAMBDA': [0.125],
+    'SIGMA': [0.175]
 }
 
 dict_additions = {
@@ -201,6 +203,8 @@ dict_instantiate.update({'DISTANCE': static_data['DISTANCE'].loc[static_data['DI
 
 static_data.update({'CAPCOST_X': pd.DataFrame.from_dict(dict_static['CAPCOST_X'], orient='index', columns=['Value'])})
 static_data.update({'VALUE_NSE': pd.DataFrame.from_dict(dict_static['VALUE_NSE'], orient='index', columns=['Value'])})
+static_data.update({'LAMBDA': pd.DataFrame(dict_static['LAMBDA'], columns=['Value'])})
+static_data.update({'SIGMA': pd.DataFrame(dict_static['SIGMA'], columns=['Value'])})
 
 # --------------------------------------------------------------------------- #
 # %% preprocessing plant data
@@ -371,6 +375,9 @@ dict_instantiate.update({'PEAK_LOAD': ts_data['zonal'].loc[:, idx[:, 'el', 'load
 dict_instantiate.update({'PEAK_PROFILE': ts_data['zonal'].loc[:, idx[:, :, 'profile']].max().unstack(2).drop(
     'ror', axis=0, level=1)})
 
+# drop rows with all zeros
+plant_data['storage_clusters'] = \
+    plant_data['storage_clusters'].loc[~(plant_data['storage_clusters'] == 0).all(axis=1), :].copy()
 
 # --------------------------------------------------------------------------- #
 # %% limits on investment - long-run vs short-run & # TODO: potentials

### 2019-09-10 doc/mathematical_description.tex
consistent naming across GAMS and gdx instantiation
further improvements in model consistency, including application generation

@@
 % LOWER-CASE GREEK LETTERS
-	distance between countries & $\delta_{z,zz}$ & \texttt{DISTANCE(z,zz)} & km \\ \hline
+	distance between countries' center of gravity & $\delta_{z,zz}$ & \texttt{DISTANCE(z,zz)} & km \\ \hline
 	fuel emission intensity & $\varepsilon_{f}$ & \texttt{CO2\_INTENSITY(f)} & $\text{t}_{\ce{CO2}}$ / MWh \\ \hline
 	power plant efficiency & $\eta_{i,m,f}$ & \texttt{EFFICIENCY\_G(i,m,f)} & MWh / MWh \\ \hline
-	efficiency power out & $\eta^{out}_{z,k}$ & \makecell[l]{\texttt{EFFICIENCY\_S\_OUT(k)}} &  \\ \hline
-	efficiency power in & $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{EFFICIENCY\_S\_IN(k)}} &  \\ \hline	
+	discharging efficiency & $\eta^{out}_{z,k}$ & \makecell[l]{\texttt{EFFICIENCY\_S\_OUT(k)}} &  \\ \hline
+	charging efficiency & $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{EFFICIENCY\_S\_IN(k)}} &  \\ \hline	
 	scaling factor for peak load & $\lambda_{z}$ & \texttt{LAMBDA(z)} &  \\ \hline
 	value of lost load & $\mu_{z}$ & \texttt{VALUE\_NSE(z)} & \EUR / MWh \\ \hline
-	reservoir inflows & $\rho_{z,t,k}$ & \makecell[l]{\texttt{INFLOWS(z,t,k)}} & MW \\ \hline	
+	inflows to storage reservoirs & $\rho_{z,t,k}$ & \makecell[l]{\texttt{INFLOWS(z,t,k)}} & MW \\ \hline	
 	scaling factor for peak intermittent generation & $\sigma_{z}$ & \texttt{SIGMA(z)} &  \\ \hline
@@
 	capital cost of storages - energy (specific, annuity) & $C^{v}_{z,k}$ & \makecell[l]{\texttt{CAPITALCOST\_V(z,k)}} & k\EUR / GW \\ \hline
-	capital cost of transmission capacity & $C^{x}$ & \texttt{CAPITALCOST\_X} & k\EUR / GW \\ \hline
+	capital cost of transmission capacity (specific, annuity) & $C^{x}$ & \texttt{CAPITALCOST\_X} & k\EUR / GW \\ \hline
 	energy demand & $D_{z,t,m}$ & \texttt{DEMAND(z,t,m)} & GW \\ \hline
@@
 	initial capacity of intermittent generators & $\widetilde{R}_{z,n}$ & \texttt{INITIAL\_CAP\_R(z,n)} & GW \\ \hline
-	max power out & $\widetilde{S}^{out}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_S\_OUT(z,k)}} & GW \\ \hline
-	max power in & $\widetilde{S}^{in}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_S\_IN(z,k)}} & GW \\ \hline
-	max energy stored & $\widetilde{V}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_V(z,k)}} &  \\ \hline
-	installed available transmission capacity & $\widetilde{X}_{z,zz}$ & \texttt{INITIAL\_CAP\_X(z,zz)} & GW \\ \hline \hline	
+	initial discharging capacity of storages & $\widetilde{S}^{out}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_S\_OUT(z,k)}} & GW \\ \hline
+	initial charging capacity of storages & $\widetilde{S}^{in}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_S\_IN(z,k)}} & GW \\ \hline
+	initial energy storage capacity & $\widetilde{V}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_V(z,k)}} &  \\ \hline
+	initial transmission capacity & $\widetilde{X}_{z,zz}$ & \texttt{INITIAL\_CAP\_X(z,zz)} & GW \\ \hline \hline	
 \end{longtable}
@@
 \paragraph{Energy generation}
-Energy generation is constrained by available installed capacity, which can be adjusted through investment and decommissioning.
+Energy generation $g_{z,t,i,m} \geq 0$ is constrained by available installed capacity, which can be adjusted through investment ($\widetilde{g}^{+}_{z,i} \geq 0$) and decommissioning $\widetilde{g}^{-}_{z,i} \geq 0$.
 \begin{align}
@@
 \end{align}
-Generator efficiency $\eta$ determines the amount of fuel that needs to be spent in order to generate a given amount of energy.
+Generator efficiency $\eta$ determines the amount of fuel $b_{z,t,i,f} \geq 0$ that needs to be spent in order to generate a given amount of energy.
 \begin{align}
-g_{z,t,i,m} = \sum_{f} \eta_{i,m,f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \notin J
+g_{z,t,i,m} \leq \sum_{f} \eta_{i,m,f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \notin J
 \end{align}
@@
 The elements $l \in L$ span up a three-dimensional, convex feasible operating region for each co-generation technology. 
-The weights $w$ form a convex combination of the corners $l$, which are scaled to the available installed capacity of each co-generation technology.
+The weights $w_{z,t,i,l} \geq 0$ form a convex combination of the corners $l$, which are scaled to the available installed capacity of each co-generation technology.
 Heat and electricity output along with the corresponding fuel requirement is then set according to the chosen weights.
@@
 \paragraph{Intermittent electricity generation}
-Electricity generation from intermittent sources wind (on-shore and off-shore), solar irradiation, and river runoff follows generation profiles $\phi_{z,t,n} \in [0,1]$ and is scaled according to corresponding installed and added capacity.
+Electricity generation from intermittent sources wind (on-shore and off-shore), solar irradiation, and river runoff follows generation profiles $\phi_{z,t,n} \in [0,1]$ and is scaled according to corresponding installed ($ \widetilde{R}_{z,n}$) and added ($\widetilde{r}^{+}_{z,n} \geq 0$) capacity.
 \begin{align}
@@
 \paragraph{Electricity storages}
-Charging and discharging of storages is constrained by the storages' power capacity $\widetilde{s}^{in}_{z,k}$ and $\widetilde{s}^{out}_{z,k}$, respectively.
-Similarly, the total energy that can be stored is constrained by the storage technology's energy capacity $\widetilde{v}_{z,k}$.
+Charging ($s^{in}_{z,t,k} \geq 0$) and discharging ($s^{out}_{z,t,k} \geq 0$) of storages is constrained by the storages' installed ($\widetilde{S}^{in}_{z,k}, \widetilde{S}^{out}_{z,k}$) and added ($\widetilde{s}^{+}_{z,k} \geq 0$) charging and discharging power, respectively.
+Similarly, the total energy that can be stored is constrained by the storage technology's initial ($\widetilde{V}_{z,k}$) and added ($\widetilde{v}^{+}_{z,k} \geq 0$) energy capacity.
 \begin{align}
@@
 \begin{align}
-v_{z,t,k} = \rho_{z,t,k} + \eta^{in}_{z,k} \: s^{in}_{z,t,k} - (\eta^{out}_{z,k})^{-1} \: s^{out}_{z,t,k} + v_{z,t-1,k} \qquad \qquad \forall z,t,k
+v_{z,t,k} = \rho_{z,t,k} + \eta^{in}_{z,k} \: s^{in}_{z,t,k} - (\eta^{out}_{z,k})^{-1} \: s^{out}_{z,t,k} + v_{z,t-1,k} \qquad \qquad \forall z,t,k: t>1, \: \eta^{out}_{z,k} > 0
 \end{align}

### 2019-09-10 applications/wind_vs_pv/results_wind_vs_pv.py

@@ -80,4 +80,4 @@ for peua in eua_range:
     df_all = df_all.dropna(axis=1, how='all')
     df_all = df_all.sort_index(axis=1, level=1)
     df_all.to_csv(os.path.join(cfg.folder, 'applications', project, 'results',
-                               f'out_costreal_{scenario}_EUA_{peua}.csv'), sep=';', encoding='utf-8', decimal=',')
+                               f'out_costopus_{scenario}_EUA_{peua}.csv'), sep=';', encoding='utf-8', decimal=',')

### 2019-09-10 medea/prepare_data.py

@@ -239,7 +239,11 @@ for zone in cfg.zones:
     tec_props.loc[:, 'eta'].update(pd.DataFrame.from_dict(dict_static['eta'], orient='index', columns=[zone]),
                                    overwrite=False)
 # index by plant element names instead of medea_type-numbers
-tec_props.index = static_data['tec'].loc[static_data['tec']['medea_type'].isin(tec_props.index), 'set_element'].values
+# TODO: massive error when matching medea-numbers to plant names
+type_plant_match = static_data['tec'][['medea_type', 'set_element']].copy()
+type_plant_match.set_index('medea_type', inplace=True)
+tec_props['set_elements'] = type_plant_match.loc[tec_props.index, :].values
+tec_props.set_index('set_elements', inplace=True)
 tec_props = tec_props.stack(-1).swaplevel(axis=0)
 tec_props = tec_props.dropna()
 dict_instantiate.update({'tec_props': tec_props})

### 2019-09-10 medea/templates/results_template.py

@@ -16,8 +16,8 @@ ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
 # specify GAMS symbols to be read from output .gdx-file along with the dimensions of the corresponding output DataFrame
 # example: generate pandas DataFrame holding system cost and co2 emissions for each CO2 price scenario
 symbols_to_read = {
-    'cost': ([], ['r']),
-    'emissions': ([], ['r'])
+    'cost_system': ([], ['z']),
+    'emission_co2': ([], ['z'])
 }
 
 # iterate over all output .gdx-files

### 2019-09-10 medea/templates/run_template.py

@@ -23,18 +23,18 @@ db_input = ws.add_database_from_gdx(os.path.join(cfg.folder, 'applications', pro
 # dict_set = {rec.keys[0] for rec in db_input['set_name']}
 # ---
 # read sets for calibration of power plant efficiencies
-dict_prd = {rec.keys[0] for rec in db_input['prd']}
+dict_prd = {rec.keys[0] for rec in db_input['m']}
 dict_fuel = {rec.keys[0] for rec in db_input['f']}
-dict_tec = {rec.keys[0] for rec in db_input['tec']}
+dict_tec = {rec.keys[0] for rec in db_input['i']}
 
 # *** read in all parameters to be adjusted ***
 # general example for reading parameter 'PARAMETER_NAME' defined over 'set_name' to pandas DataFrame df
 # df = gdx2df(db_input, 'PARAMETER_NAME', ['set_name'], [])
 # ---
 # read power plant efficiency data
-df_eff = gdx2df(db_input, 'EFFICIENCY', ['tec', 'prd', 'f'], [])
+df_eff = gdx2df(db_input, 'EFFICIENCY_G', ['i', 'm', 'f'], [])
 # read fuel requirement of CHP plants-data
-df_fuelreq = gdx2df(db_input, 'FEASIBLE_INPUT', ['tec', 'l', 'f'], [])
+df_fuelreq = gdx2df(db_input, 'FEASIBLE_INPUT', ['i', 'l', 'f'], [])
 
 
 # %% generate 'static' parameter variations (modifications that remain the same across all scenarios)
@@ -49,7 +49,7 @@ df_eff_mod = df_eff.copy()
 for fl in fuel_thermal:
     df_eff_mod.loc[idx[:, :, fl], :] = \
         df_eff.loc[idx[:, :, fl], :] * efficiency[f'e_{fl}'][0]
-reset_parameter(db_input, 'EFFICIENCY', df_eff_mod)
+reset_parameter(db_input, 'EFFICIENCY_G', df_eff_mod)
 
 # modify fuel requirement of co-generation plants
 df_fuelreq_mod = df_fuelreq.copy()
@@ -65,7 +65,7 @@ os.chdir(os.path.join(cfg.folder, 'applications', project_name, 'opt'))
 
 # create empty scenario parameter in GAMS database so that it can be modified subsequently
 # example: changing CO2 price
-scenario_co2 = df2gdx(db_input, pd.DataFrame(data=[0]), 'EUA_SCENARIO', 'par', 0, 'CO2 price scenario')
[…]

### 2019-09-20 doc/mathematical_description.tex
increased model consistency, enabled disinvestment for intermittents

@@
 	\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\ \hline \hline
-	quasi-fixed O\&M cost & $O^{q}_{i}$ & \texttt{OM\_COST\_QFIX(i)} & k\EUR / GW \\ \hline
-	variable O\&M cost & $O^{v}_{i}$ & \texttt{OM\_COST\_VAR(i)} & \EUR / MWh \\ \hline
+	variable O\&M cost of dispatchable generators & $O^{g}_{i}$ & \texttt{OM\_COST\_G\_VAR(i)} & \EUR / MWh \\ \hline
+	variable O\&M cost of intermittent generators & $O^{r}_{z,n}$ & \texttt{OM\_COST\_R\_VAR(z,n)} & \EUR / MWh \\ \hline	
 	\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_CO2(t,z)} & \EUR / $\text{t}_{\ce{CO2}}$ \\ \hline
 	fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & \EUR / MWh \\ \hline
+	quasi-fixed O\&M cost of dispatchable generators & $Q^{g}_{i}$ & \texttt{OM\_COST\_G\_QFIX(i)} & k\EUR / GW \\ \hline
+	quasi-fixed O\&M cost of intermittent generators & $Q^{r}_{z,n}$ & \texttt{OM\_COST\_R\_QFIX(z,n)} & k\EUR / GW \\ \hline	
 	initial capacity of intermittent generators & $\widetilde{R}_{z,n}$ & \texttt{INITIAL\_CAP\_R(z,n)} & GW \\ \hline
@@
 
-\begin{longtable}{p{4.1cm} c l c}
+\begin{longtable}{p{5.6cm} c l c}
 %\caption{Variables}\label{variables} \\ 
@@
 emission cost & $c^{e}_{z,t,i}$ & \texttt{cost\_co2(z,t,i)} & k\EUR \\ \hline
-total o\&m cost & $c^{om}_{z,i}$ & \texttt{cost\_om(z,i)} & k\EUR \\ \hline
+total o\&m cost of dispatchable generators & $c^{om}_{z,i}$ & \texttt{cost\_om\_g(z,i)} & k\EUR \\ \hline
+total o\&m cost of intermittent generators & $c^{om}_{z,n}$ & \texttt{cost\_om\_r(z,n)} & k\EUR \\ \hline
 capital cost of generators & $c^{g}_{z}$ & \texttt{cost\_invest\_g(z)} & k\EUR \\ \hline
@@
 added capacity of intermittents & $\widetilde{r}^{+}_{z,n}$ & \texttt{add\_r(z,n)} & GW \\ \hline
+decommissioned capacity of intermittents & $\widetilde{r}^{-}_{z,n}$ & \texttt{deco\_r(z,n)} & GW \\ \hline
 electricity generated by intermittents & $r_{z,t,n}$ & \texttt{r(z,t,n)} & GW \\ \hline
@@
 	\makecell[l]{intermittent \\units} &  $\widetilde{R}_{z,n}$ & $\widetilde{r}^{+}_{z,n}$ & $\widetilde{r}^{-}_{z,n}$ & $C^{r}_{z,n}$ & $r_{z,t,n}$ \ML
-	\makecell[l]{storages \\(power)} & $\widetilde{S}_{z,k}$ & $\widetilde{s}^{+}_{z,k}$ & $\widetilde{s}^{-}_{z,k}$ & $C^{s}_{z,k}$ & $s_{z,t,k}$ \ML
-	\makecell[l]{storages \\(energy)} & $\widetilde{V}_{z,k}$ & $\widetilde{v}^{+}_{z,k}$ & $\widetilde{v}^{-}_{z,t,k}$ & $C^{v}_{z,k}$ & na \ML
-	transmission &  $\widetilde{X}_{z,zz}$ & $\widetilde{x}^{+}_{z,zz}$ & $\widetilde{x}^{-}_{z,zz}$ & $C^{x}_{z,zz}$ & $x_{z,zz,t}$ \LL
+	\makecell[l]{storages \\(power)} & $\widetilde{S}_{z,k}$ & $\widetilde{s}^{+}_{z,k}$ & na & $C^{s}_{z,k}$ & $s_{z,t,k}$ \ML  % $\widetilde{s}^{-}_{z,k}$
+	\makecell[l]{storages \\(energy)} & $\widetilde{V}_{z,k}$ & $\widetilde{v}^{+}_{z,k}$ & na & $C^{v}_{z,k}$ & na \ML  % $\widetilde{v}^{-}_{z,t,k}$
+	transmission &  $\widetilde{X}_{z,zz}$ & $\widetilde{x}^{+}_{z,zz}$ & na & $C^{x}_{z,zz}$ & $x_{z,zz,t}$ \LL  % $\widetilde{x}^{-}_{z,zz}$
 }
@@
 \end{align}
-Zonal system costs $c_{z}$ are the sum of fuel cost $c^{b}_{z,t,i}$, emission cost $c^{e}_{z,t,i}$, operation and maintenance cost $c^{om}_{z,i}$, capital costs of investment in conventional and intermittent generation ($c^{g}_{z}$, $c^{r}_{z}$), storage ($c^{s,v}_{z}$) and transmission ($c^{x}_{z}$) equipment, and the cost of non-served load ($c^{q}_{z}$) that accrues when demand is not met, e.g.
+Zonal system costs $c_{z}$ are the sum of fuel cost $c^{b}_{z,t,i}$, emission cost $c^{e}_{z,t,i}$, operation and maintenance cost, capital costs of investment in conventional and intermittent generation ($c^{g}_{z}$, $c^{r}_{z}$), storage ($c^{s,v}_{z}$) and transmission ($c^{x}_{z}$) equipment, and the cost of non-served load ($c^{q}_{z}$) that accrues when demand is not met, e.g.
 when there is a power outage.
@@
 %\begin{split}
-c_{z} = \sum_{t,i}  c^{b}_{z,t,i} + \sum_{t,i} c^{e}_{z,t,i} + \sum_{i} c^{om}_{z,i} + c^{g}_{z} + 
+c_{z} = \sum_{t,i}  c^{b}_{z,t,i} + \sum_{t,i} c^{e}_{z,t,i} + \sum_{i} c^{om}_{z,i} + \sum_{n} c^{om}_{z,n} + c^{g}_{z} + 
  c^{r}_{z} + c^{s,v}_{z} + c^{x}_{z} + c^{q}_{z} \qquad \qquad \forall z
@@
 &c^{b}_{z,t,i}& =&\ \sum_{f} \left( P_{t,z,f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i \label{fuel_cost} \\
-&c^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: \varepsilon_{f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i\\
-&c^{om}_{z,i}& =&\ O^{q}_{i} \left(\widetilde{G}_{z,i} - \widetilde{g}^{-}_{z,i} + \widetilde{g}^{+}_{z,i}\right) + \sum_{t,m} \left(O^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
+&c^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: e_{z,t,i} \right) \qquad \qquad &\forall z,t,i\\
+&c^{om}_{z,i}& =&\ Q^{g}_{i} \left(\widetilde{G}_{z,i} - \widetilde{g}^{-}_{z,i} + \widetilde{g}^{+}_{z,i}\right) + \sum_{t,m} \left(O^{g}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
+&c^{om}_{z,n}& =&\ Q^{r}_{n} \left(\widetilde{R}_{z,n} - \widetilde{r}^{-}_{z,n} + \widetilde{r}^{+}_{z,n}\right) + \sum_{t} \left(O^{r}_{n} \: r_{z,t,n}\right) \qquad \qquad &\forall z,n \\
 &c^{g}_{z}& =&\ \sum_{i} \left( C^{g}_{z,i} \: \widetilde{g}^{+}_{z,i} \right) \qquad \qquad &\forall z\\
@@
 \begin{align}
-r_{z,t,n} = \phi_{z,t,n} \: \left( \widetilde{R}_{z,n} + \widetilde{r}^{+}_{z,n} \right) \qquad \qquad \forall z,t,n
+r_{z,t,n} = \phi_{z,t,n} \: \left( \widetilde{R}_{z,n} - \widetilde{r}^{-}_{z,n} + \widetilde{r}^{+}_{z,n} \right) \qquad \qquad \forall z,t,n
 \end{align}
@@
 \widetilde{v}^{+}_{z,k} \geq \widetilde{s}^{+}_{z,k} \qquad \qquad \forall z,k
+\end{align}
+
+\paragraph{Emission accounting}
+Burning fossil fuels for energy generation produces emissions of carbon dioxide (\ce{CO2}). The amount of \ce{CO2} emitted is tracked by the following equation
+\begin{align}
+e_{z,t,i} = \sum_{f} \left( \varepsilon_{f} \: b_{z,t,i,f} \right) \qquad \qquad \forall z,t,i
 \end{align}
@@
 
-\paragraph{Decommissioning of thermal units}
-Keeping a plant available for dispatch gives rise to quasi-fixed operation and maintenance costs. 
+\paragraph{Decommissioning}
+Keeping plant available for generation gives rise to quasi-fixed operation and maintenance costs. 
 Such cost can be avoided by decommissioning an energy generator. This is modelled as a reduction in generation capacity, which cannot exceed installed capacity.
 \begin{align}
-\widetilde{g}^{-}_{z,i} \leq \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} \qquad \qquad \forall z,i
+\widetilde{g}^{-}_{z,i} &\leq \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} \qquad \qquad \forall z,i \\
+\widetilde{r}^{-}_{z,n} &\leq \widetilde{R}_{z,n} + \widetilde{r}^{+}_{z,n} \qquad \qquad \forall z,n
 \end{align}

### 2019-09-20 applications/pass_through/ggplots_pass_through.R

@@ -1,303 +1,0 @@
-# Title     : ggplots_pass_through
-# Objective : Create publication-ready plots for pass-through paper using ggplot2
-# Created by: Sebastian
-# Created on: 02.06.2
-
-library(tidyr)
-library(ggplot2)
-library(readxl)
-library(RColorBrewer)
-
-df_path <- 'D:/git_repos/medea/applications/pass_through'
-
-# redefine scale
-ggplot <- function(...) ggplot2::ggplot(...) +
-    scale_color_brewer(palette = "Accent") +
-    scale_fill_brewer(palette = "Accent")
-
-# redefine theme
-theme_set(theme_bw())
-
-# -------------------------------------------------------------------------------------------
-# FIGURE 2
-
-df_base_Figure2 <- read_xlsx(file.path(df_path, 'ActGenByFuel_2017.xlsx'))
-df_base_Figure2 <- df_base_Figure2[c(1, 2), 1 : 7]
-names(df_base_Figure2) <- c("Source", "Nuclear", "Lignite", "Coal", "Gas", "Oil", "Biomass")
-
-barz = c(
-'#70B22F',
-'#8249F4')
-
-df_base_Figure2 %>%
-    gather(Technology, Generation, - Source) %>%
-    ggplot(aes(x = Technology, y = Generation)) +
-    geom_bar(stat = "identity", aes(col = Source, fill = Source), position = position_dodge()) +
-    scale_fill_manual(values = barz) +
-    xlab("Generation Technology") +
-    ylab("Generation [TWh/a]")
-ggsave(file.path(df_path, 'Figure2_calibration.png'), width = 6.6, height = 3.7125, dpi = 600)
-
-# -------------------------------------------------------------------------------------------
-# data preparation
-
-df_base <- read_xlsx(file.path(df_path, 'PsTru_Base_compile.xlsx'), sheet = 'price')
-df_NImp075 <- read_xlsx(file.path(df_path, 'PsTru_in_NImp075_compile.xlsx'), sheet = 'price')
-df_NImp125 <- read_xlsx(file.path(df_path, 'PsTru_in_NImp125_compile.xlsx'), sheet = 'price')
-df_Coal075 <- read_xlsx(file.path(df_path, 'PsTru_in_Coal075_compile.xlsx'), sheet = 'price')
[…]

### 2019-09-20 applications/pass_through/results_pass_through.py

@@ -1,58 +1,0 @@
-import os
-
-import pandas as pd
-from gams import *
-
-import config as cfg
-from applications.pass_through.settings_pass_through import project, eua_range, capacity_scenario_name, output_naming
-from medea.gams_io import gdx2df
-
-# %% Preparations & Settings
-ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
-
-read_variables = {
-    'cost': ([], ['r']),
-    'emissions': ([], ['r']),
-    'invest_res': (['tec_itm'], ['r']),
-    'invest_thermal': (['tec'], ['r']),
-    'invest_storage_power': (['tec_strg'], ['r']),
-    'invest_storage_energy': (['tec_strg'], ['r']),
-    'invest_ntc': (['r'], ['r']),
-    'annual_price_el': ([], ['r']),
-    # 'annual_price_ht': ([], ['r']),
-    'annual_pumping': ([], ['r']),
-    'annual_turbining': ([], ['r']),
-    'annual_generation': (['prd'], ['r']),
-    # 'annual_curtail': ([], ['r']),
-    'annual_netflow': ([], ['r']),
-    'producer_surplus': ([], ['r']),
-    'hourly_price_el': (['t'], ['r'])
-}
-
-df_all = pd.DataFrame()
-for scenario in capacity_scenario_name:
-
-    for peua in eua_range:
-        filename = f'medea_out_{output_naming}.gdx'.format(scenario, peua)
-        db_output = ws.add_database_from_gdx(os.path.join(cfg.folder, 'applications', project, 'opt', filename))
-
-        result_dict = {key: gdx2df(db_output, key, value[0], value[1])
-                       for key, value in read_variables.items()}
-
-        df = pd.DataFrame(columns=cfg.zones)
-        for key, value in result_dict.items():
-            if value.index.any() == 'Value':
-                value.index = [key]
-            df = df.append(value)
-#        df.columns = pd.MultiIndex.from_product([[f'WOn_{wind_limit}'], df.columns])
-#        df_all = pd.concat([df_all, df], axis=1)
-
-### plotting function
-# convert t1 to t8760 back to time units
-
-
-"""
-df_all = df_all.sort_index(axis=1, level=1)
-df_all.to_csv(os.path.join(cfg.folder, 'applications', project, 'output',
-                           f'out_{scenario}_{peua}.csv'), sep=';', encoding='utf-8')
-"""

### 2019-09-20 applications/pass_through/run_pass_through.py

@@ -1,115 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-use medea to estimate emission cost pass-through to electricity prices in the current and future electricity systems of Germany and Austria
-@author: sweer
-"""
-
-import os
-import subprocess
-from shutil import copyfile
-
-import pandas as pd
-from gams import *
-
-import config as cfg
-from applications.pass_through.settings_pass_through import *
-from medea.gams_io import reset_parameter, gdx2df, df2gdx
-
-# --------------------------------------------------------------------------- #
-# %% initialize GAMS, GAMS workspace and load model data
-# --------------------------------------------------------------------------- #
-# fetch main gams model, if required
-if not os.path.isfile(os.path.join(cfg.folder, 'applications', project, 'opt', 'medea_main.gms')):
-    copyfile(os.path.join(cfg.folder, 'medea', 'medea_main.gms'),
-             os.path.join(cfg.folder, 'applications', project, 'opt', 'medea_main.gms'))
-# fetch main gams data, if required
-if not os.path.isfile(os.path.join(cfg.folder, 'applications', project, 'opt', 'medea_main_data.gdx')):
-    copyfile(os.path.join(cfg.folder, 'medea', 'data', 'input', 'medea_main_data.gdx'),
-             os.path.join(cfg.folder, 'applications', project, 'opt', 'medea_main_data.gdx'))
-
-ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
-
-db_input = ws.add_database_from_gdx(os.path.join(cfg.folder, 'applications', project, 'opt', 'medea_main_data.gdx'))
-data_yr = db_input['year'].first_record().value
-
-# read sets for clusters, hydro storage plants and products from db_input
-clust_dict = {rec.keys[0] for rec in db_input['tec']}
-chp_dict = {rec.keys[0] for rec in db_input['tec_chp']}
-itm_dict = {rec.keys[0] for rec in db_input['tec_itm']}
-hsp_dict = {rec.keys[0] for rec in db_input['props']}
-prd_dict = {rec.keys[0] for rec in db_input['prd']}
-fuel_dict = {rec.keys[0] for rec in db_input['f']}
-
-# read parameters
[…]

### 2019-09-20 applications/pass_through/settings_pass_through.py

@@ -1,51 +1,0 @@
-# ---------------------------------------------------------------
-# %% Settings & Scenario Assumptions
-project = 'pass_through'
-fuels = ['Nuclear', 'Lignite', 'Gas', 'Oil', 'Coal', 'Biomass']
-techs = ['bio', 'coal', 'lig', 'ng', 'nuc', 'oil', 'wind_on', 'wind_off', 'pv']
-
-capacity_scenario_name = ['cap_avail', 'cap_inst', 'cap_2030']
-# naming of output gdx
-output_naming = '{}_eua{}'
-
-# ---------------------------------------------------------------
-# Scenario Assumptions
-# calibration of power plant efficiencies
-efficiency = {
-    # [plant efficiencies] -------------- #
-    'e_Nuclear': [1],
-    'e_Biomass': [1],
-    'e_Lignite': [1.15],
-    'e_Coal': [1.225],
-    'e_Gas': [1.15],
-    'e_Oil': [1]
-}
-
-capacity_scenarios = {
-    'AT': {
-        'bio': [0.8, 1, 0.8],
-        'coal': [0.75, 1, 0],
-        'heatpump': [1, 1, 1],
-        'lig': [0.8, 1, 0],
-        'ng': [0.85, 1, 0.85],
-        'nuc': [0, 0, 0],
-        'oil': [0.85, 1, 0.85],
-        'wind_on': [2, 2, 8],
-        'wind_off': [0, 0, 0],
-        'pv': [1, 1, 5]
-    },
-    'DE': {
-        'bio': [0.8, 1, 0.85],
-        'coal': [0.75, 1, 0.35],
-        'heatpump': [1, 1, 1],
-        'lig': [0.88, 1, 0.45],
-        'ng': [0.9, 1, 0.9],
-        'nuc': [0.72, 1, 0],
-        'oil': [0.9, 1, 0.9],
-        'wind_on': [50, 50, 90.8],
-        'wind_off': [5, 5, 15],
-        'pv': [50, 50, 73]
-    }
-}
-
-eua_range = range(5, 76, 5)

### 2019-09-20 applications/wind_vs_pv/results_wind_vs_pv.py

@@ -1,83 +1,0 @@
-import os
-
-import numpy as np
-import pandas as pd
-from gams import *
-
-import config as cfg
-from applications.wind_vs_pv.settings_wind_vs_pv import project, eua_range, scenario_2030, output_naming
-from medea.gams_io import gdx2df
-
-# %%
-ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
-
-read_variables = {
-    'cost': ([], ['z']),
-    'emissions': ([], ['z']),
-    'invest_res': (['tec_itm'], ['z']),
-    'invest_thermal': (['tec'], ['z']),
-    'invest_storage_power': (['tec_strg'], ['z']),
-    'invest_storage_energy': (['tec_strg'], ['z']),
-    'invest_atc': (['z'], ['z']),
-    'annual_price_el': ([], ['z']),
-    'annual_price_ht': ([], ['z']),
-    'annual_pumping': ([], ['z']),
-    'annual_turbining': ([], ['z']),
-    'annual_generation': (['prd'], ['z']),
-    'annual_curtail': ([], ['z']),
-    'annual_netflow': ([], ['z']),
-    'producer_surplus': ([], ['z']),
-}
-
-
-# Autarky results
-df_aut = pd.DataFrame()
-scenario = 'Autarky'
-for peua in eua_range:
-
-    wind_limit = float('inf')
-    filename = f'medea_out_{output_naming.format(scenario, peua, wind_limit)}.gdx'
-    db_output = ws.add_database_from_gdx(os.path.join(cfg.folder, 'applications', project, 'opt', filename))
-
-    result_dict = {key: gdx2df(db_output, key, value[0], value[1])
-                   for key, value in read_variables.items()}
-
-    df = pd.DataFrame(columns=cfg.zones)
-    for key, value in result_dict.items():
-        if value.index.any() == 'Value':
-            value.index = [key]
-        df = df.append(value, sort=True)
-    df.columns = pd.MultiIndex.from_product([[f'EUA_{peua}'], df.columns])
-    df_aut = pd.concat([df_aut, df], axis=1)
-
-df_aut = df_aut.replace(False, np.nan)
-df_aut = df_aut.dropna(axis=1, how='all')
-df_aut = df_aut.sort_index(axis=1, level=1)
-df_aut.to_csv(os.path.join(cfg.folder, 'applications', project, 'results',
[…]

### 2019-09-20 applications/wind_vs_pv/run_wind_vs_pv.py

@@ -1,120 +1,0 @@
-import os
-import subprocess
-from shutil import copyfile
-
-import pandas as pd
-from gams import *
-
-import config as cfg
-from applications.wind_vs_pv.settings_wind_vs_pv import *
-from medea.gams_io import reset_parameter, gdx2df, df2gdx
-
-# ---------------------------------------------------------------
-# %% initialize GAMS, GAMS workspace & load model data
-
-# fetch main gams model, if required
-if not os.path.isfile(os.path.join(cfg.folder, 'applications', project, 'opt', 'medea_main.gms')):
-    copyfile(os.path.join(cfg.folder, 'medea', 'medea_main.gms'),
-             os.path.join(cfg.folder, 'applications', project, 'opt', 'medea_main.gms'))
-# fetch main gams data, if required
-if not os.path.isfile(os.path.join(cfg.folder, 'applications', project, 'opt', 'medea_main_data.gdx')):
-    copyfile(os.path.join(cfg.folder, 'medea', 'data', 'input', 'medea_main_data.gdx'),
-             os.path.join(cfg.folder, 'applications', project, 'opt', 'medea_main_data.gdx'))
-
-ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
-db_input = ws.add_database_from_gdx(os.path.join(cfg.folder, 'applications', project, 'opt', 'medea_main_data.gdx'))
-data_yr = db_input['year'].first_record().value
-
-# read sets for clusters, hydro storage plants and products from db_input
-clust_dict = {rec.keys[0] for rec in db_input['tec']}
-chp_dict = {rec.keys[0] for rec in db_input['tec_chp']}
-itm_dict = {rec.keys[0] for rec in db_input['tec_itm']}
-hsp_dict = {rec.keys[0] for rec in db_input['props']}
-prd_dict = {rec.keys[0] for rec in db_input['prd']}
-fuel_dict = {rec.keys[0] for rec in db_input['f']}
-
-# read parameters
-df_eua = gdx2df(db_input, 'PRICE_CO2', ['t'], [])
-df_capitm = gdx2df(db_input, 'INSTALLED_CAP_ITM', ['z', 'tec_itm'], [])
-df_eff = gdx2df(db_input, 'EFFICIENCY', ['tec', 'prd', 'f'], [])
[…]

### 2019-09-20 applications/wind_vs_pv/settings_wind_vs_pv.py

@@ -1,62 +1,0 @@
-# ---------------------------------------------------------------
-# %% Settings & Scenario Assumptions
-project = 'wind_vs_pv'
-fuels = ['Nuclear', 'Lignite', 'Gas', 'Oil', 'Coal', 'Biomass']
-# naming of output gdx
-output_naming = '{}_eua-{}_won-{}'
-
-# ---------------------------------------------------------------
-# Scenario Assumptions
-# calibration of power plant efficiencies
-efficiency = {
-    # [plant efficiencies]
-    'e_Nuclear': [1],
-    'e_Biomass': [1],
-    'e_Lignite': [1.15],
-    'e_Coal': [1.225],
-    'e_Gas': [1.15],
-    'e_Oil': [1]
-}
-
-# [capacities in 2030]
-# - thermal capacities relative to installed capacities in 2016
-# - intermittent capacities in GW(p)
-x = list(range(15, -1, -1))
-x.insert(0, float('inf'))
-scenario_2030 = {
-    'AT': {
-        'bio': [0.8],
-        'coal': [0],
-        'heatpump': [1],
-        'lig': [0],
-        'ng': [0.85],
-        'nuc': [0.8],
-        'oil': [0.85],
-        'wind_on': [2.6],
-        'wind_off': [0],
-        'pv': [1.1],
-        'd_power': [1.18276362],
-        'lim_wind_on': x
-    },
-    'DE': {
-        'bio': [0.85],
-        'coal': [0.35],
-        'heatpump': [1],
-        'lig': [0.45],
-        'ng': [0.9],
-        'nuc': [0.0],
-        'oil': [0.9],
-        'wind_on': [90.8],
-        'wind_off': [15],
-        'pv': [73]
-    }
-}
-
-# [CO2 emission prices]
-eua_range = range(10, 81, 10)
-
-# [electricity exchange]
-electricity_exchange = {
-    'Autarky': [0],
-    'Openness': [float('inf')]
-}

### 2019-09-20 medea/gams_io.py

@@ -168,9 +168,9 @@ def run_medea_project(project_name, scenario_id):
     :return:
     """
     # generate file names
-    gms_model_fname = os.path.join(cfg.MEDEA_ROOT_DIR, 'applications', project_name, 'opt', 'medea_main.gms')
+    gms_model_fname = os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', project_name, 'opt', 'medea_main.gms')
     gdx_out_fname = f'gdx=medea_out_{scenario_id}.gdx'
-    input_fname = os.path.join(cfg.MEDEA_ROOT_DIR, 'applications', project_name, 'opt', f'medea_{scenario_id}_data.gdx')
+    input_fname = os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', project_name, 'opt', f'medea_{scenario_id}_data.gdx')
 
     # call GAMS to solve model / scenario
     subprocess.run(

### 2019-09-20 medea/instantiate_gdx.py

@@ -39,8 +39,8 @@ logging.info('medea sets instantiated')
 # --------------------------------------------------------------------------- #
 CAPITALCOST_G = df2gdx(db, static_data['tec']['annuity'].round(4),
                             'CAPITALCOST_G', 'par', [i_set], '[kEUR per GW]')
-CAPITALCOST_R = df2gdx(db, static_data['CAPCOST_R'].stack().reorder_levels([1, 0]).round(4), 'CAPITALCOST_R', 'par',
-                       [z_set, n_set], '[kEUR per GW]')
+CAPITALCOST_R = df2gdx(db, static_data['CAPCOST_R'].stack().reorder_levels([1, 0]).round(4).loc[:, 'annuity'],
+                       'CAPITALCOST_R', 'par', [z_set, n_set], '[kEUR per GW]')
 CAPITALCOST_S = df2gdx(db, plant_data['storage_clusters']['cost_power'].reorder_levels((1, 0)), 'CAPITALCOST_S', 'par',
                        [z_set, k_set], '[GW]')
 CAPITALCOST_V = df2gdx(db, plant_data['storage_clusters']['cost_energy'].reorder_levels((1, 0)), 'CAPITALCOST_V', 'par',
@@ -79,8 +79,13 @@ INITIAL_CAP_X = df2gdx(db, dict_instantiate['CAP_X'].stack(), 'INITIAL_CAP_X', '
 # TODO: Estimate LAMBDA, i.e. the share of must-run capacity in conventional capacity (controlling for renewables capacity)
 LAMBDA = df2gdx(db, static_data['LAMBDA'], 'LAMBDA', 'par', 0, '[]')
 
-OM_COST_QFIX = df2gdx(db, static_data['tec']['om_fix'], 'OM_COST_QFIX', 'par', [i_set], '[kEUR per GW]')
-OM_COST_VAR = df2gdx(db, static_data['tec']['om_var'], 'OM_COST_VAR', 'par', [i_set], '[kEUR per GWh]')
+OM_COST_G_QFIX = df2gdx(db, static_data['tec']['om_fix'], 'OM_COST_G_QFIX', 'par', [i_set], '[kEUR per GW]')
+OM_COST_G_VAR = df2gdx(db, static_data['tec']['om_var'], 'OM_COST_G_VAR', 'par', [i_set], '[kEUR per GWh]')
+
+OM_COST_R_QFIX = df2gdx(db, static_data['CAPCOST_R'].stack().reorder_levels([1, 0]).loc[:, 'om_qfix'],
[…]

### 2019-09-20 medea/prepare_data.py

@@ -1,3 +1,4 @@
+# %% imports
 import os
 
 import numpy as np

### 2019-09-26 doc/mathematical_description.tex
restructuring of directories

@@
 decommissioned capacity of dispatchables & $\widetilde{g}^{-}_{z,i}$ & \texttt{deco\_g(z,i)} & GW \\ \hline
-energy generated by conventionals & $g_{z,t,i,m}$ & \texttt{g(z,t,i,m)} & GW \\ \hline
+energy generated by conventionals & $g_{z,t,i,m,f}$ & \texttt{g(z,t,i,m,f)} & GW \\ \hline
 curtailed energy & $q^{+}_{z,t}$ & \texttt{q\_curtail(z,t)} & GW \\ \hline
@@
 storage energy content & $v_{z,t,k}$ & \texttt{v(z,t,k)} & GWh \\ \hline
-operating region weight & $w_{z,t,i,l}$ & \texttt{w(z,t,i,l)} & \\ \hline
+operating region weight & $w_{z,t,i,l,f}$ & \texttt{w(z,t,i,l,f)} & \\ \hline
 added transmission capacity & $\widetilde{x}^{+}_{z,zz}$ & \texttt{add\_x(z,zz)} & GW \\ \hline
@@
 	 &  initial capacity\tmark[$\dagger$] & added capacity\tmark[$\ddagger$] & \makecell[l]{decom-\\missioned \\capacity\tmark[$\ddagger$]} & \makecell[l]{specific \\investment \\cost\tmark[$\dagger$]} & dispatch\tmark[$\ddagger$] \ML
-	thermal units & $\widetilde{G}_{z,i}$ & $\widetilde{g}^{+}_{z,i}$ & $\widetilde{g}^{-}_{z,i}$ & $C^{g}_{z,i}$ & $g_{z,t,i,m}$ \ML
+	thermal units & $\widetilde{G}_{z,i}$ & $\widetilde{g}^{+}_{z,i}$ & $\widetilde{g}^{-}_{z,i}$ & $C^{g}_{z,i}$ & $g_{z,t,i,m,f}$ \ML
 	\makecell[l]{intermittent \\units} &  $\widetilde{R}_{z,n}$ & $\widetilde{r}^{+}_{z,n}$ & $\widetilde{r}^{-}_{z,n}$ & $C^{r}_{z,n}$ & $r_{z,t,n}$ \ML
@@
 &c^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: e_{z,t,i} \right) \qquad \qquad &\forall z,t,i\\
-&c^{om}_{z,i}& =&\ Q^{g}_{i} \left(\widetilde{G}_{z,i} - \widetilde{g}^{-}_{z,i} + \widetilde{g}^{+}_{z,i}\right) + \sum_{t,m} \left(O^{g}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
+&c^{om}_{z,i}& =&\ Q^{g}_{i} \left(\widetilde{G}_{z,i} - \widetilde{g}^{-}_{z,i} + \widetilde{g}^{+}_{z,i}\right) + \sum_{t} \sum_{m} \sum_{f} \left(O^{g}_{i} \: g_{z,t,i,m,f}\right) \qquad \qquad &\forall z,i \\
 &c^{om}_{z,n}& =&\ Q^{r}_{n} \left(\widetilde{R}_{z,n} - \widetilde{r}^{-}_{z,n} + \widetilde{r}^{+}_{z,n}\right) + \sum_{t} \left(O^{r}_{n} \: r_{z,t,n}\right) \qquad \qquad &\forall z,n \\
@@
 &c^{x}_{z}& =&\ \frac{1}{2} \: \sum_{zz} (C^{x} \: \delta_{z,zz} \: \widetilde{x}^{+}_{z,zz}) \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
-&c^{q}_{z}& =&\ \mu \sum_{t,m} q^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
+&c^{q}_{z}& =&\ \mu \sum_{t} \sum_{m} q^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
 \end{align}
@@
 \begin{split}
-\sum_{i} g_{z,t,i,\text{el}} + \sum_{k} s^{out}_{z,t,k} + \sum_{n} r_{z,t,n} &=  \\ D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + & \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} - q^{-}_{z,t,\text{el}} + q^{+}_{z,t} \qquad \forall z,t
+\sum_{i} \sum_{f} g_{z,t,i,\text{el},f} + \sum_{k} s^{out}_{z,t,k} + \sum_{n} r_{z,t,n} &=  \\ D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + & \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} - q^{-}_{z,t,\text{el}} + q^{+}_{z,t} \qquad \forall z,t
 \end{split}
@@
 \begin{align}
-\sum_{i} g_{z,t,i,\text{ht}} = D_{z,t,\text{ht}} - q^{-}_{z,t,\text{ht}} \qquad \forall z,t \label{market_clearing_ht}
+\sum_{i} \sum_{f} g_{z,t,i,\text{ht},f} = D_{z,t,\text{ht}} - q^{-}_{z,t,\text{ht}} \qquad \forall z,t \label{market_clearing_ht}
 \end{align}
@@
 \paragraph{Energy generation}
-Energy generation $g_{z,t,i,m} \geq 0$ is constrained by available installed capacity, which can be adjusted through investment ($\widetilde{g}^{+}_{z,i} \geq 0$) and decommissioning $\widetilde{g}^{-}_{z,i} \geq 0$.
+Energy generation $g_{z,t,i,m,f} \geq 0$ is constrained by available installed capacity, which can be adjusted through investment ($\widetilde{g}^{+}_{z,i} \geq 0$) and decommissioning $\widetilde{g}^{-}_{z,i} \geq 0$.
 \begin{align}
-g_{z,t,i,m} \leq \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} - \widetilde{g}^{-}_{z,i} \qquad \qquad \forall z,t,i,m
+\sum_{f} g_{z,t,i,m,f} \leq \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} - \widetilde{g}^{-}_{z,i} \qquad \qquad \forall z,t,i,m
 \end{align}
@@
 \begin{align}
-g_{z,t,i,m} \leq \sum_{f} \eta_{i,m,f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \notin J
+g_{z,t,i,m,f} = \sum_{f} \eta_{i,m,f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \notin J, f
 \end{align}
@@
 The elements $l \in L$ span up a three-dimensional, convex feasible operating region for each co-generation technology. 
-The weights $w_{z,t,i,l} \geq 0$ form a convex combination of the corners $l$, which are scaled to the available installed capacity of each co-generation technology.
+The weights $w_{z,t,i,l,f} \geq 0$ form a convex combination of the corners $l$, which are scaled to the available installed capacity of each co-generation technology.
+Defining weights over fuels allows co-generation units to switch fuels between multiple alternatives.
 Heat and electricity output along with the corresponding fuel requirement is then set according to the chosen weights.
 \begin{align}
-\sum_{l} w_{z,t,i,l} = \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} - \widetilde{g}^{-}_{z,i} \qquad \qquad \forall z,t,i \in J \\
-g_{z,t,i,m} \leq \sum_{l} \psi_{i,l,m} \: w_{z,t,i,l} \qquad \qquad \forall z,t,i \in J, m \\
-b_{z,t,i,f} \geq \sum_{l} \chi_{i,l,f} \: w_{z,t,i,l} \qquad \qquad \forall z,t,i \in J, f
+\sum_{l} \sum_{f} w_{z,t,i,l,f} = \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} - \widetilde{g}^{-}_{z,i} \qquad \qquad \forall z,t,i \in J \\
+g_{z,t,i,m,f} = \sum_{l} \sum_{f} \psi_{i,l,m} \: w_{z,t,i,l,f} \qquad \qquad \forall z,t,i \in J, m \\
+b_{z,t,i,f} = \sum_{l} \chi_{i,l,f} \: w_{z,t,i,l,f} \qquad \qquad \forall z,t,i \in J, f\\
+w(z,t,i,l,f) = 0 \qquad \qquad \forall z,t,i,k,f: \chi_{i,l,f} = 0
 \end{align}
@@
 \begin{align}
-\sum_{i}\left( g_{z,t,i,\text{el}} \right) + r_{z,t,\text{ror}} + \sum_{k} \left( s^{out}_{z,t,k} + s^{in}_{z,t,k} \right) \geq \lambda_{z} \widehat{D}_{z,\text{el}} + \sigma_{z} \sum_{n \setminus \{ \text{ror}\}} \widehat{\phi}_{z,n} (\widetilde{R}_{z,n} + \widetilde{r}^{+}_{z,n}) \qquad \forall z,t
+\sum_{i} \sum_{f} \left( g_{z,t,i,\text{el},f} \right) + r_{z,t,\text{ror}} + \sum_{k} \left( s^{out}_{z,t,k} + s^{in}_{z,t,k} \right) \geq \lambda_{z} \widehat{D}_{z,\text{el}} + \sigma_{z} \sum_{n \setminus \{ \text{ror}\}} \widehat{\phi}_{z,n} (\widetilde{R}_{z,n} + \widetilde{r}^{+}_{z,n}) \qquad \forall z,t
 \end{align}

### 2019-09-26 legacy/analyse_results.py

@@ -8,9 +8,9 @@ from legacy.gams_wrappers import gdx2df
 # df = pd.read_csv(os.path.join(cfg.folder, 'medea', 'opt'))
 run_string = 'PoBu_EUA20_curt_htpmp_1'
 
-ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
+ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
 filename = f'medea_out_{run_string}.gdx'
-db_output = ws.add_database_from_gdx(os.path.join(cfg.folder, 'medea', 'opt', filename))
+db_output = ws.add_database_from_gdx(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', filename))
 
 read_variables = {
    'q_curtail': (['t'], ['r', 'prd']),

### 2019-09-26 legacy/compile_results.py

@@ -8,9 +8,9 @@ from legacy.gams_wrappers import gdx2df
 
 output_folder = 'PolicyBurden'
 campaign_string = 'PoBu_EUA40_curt_htpmp'
-directory = os.path.join(cfg.folder, 'medea', 'opt')
+directory = os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt')
 
-ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
+ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
 # for file in os.listdir(directory):
 #     filename = os.fsdecode(file)
 #     if campaign_string in filename:
@@ -18,7 +18,7 @@ ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
 df_all = pd.DataFrame()
 for num in range(0, 16):
     filename = f'medea_out_{campaign_string}_{num}.gdx'
-    db_output = ws.add_database_from_gdx(os.path.join(cfg.folder, 'medea', 'opt', filename))
+    db_output = ws.add_database_from_gdx(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', filename))
 
     read_variables = {
         'cost': ([], ['r']),
@@ -48,5 +48,5 @@ for num in range(0, 16):
     df_all = pd.concat([df_all, df], axis=1)
 
 df_all = df_all.sort_index(axis=1, level=1)
-df_all.to_csv(os.path.join(cfg.folder, 'medea', 'output', output_folder, f'results_{campaign_string}.csv'),
+df_all.to_csv(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'output', output_folder, f'results_{campaign_string}.csv'),
               sep=';', encoding='utf-8')

### 2019-09-26 legacy/scenario_results.py

@@ -99,9 +99,9 @@ scenario_AT2030 = {
 # --------------------------------------------------------------------------- #
 # %% initialize GAMS workspace and load model data
 # --------------------------------------------------------------------------- #
-ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
+ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
 
-db_input = ws.add_database_from_gdx(os.path.join(cfg.folder, 'medea', 'opt', 'medea_data.gdx'))
+db_input = ws.add_database_from_gdx(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', 'medea_data.gdx'))
 data_yr = db_input['year'].first_record().value
 
 # read sets for clusters, hydro storage plants and products from db_input
@@ -187,7 +187,7 @@ WONLIM = df2gdx(db_input, df_wonlim, 'WON_LIMIT', 'par', 0, 'upper limit on onsh
 df_euafixed = pd.DataFrame(data=[0])
 EUA_SCENARIO = df2gdx(db_input, df_euafixed, 'EUA_SCENARIO', 'par', 0, 'EUA price - scenario')
 
-os.chdir(os.path.join(cfg.folder, 'medea', 'opt'))
+os.chdir(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt'))
 
 for eua in range(50, 61, 10):
     reset_parameter(db_input, 'EUA_SCENARIO', pd.DataFrame(data=[eua]))
@@ -196,13 +196,13 @@ for eua in range(50, 61, 10):
         # modify wind_on limit
         reset_parameter(db_input, 'WON_LIMIT', pd.DataFrame(data=[it]))
         # export gdx
-        export_location = os.path.join(cfg.folder, 'medea', 'opt', f'MEDEA_{scenario_name}_data.gdx')
+        export_location = os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', f'MEDEA_{scenario_name}_data.gdx')
         db_input.export(export_location)
         # --------------------------------------------------------------------------- #
         # call GAMS
-        gms_model = os.path.join(cfg.folder, 'medea', 'opt', 'medea_austria_htpmp.gms')
+        gms_model = os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', 'medea_austria_htpmp.gms')
[…]

### 2019-09-26 legacy/scenario_solve.py

@@ -200,9 +200,9 @@ scenario_CoalCommission = {
 # --------------------------------------------------------------------------- #
 # %% initialize GAMS workspace and load model data
 # --------------------------------------------------------------------------- #
-ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
+ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
 
-db_input = ws.add_database_from_gdx(os.path.join(cfg.folder, 'medea', 'opt', 'medea_data.gdx'))
+db_input = ws.add_database_from_gdx(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', 'medea_data.gdx'))
 data_yr = db_input['year'].first_record().value
 
 # read sets for clusters, hydro storage plants and products from db_input
@@ -228,7 +228,7 @@ df_load = gdx2df(db_input, 'CONSUMPTION', ['r', 't', 'prd'], [])
 # --------------------------------------------------------------------------- #
 # %% scenario generation
 # --------------------------------------------------------------------------- #
-os.chdir(os.path.join(cfg.folder, 'medea', 'opt'))
+os.chdir(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt'))
 idx = pd.IndexSlice
 traded_fuels = ['Nuclear', 'Lignite', 'Gas', 'Oil', 'Coal', 'Biomass']
 
@@ -373,7 +373,7 @@ for scn in scenario_set:
         reset_parameter(db_input, 'CONSUMPTION', df_load_mod)
 
         # export scenario data to database
-        export_location = os.path.join(cfg.folder, 'medea', 'opt', f'MEDEA_{scenario_name}_data.gdx')
+        export_location = os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', f'MEDEA_{scenario_name}_data.gdx')
         db_input.export(export_location)
 
         # call iterative model solution
@@ -383,8 +383,8 @@ for scn in scenario_set:
 #            goodness_of_fit.loc[scenario_iteration, key] = goodfit[key]
 
         # delete scenario data file
-    if os.path.isfile(os.path.join(cfg.folder, 'medea', 'opt', f'MEDEA_{scenario_name}_data.gdx')):
-        os.remove(os.path.join(cfg.folder, 'medea', 'opt', f'MEDEA_{scenario_name}_data.gdx'))
[…]

### 2019-09-26 legacy/solve.py

@@ -16,11 +16,11 @@ def solve(output_folder, output_name, scenario_iteration, gen_target):
     # --------------------------------------------------------------------------- #
     # initialize GAMS workspace
     # --------------------------------------------------------------------------- #
-    ws = GamsWorkspace(system_directory=cfg.gams_sysdir)
+    ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
 
     # read input data from MEDEA_data.gdx
     db_input = ws.add_database_from_gdx(
-        os.path.join(cfg.folder, f'medea', 'opt', 'medea_data.gdx'))
+        os.path.join(cfg.MEDEA_ROOT_DIR, f'medea', 'opt', 'medea_data.gdx'))
     # read sets for clusters, hydro storage plants and products from db_input
     clust_dict = {rec.keys[0] for rec in db_input['tec']}
     hsp_dict = {rec.keys[0] for rec in db_input['tec_hsp']}
@@ -93,7 +93,7 @@ def solve(output_folder, output_name, scenario_iteration, gen_target):
     # --------------------------------------------------------------------------- #
     # generate iterations
     # --------------------------------------------------------------------------- #
-    os.chdir(os.path.join(cfg.folder, 'medea', 'opt'))
+    os.chdir(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt'))
     for iteration in range(1, cfg.iter_num + 1):
         print(f'Iteration {iteration} in progress')
 
@@ -132,20 +132,22 @@ def solve(output_folder, output_name, scenario_iteration, gen_target):
         """
         # data export for first iteration
         # if iteration == 1:
-        db_init.export(os.path.join(cfg.folder, 'medea', 'opt', f'medea_{output_folder}_{output_name}_iterdata.gdx'))
+        db_init.export(
+            os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', f'medea_{output_folder}_{output_name}_iterdata.gdx'))
 
         # --------------------------------------------------------------------------- #
         # call GAMS
         # --------------------------------------------------------------------------- #
-        gms_in = os.path.join(cfg.folder, 'medea', 'opt', f'medea_{cfg.model_type}.gms')
+        gms_in = os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', f'medea_{cfg.model_type}.gms')
[…]

### 2019-09-26 src/data/compile_timeseries.py

@@ -3,7 +3,7 @@ import os
 import pandas as pd
 
 import config as cfg
-from data.tools.functions import download_file
+from src.tools.functions import download_file
 
 # ======================================================================================================================
 # download and process opsd time series

### 2019-09-26 src/data/get_heatload.py

@@ -5,7 +5,7 @@ import numpy as np
 import pandas as pd
 
 import config as cfg
-from data.tools.functions import heat_yr2day, heat_day2hr, resample_index, download_file
+from src.tools.functions import heat_yr2day, heat_day2hr, resample_index, download_file
 
 # ----------------------------------------------------------------------------
 # set current working directory

### 2019-09-26 src/data/get_pricedata.py

@@ -3,7 +3,7 @@ import os
 import pandas as pd
 
 import config as cfg
-from data.tools.functions import download_file
+from src.tools.functions import download_file
 
 # ======================================================================================================================
 # IMF commodity price data

### 2019-09-26 src/templates/results_template.py

@@ -5,8 +5,8 @@ import pandas as pd
 from gams import *
 
 import config as cfg
-from medea.gams_io import gdx2df
-from medea.templates.settings_template import *
+from src.templates.settings_template import *
+from src.tools.gams_io import gdx2df
 
 # -------------------------------------------------------------------------------------------------------------------- #
 # prepare GAMS workspace

### 2019-09-26 src/templates/run_template.py

@@ -5,8 +5,8 @@ import pandas as pd
 from gams import *
 
 import config as cfg
-from medea.gams_io import reset_symbol, gdx2df, df2gdx
-from medea.templates.settings_template import *
+from src.templates.settings_template import *
+from src.tools.gams_io import reset_symbol, gdx2df, df2gdx
 
 # -------------------------------------------------------------------------------------------------------------------- #
 # %% initialize GAMS, GAMS workspace and load model data
@@ -14,7 +14,7 @@ from medea.templates.settings_template import *
 # import base data from gdx
 ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
 db_input = ws.add_database_from_gdx(
-    os.path.join(cfg.MEDEA_ROOT_DIR, 'applications', project_name, 'opt', 'medea_main_data.gdx'))
+    os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', project_name, 'opt', 'medea_main_data.gdx'))
 
 
 # %% read parameters that change in scenarios (and corresponding sets)
@@ -62,7 +62,7 @@ reset_symbol(db_input, 'FEASIBLE_INPUT', df_fuelreq_mod)
 # %% generate 'dynamic' parameter variations (modifications that constitute the scenarios, i.e. that change each run)
 # -------------------------------------------------------------------------------------------------------------------- #
 # ensure that we are in the correct model directory
-os.chdir(os.path.join(cfg.MEDEA_ROOT_DIR, 'applications', project_name, 'opt'))
+os.chdir(os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', project_name, 'opt'))
 
 # create empty scenario parameter in GAMS database so that it can be modified subsequently
 # example: changing CO2 price
@@ -78,12 +78,12 @@ for price_co2 in range_co2price:
     reset_symbol(db_input, 'CO2_SCENARIO', pd.DataFrame(data=[price_co2]))
 
     # export modified GAMS database to a .gdx-file that is then being read by the GAMS model
-    export_location = os.path.join(cfg.MEDEA_ROOT_DIR, 'applications', project_name, 'opt',
+    export_location = os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', project_name, 'opt',
                                    f'medea_{identifier}_data.gdx')
     db_input.export(export_location)
 
     # generate path to medea model
[…]

### 2019-09-26 src/tools/functions.py

@@ -1,3 +1,4 @@
+import ftplib
 import shutil
 
 import numpy as np
@@ -123,3 +124,18 @@ def download_file(url, save_to):
     http = urllib3.PoolManager()
     with http.request('GET', url, preload_content=False) as r, open(save_to, 'wb') as out_file:
         shutil.copyfileobj(r, out_file)
+
+
+pwd = 'ausmaus76'
+
+
+def get_from_entsoe_ftp(user, pwd):
+    ftp = ftplib.FTP('sftp-transparency.entsoe.eu')
+    ftp.login(user, pwd)
+    data = []
+    ftp.dir(data.append)
+    ftp.quit()
+
+    for line in data:
+        print
+        "-", line

### 2019-09-26 src/tools/helpers.py

@@ -2,7 +2,7 @@ import subprocess
 
 import pandas as pd
 
-from medea.gams_io import gdx2df
+from src.tools.gams_io import gdx2df
 
 
 def is_leapyear(year):

### 2019-09-26 src/tools/instantiate_gdx.py

@@ -5,8 +5,8 @@ import pandas as pd
 from gams import *
 
 import config as cfg
-from medea.gams_io import df2gdx
-from medea.prepare_data import dict_sets, dict_instantiate, static_data, plant_data, ts_data, invest_limits
+from src.tools.gams_io import df2gdx
+from src.tools.prepare_data import dict_sets, dict_instantiate, static_data, plant_data, ts_data, invest_limits
 
 # TODO: Add energy stored in hydro reservoirs - STORAGE_LEVEL
 

### 2019-09-26 src/tools/prepare_data.py

@@ -5,7 +5,7 @@ import numpy as np
 import pandas as pd
 
 import config as cfg
-from medea.helpers import hours_in_year
+from src.tools.helpers import hours_in_year
 
 # TODO: PerformanceWarning: indexing past lexsort-depth may impact performance. need to sort index of affected
 # dataframes. Possibly by reindex() or sort_index(). lexsort can be checked with df.index.is_lexsorted()
@@ -59,7 +59,8 @@ dict_sets = {
         'Solar': [80],
         'Wind': [90],
         'Power': [100],
-        'Heat': [110]
+        'Heat': [110],
+        'Syngas': [120]
     },
     'l': {f'l{x}': [True] for x in range(1, 5)},
     'm': {
@@ -103,7 +104,8 @@ dict_static = {
         'Solar': [0],
         'Wind': [0],
         'Power': [0],
-        'Heat': [0]
+        'Heat': [0],
+        'Syngas': [0]
     },
     'eta': {
         'nuc': [0.34],

### 2019-09-26 src/tools/projects.py

@@ -4,14 +4,14 @@ from shutil import copyfile
 
 def create_project(project_name, medea_dir):
     """
-    creates application folder and subfolders according to medea-conventions and copies required model and data files.
+    creates project folder and subfolders according to medea-conventions and copies required model and data files.
     Additionally, also fetches some templates to set up scenarios with medea
     :param project_name: name of the project to be initialised
     :param medea_dir: base directory where medea is installed / located
     :return:
     """
 
-    project_dir = os.path.join(medea_dir, 'applications', project_name)
+    project_dir = os.path.join(medea_dir, 'projects', project_name)
     if os.path.exists(project_dir):
         raise NameError(f'A project called {project_name} already exists.')
     else:
@@ -22,47 +22,47 @@ def create_project(project_name, medea_dir):
         os.makedirs(os.path.join(project_dir, 'results'))
 
         # fetch main gams model, if required
-        if not os.path.isfile(os.path.join(medea_dir, 'applications', project_name, 'opt', 'medea_main.gms')):
+        if not os.path.isfile(os.path.join(medea_dir, 'projects', project_name, 'opt', 'medea_main.gms')):
             copyfile(os.path.join(medea_dir, 'medea', 'medea_main.gms'),
-                     os.path.join(medea_dir, 'applications', project_name, 'opt', 'medea_main.gms'))
+                     os.path.join(medea_dir, 'projects', project_name, 'opt', 'medea_main.gms'))
         # fetch template for custom gams model modifications, if required
-        if not os.path.isfile(os.path.join(medea_dir, 'applications', project_name, 'opt', f'medea_{project_name}.gms')):
+        if not os.path.isfile(os.path.join(medea_dir, 'projects', project_name, 'opt', f'medea_{project_name}.gms')):
             copyfile(os.path.join(medea_dir, 'medea', 'medea_custom.gms'),
-                     os.path.join(medea_dir, 'applications', project_name, 'opt', f'medea_{project_name}.gms'))
+                     os.path.join(medea_dir, 'projects', project_name, 'opt', f'medea_{project_name}.gms'))
         # fetch main gams data, if required
-        if not os.path.isfile(os.path.join(medea_dir, 'applications', project_name, 'opt', 'medea_main_data.gdx')):
+        if not os.path.isfile(os.path.join(medea_dir, 'projects', project_name, 'opt', 'medea_main_data.gdx')):
             copyfile(os.path.join(medea_dir, 'medea', 'data', 'medea_main_data.gdx'),
[…]

### 2020-02-21 projects/asparagus/doc/slides.tex
initial commit of project asparagus

\documentclass[aspectratio=169, xcolor=dvipsnames]{beamer}
\usetheme{Pittsburgh}
\usepackage{tikz}
\usepackage{booktabs}
\usepackage[version=4]{mhchem}
\usepackage{animate}
\usepackage{biblatex}
\usepackage[absolute,overlay]{textpos}
\usepackage{fancyvrb}
\usepackage{xcolor}
\usepackage{multicol}
\usepackage{setspace}
\usepackage{siunitx}
\DeclareSIUnit\year{a}
\usepackage{qrcode}
\usepackage{eurosym}
\usepackage{xfrac}

% Thin fonts
\usepackage{cmbright}
\usepackage[T1]{fontenc}

\usetikzlibrary{arrows.meta}
\usetikzlibrary{shapes.geometric, arrows}
\urlstyle{same}

%gets rid of bottom navigation bars
\setbeamertemplate{footline}[frame number]{}
%gets rid of bottom navigation symbols
\setbeamertemplate{navigation symbols}{}

% colors
\definecolor{bokugreen}{rgb}{0.302, 0.675, 0.149}
\setbeamercolor{title}{fg=bokugreen, bg=white}
\setbeamercolor{frametitle}{fg=bokugreen}
\setbeamercolor{itemize item}{fg=bokugreen}
\setbeamercolor{enumerate item}{fg=bokugreen}
\setbeamercolor{local structure}{fg=bokugreen}

\newcommand{\backupbegin}{
   \newcounter{framenumberappendix}
   \setcounter{framenumberappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
   \addtocounter{framenumberappendix}{-\value{framenumber}}
   \addtocounter{framenumber}{\value{framenumberappendix}} 
}

% title page
\setbeamertemplate{title page}[default][colsep=-4bp,rounded=true]
\title[The cost of undisturbed landscapes]{The cost of undisturbed landscapes}
\subtitle{Assessing systemic effects of renewables expansion in Austria}
\author{Sebastian Wehrle, Johannes Schmidt}
\institute{University of Natural Resources and Life Sciences, Vienna}
\date{February 25, 2020\\ \medskip NOeG Annual Meeting 2020}

\setbeamertemplate{section page}{
	\begin{textblock*}{4cm}(11cm, 4cm) % {block width} (coords)
	{\LARGE \insertsection}
	\end{textblock*}
}

\begin{document}
    %%%%Title Page
    {
        \usebackgroundtemplate{
	        \begin{picture}(320,300)
    	        \hspace{9.0cm}
	              \includegraphics[width=0.7\linewidth]{./logos/refuel_logo_with_text}
            \end{picture}
        }

        \begin{frame}[noframenumbering,plain]
          \maketitle
          \vspace{1.15cm}
          \includegraphics[height=0.7cm]{./logos/creative-commons-by}
          \hspace{4.0cm}\includegraphics[height=1.7cm]{./logos/boku-logo}\\
         \end{frame}
    }

\section{Introduction}
\begin{frame}
\frametitle{Austrian energy policy objectives}
\framesubtitle{according to government programme 2020-2024}
\begin{itemize}
\item 100\%\footnote[frame]{excluding system services and industry own consumption. At current levels this equals 10\% of consumption, i.e. actual target is around 90\%.} of electricity demand from domestic renewable sources on annual balance by 2030 \pause
\item additional annual electricity generation of \SI{27}{\tera\watt\hour} is expected to suffice \pause
\item technology-specific additions:
\end{itemize}

\begin{table}
\centering
\begin{tabular}{| l | r | r | r | r | r |}
\hline
 & 2018 & 2018\footnote[frame]{meteorological conditions as in 2016} & Policy & 2030 & 2030 \\ 
 & [GW] & [TWh] & [TWh] & [TWh] & [GW] \\ \hline \hline
Solar PV & 1.44 & 1.23 & +11 & 12.23 &  14.27 \\ \hline
Wind (onshore) & 3.05 & 6.14 & +10 & 16.14 & 8 \\ \hline
Hydro (run-of-river) & 5.72 & 28.34 & +5 & 33.34 & 6.73 \\ \hline
%Biomass & 1 & 5 & +1 & 6 & 2 \\ \hline
\end{tabular}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Stylized Facts}
\begin{itemize}
\item Apart from wind and solar, potentials for renewable electricity generation in Austria largely exhausted
\item Under announced policies, solar PV is only large-scale substitute to wind power
\item Increasing social conflict around the large-scale expansion of onshore wind power
\item Traditional power system models do not account for local negative externalities of renewable energy generators, such as:
	\begin{itemize}
	\item visual impact on landscape
	\item harm to wildlife
	\item noise, flickering, glaring
	\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Why wind is not solar}
\framesubtitle{A side note on imperfect substitutes}
\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
\centering
Aggregate generation profile in Austria
\includegraphics[width=\textwidth, keepaspectratio]{./figures/wind_not_pv}
\end{textblock*}
\begin{textblock*}{4cm}(11cm, 2.25cm)
%Aggregate profile for Austria
\bigskip
\begin{itemize}
\item secure system operation requires $S=D$ at any point in time
\item electricity not easily storable
\item grid operators >>transform<< electricity feed-in to end-use electricity
\end{itemize}
\end{textblock*}
\end{frame}

\begin{frame}
\frametitle{The problem from a social planner's perspective}
How to account for local negative externalities in renewables expansion planning?

\bigskip
\textcolor{bokugreen}{The cost of undisturbed landscapes}
\begin{itemize}
\item assess energy system effects and costs of substituting wind power with solar PV
\end{itemize}

\bigskip
\textcolor{bokugreen}{The value of undisturbed landscapes}
\begin{itemize}
\item estimates of the negative external effect of wind turbines reported in literature
\end{itemize}
\end{frame}

\section{Research agenda}
\begin{frame}
\frametitle{The cost of undisturbed landscapes}
\framesubtitle{Approach}
\begin{itemize}
\item Resemble Austrian power system in 2030
\item include most important electricity trading partner Germany
\item set policy target of meeting $90\%$ of demand in 2030 from domestic renewable sources
\item incorporate announced electricity system targets for Germany in 2030
	\begin{itemize}
	\item nuclear phase-out
	\item partial coal exit
	\item expansion of renewable capacities in line with EEG 2017
	\end{itemize}
\item simulate prospective electricity system with \emph{medea}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Power system model \emph{medea}}
%\framesubtitle{Marketing slide}
\begin{footnotesize}
\begin{columns}[t]
\begin{column}{0.325\textwidth}
\textbf{Objective}
\begin{itemize}
\item minimize total system cost
	\begin{itemize}
	\begin{footnotesize}
	\item fuel and \ce{CO2} cost
	\item O\&M cost
	\item capital cost
	\end{footnotesize}
	\end{itemize}
\end{itemize}
\end{column}

\begin{column}{0.32\textwidth}  %%<--- here
\textbf{Decision variables}
\begin{itemize}
	\item hourly dispatch
	\item inter-zonal electricity trade
	\item investment in power plants, storages, and transmission
\end{itemize}
\end{column}

\begin{column}{0.32\textwidth}  %%<--- here
\textbf{Constraints}
\begin{itemize}
\item market clearing
\item capacity constraints
\item co-generation \& fuel use
%\item storage volume
\item system service requirement
\item inter-zonal electricity trade
\end{itemize}
\end{column}
\end{columns}
\end{footnotesize}
\bigskip
\begin{footnotesize}
\begin{columns}[t]
\begin{column}{0.325\textwidth}
\textbf{Economic assumptions}
\begin{itemize}
\item perfect competition
\item perfect foresight
\item price-inelastic demand
\end{itemize}
\end{column}

\begin{column}{0.325\textwidth}
\textbf{Resolution}
\begin{itemize}
\item hours (one year)
\item bidding zones
\item 41 technologies
\end{itemize}
\end{column}

\begin{column}{0.325\textwidth}
\textbf{Implementation}
\begin{itemize}
\item linear program
\item python \& GAMS
\end{itemize}
\end{column}
\end{columns}
\end{footnotesize}
\medskip
\begin{center}
\emph{medea} is available on \href{https://github.com/inwe-boku/medea}{github.com/inwe-boku/medea} under an open MIT license
\end{center}
\end{frame}
\note{
medea is a partial-equilibrium model of the coupled electricity and district heating sectors.
Operation of and investment in conventional and renewable power plants, storages and transmission lines are determined simultaneously for one year.
Results include hourly dispatch, investment, inter-zonal trade, electricity and heat prices, profits, and emissions of air pollutants.
The model accounts for various technical characteristics of power systems, such as co-generation of heat and power, balancing requirements, 
}

\begin{frame}
\frametitle{The cost of undisturbed landscapes}
\framesubtitle{Estimating the opportunity cost of wind power}
\begin{itemize}
\item[1)] derive unrestricted optimal deployment of wind and solar power
\item[2)] restrict deployment of wind power by a small margin \\($\rightarrow$ solar PV substitutes for wind)
\item[3)] repeat till no wind power can be deployed
\end{itemize}

\bigskip
We approximate the cost of undisturbed landscapes (i.e. the forgone value of wind power $w$) by the change in net cost of the electricity system including air pollution cost $c_{net}$ in response to a change in deployed wind power $w$, i.e.
\begin{align*}
oc_w = \frac{\Delta c_{net}}{\Delta w}
\end{align*}

\end{frame}

\section{Results}
\begin{frame}
\frametitle{The cost of undisturbed landscapes}

\begin{textblock*}{10cm}(0.75cm, 1.5cm) % {block width} (coords)
\includegraphics[width=\textwidth, keepaspectratio]{./figures/undisturbed_base}
\end{textblock*}

\begin{textblock*}{4.5cm}(10.9cm, 1.5cm)
\begin{small}
\begin{table}
\begin{tabular}{p{1.8cm} | r l}
Capital cost of solar PV & $630$ & \SI{}{\text{\euro}\per\kilo\watt\text{p}}
\end{tabular}
\end{table}
\smallskip
\quad $\rightarrow$ about $\sfrac{2}{3}$ rooftop PV, \\
\quad $\sfrac{1}{3}$ open space PV

\vspace{3.75cm}
\quad wind power and solar PV \\ \quad are substitutes
\end{small}
\end{textblock*}

\begin{textblock*}{9.5cm}(1.5cm, 7.7cm)
\begin{tikzpicture}%[thick]
\draw (5.75,0) -- (9,0);
\draw[{Latex[length=2mm]}-] (0,0) -- (3.25,0);
\draw (4.5,0.0) node[] {\scriptsize more wind power};
\end{tikzpicture}

\begin{tikzpicture}%[thick]
\draw[-{Latex[length=2mm]}] (5.5,0) -- (9,0);
\draw (0,0) -- (3.5,0);
\draw (4.5,0.0) node[] {\scriptsize more solar PV};
\end{tikzpicture}
\end{textblock*}

\end{frame}

\begin{frame}
\frametitle{The cost of undisturbed landscapes}
\framesubtitle{Cost with restricted wind power}
\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
\includegraphics[width=\textwidth, keepaspectratio]{./figures/cost}
\end{textblock*}

\begin{textblock*}{4cm}(11cm, 2.25cm)
\begin{small}
\begin{table}
\begin{tabular}{p{1.8cm} | p{0.3cm} l}
Capital cost of solar PV & $630$ & \SI{}{\text{\euro}\per\kilo\watt\text{p}} \\
\ce{CO2} price & $90$ & \SI{}{\text{\euro}\per\mega\watt\hour}
\end{tabular}
\end{table}
\end{small}
\end{textblock*}

\end{frame}

\begin{frame}
\frametitle{The cost of undisturbed landscapes}
\framesubtitle{System operation with restricted wind power}
\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
\includegraphics[width=\textwidth, keepaspectratio]{./figures/sysops}
\end{textblock*}

\begin{textblock*}{4cm}(11cm, 2.25cm)
\begin{small}
\begin{table}
\begin{tabular}{p{1.8cm} | p{0.3cm} l}
Capital cost of solar PV & $630$ & \SI{}{\text{\euro}\per\kilo\watt\text{p}} \\
\ce{CO2} price & $90$ & \SI{}{\text{\euro}\per\mega\watt\hour}
\end{tabular}
\end{table}
\end{small}
\end{textblock*}

\end{frame}

\section{Sensitivity}
\begin{frame}
\frametitle{The cost of undisturbed landscapes}
\framesubtitle{Sensitivity to capital cost of solar PV}
\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
\includegraphics[width=\textwidth, keepaspectratio]{./figures/sens_nrg}
\end{textblock*}

\begin{textblock*}{4cm}(11cm, 2.25cm)
\begin{small}
Capital cost estimates for solar PV in 2030
\begin{table}
\begin{tabular}{p{1.8cm} | p{0.3cm} l}
Small-scale rooftop & $830$ & \SI{}{\text{\euro}\per\kilo\watt\text{p}} \\
Utility-scale open space & $280$ & \SI{}{\text{\euro}\per\kilo\watt\text{p}} \\
\end{tabular}
\end{table}
\end{small}
\end{textblock*}
\end{frame}

\begin{frame}
\frametitle{The cost of undisturbed landscapes}
\framesubtitle{Sensitivity to capital cost of solar PV}
\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
\includegraphics[width=\textwidth, keepaspectratio]{./figures/undisturbed_low}
\end{textblock*}

\begin{textblock*}{4cm}(11cm, 2.25cm)
\begin{small}
\begin{table}
\begin{tabular}{p{1.8cm} | p{0.3cm} l}
Capital cost of solar PV & $280$ & \SI{}{\text{\euro}\per\kilo\watt\text{p}}
\end{tabular}
\end{table}
\end{small}
\end{textblock*}
\end{frame}

\section{Discussion}
\begin{frame}
\frametitle{The cost of undisturbed landscapes}
\framesubtitle{Discussion of results}
\begin{itemize}
\item Renewable resource quality held constant as capacity is expanded
\item Sub-national electricity transmission and distribution grids neglected
\item Technical operation of generators not fully represented \\(e.g. no unit commit, simplified balancing)
\item Electricity market-splitting has increased market concentration
\item Announced policy necessarily turns Austria into a net exporter of electricity\\
$\rightarrow$ "loop-flows" potentially avoided \\
$\rightarrow$ artificial transmission restriction between DE and AT could be eliminated
%\item use Austrian landscape for "overinvesting" in wind and exporting to Germany?
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item If we value \ce{CO2} emissions at 30 \SI{}{\text{\euro}\per\mega\watt\hour} or lower, onshore wind power can be substituted by open-space utility-scale solar PV at little loss
% and approve of open-space utility-scale solar PV, onshore wind power is sub-optimal

\item \ce{CO2} valuation above 30 \SI{}{\text{\euro}\per\mega\watt\hour} or a preference for rooftop PV allows for gains to be made from wind power deployment

\item Gains from wind power could be used to compensate the ones affected by local negative externalities of wind turbines

\item Complementing our analysis with spatially resolved estimates of wind turbine impacts, one could derive a spatially explicit plan for the socially optimal expansion of wind power in Austria
\end{itemize}
\end{frame}

 %%%Last Frame
 {
    \usebackgroundtemplate{
        \begin{picture}(320,290)
        \hspace{9.5cm}
          \includegraphics[width=0.6\linewidth]{./logos/refuel_logo_with_text}
%        \begin{picture}(200,305)
%        \hspace{7.7cm}
%          \includegraphics[width=0.6\linewidth]{refuel_logo_with_text}
        \end{picture}
    }
    \begin{frame}
        \frametitle{Thank you!}
        \vspace{0.5cm}
        \url{https://refuel.world}\\
        \url{https://github.com/inwe-boku/medea}\\
        \url{sebastian.wehrle@boku.ac.at}

		\vspace{0.6cm}
        \qrcode[height=2cm]{http://github.com/inwe-boku/medea}\\

        \vspace{1.65cm}
        \includegraphics[height=0.7cm]{./logos/creative-commons-by}

        \smallskip
        \tiny{We gratefully acknowledge support from the European Research Council \\(“reFUEL” ERC-2017-STG 758149).}

    \end{frame}
}   
\appendix
\backupbegin

\begin{frame}
\frametitle{The cost of undisturbed landscapes}
\framesubtitle{Unconstrained deployment of wind and solar power}

\begin{textblock*}{10cm}(0.75cm, 1.8cm) % {block width} (coords)
\begin{table}
\begin{tabular}{l r l}
\emph{Results for Austria} \\ \hline \hline
Added Solar PV & 0.0 & \SI{}{\tera\watt\hour} \\ \hline
Added Wind & 31.0 & \SI{}{\tera\watt\hour} \\ \hline \hline
System Cost\footnote[frame]{excluding cost for domestic transmission and distribution grids} & $3095.2$ & \SI{}{\mega\text{\euro}} \\ \hline
Trade balance & $746.8$ & \SI{}{\mega\text{\euro}} \\ \hline
Cost of air pollution & $184.4$ & \SI{}{\mega\text{\euro}} \\ \hline
\ce{CO2} emissions & $5227.3$ & \SI{}{\kilo\tonne} \\ \hline
Electricity generation & $92.8$ & \SI{}{\tera\watt\hour} \\ \hline
Curtailment & $2.4$ & \SI{}{\tera\watt\hour} \\ \hline
Net exports & $5.9$ & \SI{}{\tera\watt\hour} \\ \hline
Wholesale electricity price & $56.05$ & \SI{}{\text{\euro}\per\mega\watt\hour} \\ \hline
\end{tabular}
%\caption{Results for Austria}
\end{table}
\end{textblock*}

\begin{textblock*}{4cm}(11cm, 2.05cm)
\begin{small}
\begin{table}
\begin{tabular}{p{1.8cm} | p{0.3cm} l}
Capital cost of solar PV & $630$ & \SI{}{\text{\euro}\per\kilo\watt\text{p}} \\
\ce{CO2} price & $90$ & \SI{}{\text{\euro}\per\mega\watt\hour}
\end{tabular}
\end{table}
\end{small}
\end{textblock*}

\end{frame}

\begin{frame}
\frametitle{The value of undisturbed landscapes}
\framesubtitle{Local Negative Externalities of Wind Turbines}
%\begin{table}[h!]
%\caption{Estimates of Wind Turbine Impact on Property Prices}
\begin{table}
\centering
\begin{scriptsize}
	\begin{tabular}{p{3.2cm} l l l p{1.1cm} p{2.4cm}}
	\hline
	Authors & Country & Sample & Framework\footnote[frame]{DID $\ldots$ difference-in-differences, SAR $\ldots$ spatial autoregressive, SFE $\ldots$ spatial fixed effects} & Impact Proxy\footnote[frame]{d $\ldots$ distance, v $\ldots$ visibility} & Estimated Impact \\
	\hline \hline
%	\cite{Sims2008} & Cornwall, UK & 2000 - 2007 & hedonic & visibility dummies & insignificant \\ \hline
	Heintzelmann and Tuttle, 2012 & USA, NY & 2000 - 2009 & hedonic, SFE & d & $-3.6\%$ to $-9.6\%$, insignificant \\ \hline
	Jensen et al., 2014 & DNK & 2000 - 2011 & hedonic, SAR & d, v & $-4.35\%$ \\ \hline
	Lang et al., 2014 & USA, RI & 2000 - 2013 & DID & d, v & insignificant  \\ \hline
	Gibbons, 2015 & GBR, EAW & 2000 - 2011 & DID & d, v & $-6\%$ \\ \hline
	Hoen et al., 2015 & USA & 1996 - 2011 & DID & d & insignificant \\ \hline
	Dröes and Koster, 2016 & NED & 1985 - 2011 & DID & d & $-1.4\%$\\ \hline
	Sunak and Madlener, 2016 & GER, NRW & 1992 - 2010 & DID & v & $-9\%$ to $-14\%$ \\ \hline
	Jensen et al., 2018 & DNK & - & hedonic, SAR & d, \# of turbines & $-0.2\%$ to $-1.1\%$ \\ \hline
	Kussel et al., 2019 & GER & 2007 - 2015 & DID & d & $-6.0\%$ \\ \hline \hline
	\end{tabular}
%\label{tab:negex}
\end{scriptsize}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Compensating for local negative externalities}
\framesubtitle{An example}
\begin{textblock*}{7cm}(0.75cm, 2.0cm) % {block width} (coords)
OC of wind turbines in Austria
\vspace{-0.41cm}
\begin{table}
%\centering
\begin{footnotesize}
	\begin{tabular}{c | c | c | c}
	$oc_w$ & $\delta$ & NPV $oc_w$ & NPV $oc_w$ \\
	$[\text{\euro}\, \text{MW}^{-1} \text{a}^{-1}]$ & & $[\text{\euro}\, \text{MW}^{-1}]$ & $[\text{\euro} \, \text{turbine}^{-1}]$ \\ \hline \hline
	$40\,000$ & $0.05$ & $563\,758$ & $1\,973\,152$ \\ \hline
	$40\,000$ & $0.03$ & $696\,526$ & $2\,437\,841$ \\ \hline
	$10\,000$ & $0.05$ & $140\,939$ & $493\,288$ \\ \hline
	$10\,000$ & $0.03$ & $171\,131$ & $609\,460$ \\ \hline
	\end{tabular}
\end{footnotesize}
\end{table}
\end{textblock*}

\begin{textblock*}{7cm}(8.25cm, 2.0cm) % {block width} (coords)
Property value loss in Germany
%\vspace{-0.5cm}
\begin{table}
%\centering
\begin{footnotesize}
	\begin{tabular}{c | c | c | c}
	  & value & impact & loss \\
	  & \euro & \% & \euro \\ \hline \hline
	mean & $273\,786$ & $0.06$ & $16\,427$ \\ \hline
	mean & $273\,786$ & $0.10$ & $27\,379$ \\ \hline
	mean + 1 s.d. & $476\,923$ & $0.06$ & $28\,615$ \\ \hline
	mean + 1 s.d. & $476\,923$ & $0.10$ & $47\,692$ \\ \hline
	\end{tabular}
\end{footnotesize}
\end{table}
\end{textblock*}

\begin{textblock*}{14.5cm}(0.75cm, 5.6cm) % {block width} (coords)
\centering
Compensation potential
\vspace{-0.1cm}
\begin{table}
\begin{footnotesize}
	\begin{tabular}{c | c | c | c | c}
	$oc_w$  & $\delta$ & value & impact & \# property owners \\ \hline \hline
	$40\,000$ & $0.05$ & $273\,786$ & $0.06$ & $120$ \\ \hline
	$40\,000$ & $0.05$ & $273\,786$ & $0.10$ & $72$ \\ \hline
	$40\,000$ & $0.03$ & $476\,923$ & $0.06$ & $69$ \\ \hline
	$40\,000$ & $0.03$ & $476\,923$ & $0.10$ & $41$ \\ \hline
%	$10\,000$ & $0.05$ & $476\,923$ & $0.06$ & $1$ \\ \hline
	$10\,000$ & $0.03$ & $476\,923$ & $0.10$ & $18$ \\ \hline
	\end{tabular}
\end{footnotesize}
\end{table}
\end{textblock*}

\end{frame}
\backupend
\end{document}

### 2020-02-21 projects/asparagus/doc/plots_asparagus.py

@@ -1,0 +1,234 @@
+import itertools
+import os
+
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+
+import config as cfg
+
+
+# %% define plotting functions
+def plot_subn(df, fname, width=2, xlim=None, ylim=None, xlabel=None, ylabel=None, color=None):
+    """
+    plots each column of a pandas dataframe on a seperate subplot. Uses column names as subplot-titles.
+    :param df:
+    :param fname:
+    :param width:
+    :param xlim:
+    :param ylim:
+    :param xlabel:
+    :param ylabel:
+    :param color:
+    :return:
+    """
+    ncols = len(df.columns)
+    sub_length = np.ceil(ncols / width).astype(int)
+    sub_boxes = list(itertools.product(range(0, width), range(0, sub_length)))
+
+    figure, axis = plt.subplots(width, sub_length, figsize=(8, 5))
+    for col in range(0, ncols):
+        a = sub_boxes[col][0]
+        b = sub_boxes[col][1]
+
+        if color is not None:
+            axis[a, b].plot(df.iloc[:, col], color=color[col])
+        else:
+            axis[a, b].plot(df.iloc[:, col])
+
+        if xlabel is not None:
+            axis[a, b].set_xlabel(xlabel)
+
+        if (ylabel is not None) & (len(ylabel) == 1):
+            axis[a, b].set_ylabel(*ylabel)
+        elif (ylabel is not None) & (len(ylabel) == ncols):
+            axis[a, b].set_ylabel(ylabel[col])
+
+        if xlim is not None:
+            axis[a, b].set_xlim(xlim[0], xlim[1])
+
+        if ylim is not None:
+            axis[a, b].set_ylim(ylim[0], ylim[1])
+
+        axis[a, b].grid()
+        axis[a, b].set_title(df.columns[col])
+    plt.tight_layout()
+    plt.savefig(fname)
+    plt.close()
+
+
+def plot_lines(df, fname, xlim=None, ylim=None, xlabel=None, ylabel=None, color=None):
+    """
+    line plot of each column in df.
+    :param df:
+    :param fname:
+    :param xlim:
+    :param ylim:
+    :param xlabel:
+    :param ylabel:
+    :param color:
+    :return:
+    """
+    ncols = len(df.columns)
[…]

### 2020-02-21 projects/asparagus/results_asparagus.py

@@ -1,0 +1,82 @@
+# %% imports
+import os
+
+import pandas as pd
+from gams import *
+
+import config as cfg
+from projects.asparagus.settings_asparagus import *
+from src.tools.gams_io import gdx2df
+
+# -------------------------------------------------------------------------------------------------------------------- #
+# %% prepare GAMS workspace
+# -------------------------------------------------------------------------------------------------------------------- #
+ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
+
+# specify GAMS symbols to be read from output .gdx-file along with the dimensions of the corresponding output DataFrame
+# example: generate pandas DataFrame holding system cost and co2 emissions for each CO2 price scenario
+symbols_to_read = {
+    'cost_zonal': ([], ['z']),
+    'AnnCO2Emissions': ([], ['z']),
+    'AnnCurtail': ([], ['z']),
+    'AnnG': (['m'], ['z']),
+    'AnnGBiomass': ([], ['z']),
+    'AnnR': ([], ['z']),
+    'AnnSpendingEl': ([], ['z']),
+    'AnnProdSurplus': ([], ['z']),
+    'AnnX': ([], ['z']),
+    'AnnValueX': ([], ['z']),
+    'AnnValueI': ([], ['z']),
+    'AnnSIn': ([], ['z']),
+    'AnnSOut': ([], ['z']),
+    'AvgPriceEl': ([], ['z']),
+    'AvgPriceHt': ([], ['z']),
+    'add_r': (['n'], ['z']),
+    'add_g': (['i'], ['z']),
+}
+# cost_air_pol(z,f)
+var_names = ['syscost', 'co2emission', 'curtail', 'gen_el', 'gen_ht', 'gen_biomass', 'gen_renew', 'cons_spend',
+             'prod_surplus', 'exports', 'export_value', 'import_value', 'in_storages',
+             'gen_storages', 'price_el', 'price_ht', 'add_pv', 'add_ror', 'add_wind_off', 'add_wind_on', 'add_nuc',
+             'add_lig_stm', 'add_lig_stm_chp', 'add_lig_boa', 'add_lig_boa_chp', 'add_coal_sub', 'add_coal_sub_chp',
+             'add_coal_sc', 'add_coal_sc_chp', 'add_ng_stm', 'add_ng_stm_chp', 'add_ng_cbt_lo', 'add_ng_cbt_lo_chp',
[…]

### 2020-02-21 projects/asparagus/run_asparagus.py

@@ -1,0 +1,122 @@
+import os
+
+import pandas as pd
+from gams import *
+
+import config as cfg
+from projects.asparagus.settings_asparagus import *
+from src.tools.gams_io import reset_symbol, gdx2df, df2gdx, run_medea_project
+
+# path to main medea model to use
+GMS_MODEL = os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', PROJECT_NAME, 'opt', 'medea_main.gms')
+# -------------------------------------------------------------------------------------------------------------------- #
+# %% initialize GAMS, GAMS workspace and load model data
+# -------------------------------------------------------------------------------------------------------------------- #
+# import base data from gdx
+ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
+db_input = ws.add_database_from_gdx(
+    os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', PROJECT_NAME, 'opt', 'medea_main_data.gdx'))
+
+# %% read parameters that change in scenarios (and corresponding sets)
+# -------------------------------------------------------------------------------------------------------------------- #
+# *** read in all sets over which adjusted parameters are defined ***
+# general example for reading a GAMS set from the gdx-database to a python dictionary:
+# dict_set = {rec.keys[0] for rec in db_input['set_name']}
+# ---
+# read sets for calibration of power plant efficiencies
+dict_m = {rec.keys[0] for rec in db_input['m']}
+dict_f = {rec.keys[0] for rec in db_input['f']}
+dict_i = {rec.keys[0] for rec in db_input['i']}
+dict_r = {rec.keys[0] for rec in db_input['n']}
+
+# *** read in all parameters to be adjusted ***
+# general example for reading parameter 'PARAMETER_NAME' defined over 'set_name' to pandas DataFrame df
+# df = gdx2df(db_input, 'PARAMETER_NAME', ['set_name'], [])
+# ---
+EFFICIENCY_G = gdx2df(db_input, 'EFFICIENCY_G', ['i', 'm', 'f'], [])
+FEASIBLE_INPUT = gdx2df(db_input, 'FEASIBLE_INPUT', ['i', 'l', 'f'], [])
+INITIAL_CAP_G = gdx2df(db_input, 'INITIAL_CAP_G', ['z', 'i'], [])
[…]

### 2020-02-21 projects/asparagus/settings_asparagus.py

@@ -1,0 +1,94 @@
+# This file holds all scenario-specific settings and assumptions
+
+PROJECT_NAME = 'asparagus'
+# -------------------------------------------------------------------------------------------------------------------- #
+"""
+Description:
+determine optimal investment in Austrian dispatchable and intermittent energy generation and storage technologies 
+in 2030, given
+    a) the targets of the Austrian energy- and climate strategy ("mission 2030"), notably
+        a1) 100% of electricity consumption met from domestic, renewable sources (annual average, 
+            excluding industry own-consumption, i.e. effectively 90% of consumption)
+        a2) balanced international electricity exchange (on annual average)
+        a3) reduction of transport sector CO2-emissions by 7.2 mn t, translating in a fuel consumption reduction of 
+            27.3 TWh, which, if substituted by battery-electric mobility, raises electricity demand by 12.5 TWh.
+    b) restrictions on land availability for wind turbines due to social conflicts
+Our analysis allows for electricity exchange with Germany, Austria's most important electricity trading partner and 
+accounts for Germany's nuclear exit and its proposed coal phase-out.
+As restrictions on land availability for wind turbines are not specified, we generate scenarios of deployable wind 
+turbine capacity in the range between 0 and unconstrained deployment. This is repeated for a range of CO2 price 
+assumptions.
+For sensitivity analysis, we also vary the capitalcost of solar PV. 
+"""
+
+# campaigns
+dict_campaigns = {
+    'base': {
+        'co2_price': range(120, 29, -30),
+        'wind_cap': range(18, -1, -2),
+        'pv_cost': [36715, 16715]
+    },
+    'cheappv': {
+        'co2_price': [30, 60, 120],
+        'wind_cap': range(18, 17, -2),
+        'pv_cost': range(36715, 16026, -1000)
+    }
+}
+
+"""
+CAPACITY ASSUMPTIONS
+- GERMANY:
+  - nuclear exit
+  - proposal of coal commission is followed, i.e. reduction of 9 GW lignite and 11 GW coal till 2030
+  - oil and natural gas-fired capacities remain unchanged
+  - biomass additions according to EEG 2017 (+1.05 GW)
+- AUSTRIA:
[…]

### 2020-02-21 legacy/analyse_results.py
drop legacy code

@@ -1,45 +1,0 @@
-import os
-
-from gams import *
-
-import config as cfg
-from legacy.gams_wrappers import gdx2df
-
-# df = pd.read_csv(os.path.join(cfg.folder, 'medea', 'opt'))
-run_string = 'PoBu_EUA20_curt_htpmp_1'
-
-ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
-filename = f'medea_out_{run_string}.gdx'
-db_output = ws.add_database_from_gdx(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', filename))
-
-read_variables = {
-   'q_curtail': (['t'], ['r', 'prd']),
-    'flow': (['t'], ['r', 'rr']),
-    'invest_res': (['tec_itm'], ['r']),
-    'GEN_PROFILE': (['t'], ['r', 'tec_itm']),
-    'CONSUMPTION': (['t'], ['r', 'prd']),
-    'INSTALLED_CAP_ITM': (['tec_itm'], ['r'])
-        # 'ntc_invest': (['r'], ['r'])
-    }
-result_dict = {key: gdx2df(db_output, key, value[0], value[1])
-               for key, value in read_variables.items()}
-
-# #% data analysis
-
-df_windat = result_dict['GEN_PROFILE'].loc[:, ('AT', 'wind_on')] * (
-    result_dict['invest_res'].loc['wind_on', 'AT'] + result_dict['INSTALLED_CAP_ITM'].loc['wind_on', 'AT'])
-df_pvat = result_dict['GEN_PROFILE'].loc[:, ('AT', 'pv')] * (
-    result_dict['invest_res'].loc['pv', 'AT'] + result_dict['INSTALLED_CAP_ITM'].loc['pv', 'AT'])
-df_rorat = result_dict['GEN_PROFILE'].loc[:, ('AT', 'ror')] * (
-    result_dict['invest_res'].loc['ror', 'AT'] + result_dict['INSTALLED_CAP_ITM'].loc['ror', 'AT'])
-df_itm = df_windat + df_pvat
-df_consat = result_dict['CONSUMPTION'].loc[:, ('AT', 'Power')]
-df_curt = result_dict['q_curtail'].loc[:, ('AT', 'Power')]
-
-df_surpat = df_windat + df_pvat + df_rorat - df_consat
-df_surpat_pos = df_surpat.copy()
[…]

### 2020-02-21 legacy/compile_results.py

@@ -1,52 +1,0 @@
-import os
-
-import pandas as pd
-from gams import *
-
-import config as cfg
-from legacy.gams_wrappers import gdx2df
-
-output_folder = 'PolicyBurden'
-campaign_string = 'PoBu_EUA40_curt_htpmp'
-directory = os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt')
-
-ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
-# for file in os.listdir(directory):
-#     filename = os.fsdecode(file)
-#     if campaign_string in filename:
-
-df_all = pd.DataFrame()
-for num in range(0, 16):
-    filename = f'medea_out_{campaign_string}_{num}.gdx'
-    db_output = ws.add_database_from_gdx(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'opt', filename))
-
-    read_variables = {
-        'cost': ([], ['r']),
-        'emissions': ([], ['r']),
-        'invest_res': (['tec_itm'], ['r']),
-        'invest_thermal': (['tec'], ['r']),
-        'ntc_invest': (['r'], ['r']),
-        'annual_price_el': ([], ['r']),
-        'annual_price_ht': ([], ['r']),
-        'annual_pumping': ([], ['r']),
-        'annual_turbining': ([], ['r']),
-        'annual_generation': (['prd'], ['r']),
-        'annual_curtail': ([], ['r']),
-        'annual_netflow': ([], ['r']),
-        'producer_surplus': ([], ['r']),
-    }
-
-    result_dict = {key: gdx2df(db_output, key, value[0], value[1])
-                   for key, value in read_variables.items()}
-
-    df = pd.DataFrame(columns=['AT', 'DE'])
-    for key, value in result_dict.items():
-        if value.index.any() == 'Value':
-            value.index = [key]
-        df = df.append(value)
-    df.columns = pd.MultiIndex.from_product([[f'WOn_{num}'], df.columns])
-    df_all = pd.concat([df_all, df], axis=1)
-
-df_all = df_all.sort_index(axis=1, level=1)
-df_all.to_csv(os.path.join(cfg.MEDEA_ROOT_DIR, 'medea', 'output', output_folder, f'results_{campaign_string}.csv'),
-              sep=';', encoding='utf-8')

### 2020-02-21 legacy/gams_wrappers.py

@@ -1,88 +1,0 @@
-import pandas as pd
-from gams import *
-
-
-def reset_parameter(gams_db, parameter_name, df):
-    """
-    writes values in df to the already existing parameter "parameter name" in GAMS-database gams_db
-    :param gams_db: a GAMS database object
-    :param parameter_name: a string with the parameter name
-    :param df: a pandas dataframe with one line per value and all correspondong dimensions in the index
-    :return: modifies gams database, does not return anything
-    """
-    gams_parameter = gams_db.get_parameter(parameter_name)
-    gams_parameter.clear()
-    if gams_parameter.get_dimension() > 0:
-        for row in df.itertuples():
-            gams_parameter.add_record(row[0]).value = row[1]
-    elif gams_parameter.get_dimension() == 0:
-        for row in df.itertuples():
-            gams_parameter.add_record().value = row[1]
-    else:
-        raise ValueError('dimension_list must be list or integer')
-
-
-def gdx2df(db_gams, symbol, index_list, column_list):
-    """
-    writes data from a GAMS gdx to a pandas dataframe.
-    :param db_gams: a GAMS database object
-    :param symbol: string of the GAMS symbol name
-    :param index_list:
-    :param column_list:
-    :return:
-    """
-    sym = db_gams.get_symbol(symbol)
-    if isinstance(sym, GamsParameter):
-        gdx_dict = {tuple(obj.keys): obj.value for obj in sym}
-    elif isinstance(sym, GamsVariable):
-        gdx_dict = {tuple(obj.keys): obj.level for obj in sym}
-    elif isinstance(sym, GamsEquation):
-        gdx_dict = {tuple(obj.keys): obj.marginal for obj in sym}
-    elif isinstance(sym, GamsSet):
-        raise ValueError('gdx2df for sets not yet implemented')
-
-    gdx_df = pd.DataFrame(list(gdx_dict.values()), index=pd.MultiIndex.from_tuples(gdx_dict.keys()), columns=['Value'])
-    gdx_df.index.names = db_gams.get_symbol(symbol).domains_as_strings
-    gdx_df = pd.pivot_table(gdx_df, values='Value', index=index_list, columns=column_list)
-    if 't' in index_list:
-        gdx_df.reset_index(inplace=True)
-        gdx_df['tix'] = pd.to_numeric(gdx_df['t'].str.split(pat='t').str.get(1))
-        gdx_df.sort_values(by=['tix'], inplace=True)
-        gdx_df.set_index(index_list, drop=True, inplace=True)
[…]

### 2020-02-21 legacy/main.py

@@ -1,13 +1,0 @@
-
-# %% compilation of raw data
-data/...
-
-# %% check config
-config.py
-
-# %% instantiate model with data (i.e. write data to gdx)
-instantiation/instantiate.py
-
-# %% execute model
-#
-scenario_solve.py

### 2020-02-21 legacy/scenario_results.py

@@ -1,231 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-use medea to analyze Austrian energy and climate policy
-@author: Sebastian Wehrle
-"""
-
-import os
-import subprocess
-
-import numpy as np
-import pandas as pd
-from gams import *
-
-import config as cfg
-from legacy.gams_wrappers import reset_parameter, gdx2df, df2gdx
-
-# %% scenario assumptions
-# Calibration of model (via efficiencies and capacities) to 2017 prices, fuel burn and emissions
-baseline = {
-    # [plant efficiencies] -------------- #
-    'e_Nuclear': [1],
-    'e_Biomass': [1],
-    'e_Lignite': [1.15],
-    'e_Coal': [1.225],
-    'e_Gas': [1.15],
-    'e_Oil': [1],
-    # [capacity availability of plants ] ------ #
-    'av_Nuclear': [0.72],
-    'av_Lignite': [0.88],
-    'av_Coal': [0.75],
-    'av_Coal_chp': [0.775],
-    'av_Gas': [0.9],
-    'av_Gas_chp': [0.9],
-    'av_Oil': [1.0],
-    'av_Biomass': [0.8]
-}
-capcy = {
-    'AT': {
-        'bio': 0.8,
-        'coal': 0.75,
-        'lig': 0.8,
-        'ng': 0.85,
-        'nuc': 0.8,
-        'oil': 0.85
-    },
-    'DE': {
-        'bio': 0.8,
-        'coal': 0.75,
-        'lig': 0.88,
-        'ng': 0.9,
-        'nuc': 0.72,
-        'oil': 0.9
-    }
-}
-
-capcy_2030 = {
-    'AT': {
-        'bio': 0.8,
-        'coal': 0,
-        'lig': 0,
-        'ng': 0.85,
-        'nuc': 0.8,
-        'oil': 0.85
-    },
-    'DE': {
-        'bio': 0.85,
-        'coal': 0.35,
-        'lig': 0.45,
-        'ng': 0.9,
-        'nuc': 0.0,
-        'oil': 0.9
-    }
-}
-
-# Germany 2030: nuclear exit, coal phase out, renewables expansion as in EEG 2017
-scenario_RES_DE2030 = {
-    # nuclear phase-out in Germany
-    # 'av_Nuclear': [0.0],
-    # reduction of coal and lignite capacities in line with coal phase-out
-    # 'av_Lignite': [0.45],
[…]

### 2020-02-21 legacy/scenario_solve.py

@@ -1,393 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-use medea_lin to estimate pass-through from emission prices to power prices
-@author: Sebastian Wehrle
-"""
-
-import os
-from itertools import product
-
-import pandas as pd
-from gams import *
-
-import config as cfg
-from legacy.gams_wrappers import reset_parameter, gdx2df
-from legacy.postprocessing.postprocess import postprocess
-from legacy.solve import solve
-
-# --------------------------------------------------------------------------- #
-# %% definition of scaling, carbon pricing and 'setting'
-# --------------------------------------------------------------------------- #
-campaign = 'PolicyBurden'
-# currently implemented res scalings:
-# * res_pen - renewable capacity is scaled to a generation level relative to electricity demand ('penetration')
-# * res_base - renewable capacity is scaled relative to res capacity in base year
-scaling = 'res_base'
-# currently implemented carbon price scenarios:
-# * range - use a range of carbon prices from 0 to 100 EUR per t CO2 in 5 EUR per t CO2 increments
-# * scaling - scale actual carbon price as in base year; default scaling factor 1
-carbon_pricing = 'range'
-
-# --------------------------------------------------------------------------- #
-# %% scenario definition
-# --------------------------------------------------------------------------- #
-# target level for generation in 2017 [TWh]
-tgt = pd.DataFrame(
-    {'Nuclear': [76.3], 'Lignite': [147.5], 'Coal': [94.1], 'Gas': [89.4], 'Oil': [6.1], 'Biomass': [47.2]})
-
-# implemented switches:
-# '2030', 'PriceCoal075', 'PriceCoal125', 'CapStorage_2', 'CapStorage_10', 'CapCoal05', 'CapCoal0', 'CapLignite05',
-# 'CapLignite0', 'CapGas175', 'CapGas_2', 'DemAncillary05', 'DemAncillary_0', 'DemDistHeat_0'
-# all switches may be combined. For example, 'PriceCoal075_2030' projects coal prices into 2030
-
[…]

### 2020-02-21 legacy/solve.py

@@ -1,325 +1,0 @@
-import os
-import subprocess
-
-import pandas as pd
-from gams import *
-
-import config as cfg
-from legacy.gams_wrappers import gdx2df
-
-"""
-solves medea regional models (iteratively), compiles and writes output to csv-files 
-"""
-
-
-def solve(output_folder, output_name, scenario_iteration, gen_target):
-    # --------------------------------------------------------------------------- #
-    # initialize GAMS workspace
-    # --------------------------------------------------------------------------- #
-    ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
-
-    # read input data from MEDEA_data.gdx
-    db_input = ws.add_database_from_gdx(
-        os.path.join(cfg.MEDEA_ROOT_DIR, f'medea', 'opt', 'medea_data.gdx'))
-    # read sets for clusters, hydro storage plants and products from db_input
-    clust_dict = {rec.keys[0] for rec in db_input['tec']}
-    hsp_dict = {rec.keys[0] for rec in db_input['tec_hsp']}
-    prd_dict = {rec.keys[0] for rec in db_input['prd']}
-    reg_dict = {rec.keys[0] for rec in db_input['r']}
-
-    # TODO: Activate energy stored in reservoirs - STORAGE_LEVEL
-    """
-    # read stored energy from db_input
-    q_enstor_dict = {tuple(rec.keys): rec.value for rec in db_input['STORAGE_LEVEL']}
-    df_enstor = pd.DataFrame(list(q_enstor_dict.values()), index=pd.MultiIndex.from_tuples(q_enstor_dict.keys()))
-    del q_enstor_dict
-    """
-    # read storage props from db_input
-    df_storprop = gdx2df(db_input, 'HSP_PROPERTIES', ['tec_hsp'], ['r', 'props'])
-
-    # read mapping of fuels to plants from EFFICIENCY parameter
-    df_eff = gdx2df(db_input, 'EFFICIENCY', ['tec', 'f'], ['prd'])
-
-    # read emission factors
-    df_emf = gdx2df(db_input, 'EMISSION_INTENSITY', ['f'], [])
-
-    # read day-ahead electricity price
-    df_pda = gdx2df(db_input, 'PRICE_DA', ['t'], [])
-    df_pda.columns = ['price_da']
-
-    # --------------------------------------------------------------------------- #
-    # initialize db and sets for iterations, declare parameters
-    # --------------------------------------------------------------------------- #
[…]

### 2020-04-07 config.py
automatic recognition of MEDEA_ROOT_DIR

@@ -43,12 +43,12 @@ is_64bit = sys.maxsize > 2 ** 32
 if not is_64bit:
     print('System appears to be 32-bit. 64-bit system recommended.')
 else:
+    rpath = os.path.normpath(os.getcwd())
+    pos_medea = rpath.find('medea')
+    MEDEA_ROOT_DIR = rpath[0:pos_medea + 6]
     if sys_id == 'win32 WINP1218':
-        MEDEA_ROOT_DIR = r'Y:\git_repos\medea'
         GMS_SYS_DIR = os.path.join(r'C:\GAMS\win64', version[0:4])
     elif sys_id == 'darwin Sebastians-MBP':
-        MEDEA_ROOT_DIR = '/Users/seb/git_repos/medea'
         GMS_SYS_DIR = os.path.join(f'/Applications/GAMS{version[0:4]}', 'sysdir')
     else:
-        MEDEA_ROOT_DIR = r'D:\git_repos\medea'
         GMS_SYS_DIR = os.path.join(r'C:\GAMS\win64', version[0:4])

### 2020-04-07 projects/asparagus/doc/slides.tex
updated slides.tex in projects/asparagus/doc

@@
-\documentclass[aspectratio=169, xcolor=dvipsnames]{beamer}
+\documentclass[aspectratio=1610, xcolor=dvipsnames,handout]{beamer} %aspectratio=169, 
 \usetheme{Pittsburgh}
 \usepackage{tikz}
+\usepackage{pgfplots}
 \usepackage{booktabs}
@@
 \usepackage{xfrac}
+\usepackage[export]{adjustbox}
 
@@
 \usepackage[T1]{fontenc}
+\usepackage{lmodern}
+\usepackage{graphicx}
+\usepackage{amsmath}
 
+\pgfplotsset{compat = 1.17}
 \usetikzlibrary{arrows.meta}
@@
 \title[The cost of undisturbed landscapes]{The cost of undisturbed landscapes}
-\subtitle{Assessing systemic effects of renewables expansion in Austria}
+\subtitle{Towards designing energy systems from a social perspective}
 \author{Sebastian Wehrle, Johannes Schmidt}
 \institute{University of Natural Resources and Life Sciences, Vienna}
-\date{February 25, 2020\\ \medskip NOeG Annual Meeting 2020}
+\date{May 7, 2020\\ \medskip EGU General Assembly 2020}
 
-\setbeamertemplate{section page}{
-	\begin{textblock*}{4cm}(11cm, 4cm) % {block width} (coords)
-	{\LARGE \insertsection}
-	\end{textblock*}
+%\setbeamertemplate{section page}{
+%	\begin{center}
+%	{\usebeamerfont[fg]{section title} \insertsection}
+%	\end{center}
+%}
+\AtBeginSection[]{
+  \begin{frame}
+  \vfill
+%  \centering
+%  \begin{beamercolorbox}[sep=8pt,center,shadow=false,rounded=true]{title}
+\tableofcontents[currentsection]
+%\begin{textblock*}{10cm}(7cm, 2.0cm) % {block width} (coords)
+%\adjincludegraphics[height=6cm,trim={0 0 {.3\width} 0},clip]{./figures/pv}
+%\end{textblock*}
+ %   \usebeamerfont{title} \usebeamercolor{title} \insertsection\par%
+%  \end{beamercolorbox}
+  \vfill
+  \end{frame}
 }
@@
         \usebackgroundtemplate{
-	        \begin{picture}(320,300)
+	        \begin{picture}(320,325)
     	        \hspace{9.0cm}
@@
 
-\section{Introduction}
+    \section{Background \& Motivation}\label{sec:background}
     \begin{frame}
-\frametitle{Austrian energy policy objectives}
-\framesubtitle{according to government programme 2020-2024}
+        \frametitle{State of the Art}
+        Power system models used to
         \begin{itemize}
-\item 100\%\footnote[frame]{excluding system services and industry own consumption.
-At current levels this equals 10\% of consumption, i.e.
-actual target is around 90\%.} of electricity demand from domestic renewable sources on annual balance by 2030 \pause
-\item additional annual electricity generation of \SI{27}{\tera\watt\hour} is expected to suffice \pause
-\item technology-specific additions:
+            \item demonstrate the feasibility of highly renewable energy systems
+            \item determine the system cost-minimizing mix of renewable energy technologies
         \end{itemize}
+        find significant capacities of renewable energy technologies (RET), largely wind power, need to be added.
+        \medskip
 
+        Researchers are increasingly aware that social and political restrictions on such large-scale RET expansions must be
+        recognized.
+        \begin{itemize}
+            \item RET deployment in power system models often limited on the upside by results from analyses of land
+            availability, e.g. by excluding areas for technical or legal reasons, or because they are constested by stakeholders
+        \end{itemize}
+        \medskip
+
+        Yet, such `binary' analysis neglects the trade-offs faced by a social planner deciding on RET expansion.
+    \end{frame}
+
+    \begin{frame}
+        \frametitle{Motivation}
+        \begin{itemize}
+            \item So far, analysis of \emph{system cost} minimizing energy systems
+            \begin{itemize}
+                \item Investment cost
+                \item Operation and maintenance cost (including fuel and emission cost)
+                \smallskip
+                \item[$\rightarrow$] includes external effects of RET on power system, such as
+                \begin{itemize}
+                    \item Profile cost
+                    \item Transmission expansion cost
+                    \item Balancing cost
+                \end{itemize}
+            \end{itemize}
+        \end{itemize}
+        \medskip
+        \begin{itemize}
+            \item We need analysis of \textbf{\emph{social cost}} minimizing energy systems, including
+            %\item We need to develop tools to minimize \textbf{social cost} of power systems, including
+            \begin{itemize}
+                \item Local external cost, such as visual impact on landscape, harm to wildlife, noise, glaring, \ldots
+                \item Benefits of system transition, such as avoided cost of local air pollution
+                \item[$\rightarrow$] includes external effects of RET beyond power system
+            \end{itemize}
+        \end{itemize}
+    \end{frame}
+
+    \begin{frame}
+        \frametitle{Framework}
+
+        \begin{textblock*}{5cm}
+            (1.0cm,2cm)
+            Consider a renewable energy technology that
+            \begin{itemize}
+                \item saves system cost (relative to its best alternative), but
+                \item comes at social cost (e.g. due to local negative external effects)
+            \end{itemize}
+            \medskip
+
+            A social planner should deploy this RET (instead of another one) up to the capacity at which
+            \begin{itemize}
+                \item system cost saving equals
+                \item added social cost
+            \end{itemize}
+        \end{textblock*}
+
+        \begin{textblock*}{8.0cm}
+            (6.5cm,2cm)
+            \begin{tikzpicture}
+                \begin{axis}[
+                    width=\textwidth,
+                    clip=false,
+                    axis lines=left,
+                    ticks=none,
+                    xlabel={capacity substituted},
+                    ylabel={cost},
+                    domain=0.5:1.75,
+                    xmin=0.5, xmax=2,
+                    ymin=0, ymax=4,
+                    samples=200
+                ]
+                    \addplot[blue, thick] (x, 1.3*x^1.1+0.6);
+                    \node at (axis cs: 1.925,3) [blue, align=left] {saved\\ system\\ cost};
+
+                    \addplot[bokugreen, thick] (x, -2*x^0.8+4);
+                    \node at (axis cs: 2.0,1) [bokugreen, align=left]{harm \\(from RET\\ externality)};
+                \end{axis}
+            \end{tikzpicture}
+        \end{textblock*}
+    \end{frame}
+
+    \begin{frame}
+        \frametitle{Approach}
+        Focus on the social cost of power systems, including local externalities of \emph{wind power}
+        \begin{itemize}
+            \item Seems to be at center of societal debate currently
+            \item Local externalities well researched (compared to externalities of open-space PV)
+        \end{itemize}
+        \medskip
+        An assessment of socially optimal renewables expansion must be based on two pillars
+        \begin{enumerate}
+            \item \textbf{\textcolor{bokugreen}{The cost of undisturbed landscapes}}
+            \begin{itemize}
+                \item[--] energy system costs of substituting wind power with other RET
+                \item[--] the \emph{opportunity cost} of wind power (versus its best alternative)
+            \end{itemize}
+            \item \textcolor{bokugreen}{The benefit of undisturbed landscapes}
+            \begin{itemize}
+                \item[--] avoided (local) negative external effects of wind turbines
+                \item[--] avoided (local) external cost of wind power
+            \end{itemize}
+        \end{enumerate}
+    \end{frame}
+
+    \section{Method}\label{sec:method}
+
+    \begin{frame}
+        \frametitle{The cost of undisturbed landscapes}
+        \framesubtitle{Estimating the opportunity cost of wind power}
+        To determine the opportunity cost of substituting wind power with solar PV, we
+        \begin{enumerate}
+            \item derive the unrestricted system cost-minimizing deployment of wind and solar power \label{step1}
+            \item restrict deployment of wind power by a small margin (so that the next best RET substitutes for wind power) and
+            observe net system cost $c_{net}$ (calculated as total system cost including air pollution cost net of trade balance) \label{step2}
+            \item repeat \ref{step2} till no wind power can be deployed
+            \item the opportunity cost of wind power is approximated by the change in $c_{net}$ in response to a change in wind
+            power capacity $w$ deployed.
+            \begin{align*}
+                OC_w = \frac{\Delta c_{net}}{\Delta w}
+            \end{align*}
+        \end{enumerate}
+    \end{frame}
+
+    \begin{frame}
+        \frametitle{The case of Austria}
+        \begin{itemize}
+            \item Policy goal: \SI{100}{\percent}\footnote[frame]{excluding system services and industry own consumption.
+            At current levels this equals 10\% of consumption, i.e. actual target is around 90\%.} of electricity demand from
+            domestic renewable sources on annual balance by 2030
+            \begin{itemize}
+                \item[$\rightarrow$] Ideal frame for studying substitution / opportunity cost
+            \end{itemize}
+            \item necessarily turns Austria into a net exporter of electricity
+            \item Official expectation: additional generation of \SI{27}{\tera\watt\hour} sufficient
+        \end{itemize}
         \begin{table}
@@
                 Hydro (run-of-river) & 5.72 & 28.34 & +5 & 33.34 & 6.73 \\ \hline
-%Biomass & 1 & 5 & +1 & 6 & 2 \\ \hline
+                Biomass & $\sim 1$ & 4.78 & +1 & 5.78 & $<2$ \\ \hline
             \end{tabular}
@@
 
+%\begin{frame}
+%\frametitle{Austrian energy policy objectives}
+%\framesubtitle{according to government programme 2020--2024}
+%\begin{itemize}
+%\item 100\%\footnote[frame]{excluding system services and industry own consumption.
+%    At current levels this equals 10\% of consumption, i.e. actual target is around 90\%.} of electricity demand from
+%    domestic renewable sources on annual balance by 2030 \pause
+%\begin{itemize}
+%\item[$\rightarrow$] necessarily turns Austria into a net exporter of electricity \pause
+%\end{itemize}
+%\item additional annual electricity generation of \SI{27}{\tera\watt\hour} is expected to suffice \pause
+%\item technology-specific additions:
+%\end{itemize}
+%
+%\begin{table}
+%\centering
+%\begin{tabular}{| l | r | r | r | r | r |}
+%\hline
+% & 2018 & 2018\footnote[frame]{meteorological conditions as in 2016} & Policy & 2030 & 2030 \\
+% & [GW] & [TWh] & [TWh] & [TWh] & [GW] \\ \hline \hline
+%Solar PV & 1.44 & 1.23 & +11 & 12.23 &  14.27 \\ \hline
+%Wind (onshore) & 3.05 & 6.14 & +10 & 16.14 & 8 \\ \hline
+%Hydro (run-of-river) & 5.72 & 28.34 & +5 & 33.34 & 6.73 \\ \hline
+%Biomass & $\sim 1$ & 4.78 & +1 & 5.78 & $>2$ \\ \hline
+%\end{tabular}
+%\end{table}
+%\end{frame}
+%
+%
+%\begin{frame}
+%\frametitle{Why wind is not solar}
+%\framesubtitle{A side note on imperfect substitutes}
+%%\begin{textblock*}{4cm}(0.75cm, 2.75cm)
+%%Aggregate profile for Austria
+%\begin{textblock*}{4cm}(0.75cm, 2.75cm)
+%\begin{itemize}
+%\item secure system operation requires $S=D$ at any point in time
+%\item electricity not easily storable
+%\item >>transforming<< electricity feed-in to end-use electricity is costly
+%\end{itemize}
+%\end{textblock*}
+%
+%\only<2>{
+%\begin{textblock*}{10cm}(5.5cm, 2.25cm) % {block width} (coords)
+%\centering
+%Aggregate generation profile in Austria
+%\includegraphics[width=\textwidth, keepaspectratio]{./figures/wind_not_pv}
+%\end{textblock*}
+%}
+%\end{frame}
+%
+
     \begin{frame}
-\frametitle{Stylized Facts}
+        \frametitle{The case of Austria}
+        \framesubtitle{Some Stylized Facts}
         \begin{itemize}
-\item Apart from wind and solar, potentials for renewable electricity generation in Austria largely exhausted
-\item Under announced policies, solar PV is only large-scale substitute to wind power
-\item Increasing social conflict around the large-scale expansion of onshore wind power
-\item Traditional power system models do not account for local negative externalities of renewable energy generators, such as:
+            \item Social conflict around the large-scale expansion of onshore wind power
+            \item Solar PV is the only feasible large-scale substitute for wind power
             \begin{itemize}
-	\item visual impact on landscape
-	\item harm to wildlife
-	\item noise, flickering, glaring
+                \item Use of nuclear power banned in a 1978 referendum
+                \item Hydro power potential largely exploited
+                \item Biomass neither ecologically nor economically sustainable
             \end{itemize}
+            \item Germany is Austria's most important electricity trading partner, by far
         \end{itemize}
     \end{frame}
+%
+%\begin{frame}
+%\frametitle{The problem from a social planner's perspective}
+%How to account for local negative externalities in renewables expansion planning?
+%
+%\bigskip
+%\textcolor{bokugreen}{The cost of undisturbed landscapes}
+%\begin{itemize}
+%\item assess energy system effects and costs of substituting wind power with solar PV
+%\end{itemize}
+%
+%\bigskip
+%\textcolor{bokugreen}{The value of undisturbed landscapes}
+%\begin{itemize}
+%\item estimates of the negative external effect of wind turbines reported in literature
+%\end{itemize}
+%\end{frame}
 
     \begin{frame}
-\frametitle{Why wind is not solar}
-\framesubtitle{A side note on imperfect substitutes}
-\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
-\centering
-Aggregate generation profile in Austria
-\includegraphics[width=\textwidth, keepaspectratio]{./figures/wind_not_pv}
-\end{textblock*}
-\begin{textblock*}{4cm}(11cm, 2.25cm)
-%Aggregate profile for Austria
-\bigskip
+        \frametitle{The case of Austria}
+        \framesubtitle{Modeling the Austrian Power System}
         \begin{itemize}
-\item secure system operation requires $S=D$ at any point in time
-\item electricity not easily storable
-\item grid operators >>transform<< electricity feed-in to end-use electricity
-\end{itemize}
-\end{textblock*}
-\end{frame}
-
-\begin{frame}
-\frametitle{The problem from a social planner's perspective}
-How to account for local negative externalities in renewables expansion planning?
-
-\bigskip
-\textcolor{bokugreen}{The cost of undisturbed landscapes}
-\begin{itemize}
-\item assess energy system effects and costs of substituting wind power with solar PV
-\end{itemize}
-
-\bigskip
-\textcolor{bokugreen}{The value of undisturbed landscapes}
-\begin{itemize}
-\item estimates of the negative external effect of wind turbines reported in literature
-\end{itemize}
-\end{frame}
-
-\section{Research agenda}
-\begin{frame}
-\frametitle{The cost of undisturbed landscapes}
-\framesubtitle{Approach}
-\begin{itemize}
-\item Resemble Austrian power system in 2030
-\item include most important electricity trading partner Germany
-\item set policy target of meeting $90\%$ of demand in 2030 from domestic renewable sources
-\item incorporate announced electricity system targets for Germany in 2030
+            \item Resemble prospective power system in Austria in 2030
+            \item Set policy target of meeting at least $90\%$ of demand in 2030 from domestic renewable sources
+            \medskip
+            \item Include most important electricity trading partner Germany
+            \item Incorporate announced electricity system targets for Germany in 2030
             \begin{itemize}
@@
             \end{itemize}
-\item simulate prospective electricity system with \emph{medea}
+            \medskip
+            \item Simulate operation of and investment in prospective electricity system with \emph{medea}
         \end{itemize}
@@
 \end{footnotesize}
+\bigskip
+
 \medskip
 \begin{center}
-\emph{medea} is available on \href{https://github.com/inwe-boku/medea}{github.com/inwe-boku/medea} under an open MIT license
+    \emph{medea} is available on \textcolor{blue}{\href{https://github.com/inwe-boku/medea}{github.com/inwe-boku/medea}}
+    under an open MIT license
 \end{center}
@@
 
-\begin{frame}
-\frametitle{The cost of undisturbed landscapes}
-\framesubtitle{Estimating the opportunity cost of wind power}
-\begin{itemize}
-\item[1)] derive unrestricted optimal deployment of wind and solar power
-\item[2)] restrict deployment of wind power by a small margin \\($\rightarrow$ solar PV substitutes for wind)
-\item[3)] repeat till no wind power can be deployed
-\end{itemize}
+\section{Results}
 
-\bigskip
-We approximate the cost of undisturbed landscapes (i.e.
-the forgone value of wind power $w$) by the change in net cost of the electricity system including air pollution cost $c_{net}$ in response to a change in deployed wind power $w$, i.e.
-\begin{align*}
-oc_w = \frac{\Delta c_{net}}{\Delta w}
-\end{align*}
-
-\end{frame}
-
-\section{Results}
 \begin{frame}
@@
 
-\vspace{3.75cm}
+\vspace{4.25cm}
 \quad wind power and solar PV \\ \quad are substitutes
@@
 
-\begin{textblock*}{9.5cm}(1.5cm, 7.7cm)
+\begin{textblock*}{9.5cm}(1.5cm, 8.2cm)
 \begin{tikzpicture}%[thick]
@@
 
-\section{Discussion}
+\section{Discussion \& Conclusion}
 \begin{frame}
-\frametitle{The cost of undisturbed landscapes}
-\framesubtitle{Discussion of results}
+    \frametitle{Discussion of results}
     \begin{itemize}
@@
     \begin{itemize}
-\item If we value \ce{CO2} emissions at 30 \SI{}{\text{\euro}\per\mega\watt\hour} or lower, onshore wind power can be substituted by open-space utility-scale solar PV at little loss
-% and approve of open-space utility-scale solar PV, onshore wind power is sub-optimal
-
-\item \ce{CO2} valuation above 30 \SI{}{\text{\euro}\per\mega\watt\hour} or a preference for rooftop PV allows for gains to be made from wind power deployment
-
-\item Gains from wind power could be used to compensate the ones affected by local negative externalities of wind turbines
-
-\item Complementing our analysis with spatially resolved estimates of wind turbine impacts, one could derive a spatially explicit plan for the socially optimal expansion of wind power in Austria
+        \item We demonstrated that restricting wind power can come at significant opportunity cost
+        \item Socially optimal wind power deployment depends on cost and benefits of undisturbed landscapes
+        \item Our analysis constitutes one of two pillars needed to determine a spatially explicit, socially
+        optimal renewables expansion
+        \begin{itemize}
+            \item[$\rightarrow$] need to complement findings with a spatially resolved assessment of the negative
+            externalities of wind turbines, for example approximated by wind turbine impact on property prices
+        \end{itemize}
+        \medskip
+        \item For the case of Austria, we find
+        \begin{itemize}
+            \item little social value lost from substituting onshore wind power with utility-scale, open-space solar PV,
+            \emph{if} \ce{CO2} emissions are valued at \SI{30}{\text{\euro}\per\mega\watt\hour} or lower
+            \item potential gains from wind power \emph{if} (i) there is a preference for rooftop solar PV or (ii)
+            the cost of \ce{CO2} emissions exceeds \SI{30}{\text{\euro}\per\mega\watt\hour}
+            \item Gains from wind power deployment could be used to compensate individuals affected by local wind turbine
+            externalities
+        \end{itemize}
     \end{itemize}
@@
         \smallskip
-        \tiny{We gratefully acknowledge support from the European Research Council \\(“reFUEL” ERC-2017-STG 758149).}
+        \tiny{We gratefully acknowledge support from the European Research Council \\(``reFUEL'' ERC-2017-STG 758149).}
 
@@
 }   
-\appendix
-\backupbegin
-
-\begin{frame}
-\frametitle{The cost of undisturbed landscapes}
-\framesubtitle{Unconstrained deployment of wind and solar power}
-
-\begin{textblock*}{10cm}(0.75cm, 1.8cm) % {block width} (coords)
-\begin{table}
-\begin{tabular}{l r l}
-\emph{Results for Austria} \\ \hline \hline
-Added Solar PV & 0.0 & \SI{}{\tera\watt\hour} \\ \hline
-Added Wind & 31.0 & \SI{}{\tera\watt\hour} \\ \hline \hline
-System Cost\footnote[frame]{excluding cost for domestic transmission and distribution grids} & $3095.2$ & \SI{}{\mega\text{\euro}} \\ \hline
-Trade balance & $746.8$ & \SI{}{\mega\text{\euro}} \\ \hline
-Cost of air pollution & $184.4$ & \SI{}{\mega\text{\euro}} \\ \hline
-\ce{CO2} emissions & $5227.3$ & \SI{}{\kilo\tonne} \\ \hline
-Electricity generation & $92.8$ & \SI{}{\tera\watt\hour} \\ \hline
-Curtailment & $2.4$ & \SI{}{\tera\watt\hour} \\ \hline
-Net exports & $5.9$ & \SI{}{\tera\watt\hour} \\ \hline
-Wholesale electricity price & $56.05$ & \SI{}{\text{\euro}\per\mega\watt\hour} \\ \hline
-\end{tabular}
-%\caption{Results for Austria}
-\end{table}
-\end{textblock*}
-
-\begin{textblock*}{4cm}(11cm, 2.05cm)
-\begin{small}
-\begin{table}
-\begin{tabular}{p{1.8cm} | p{0.3cm} l}
-Capital cost of solar PV & $630$ & \SI{}{\text{\euro}\per\kilo\watt\text{p}} \\
-\ce{CO2} price & $90$ & \SI{}{\text{\euro}\per\mega\watt\hour}
-\end{tabular}
-\end{table}
-\end{small}
-\end{textblock*}
-
-\end{frame}
-
-\begin{frame}
-\frametitle{The value of undisturbed landscapes}
-\framesubtitle{Local Negative Externalities of Wind Turbines}
-%\begin{table}[h!]
-%\caption{Estimates of Wind Turbine Impact on Property Prices}
-\begin{table}
-\centering
-\begin{scriptsize}
-	\begin{tabular}{p{3.2cm} l l l p{1.1cm} p{2.4cm}}
-	\hline
-	Authors & Country & Sample & Framework\footnote[frame]{DID $\ldots$ difference-in-differences, SAR $\ldots$ spatial autoregressive, SFE $\ldots$ spatial fixed effects} & Impact Proxy\footnote[frame]{d $\ldots$ distance, v $\ldots$ visibility} & Estimated Impact \\
-	\hline \hline
-%	\cite{Sims2008} & Cornwall, UK & 2000 - 2007 & hedonic & visibility dummies & insignificant \\ \hline
-	Heintzelmann and Tuttle, 2012 & USA, NY & 2000 - 2009 & hedonic, SFE & d & $-3.6\%$ to $-9.6\%$, insignificant \\ \hline
-	Jensen et al., 2014 & DNK & 2000 - 2011 & hedonic, SAR & d, v & $-4.35\%$ \\ \hline
-	Lang et al., 2014 & USA, RI & 2000 - 2013 & DID & d, v & insignificant  \\ \hline
-	Gibbons, 2015 & GBR, EAW & 2000 - 2011 & DID & d, v & $-6\%$ \\ \hline
-	Hoen et al., 2015 & USA & 1996 - 2011 & DID & d & insignificant \\ \hline
-	Dröes and Koster, 2016 & NED & 1985 - 2011 & DID & d & $-1.4\%$\\ \hline
-	Sunak and Madlener, 2016 & GER, NRW & 1992 - 2010 & DID & v & $-9\%$ to $-14\%$ \\ \hline
-	Jensen et al., 2018 & DNK & - & hedonic, SAR & d, \# of turbines & $-0.2\%$ to $-1.1\%$ \\ \hline
-	Kussel et al., 2019 & GER & 2007 - 2015 & DID & d & $-6.0\%$ \\ \hline \hline
-	\end{tabular}
-%\label{tab:negex}
-\end{scriptsize}
-\end{table}
-\end{frame}
-
-\begin{frame}
-\frametitle{Compensating for local negative externalities}
-\framesubtitle{An example}
-\begin{textblock*}{7cm}(0.75cm, 2.0cm) % {block width} (coords)
-OC of wind turbines in Austria
-\vspace{-0.41cm}
-\begin{table}
+%\appendix
+%\backupbegin
+%
+%
+%\begin{frame}
+%\frametitle{The cost of undisturbed landscapes}
+%\framesubtitle{Unconstrained deployment of wind and solar power}
+%
+%\begin{textblock*}{10cm}(0.75cm, 1.8cm) % {block width} (coords)
+%\begin{table}
+%\begin{tabular}{l r l}
+%\emph{Results for Austria} \\ \hline \hline
+%Added Solar PV & 0.0 & \SI{}{\tera\watt\hour} \\ \hline
+%Added Wind & 31.0 & \SI{}{\tera\watt\hour} \\ \hline \hline
+%System Cost\footnote[frame]{excluding cost for domestic transmission and distribution grids} & $3095.2$ & \SI{}{\mega\text{\euro}} \\ \hline
+%Trade balance & $746.8$ & \SI{}{\mega\text{\euro}} \\ \hline
+%Cost of air pollution & $184.4$ & \SI{}{\mega\text{\euro}} \\ \hline
+%\ce{CO2} emissions & $5227.3$ & \SI{}{\kilo\tonne} \\ \hline
+%Electricity generation & $92.8$ & \SI{}{\tera\watt\hour} \\ \hline
+%Curtailment & $2.4$ & \SI{}{\tera\watt\hour} \\ \hline
+%Net exports & $5.9$ & \SI{}{\tera\watt\hour} \\ \hline
+%Wholesale electricity price & $56.05$ & \SI{}{\text{\euro}\per\mega\watt\hour} \\ \hline
+%\end{tabular}
+%%\caption{Results for Austria}
+%\end{table}
+%\end{textblock*}
+%
+%\begin{textblock*}{4cm}(11cm, 2.05cm)
+%\begin{small}
+%\begin{table}
+%\begin{tabular}{p{1.8cm} | p{0.3cm} l}
+%Capital cost of solar PV & $630$ & \SI{}{\text{\euro}\per\kilo\watt\text{p}} \\
+%\ce{CO2} price & $90$ & \SI{}{\text{\euro}\per\mega\watt\hour}
+%\end{tabular}
+%\end{table}
+%\end{small}
+%\end{textblock*}
+%
+%\end{frame}
+%
+%
+%\begin{frame}
+%\frametitle{The value of undisturbed landscapes}
+%\framesubtitle{Local Negative Externalities of Wind Turbines}
+%%\begin{table}[h!]
+%%\caption{Estimates of Wind Turbine Impact on Property Prices}
+%\begin{table}
 %\centering
-\begin{footnotesize}
-	\begin{tabular}{c | c | c | c}
-	$oc_w$ & $\delta$ & NPV $oc_w$ & NPV $oc_w$ \\
-	$[\text{\euro}\, \text{MW}^{-1} \text{a}^{-1}]$ & & $[\text{\euro}\, \text{MW}^{-1}]$ & $[\text{\euro} \, \text{turbine}^{-1}]$ \\ \hline \hline
-	$40\,000$ & $0.05$ & $563\,758$ & $1\,973\,152$ \\ \hline
-	$40\,000$ & $0.03$ & $696\,526$ & $2\,437\,841$ \\ \hline
-	$10\,000$ & $0.05$ & $140\,939$ & $493\,288$ \\ \hline
-	$10\,000$ & $0.03$ & $171\,131$ & $609\,460$ \\ \hline
-	\end{tabular}
-\end{footnotesize}
-\end{table}
-\end{textblock*}
-
-\begin{textblock*}{7cm}(8.25cm, 2.0cm) % {block width} (coords)
-Property value loss in Germany
-%\vspace{-0.5cm}
-\begin{table}
+%\begin{scriptsize}
+%	\begin{tabular}{p{3.2cm} l l l p{1.1cm} p{2.4cm}}
+%	\hline
+%	Authors & Country & Sample & Framework\footnote[frame]{DID $\ldots$ difference-in-differences, SAR $\ldots$ spatial autoregressive, SFE $\ldots$ spatial fixed effects} & Impact Proxy\footnote[frame]{d $\ldots$ distance, v $\ldots$ visibility} & Estimated Impact \\
+%	\hline \hline
+%%	\cite{Sims2008} & Cornwall, UK & 2000 - 2007 & hedonic & visibility dummies & insignificant \\ \hline
+%	Heintzelmann and Tuttle, 2012 & USA, NY & 2000 - 2009 & hedonic, SFE & d & $-3.6\%$ to $-9.6\%$, insignificant \\ \hline
+%	Jensen et al., 2014 & DNK & 2000 - 2011 & hedonic, SAR & d, v & $-4.35\%$ \\ \hline
+%	Lang et al., 2014 & USA, RI & 2000 - 2013 & DID & d, v & insignificant  \\ \hline
+%	Gibbons, 2015 & GBR, EAW & 2000 - 2011 & DID & d, v & $-6\%$ \\ \hline
+%	Hoen et al., 2015 & USA & 1996 - 2011 & DID & d & insignificant \\ \hline
+%	Dröes and Koster, 2016 & NED & 1985 - 2011 & DID & d & $-1.4\%$\\ \hline
+%	Sunak and Madlener, 2016 & GER, NRW & 1992 - 2010 & DID & v & $-9\%$ to $-14\%$ \\ \hline
+%	Jensen et al., 2018 & DNK & - & hedonic, SAR & d, \# of turbines & $-0.2\%$ to $-1.1\%$ \\ \hline
+%	Kussel et al., 2019 & GER & 2007 - 2015 & DID & d & $-6.0\%$ \\ \hline \hline
+%	\end{tabular}
+%%\label{tab:negex}
+%\end{scriptsize}
+%\end{table}
+%\end{frame}
+%
+%
+%\begin{frame}
+%\frametitle{Compensating for local negative externalities}
+%\framesubtitle{An example}
+%\begin{textblock*}{7cm}(0.75cm, 2.0cm) % {block width} (coords)
+%OC of wind turbines in Austria
+%\vspace{-0.41cm}
+%\begin{table}
+%%\centering
+%\begin{footnotesize}
+%	\begin{tabular}{c | c | c | c}
+%	$oc_w$ & $\delta$ & NPV $oc_w$ & NPV $oc_w$ \\
+%	$[\text{\euro}\, \text{MW}^{-1} \text{a}^{-1}]$ & & $[\text{\euro}\, \text{MW}^{-1}]$ & $[\text{\euro} \, \text{turbine}^{-1}]$ \\ \hline \hline
+%	$40\,000$ & $0.05$ & $563\,758$ & $1\,973\,152$ \\ \hline
+%	$40\,000$ & $0.03$ & $696\,526$ & $2\,437\,841$ \\ \hline
+%	$10\,000$ & $0.05$ & $140\,939$ & $493\,288$ \\ \hline
+%	$10\,000$ & $0.03$ & $171\,131$ & $609\,460$ \\ \hline
+%	\end{tabular}
+%\end{footnotesize}
+%\end{table}
+%\end{textblock*}
+%
+%\begin{textblock*}{7cm}(8.25cm, 2.0cm) % {block width} (coords)
+%Property value loss in Germany
+%%\vspace{-0.5cm}
+%\begin{table}
+%%\centering
+%\begin{footnotesize}
+%	\begin{tabular}{c | c | c | c}
+%	  & value & impact & loss \\
+%	  & \euro & \% & \euro \\ \hline \hline
+%	mean & $273\,786$ & $0.06$ & $16\,427$ \\ \hline
+%	mean & $273\,786$ & $0.10$ & $27\,379$ \\ \hline
+%	mean + 1 s.d. & $476\,923$ & $0.06$ & $28\,615$ \\ \hline
+%	mean + 1 s.d. & $476\,923$ & $0.10$ & $47\,692$ \\ \hline
+%	\end{tabular}
+%\end{footnotesize}
+%\end{table}
+%\end{textblock*}
+%
+%\begin{textblock*}{14.5cm}(0.75cm, 6.0cm) % {block width} (coords)
 %\centering
-\begin{footnotesize}
-	\begin{tabular}{c | c | c | c}
-	  & value & impact & loss \\
-	  & \euro & \% & \euro \\ \hline \hline
-	mean & $273\,786$ & $0.06$ & $16\,427$ \\ \hline
-	mean & $273\,786$ & $0.10$ & $27\,379$ \\ \hline
-	mean + 1 s.d. & $476\,923$ & $0.06$ & $28\,615$ \\ \hline
-	mean + 1 s.d. & $476\,923$ & $0.10$ & $47\,692$ \\ \hline
-	\end{tabular}
-\end{footnotesize}
-\end{table}
-\end{textblock*}
-
-\begin{textblock*}{14.5cm}(0.75cm, 5.6cm) % {block width} (coords)
-\centering
-Compensation potential
-\vspace{-0.1cm}
-\begin{table}
-\begin{footnotesize}
-	\begin{tabular}{c | c | c | c | c}
-	$oc_w$  & $\delta$ & value & impact & \# property owners \\ \hline \hline
-	$40\,000$ & $0.05$ & $273\,786$ & $0.06$ & $120$ \\ \hline
-	$40\,000$ & $0.05$ & $273\,786$ & $0.10$ & $72$ \\ \hline
-	$40\,000$ & $0.03$ & $476\,923$ & $0.06$ & $69$ \\ \hline
-	$40\,000$ & $0.03$ & $476\,923$ & $0.10$ & $41$ \\ \hline
-%	$10\,000$ & $0.05$ & $476\,923$ & $0.06$ & $1$ \\ \hline
-	$10\,000$ & $0.03$ & $476\,923$ & $0.10$ & $18$ \\ \hline
-	\end{tabular}
-\end{footnotesize}
-\end{table}
-\end{textblock*}
-
-\end{frame}
-\backupend
+%Compensation potential
+%\vspace{-0.1cm}
+%\begin{table}
+%\begin{footnotesize}
+%	\begin{tabular}{c | c | c | c | c}
+%	$oc_w$  & $\delta$ & value & impact & \# property owners \\ \hline \hline
+%	$40\,000$ & $0.05$ & $273\,786$ & $0.06$ & $120$ \\ \hline
+%	$40\,000$ & $0.05$ & $273\,786$ & $0.10$ & $72$ \\ \hline
+%	$40\,000$ & $0.03$ & $476\,923$ & $0.06$ & $69$ \\ \hline
+%	$40\,000$ & $0.03$ & $476\,923$ & $0.10$ & $41$ \\ \hline
+%%	$10\,000$ & $0.05$ & $476\,923$ & $0.06$ & $1$ \\ \hline
+%	$10\,000$ & $0.03$ & $476\,923$ & $0.10$ & $18$ \\ \hline
+%	\end{tabular}
+%\end{footnotesize}
+%\end{table}
+%\end{textblock*}
+%
+%\end{frame}
+%\backupend
 \end{document}

### 2020-04-07 projects/asparagus/doc/plots_asparagus.py

@@ -56,7 +56,7 @@ def plot_subn(df, fname, width=2, xlim=None, ylim=None, xlabel=None, ylabel=None
 
         axis[a, b].grid()
         axis[a, b].set_title(df.columns[col])
-    plt.rcParams.update({'font.size': 16})
+    # plt.rcParams.update({'font.size': 12})
     plt.tight_layout()
     plt.savefig(fname)
     plt.close()
@@ -103,7 +103,7 @@ def plot_lines(df, fname, xlim=None, ylim=None, xlabel=None, ylabel=None, color=
     axis.set_ylim(ylim)
     axis.set_xlim(xlim)
     axis.grid()
-    plt.rcParams.update({'font.size': 16})
+    # plt.rcParams.update({'font.size': 16})
     plt.tight_layout()
     plt.savefig(fname)
     plt.close()
@@ -180,6 +180,8 @@ restrict_cost.index = restrict_cost.index.droplevel(0)
 # plot data
 plot_subn(restrict_cost / 1000, os.path.join(FPATH, 'cost.pdf'), xlim=[18, 0], color=REFUEL_COLORS,
           xlabel='Added Capacity of Wind [GW]', ylabel=['million €'])
+plot_subn(restrict_sysops / 1000, os.path.join(FPATH, 'sysops.pdf'), xlim=[18, 0], color=REFUEL_COLORS,
+          xlabel='Added Capacity of Wind [GW]', ylabel=['TWh', 'TWh', 'TWh', 'million t'])
 
 plot_lines(restrict_cost.loc[:, 'Cost of air pollution (SOx, NOx, PM)'] / 1000,
            os.path.join(FPATH, 'cost_airpollution.pdf'), xlim=[18, 0], color=[REFUEL_COLORS[2]],

### 2020-04-21 projects/asparagus/doc/slides.tex

--- 2020-02-21 projects/asparagus/doc/slides.tex
@@
-\documentclass[aspectratio=169, xcolor=dvipsnames]{beamer}
+\documentclass[aspectratio=1610, xcolor=dvipsnames,handout]{beamer} %aspectratio=169, 
 \usetheme{Pittsburgh}
@@
 
-\setbeamertemplate{section page}{
-	\begin{textblock*}{4cm}(11cm, 4cm) % {block width} (coords)
-	{\LARGE \insertsection}
-	\end{textblock*}
+%\setbeamertemplate{section page}{
+%	\begin{center}
+%	{\usebeamerfont[fg]{section title} \insertsection}
+%	\end{center}
+%}
+\AtBeginSection[]{
+  \begin{frame}
+  \vfill
+%  \centering
+%  \begin{beamercolorbox}[sep=8pt,center,shadow=false,rounded=true]{title}
+\tableofcontents[currentsection]
+%\begin{textblock*}{10cm}(7cm, 2.0cm) % {block width} (coords)
+%\adjincludegraphics[height=6cm,trim={0 0 {.3\width} 0},clip]{./figures/pv}
+%\end{textblock*}
+ %   \usebeamerfont{title} \usebeamercolor{title} \insertsection\par%
+%  \end{beamercolorbox}
+  \vfill
+  \end{frame}
 }
@@
 
-\section{Introduction}
+\section{Background}
 \begin{frame}
@@
 actual target is around 90\%.} of electricity demand from domestic renewable sources on annual balance by 2030 \pause
+\begin{itemize}
+\item[$\rightarrow$] necessarily turns Austria into a net exporter of electricity \pause
+\end{itemize}
 \item additional annual electricity generation of \SI{27}{\tera\watt\hour} is expected to suffice \pause
@@
 Hydro (run-of-river) & 5.72 & 28.34 & +5 & 33.34 & 6.73 \\ \hline
-%Biomass & 1 & 5 & +1 & 6 & 2 \\ \hline
+Biomass & $\sim 1$ & 4.78 & +1 & 5.78 & $>2$ \\ \hline
 \end{tabular}
 \end{table}
+\end{frame}
+
+\begin{frame}
+\frametitle{Why wind is not solar}
+\framesubtitle{A side note on imperfect substitutes}
+%\begin{textblock*}{4cm}(0.75cm, 2.75cm)
+%Aggregate profile for Austria
+\begin{textblock*}{4cm}(0.75cm, 2.75cm)
+\begin{itemize}
+\item secure system operation requires $S=D$ at any point in time 
+\item electricity not easily storable 
+\item >>transforming<< electricity feed-in to end-use electricity is costly
+\end{itemize}
+\end{textblock*}
+
+\only<2>{
+\begin{textblock*}{10cm}(5.5cm, 2.25cm) % {block width} (coords)
+\centering
+Aggregate generation profile in Austria
+\includegraphics[width=\textwidth, keepaspectratio]{./figures/wind_not_pv}
+\end{textblock*}
+}
 \end{frame}
@@
 \item Apart from wind and solar, potentials for renewable electricity generation in Austria largely exhausted
-\item Under announced policies, solar PV is only large-scale substitute to wind power
+\item Under announced policies, solar PV is only large-scale substitute to wind power \pause
 \item Increasing social conflict around the large-scale expansion of onshore wind power
@@
 
-\begin{frame}
-\frametitle{Why wind is not solar}
-\framesubtitle{A side note on imperfect substitutes}
-\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
-\centering
-Aggregate generation profile in Austria
-\includegraphics[width=\textwidth, keepaspectratio]{./figures/wind_not_pv}
-\end{textblock*}
-\begin{textblock*}{4cm}(11cm, 2.25cm)
-%Aggregate profile for Austria
-\bigskip
-\begin{itemize}
-\item secure system operation requires $S=D$ at any point in time
-\item electricity not easily storable
-\item grid operators >>transform<< electricity feed-in to end-use electricity
-\end{itemize}
-\end{textblock*}
-\end{frame}
-
+\section{Motivation \& Method}
 \begin{frame}
@@
 
-\section{Research agenda}
 \begin{frame}
@@
 \item Resemble Austrian power system in 2030
-\item include most important electricity trading partner Germany
+\item include most important electricity trading partner Germany \pause
 \item set policy target of meeting $90\%$ of demand in 2030 from domestic renewable sources
@@
 	\item expansion of renewable capacities in line with EEG 2017
-	\end{itemize}
+	\end{itemize} \pause
 \item simulate prospective electricity system with \emph{medea}
@@
 \end{footnotesize}
+\bigskip
+
 \medskip
@@
 
-\section{Discussion}
+\section{Discussion \& Conclusion}
 \begin{frame}

### 2020-04-21 src/tools/instantiate_gdx.py
changes in medea_main.gms and instantiate_gdx.py to allow for multifuel conversion processes

@@ -8,7 +8,7 @@ import config as cfg
 from src.tools.gams_io import df2gdx
 from src.tools.prepare_data import dict_sets, dict_instantiate, static_data, plant_data, ts_data, invest_limits
 
-# TODO: Add energy stored in hydro reservoirs - STORAGE_LEVEL
+# TODO: Add observed energy stored in hydro reservoirs - STORAGE_LEVEL
 
 idx = pd.IndexSlice
 # --------------------------------------------------------------------------- #
@@ -17,10 +17,32 @@ idx = pd.IndexSlice
 ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
 db = ws.add_database()
 
+"""
+parameters affected by multifuel-conversion
+- e_set / f_set
+- m_set (potentially)
+- AIR_POL_COST_FIX
+- AIR_POL_COST_VAR
+- 
+
+i .. 'Elyser', 'Methn'
+e .. 'Syngas', 'Hydrogen'
+k .. 'str_hydrogn', 'str_syng'
+
+introduce:
+ - new unit (set i) 'Elyser', which converts Electricity to 'Hydrogen'
+ - new unit (set i) 'Methn', which converts 'Hydrogen' to 'Syngas'
+ - new storage unit (set k) 'str_hydrogn' which stores 'Hydrogen'
+ - new storage unit (set k) 'str_syng' which stores 'Syngas'
+ - new energy carrier (set e) 'Hydrogen'
+ - new energy carrier (set e) 'Syngas'
+
+
+"""
 # --------------------------------------------------------------------------- #
 # %% instantiate SETS
 # --------------------------------------------------------------------------- #
-f_set = df2gdx(db, dict_sets['f'], 'f', 'set', [])
+e_set = df2gdx(db, dict_sets['e'], 'e', 'set', [])
 l_set = df2gdx(db, dict_sets['l'], 'l', 'set', [])
 m_set = df2gdx(db, dict_sets['m'], 'm', 'set', [])
 z_set = df2gdx(db, dict_sets['z'], 'z', 'set', [])
@@ -37,6 +59,14 @@ logging.info('medea sets instantiated')
 # --------------------------------------------------------------------------- #
[…]

### 2020-04-21 projects/asparagus/doc/slides.tex
updates to project asparagus slides.tex

--- 2020-04-07 projects/asparagus/doc/slides.tex
@@
         \end{itemize}
-        find significant capacities of renewable energy technologies (RET), largely wind power, need to be added.
+        find significant capacities of renewable energy technologies (RET), in Central Europe largely wind power,
+        need to be added.
         \medskip
 
-        Researchers are increasingly aware that social and political restrictions on such large-scale RET expansions must be
-        recognized.
+        Researchers are increasingly aware that social and political restrictions on such large-scale RET expansions
+        must be recognized.
         \begin{itemize}
             \item RET deployment in power system models often limited on the upside by results from analyses of land
-            availability, e.g. by excluding areas for technical or legal reasons, or because they are constested by stakeholders
+            availability, e.g. by excluding areas for technical or legal reasons, or because they are constested by
+            stakeholders
         \end{itemize}
@@
 
-        Yet, such `binary' analysis neglects the trade-offs faced by a social planner deciding on RET expansion.
+        Yet, this neglects the trade-offs faced by a social planner deciding on RET expansion.
     \end{frame}
@@
 
-        \begin{textblock*}{5cm}
+        \begin{textblock*}{5.75cm}
             (1.0cm,2cm)
-            Consider a renewable energy technology that
+            Consider a renewable energy technology $\omega$ that
             \begin{itemize}
-                \item saves system cost (relative to its best alternative), but
-                \item comes at social cost (e.g. due to local negative external effects)
+                \item is \emph{system cost} minimizing, i.e.\ it saves cost relative to its best alternative $\mu$, but
+                \item comes at \emph{social cost}, e.g. due to unaccounted negative external effects
             \end{itemize}
@@
 
-            A social planner should deploy this RET (instead of another one) up to the capacity at which
-            \begin{itemize}
-                \item system cost saving equals
-                \item added social cost
-            \end{itemize}
+            A social planner should substitute $\omega$ with $\mu$ up to the point where the increase in system cost
+            equals the reduction in harm (from $\omega$'s negative externality unaccounted in system cost)
         \end{textblock*}
 
-        \begin{textblock*}{8.0cm}
-            (6.5cm,2cm)
+        \begin{textblock*}{7.5cm}
+            (7.5cm,2cm)
             \begin{tikzpicture}
@@
                     ticks=none,
-                    xlabel={capacity substituted},
+                    xlabel={$\omega$ substituted with equivalent $\mu$},
                     ylabel={cost},
@@
                 ]
-                    \addplot[blue, thick] (x, 1.3*x^1.1+0.6);
-                    \node at (axis cs: 1.925,3) [blue, align=left] {saved\\ system\\ cost};
+                    \addplot[blue, thick] (x, 2*x^1.1-0.933);
+                    \node at (axis cs: 1.925,3) [blue, align=left] {additional\\ system\\ cost};
 
@@
 
-\begin{textblock*}{10cm}(0.75cm, 1.5cm) % {block width} (coords)
+    \begin{textblock*}{10cm}
+        (1.25cm, 1.75cm) % {block width} (coords)
+        \centering
+        \begin{tikzpicture}
+            \draw (4.5,0.0) node[] {\small Opportunity Cost of Wind Power};
+        \end{tikzpicture}
+    \end{textblock*}
+
+    \begin{textblock*}{10cm}
+        (0.75cm, 2.25cm) % {block width} (coords)
         \includegraphics[width=\textwidth, keepaspectratio]{./figures/undisturbed_base}
@@
 
-\begin{textblock*}{4.5cm}(10.9cm, 1.5cm)
+    \begin{textblock*}{4.5cm}
+        (10.9cm, 2.5cm)
         \begin{small}
@@
 
-\vspace{4.25cm}
+            \vspace{3.75cm}
             \quad wind power and solar PV \\ \quad are substitutes
@@
 
-\begin{textblock*}{9.5cm}(1.5cm, 8.2cm)
+    \begin{textblock*}{9.5cm}
+        (1.5cm, 8.7cm)
         \begin{tikzpicture}%[thick]
@@
     \framesubtitle{Sensitivity to capital cost of solar PV}
-\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
+
+    \begin{textblock*}{10cm}
+        (1.25cm, 1.75cm) % {block width} (coords)
+        \centering
+        \begin{tikzpicture}
+            \draw (4.5,0.0) node[] {\small Least Cost Technology Mix};
+        \end{tikzpicture}
+    \end{textblock*}
+
+    \begin{textblock*}{10cm}
+        (0.75cm, 2.25cm) % {block width} (coords)
+%\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
         \includegraphics[width=\textwidth, keepaspectratio]{./figures/sens_nrg}
@@
 
-\begin{textblock*}{4cm}(11cm, 2.25cm)
+    \begin{textblock*}{4cm}
+        (11cm, 2.5cm)
         \begin{small}
@@
     \framesubtitle{Sensitivity to capital cost of solar PV}
-\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
+    \begin{textblock*}{10cm}
+        (1.25cm, 1.75cm) % {block width} (coords)
+        \centering
+        \begin{tikzpicture}
+            \draw (4.5,0.0) node[] {\small Opportunity Cost of Wind Power};
+        \end{tikzpicture}
+    \end{textblock*}
+
+    \begin{textblock*}{10cm}
+        (0.75cm, 2.25cm) % {block width} (coords)
+%\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
         \includegraphics[width=\textwidth, keepaspectratio]{./figures/undisturbed_low}
@@
 
-\begin{textblock*}{4cm}(11cm, 2.25cm)
+    \begin{textblock*}{4cm}
+        (11cm, 2.25cm)
         \begin{small}

### 2020-04-21 projects/asparagus/doc/plots_asparagus.py

@@ -109,12 +109,66 @@ def plot_lines(df, fname, xlim=None, ylim=None, xlabel=None, ylabel=None, color=
     plt.close()
 
 
+def plot_sublines(df, fname, width=2, midx_level=0, subtitle=True, xlim=None, ylim=None, xlabel=None, ylabel=None,
+                  color=None):
+    """
+    plots multiindex pd.DataFrame to multiple subplots
+    :param df: Pandas MultiIndex DataFrame
+    :param fname: name of output file generated
+    :param width: number of subplots in a horizontal row
+    :param midx_level: MultiIndex level to group plots by
+    :param subtitle: boolean, if True generate title for each subplot; False: generate long legend entries instead
+    :param xlim:
+    :param ylim:
+    :param xlabel:
+    :param ylabel:
+    :param color:
+    :return: file with subplots
+    """
+    # determine length and width of subfigures
+    idx_to_group = list(df.columns.get_level_values(midx_level).unique())
+    num_subfigures = len(idx_to_group)
+    sub_length = np.ceil(num_subfigures / width).astype(int)
+
+    fig = plt.figure(figsize=(8, 5))
+    for subplot in range(0, num_subfigures):
+        data_to_plot = df.xs(idx_to_group[subplot], axis=1, level=midx_level)
+        _, num_lines = data_to_plot.shape
+        ax = fig.add_subplot(width, sub_length, subplot + 1)
+
+        if not color:
+            ax.plot(data_to_plot, linewidth=2)
+        elif len(color) == num_lines:
+            for c in range(0, num_lines):
+                ax.plot(data_to_plot.iloc[:, c], linewidth=2, color=color[c])
+        else:
+            raise TypeError(f'Colors misspecified. Should be {num_lines} colors')
+
+        if subtitle:
+            ax.set_title(idx_to_group[subplot])
+            ax.legend(data_to_plot.columns)
+        if not subtitle:
+            legend_entries = [", ".join(i) for i in itertools.product(list(data_to_plot.columns),
+                                                                      [idx_to_group[subplot]])]
[…]

### 2020-04-28 projects/asparagus/doc/paper_clean.tex
adds draft of undisturbed landscapes-paper

\documentclass[final, 3p, times]{elsarticle} %review, twocolumn, 5p, 12pt, authoryear, twocolumn, 5p,

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

% improved float environment
\usepackage{float}
% improved underlines (unterstreichungen)
\usepackage[normalem]{ulem}

% plotting
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.17}

% tables
\usepackage{booktabs}
\usepackage{tabulary}
\usepackage{threeparttable}
\usepackage{longtable}
%\usepackage{ctable}
\usepackage{makecell}

% SI units
\usepackage{siunitx}
\DeclareSIUnit{\thermal}{th}
\DeclareSIUnit{\electric}{el}
\DeclareSIUnit{\year}{a}

% Euro symbol
\usepackage[gen]{eurosym}

% citations
\usepackage{natbib}

% to-do notes
\setlength {\marginparwidth }{2cm}
\usepackage[draft]{todonotes}

% graphics
\usepackage{graphicx}

\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\COO}{\ensuremath{\mathrm{CO_2}} }

\usepackage[colorlinks=true, citecolor=blue, linkcolor=blue, filecolor=blue,urlcolor=blue]{hyperref}

\begin{document}
    \begin{frontmatter}
        \title{The Cost of Undisturbed Landscapes}
        \author[1]{Sebastian Wehrle\corref{cor1}}
        \ead{sebastian.wehrle@boku.ac.at}
        \author[1]{Johannes Schmidt}
        \ead{johannes.schmidt@boku.ac.at}
        \cortext[cor1]{Corresponding author}
        \address[1]{Institute for Sustainable Economic Development, University of Natural Resources and Life Sciences,
        Feistmantelstrasse 4, 1180 Vienna, Austria}

        \begin{abstract}
            Lorem ipsum et dolores
        \end{abstract}

        \begin{keyword}
            externalities \sep wind power \sep system design \sep policy
            \JEL L94 \sep Q41 \sep C61
        \end{keyword}
    \end{frontmatter}
    \newpage

% %%%%% %%%%% %%%%% DATA AND METHODS %%%%% %%%%% %%%%% %%%%%

    \section{Introduction} \label{sec:introduction}
    Future, low carbon electricity systems should largely rely on renewable energy technologies (RET), such as solar
    photovoltaics (PV) or wind turbines, according to a substantial body of literature~\citep{Rodriguez2015}.\todo{references}
    For central Europe, it is typically found that system-cost\footnote{System cost are the sum of fuel and emission
    cost,operation and maintenance cost, and investment cost} minimizing systems to rely on large shares of onshore
    wind power~\citep{Rodriguez2015}\todo{references}.
    Such undertakings frequently implement upper bounds on the deployment of RET, which are thought to reflect the
    available potential for RET erection~\citep{references}\todo{references}.
    These potentials, in turn, are usually derived from surveys of land available and suitable for the erection of RET\@.
    The typical methodology is to exclude areas that are considered not suited because of insufficient resource
    availability or because of technical or legal restrictions such as minimum distances to settlements.
    Areas not excluded are then considered as feasible for RET deployment \citep[e.g.][]{McKenna2015, Bosch2017, Ryberg2020}.
    \todo[inline]{refer to excluded areas as black, feasible areas as white}
    However, researchers are increasingly aware that such a 'black-or-white'-approach fails to account for more nuanced
    issues such as the social acceptance of interference with landscapes.
    Indeed, several contributions sought to improve on such shortcomings, for example by participatory
    modelling~\citep{Hoeltinger2016}, by xxx~\citep{Jaeger2016} or xxx~\citep{Harper2019}\todo[inline]{add mini review paragraph}

    In contrast, the work of~\cite{Drechsler2011} % Drechsler et al. 2011: Combining spatial modelling and choice experiments ...
    \todo[inline]{mention later paper by Drechsler} \cite{Drechsler2017}
    \todo[inline]{refer to negative externalities not included in system cost}

    We improve on this work by xxx. \todo[inline]{explain what we do also in contrast to Drechsler et al.}

    \subsection{Framework}
    Similar to~\cite{Drechsler2011}, we regard the problem as one of welfare optimization, in which we seek to minimize
    the social cost (including, but not limited to system cost) of low carbon energy systems.
    Our framework is best understood by considering a benevolent social planner who is to decide on the deployment of
    RET\@.
    Such a social planner faces a trade-off between
    (i) (potential) negative externalities of a RET at the local level (recall that externalities from emissions are
    already included in system cost), such as the visual impact on landscapes and
    (ii) (potential) positive impact of a RET through lowering system cost (compared to its best alternative).

    \begin{figure}[H]
        \centering
        \begin{tikzpicture}
            \begin{axis}[
                width=0.66 \textwidth,
                clip=false,
                axis lines=left,
                ticks=none,
                xlabel={$\alpha$},
                ylabel={cost},
                domain=0.5:1.75,
                xmin=0.5, xmax=2,
                ymin=0, ymax=4,
                samples=200
            ]
                \addplot[thick] (x, x^2-0.25);
                \addplot[thick] (x, x^2-4.25*x+5);
                \node at (axis cs: 1.8,2.85) {$s$};
                \node at (axis cs: 1.8,0.65) {$h$};
                \node at (axis cs: 0.45,-0.1) {$0$};
                %\addplot (1.235, x);
            \end{axis}
        \end{tikzpicture}
        \caption{Social optimum}
        \label{figure:social-optimum}
    \end{figure}

    For illustration, consider figure~\ref{figure:social-optimum}, which depicts the effect of deviating from a
    system-cost minimizing RET allocation by substituting some share of technology $A$ with its best alternative,
    technology $B$.\footnote{Here, we consider the interesting case in which in system-cost minimum a RET causes higher
    external cost than it saves in system cost.}
    We let $\alpha$ denote the share of technology $A$ substituted by the best available alternative technology $B$.
    At $\alpha=0$, we are in the system cost minimum, while the 'harm' $h$ from otherwise unaccounted external cost is
    high.
    Now suppose that we can lower external cost by substituting $A$ with $B$, i.e.\ harm $h$ is decreasing in $\alpha$.
    A benevolent social planner will substitute $A$ with $B$ as long as the resulting increase in system cost is lower
    than the corresponding decline in harm.
    \bigskip
    \todo[inline]{the following subsection must be overhauled}
    Within this framework, we quantify the opportunity cost of wind power ($s$ in Figure~\ref{figure:social-optimum})
    for the case of Austria in the work at hand.
    We chose Austria as our case study, as the government has comitted itself to meeting \SI{100}{\percent} of
    electricity consumption from renewable sources (on annual balance, excluding industry own-consumption and system
    services) by 2030.
    This policy makes Austria an ideal case for studying the substitution of RET while keeping total renewable
    electricity generation constant.
    Moreover, the bulk of additional renewable electricity must come either from wind power or solar photovoltaics, as
    the other renewables potentials are largely exploited already (see section~\ref{section:}), which simplifies our
    analysis further.
    In the context of Austria, as in many other European countries, least-cost renewable electricity systems are largely
    based on wind power~\citep{Rodriguez2015}. \todo[inline]{references}
    Harm is inflicted by local external effects of wind turbines, such as flickering, noise, negative impacts on
    wildlife, or (perceived) disturbance of landscapes.
    \cite{Mattmann2016} identifies the latter as a main trigger of opposition against wind power at the local level.
    For a review of wind power externalities, see~\cite{Zerrahn2017}.
    \todo[inline]{1) motivate Austrian case, 2) make clear that paper only looks at opportunity cost of wind power --
    cost of undisturbed landscapes, NOT at harm, which is assessed in various ways by many papers, though not for
    Austria}

% %%%%% %%%%% %%%%% DATA AND METHODS %%%%% %%%%% %%%%% %%%%%

    \section{Data and Methods} \label{sec:data-and-methods}
    To investigate the cost of undisturbed landscapes (i.e. the opportunity cost of wind power) we use the power system
    model \emph{medea}, summarized in section~\ref{subsec:medea}, which we instantiate with scenario assumptions
    (see section~\ref{subsec:data}) and observed data of the year 2016 (see section~\ref{subsec:scenarios}).

    For determining the opportunity cost of wind power, we proceed as follows:
    \begin{enumerate}
        \item We derive the unrestricted system cost-minimizing deployment of wind and solar power,
        \item we restrict deployment of wind power by a small margin (so that the next best RET substitutes for wind power)
        and observe net system cost $c_{net}$ for Austria (calculated as total system cost including air pollution cost
        net of trade balance).\label{enumerate:restrict}
        \item We repeat step~\ref{enumerate:restrict} till no wind power can be deployed.
        \item Finally, we approximate the opportunity cost of wind power (at given wind power capacity $w$) $OC_w$ by
        the change in $c_{net}$ in response to a change in wind power capacity $w$ deployed, i.e.\
        \[OC_w = \frac{\Delta c_{net}}{\Delta w}\]
        where $\Delta$ is the difference operator.
    \end{enumerate}

% ***** medea *****

    \subsection{Power system model \emph{medea}} \label{subsec:medea}
    We make use of the power system model \emph{medea} to simulate (dis)investment in and hourly operation of the
    prospective Austrian (and German) power system.

    The model is cast as a linear optimization seeking to minimize total system cost, which consist of fuel and emission
    cost, quasi-fixed and variable operation and maintenance (O\&M) cost, the costs of investment in energy generation,
    storage, and transmission assets, and (potential) cost of non-served load.

    The system is required to meet exogenous and inelastic demand for electricity and heat in any hour of the modeled
    year.
    Energy supply, in turn, is constrained by available installed capacities of energy conversion, storage, and
    transmission units.
    Co-generation units convert fuel to heat and power subject to a feasible operating region defined by the unit's
    electrical efficiency, the electricity loss per unit of heat production, and the back pressure coefficient.
    Electricity generation from intermittent sources (wind, water, solar) is subject to exogenous hourly generation
    profiles, which are scaled according to total installed capacities.
    Electricity from these sources can be curtailed at no additional cost (free disposal).

    \todo{Energy storage}
    Electricity can be stored in reservoir and pumped hydro storages or batteries.
    Generation from hydro storage plants is constrained by turbine capacity and energy contained in reservoirs of
    limited size.
    Hydro reservoirs are filled by inflows or by pumping (pumped storages only).
    To better capture operational differences, we model daily, weekly and seasonal reservoir and pumped storage plants
    separately.

    \todo{Electricity exchange}
    Electricity can be exchanged between Austria and Germany.

    \todo{Ancillary services}
    To ensure a stable and secure operation of the electricity system, power plants must provide ancillary services
    (e.g. frequency control, voltage support).
    We assume that this requires generators with an installed capacity of at least 21 GW2 to be operational (either
    generating or pumping in case of pumped hydro storages) at any point in time.

    For a detailed, mathematical description please refer to \ref{sec:medea}.

    Data processing is implemented in python, the optimization model is based on GAMS. Running the model on an Intel
    i7-8700 machine with 16 GB RAM, using CPLEX 12.9 as a solver, takes 10--15 minutes, depending on the model
    parameters.
    \todo[inline]{perfect foresight, perfect competition, reflects energy-only market}

% ***** Data *****

    \subsection{Data}\label{subsec:data}
    We set up our model with the goal to resemble Austria's prospective electricity and district heating systems in the
    year 2030.
    In addition, we model Austria's largest electricity trading partner Germany, to account for potential effects from
    electricity trade.
    As for Austria, our scenario reflects Germany's currently announced electricity sector policies, so that we set
    generation capacities to levels consistent with policies effective by 2030.
    These assumptions are laid out in sections~\ref{subsubsec:assumptions-austria} and \ref{subsubsec:assumptions-germany}
    and summarized in table~\ref{tab:instcaps}.

    \subsubsection{Scenario Assumptions for Austria}\label{subsubsec:assumptions-austria}
    Austria has set itself the goal of generating \SI{100}{\percent} of electricity consumption from renewable sources
    on annual balance by 2030.
    However, industry own consumption and system services, which currently account for \SI{10}{\percent} of annual
    electricity consumption, are exempt.
    Hence, in our analysis we set the policy goal to generating at least \SI{90}{\percent} of annual electricity
    consumption from renewable sources.\footnote{Please note that this policy goal does not imply that the generated
    electricity must actually be consumed.
    Hence, we consider curtailed electricity as contributing to the policy goal.}
    The government plans to achieve this goal by generating an additional \SI{27}{\tera\watt\hour} of electricity
    annually from renewable sources.
    Consequently, we have set prospective electricity consumption accordingly.

    New hydro power plants are thought to contribute \SI{5}{\tera\watt\hour} annually, while additional electricity
    generation from biomass is envisaged to account for \SI{1}{\tera\watt\hour} annually.
    We add corresponding generation capacities to reach these targets.

    The remainder is projected to come from solar PV (\SI{11}{\tera\watt\hour} annually) and onshore wind turbines
    (\SI{10}{\tera\watt\hour} annually)~\cite{Regierungsprogramm2020}.

    The latter two RET are the only low-carbon electricity generation technologies with significant expansion potential.
    Austrians ruled out the use of nuclear power in a 1978 referendum~\citep{Pelinka1983}.
    Biomass lacks ecological and economical sustainability~\citep{Erb2018}, while available hydro power potentials were
    already largely exploited~\citep{Walder2014} even before adding another \SI{5}{\tera\watt\hour} of annual generation.

    As we are interested in determining the opportunity cost of wind power versus its best alternative, we allow for
    endogenous investment in wind and solar power, without enforcing announced targets.

    \subsubsection{Scenario Assumptions for Germany}\label{subsubsec:assumptions-germany}
    Germany has announced specific capacity targets for several power generation technologies.
    Following these announcements, we anticipate an end to nuclear power generation, a (partial) coal exit according to
    recommendations by~\cite{WSB2019}, and a further expansion of renewable electricity generation in line
    with the German Renewable Energy Sources Act~\cite{noauthor_renewable_2017}.
    We do, however, expect a lifting of the \SI{52}{\giga\watt} limit on solar PV\@.

    \begin{table}[H]
        \caption{Initial Generation Capacities}
        \centering
        \begin{tabulary}{0.55\textwidth}{lcc}
            \toprule
            Technology & Austria & Germany \\
            & GW & GW \\
            \midrule
            Wind Onshore & $2.6$ & $90.8$ \\
            Wind Offshore & $0.0$ & $15.0$ \\
            Solar PV & $1.1$ & $73.0$ \\
            Run-of-river Hydro & $6.7$ & $4.5$ \\
            Biomass & $0.725$ & $8.4$ \\
            Lignite & $0.0$ & $11.4$ \\
            Coal & $0.0$ & $14.0$ \\
            Natural Gas & $3.9$ & $24.2$ \\
            Mineral Oil & $0.2$ & $3.5$ \\
            Heat Pump & $1.0$ & $1.0$ \\
            Gas Boiler & $3.9$ & $25.5$ \\ \bottomrule
        \end{tabulary}
        \label{tab:instcaps}
    \end{table}

    \subsection{Energy supply}
    We represent 27\todo{compare with table!} dispatchable energy conversion and storage technologies that are expected
    to be operational by 2030.
    In addition to exogenously set, initally installed capacities (see~\ref{tab:instcaps}), the model can endogenously
    add further generation capacities that are compatible with stated policy objectives.
    All technologies can also be decommissioned, so that we adopt a long-run perspective on the power system.

    Assumed technology parameters are summarized in Table~\ref{itm:techcost}.

    \begin{table*}[h]
        \centering
        \begin{threeparttable}
            \caption{Technology Input Parameters}\label{itm:techcost}
            %\begin{longtable}{p{2.9cm} p{1.5cm} p{1.5cm} p{1.6cm} p{1.5cm} p{1.0cm} p{1.2cm} p{1.0cm}}
            \begin{tabulary}{0.5\textwidth}{lccccccl}
                \toprule
                & \multicolumn{2}{c}{Capital Cost} & \multicolumn{2}{c}{O\&M-Cost} & & & \\ \cmidrule(lr){2-3} \cmidrule(lr){4-5}
                Technology & Power & Energy & Quasi- & Variable & Life- & Efficiency & Source    \\
                & & & fixed & & time & &           \\
                & \EUR/kW & \EUR/kWh & \EUR/MWa & \EUR/MWh & a & & \\
                \midrule
                %\multicolumn{8}{c}{\emph{Renewable Energy Technologies}}\\ \midrule
                Wind Onshore & $1040$    & NA & $12600$   & $1.35$     & $30$ & NaN & 1 \\
                Wind Offshore & $1930$    & NA & $36053$   & $2.70$     & $30$ & NaN & 1 \\
                %Solar PV       & $630$ & NaN & $8844$ & $0.00$ & $40$ & NaN & 9 \\
                PV Rooftop & $870$     & NA & $10815$   & $0.00$     & $40$ & NaN & 1 \\
                PV Open-space & $330$     & NA & $6380$    & $0.00$     & $40$ & NaN & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Thermal Energy Conversion}}\\ \midrule
                Lignite Adv & -- & NA & $40500$   & $0.85$     & $40$ & 0.439 & \cite{OekoInstitut2017} \\
                Coal SC & -- & NA & $25000$   & $6.00$     & $40$ & 0.425 & 1 \\
                Coal USC & -- & NA & $31000$   & $2.90$     & $40$ & 0.485 & 1 \\
                Nat Gas ST & $400$     & NA & $15000$   & $3.00$     & $30$ & 0.407 & 1 \\
                Nat Gas GT & $590$     & NA & $19500$   & $4.40$     & $25$ & 0.420 & 1 \\
                Nat Gas CC & $880$     & NA & $29300$   & $4.40$     & $25$ & 0.610 & 1 \\
                Oil ST & $400$     & NA & $6000$    & $3.00$     & $30$ & 0.396 & 1 \\
                Oil GT & $378$     & NA & $8068$    & $4.50$     & $25$ & 0.410 & 1 \\
                Oil CC & $800$     & NA & $25000$   & $4.00$     & $25$ & 0.470 & 1 \\
                Biomass & $3400$    & NA & $140000$  & $4.50$     & $25$ & 0.298 & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Energy Storage Technologies}}\\ \midrule
                Hydro Reservoir & -- & -- & $X$       & $X$       & $80$ & $0.9$  & X \\
                Hydro PSP & -- & -- & $X$       & $X$       & $80$ & $0.9^2$ & X\\
                Battery Li-Ion & $1930$    & $321$ & $36053$   & $2.70$    & $30$ & NaN & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Heat Generation Technologies}}\\ \midrule
                Resistive Heater & $X$      & $X$   & $X$       & $X$       & $25$  & NaN & X \\
                Heatpump & $500$     & NaN & $36053$   & $2.70$    & $30$ & NaN & 1 \\
                Natural gas boiler & $60$   & NA & $1950$    & $1.00$    & $20$ & 0.90 & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Electricity Transmission}}\\ \midrule
                Transmission & $1930$    & NaN & $36053$   & $2.70$    & $30$ & NaN & 1 \\
                \bottomrule
            \end{tabulary}

            \begin{tablenotes}
                \item SC -- supercritical, USC -- ultra-supercritical, ST -- steam turbine, GT -- gas turbine,
                CC -- combined cycle, PSP -- Pumped storage plant
                %\item [a] Hydroelectric facilities are not expanded in this model and are considered to be fully amortized.
                %\item [b] The costs for distribution infrastructure and building retrofitting are approximate (see text)
                % and they are therefore not optimised or included in the presented total system costs, but calculated
                % retrospectively and analysed in the text.
            \end{tablenotes}

        \end{threeparttable}
    \end{table*}

    Hourly generation from non-dispatchable technologies solar photovoltaics, wind turbines and run-of-river hydro is
    assumed to follow hourly profiles as observed in 2016.
    \todo[inline]{more details on derivation of load profiles for Austria, source for Germany, scaling}

    Similarly, hourly inflows into hydro reservoirs (including pumped storage reservoirs) are inferred from weekly
    reservoir levels in combination with hourly generation and pumping time series published by~\cite{ENTSO-E2020}.
    However, all storage technologies (including hydro reservoir and pumped storage plant) are charged and dispatched
    endogenously.
    \todo[inline]{improve description of methodology for inflow approximation}

    \subsection{Energy demand}\label{subsec:energy-demand}
    According to projections of the Austrian government, additional renewable electricity generation of
    \SI{27.0}{\tera\watt\hour} will suffice to meet policy goals by 2030 \citep{Regierungsprogramm2020}.
    However, official documents do neither provide any information about expected electricity consumption in 2030,
    nor about expected prices for \COO emissions.
    Absent these information, we chose to set 2030 electricity demand to a level that results in \SI{27.0}{\tera\watt\hour}
    additional electricity generation from renewable sources at \COO prices of 60 \EUR per tonne in a setting without
    restrictions on renewables deployment.
    The corresponding annual electricity consumption is \SI{86.0}{\tera\watt\hour}, \SI{24}{\tera\watt\hour} above
    electricity end use observed in our base year 2016.
    The difference in annual electricity generation stems from

    Please note that the demand level consistent with generating \SI{27.0}{\tera\watt\hour} additional renewable
    electricity depends on the price of \COO.
    To keep scenarios comparable, however, we keep annual electricity consumption constant at \SI{86.0}{\tera\watt\hour}
    throughout all scenarios.
    In effect, added RET capacities fluctuate with the \COO price.

    For Germany, we assume electricity demand to remain at levels observed in 2016.
    In consequence, we scale reported hourly load to match annual electricity consumption published in national energy
    balances \citep{AGEB}.
    \todo[inline]{cross-check code}

    Annual heat consumption is derived from energy balances for Austria and Germany, respectively \citep{AGEB, OeSTAT2019}.
    Subsequently, annual heat consumption is broken down to hourly heat consumption on the basis of standard natural gas
    load profiles for space heating in the residential and commercial sectors \citep{Almbauer2008}.
    These load profiles make use of daily average temperatures to calculate daily heat demand.
    We extract spatially resolved temperature data from ERA-5 climate data sets \citep{era-5} and compute a capacity-weighted
    average of temperatures at locations of combined heat and power (CHP) generation units.
    Daily heat demand based on these weighted average temperatures is then broken down to hourly consumption on the
    basis of standardized factors accounting for weekday and time-of-day effects.
    Descriptive statistics of electricity and heat consumption are provided in table \ref{tab:time-series}.

    \begin{table*}[ht]
        \centering
        \caption{Descriptive data of time series used}\label{tab:time-series}
        \begin{tabular}{l c c c c c c l}
            \toprule
            Name & Area & Unit & mean & median & max & min & Source    \\
            \midrule
            Electricity load & AT & GW & 7.15 & 7.14 & 10.5 & 4.21 & nose      \\
            Electricity load & DE & GW & 60.2 & 59.6 & 82.8 & 34.6 & nose      \\
            District heating load & AT & GW & 2.56 & 2.43 & 5.77 & 0.91 & bladder   \\
            District heating load & DE & GW & 14.6 & 14.1 & 26.0 & 8.9 & bladder   \\
            Wind onshore profile & AT & \%    & 0.229 & 0.142 & 0.929 & 0.000 &           \\
            Wind onshore profile & DE & \%    & 0.176 & 0.133 & 0.764 & 0.003 &           \\
            Wind offshore profile & DE & \%    & 0.371 & 0.327 & 0.900 & 0.000 &           \\
            Solar PV profile & AT & \%    & 0.098 & 0.012 & 0.570 & 0.000 &           \\
            Solar PV profile & DE & \%    & 0.099 & 0.003 & 0.661 & 0.000 &           \\
            Run-of-river profile & AT & \%    & 0.564 & 0.538 & 0.971 & 0.202 &           \\
            Run-of-river profile & DE & \%    & 0.437 & 0.420 & 0.637 & 0.208 &           \\
            \bottomrule
        \end{tabular}
    \end{table*}

    Electricity transmission between the modelled market areas is constrained by net transfer capacities reported by
    ENTSO-E for the year 20XX.
    In addition, the model can expand transmission capacities at a cost of XXX EUR per MW and km.

    Monthly prices for exchange-traded fuels (hard coal, crude oil (Brent), natural gas) are retrieved from the
    International Monetary Fund's Commodity Data Portal.
    We convert these prices to \EUR/MWh based on the fuel's energy content and market exchange rates obtained from the
    European Central Bank \citep{ECB}.
    Finally, we resample prices to hourly frequency using piecewis cubic hermite interpolation.
    The price of lignite is assumed at $1.50 EUR / MWh_{th}$, as reported by \cite{OekoInstitut2017}.
    Biomass is assumed to be subsidized such that the cost of fuel is reduced to 6.5 $EUR / MWh$, which results in
    continuous operation of biomass plant in our scenarios.\footnote{Subsidies are not included in system cost, as the
    required level is uncertain. The total sum of subsidies is, however, constant across scenarios as biomass capacity
    can not be expanded.}

    \begin{table*}[h!]
        \centering
        \caption{Descriptive data of price time series}
        \begin{tabular}{l c r r r r l}
            \toprule
            Name & Unit & mean & median & max & min & Source                  \\
            \midrule
            Lignite & \EUR/MWh & 1.50 & 1.50 1.50 & 1.50 & \cite{OekoInstitut2017} \\
            Coal & \EUR/MWh & 8.58 & 8.13 & 11.90 & 6.69 & \cite{IMF}              \\
            Natural gas & \EUR/MWh & 13.62 & 12.87 & 20.18 & 12.04 & \cite{IMF}              \\
            Mineral oil & \EUR/MWh & 26.40 & 26.63 & 33.44 & 18.34 & \cite{IMF}              \\
            Biomass & \EUR/MWh & 6.50 & 6.50 & 6.50 & 6.50 & own assumption          \\
            \bottomrule
        \end{tabular}
    \end{table*}
    All price as in our baseline year 2016.
    Transport costs apply to different regions.

    The carbon intensity of fossil fuels is displayed in Table~\ref{table:carbon-intensity}, and is based on an
    analysis by the German environmental agency~\citep{Juhrich2016}.

    \begin{table}[h!]
        \centering
        \caption{\COO intensity of fossil fuels in tonnes \COO per MWh}
        \label{table:carbon-intensity}
        \begin{tabular}{c c c l}
            \toprule
            Lignite & Coal & Natural Gas & Mineral Oil   \\
            \midrule
            0.399 & 0.337 & 0.201 & 0.266         \\
            \bottomrule
        \end{tabular}
    \end{table}

    All data retrieval and processing scripts are available at medea's github repository
    \url{https://github.com/inwe-boku/medea} under an open MIT license.

    \section{Results and Discussion} \label{sec:results-discussion}
    To derive the social opportunity cost of not using wind power (but solar photovoltaics instead), we start by
    determining the long-run power market equilibrium without constraints on renewable capacity addition.
    This gives us our baseline scenario with minimal system costs.
    Starting from this baseline scenario, we evaluate further scenarios in which we gradually tighten the upper limit
    on wind power expansion, for example due to social conflict around wind turbine impacts on landscapes.
    We continue to tighten upper limits on wind power expansion till no wind power is allowed anymore.
    For each step, we observe system cost, total \COO emissions and other variables of interest.
    By relating the change in generated wind energy to the change in system cost (or other variables of interest), we
    derive a measure of the (approximate) marginal impact of wind power expansion (or contraction).

    \subsection{Baseline} \label{subsec:baseline}
    As a point of reference for our analysis, we determine the least cost power system configuration when addition of
    wind turbines and solar PV is unconstrained.

    In our baseline scenario, we assume that about two thirds of solar PV installations are mounted on roof tops, while
    the remainder is realized as open-space solar PV at utility scale.
    This gives us capacity-weighted average capital cost for solar PV of $630$ \EUR per kW installed.
    In section~\ref{subsec:sensitivity} we analyze the sensitivity of our results with respect to this assumption.
    Moreover, we assume pricing of \COO emissions, i.e.\ all otherwise external cost of \COO emissions are
    internalized through \COO pricing.
    As future \COO prices are uncertain but impact renewables deployment ~\citep{Brown2020, Kirchner2019}, we conduct
    our analysis for \COO prices of 30, 60, 90, and 120 \EUR per tonne.

    Given our baseline assumptions, the cost-minimizing system set-up that is consistent with the stated policy
    goal\footnote{Recall that the policy target is to generate at least 90\% of electricity demand from renewable sources, on annual balance}
    adds \SI{16.5}{\giga\watt} wind power and \SI{0}{\giga\watt} solar PV to initially set capacities in Austria.
    Overall, electricity generation from both sources totals \SI{38.5}{\tera\watt\hour} and
    \SI{0.94}{\tera\watt\hour} annually, respectively.
    For system services and heat generation, \SI{6.8}{\giga\watt} of fossil thermal co-generation capacities remain
    active, giving rise to \SI{8.9}{} million tons of \COO emissions per year.
    As Austria has ample hydro storage capacity in place, no further storage (e.g. from batteries) is added.
    However, the addition of \SI{1.5}{\giga\watt} of compression heat pumps allows to use electricity for heat
    generation and to limit curtailment to \SI{2.4}{\tera\watt\hour\per\year}.
    System cost in Austria amount to $3.4$ billon \EUR in this scenario.\todo{add co2 emissions}

    \subsection{The opportunity cost of wind power in Austria}\label{subsec:opportunity-cost-wind}
    With a gradual restriction of wind power potentials, the system must increasingly rely on solar PV to generate the
    amount of electricity from renewable sources that is required to meet policy goals.
    Consequently, each \SI{}{\giga\watt} of wind power not installed needs to be replaced by \SI{2.35}{\giga\watt} of
    solar PV\@.
    Due to the different generation profiles of these technologies, this induces changes in system operation and least
    cost system configuration.

    \todo[inline]{describe changes in system operation}
    Substitution of wind power with solar PV leads to a reduction of electricity exports to Germany.
    Due to the favorable complementarity of Austrian and German wind, Austria adds wind energy generation above the level
    required for its own consumption.
    As the admissible addition of wind power is reduced, electricity exports to Germany are being scaled back.
    This is reinforced by the similarities in generation profiles of solar PV across both countries.

    Due to the concentration of solar PV generation in time (both on a daily and on a seasonal scale), installing solar
    PV instead of wind power leads to an increase in dispatchable thermal generation.
    As thermal generation in Austria is predominantly by fossil thermal, this induces an increase in \COO emissions due
    to the increased use of PV.

    \begin{figure*}
        \centering
        \includegraphics[width=0.8 \textwidth]{./figures/sysops.pdf}
        \caption{Opportunity cost of wind power assuming PV overnight cost of 630 EUR/kWp}
        \label{figure:system-operation-base}
    \end{figure*}

    \todo[inline]{describe changes in system configuration}

    \todo[inline]{describe marginal change in system cost due to less wind deployed}

    \todo[inline]{insert table with key scenario results}

    \begin{figure*}
        \centering
        \includegraphics[width=0.8 \textwidth]{./figures/undisturbed_base.pdf}
        \caption{Opportunity cost of wind power assuming PV overnight cost of 630 EUR/kWp}
        \label{figure:opportunity-cost-base}
    \end{figure*}

    \subsection{Sensitivity Analysis} \label{subsec:sensitivity}
    A major factor of influence for our results are the assumed capital cost of wind turbines and solar PV.
    Thus, we complement our baseline scenario with a sensitivity analysis on investment cost.
    For this purpose, we vary the investment cost of solar PV between our baseline assumption of 630 EUR/kW and 275
    EUR/kW that \cite{Vartiainen2019} project for utility-scale solar PV in the year 2030.

    \begin{figure*}
        \centering
        \includegraphics[width=0.8 \textwidth]{./figures/undisturbed_low.pdf}
        \caption{Opportunity cost of wind power}
    \end{figure*}

%    Further uncertainty surrounds
%    •	Storage prices
%    •	Weather / full-load hours
%    •	LOOK AT WHAT HAPPENS WHEN GERMANY BUILDS MORE PV THAN ANTICIPATED (particularly in the low pv-cost scenario)

    \subsection{Limitations of the Analysis}
    To little surprise, the optimal deployment of renewable energy technologies heavily depends on underlying capital
    cost.
    Alas, uncertainty regarding future capital cost is high.
    \todo[inline]{WRITE SOMETHING ON THE IMPACT OF DIFFERENT COST ASSUMPTIONS ON TOTAL SYSTEM COST.}

    Nevertheless, several findings are remarkably robust to changing capital cost.
    A higher penetration level of PV leads to increasing emissions of CO2 in the least-cost systems we investigated.
    Due to the strong seasonal differences in electricity generation from solar PV, systems relying heavily on this technology either need to deploy excess PV capacity to boost generation in winter (resulting in heavy curtailment in summer) or, alternatively, add additional capacity to make up for the winter shortfall in electricity generation from solar PV.
    This additional capacity could be storage capacity to transfer summer excesses to winter.
    However, storages would have to balance seasonal differences in this case, implying the need for very high storage capacities, which are costly to realize.
    Typically, it will be cheapest to add (at least some) fossil-fueled backup plants, even at \COO prices of EUR 70 per ton.
    \todo[inline]{HERE, WE COULD INVESTIGATE AT WHICH CO2-PRICE FOSSIL BACKUP VANISHES.}

    Moreover, the observed increase in producer surplus in line with higher CO2 penetration is robust to capital cost uncertainty.
    Higher levels of PV penetration require more efforts for system integration.
    \todo[inline]{THIS CAN'T BE THE FULL EXPLANATION AS PROD SURPLUS DECREASES AT VERY HIGH LEVELS OF PV}

    \section{Conclusions and Policy Implication} \label{sec:conclusions-policy-implication}
    We have analyzed the social willingness-to-pay for wind power (versus solar PV).

    By combining estimates of the social willingness-to-pay for wind power with (spatially resolved) estimates of the negative externalities of wind turbines, we can determine the socially optimal deployment of wind turbines in Austria.
    This opens up several possibilities for evidence-based policy making.

    First, these estimates can inform compensation schemes for residents accepting wind turbines nearby.
    Compensation might be an economically efficient measure to increase wind turbines acceptance.

    Second, energy policies could be evaluated based on their impact on the opportunity cost of wind turbines.
    This would allow to include otherwise unaccounted negative externalities of wind turbines in decision making, for example through a direct measure of opportunity cost, or an indirect measure of avoided wind turbines in social optimum.

    Finally, our analysis also sheds some light on the distributional consequences of energy system design.

    \section{Data Availability}\label{sec:data-availability}
    Data and model is avaliable on github \url{}. Raw data is sourced from danish Energy Agency \url{}, ENTSO-E's
    transparency database \url{},

    \newpage
    \bibliography{asparagus}
    \bibliographystyle{elsarticle-harv}

    \newpage
    \appendix

    \section{Description of the power system model \emph{medea}} \label{sec:medea}

    \subsection{Sets} \label{sets}
    Sets are denoted by upper-case latin letters, while set elements are denoted by lower-case latin letters.

    \begin{table}
        \caption{Sets}
        \begin{tabulary}{\textwidth}{LLLp{6cm}}
            \toprule
            mathematical symbol & programming symbol & description & elements                                                      \\
            \midrule
            $f \in F$             & \texttt{f}         & fuels & \texttt{nuclear, lignite, coal, gas, oil, biomass, power}     \\
            $i \in I$             & \texttt{i}         & power generation technologies & \texttt{nuc, lig\_stm, lig\_stm\_chp, lig\_boa, lig\_boa\_chp,
            coal\_sub, coal\_sub\_chp, coal\_sc, coal\_sc\_chp, coal\_usc,
            coal\_usc\_chp, coal\_igcc, ng\_stm, ng\_stm\_chp, ng\_ctb\_lo,
            ng\_ctb\_lo\_chp, ng\_ctb\_hi, ng\_ctb\_hi\_chp, ng\_cc\_lo,
            ng\_cc\_lo\_chp, ng\_cc\_hi, ng\_cc\_hi\_chp, ng\_mtr,
            ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm, oil\_stm\_chp,
            oil\_ctb, oil\_ctb\_chp, oil\_cc, oil\_cc\_chp, bio,
            bio\_chp, heatpump\_pth}                                        \\
            $h \in H \subset I$     & \texttt{h(i)}    & power to heat technologies & \texttt{heatpump\_pth}                                        \\
            $j \in J \subset I$     & \texttt{j(i)}    & CHP technologies & \texttt{lig\_stm\_chp, lig\_boa\_chp, coal\_sub\_chp,
            coal\_sc\_chp, coal\_usc\_chp, ng\_stm\_chp, ng\_ctb\_lo\_chp,
            ng\_ctb\_hi\_chp, ng\_cc\_lo\_chp, ng\_cc\_hi\_chp,
            ng\_mtr\_chp, ng\_boiler\_chp,
            oil\_stm\_chp, oil\_ctb\_chp, oil\_cc\_chp, bio\_chp}           \\
            $k \in K$               & \texttt{k}       & storage technologies & \texttt{res\_day, res\_week, res\_season, psp\_day,
            psp\_week, psp\_season, battery}                                \\
            $l \in L$               & \texttt{l}       & feasible operation region limits & \texttt{l1, l2, l3, l4}                                       \\
            $m \in M$               & \texttt{m}       & energy products & \texttt{el, ht}                                               \\
            $n \in N$               & \texttt{n}       & intermittent generators & \texttt{wind\_on, wind\_off, pv, ror}                         \\
            $t \in T$               & \texttt{t}       & time periods (hours)               & \texttt{t1, t2, \ldots, t8760}                                \\
            $z \in Z$               & \texttt{z}       & market zones & \texttt{AT, DE}                                               \\
            \bottomrule
        \end{tabulary}
    \end{table}

    \newpage

    \subsection{Parameters} \label{parameters}
    Parameters are denoted either by lower-case greek letters or by upper-case latin letters.
    \begin{table}
        \caption{Parameters}
        \begin{tabulary}{\textwidth}{LLLL}
            %\begin{longtable}{p{4.1cm} c l c}
            \toprule
            mathematical symbol & programming symbol & description & unit                      \\
            %\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\
            % LOWER-CASE GREEK LETTERS
            \midrule
            $\delta_{z,zz}$               & \texttt{DISTANCE(z,zz)}                           & distance between countries' center of gravity & km                        \\
            $\varepsilon_{f}$             & \texttt{CO2\_INTENSITY(f)}                        & fuel emission intensity & $\text{t}_{\COO}$ / MWh   \\
            $\eta_{i,m,f}$                & \texttt{EFFICIENCY\_G(i,m,f)}                     & power plant efficiency & MWh / MWh                 \\
            $\eta^{out}_{z,k}$            & \makecell[l]{\texttt{EFFICIENCY\_S\_OUT(k)}}      & discharging efficiency &                           \\
            $\eta^{in}_{z,k}$             & \makecell[l]{\texttt{EFFICIENCY\_S\_IN(k)}}       & charging efficiency &                           \\
            $\lambda_{z}$                 & \texttt{LAMBDA(z)}                                & scaling factor for peak load &                           \\
            $\mu_{z}$                     & \texttt{VALUE\_NSE(z)}                            & value of lost load & \EUR / MWh                \\
            $\rho_{z,t,k}$                & \makecell[l]{\texttt{INFLOWS(z,t,k)}}             & inflows to storage reservoirs & MW                        \\
            $\sigma_{z}$                  & \texttt{SIGMA(z)}                                 & scaling factor for peak intermittent generation &                           \\
            $\phi_{z,t,n}$                & \texttt{GEN\_PROFILE(z,t,n)}                      & intermittent generation profile & $[0,1]$                   \\
            $\widehat{\phi}_{z,n}$        & \texttt{PEAK\_PROFILE(z,n)}                       & peak intermittent generation profile & $[0,1]$                   \\
            $\chi_{i,l,f}$                & \texttt{FEASIBLE\_INPUT(i,l,f)}                   & inputs of feasible operating region & $[0,1]$                   \\
            $\psi_{i,l,m}$                & \texttt{FEASIBLE\_OUTPUT(i,l,m)}                  & output tuples of feasible operating region & $[0,1]$                   \\
            % UPPER-CASE LATIN LETTERS
            $C^{r}_{z,n}$                 & \texttt{CAPITALCOST\_R(z,n)}                      & capital cost of intermittent generators (specific, annuity)     & \EUR / MW                 \\
            $C^{g}_{z,i}$                 & \texttt{CAPITALCOST\_G(z,i)}                      & capital cost of thermal generators (specific, annuity)          & \EUR / MW                 \\
            $C^{s}_{z,k}$                 & \makecell[l]{\texttt{CAPITALCOST\_S(z,k)}}        & capital cost of storages - power (specific, annuity)            & \EUR / MW                 \\
            $C^{v}_{z,k}$                 & \makecell[l]{\texttt{CAPITALCOST\_V(z,k)}}        & capital cost of storages - energy (specific, annuity)           & \EUR / MW                 \\
            $C^{x}$                       & \texttt{CAPITALCOST\_X}                           & capital cost of transmission capacity (specific, annuity)       & \EUR / MW                 \\
            $D_{z,t,m}$                   & \texttt{DEMAND(z,t,m)}                            & energy demand & GW                        \\
            $\widehat{D}_{z,m}$           & \texttt{PEAK\_LOAD(z,m)}                          & peak demand & GW                        \\
            $\widetilde{G}_{z,i}$         & \texttt{INITIAL\_CAP\_G(z,tec)}                   & initial capacity of dispatchable generators & GW                        \\
            $O^{g}_{i}$                   & \texttt{OM\_COST\_G\_VAR(i)}                      & variable O\&M cost of dispatchable generators & \EUR / MWh                \\
            $O^{r}_{z,n}$                 & \texttt{OM\_COST\_R\_VAR(z,n)}                    & variable O\&M cost of intermittent generators & \EUR / MWh                \\
            $P^{e}_{t,z}$                 & \texttt{PRICE\_CO2(t,z)}                          & \COO price & \EUR / $\text{t}_{\COO}$  \\
            $P_{t,z,f}$                   & \texttt{PRICE\_FUEL(t,z,f)}                       & fuel price & \EUR / MWh                \\
            $Q^{g}_{i}$                   & \texttt{OM\_COST\_G\_QFIX(i)}                     & quasi-fixed O\&M cost of dispatchable generators & \EUR / MW                 \\
            $Q^{r}_{z,n}$                 & \texttt{OM\_COST\_R\_QFIX(z,n)}                   & quasi-fixed O\&M cost of intermittent generators & \EUR / MW                 \\
            $\widetilde{R}_{z,n}$         & \texttt{INITIAL\_CAP\_R(z,n)}                     & initial capacity of intermittent generators & GW                        \\
            $\widetilde{S}^{out}_{z,k}$   & \makecell[l]{\texttt{INITIAL\_CAP\_S\_OUT(z,k)}}  & initial discharging capacity of storages & GW                        \\
            $\widetilde{S}^{in}_{z,k}$    & \makecell[l]{\texttt{INITIAL\_CAP\_S\_IN(z,k)}}   & initial charging capacity of storages & GW                        \\
            $\widetilde{V}_{z,k}$         & \makecell[l]{\texttt{INITIAL\_CAP\_V(z,k)}}       & initial energy storage capacity &                           \\
            $\widetilde{X}_{z,zz}$        & \texttt{INITIAL\_CAP\_X(z,zz)}                    & initial transmission capacity & GW                        \\
            \bottomrule
        \end{tabulary}
    \end{table}

    \newpage

    \subsection{Variables} \label{variables}
    Variables are denoted by lower-case latin letters.

    \begin{table*}
        \centering
        \caption{Variables}
        \begin{tabulary}{\textwidth}{LLLL}
            %\caption{Variables}\label{variables} \\
            \toprule
            mathematical symbol & programming symbol & description & unit      \\
            \midrule
            $b_{z,t,i,f}$                 & \texttt{b(z,t,i,f)}           & fuel burn for energy generation & GW        \\
            $c$                           & \texttt{cost\_system}         & total system cost & k\EUR     \\
            $c_{z}$                       & \texttt{cost\_zonal(z)}       & zonal system cost & k\EUR     \\
            $c^{b}_{z,t,i}$               & \texttt{cost\_fuel(z,t,i)}    & fuel cost & k\EUR     \\
            $c^{e}_{z,t,i}$               & \texttt{cost\_co2(z,t,i)}     & emission cost & k\EUR     \\
            $c^{om}_{z,i}$                & \texttt{cost\_om\_g(z,i)}     & total o\&m cost of dispatchable generators & k\EUR     \\
            $c^{om}_{z,n}$                & \texttt{cost\_om\_r(z,n)}     & total o\&m cost of intermittent generators & k\EUR     \\
            $c^{g}_{z}$                   & \texttt{cost\_invest\_g(z)}   & capital cost of generators & k\EUR     \\
            $c^{q}_{z}$                   & \texttt{cost\_nse(z)}         & total cost of non-served load & k\EUR     \\
            $c^{r}_{z}$                   & \texttt{cost\_invest\_r(z)}   & capital cost of intermittent generators & k\EUR     \\
            $c^{s,v}_{z}$                 & \texttt{cost\_invest\_sv(z)}  & capital cost of storages & k\EUR     \\
            $c^{x}_{z}$                   & \texttt{cost\_invest\_x(z)}   & capital cost of interconnectors & k\EUR     \\
            $e_{z}$                       & \texttt{emission\_co2(z)}     & \COO emissions & t \COO    \\
            $\widetilde{g}^{+}_{z,i}$     & \texttt{add\_g(z,i)}          & added capacity of dispatchables & GW        \\
            $\widetilde{g}^{-}_{z,i}$     & \texttt{deco\_g(z,i)}         & decommissioned capacity of dispatchables & GW        \\
            $g_{z,t,i,m,f}$               & \texttt{g(z,t,i,m,f)}         & energy generated by conventionals & GW        \\
            $q^{+}_{z,t}$                 & \texttt{q\_curtail(z,t)}      & curtailed energy & GW        \\
            $q^{-}_{z,t,m}$               & \texttt{q\_nse(z,t,m)}        & non-served energy & GW        \\
            $\widetilde{r}^{+}_{z,n}$     & \texttt{add\_r(z,n)}          & added capacity of intermittents & GW        \\
            $\widetilde{r}^{-}_{z,n}$     & \texttt{deco\_r(z,n)}         & decommissioned capacity of intermittents & GW        \\
            $r_{z,t,n}$                   & \texttt{r(z,t,n)}             & electricity generated by intermittents & GW        \\
            $\widetilde{s}^{+}_{z,k}$     & \texttt{add\_s(z,k)}          & added storage capacity (power)              & GW        \\
            $s^{in}_{z,t,k}$              & \texttt{s\_in(z,t,k)}         & energy stored in & GW        \\
            $s^{out}_{z,t,k}$             & \texttt{s\_out(z,t,k)}        & energy stored out & GW        \\
            $\widetilde{v}^{+}_{z,k}$     & \texttt{add\_v(z,k)}          & added storage capacity (energy)             & GWh       \\
            $v_{z,t,k}$                   & \texttt{v(z,t,k)}             & storage energy content & GWh       \\
            $w_{z,t,i,l,f}$               & \texttt{w(z,t,i,l,f)}         & operating region weight &           \\
            $\widetilde{x}^{+}_{z,zz}$    & \texttt{add\_x(z,zz)}         & added transmission capacity & GW        \\
            $x_{z,zz,t}$                  & \texttt{x(z,zz,t)}            & electricity net export & GW        \\
            \bottomrule
        \end{tabulary}
    \end{table*}

    \newpage

    \subsection{Naming system}
    \begin{table*}
        \centering
        \begin{threeparttable}
            \caption{Naming System}
            \begin{tabulary}{\textwidth}{LLLLLL}
                \toprule
                & initial capacity\tnote{$\dagger$}    & added capacity\tnote{$\ddagger$}  & decommissioned capacity\tnote{$\ddagger$} & specific investment cost\tnote{$\dagger$} & dispatch\tnote{$\ddagger$}    \\
                \midrule
                thermal units & $\widetilde{G}_{z,i}$                 & $\widetilde{g}^{+}_{z,i}$         & $\widetilde{g}^{-}_{z,i}$                 & $C^{g}_{z,i}$                             & $g_{z,t,i,m,f}$               \\
                intermittent units & $\widetilde{R}_{z,n}$                 & $\widetilde{r}^{+}_{z,n}$         & $\widetilde{r}^{-}_{z,n}$                 & $C^{r}_{z,n}$                             & $r_{z,t,n}$                   \\
                storages (power)    & $\widetilde{S}_{z,k}$                 & $\widetilde{s}^{+}_{z,k}$         & NA & $C^{s}_{z,k}$                             & $s_{z,t,k}$                   \\ % $\widetilde{s}^{-}_{z,k}$
                storages (energy)   & $\widetilde{V}_{z,k}$                 & $\widetilde{v}^{+}_{z,k}$         & NA & $C^{v}_{z,k}$                             & NA                            \\ % $\widetilde{v}^{-}_{z,t,k}$
                transmission & $\widetilde{X}_{z,zz}$                & $\widetilde{x}^{+}_{z,zz}$        & NA & $C^{x}_{z,zz}$                            & $x_{z,zz,t}$                  \\ % $\widetilde{x}^{-}_{z,zz}$
                \bottomrule
            \end{tabulary}

            \begin{tablenotes}
                \item [$\dagger$] parameter
                \item [$\ddagger$] variable
            \end{tablenotes}
        \end{threeparttable}
    \end{table*}

    \newpage

    \subsection{Mathematical description} \label{mathmodel}

    \paragraph{Model objective}
    \emph{medea} minimizes total system cost $c$, i.e. the total cost of generating electricity and heat from technologies and capacities adequate to meet demand, over a large number of decision variables, essentially representing investment and dispatch decisions in each market zone $z$ of the modelled energy systems.
    \begin{align}
        \min c = \sum_{z} (c_{z})
    \end{align}
    Zonal system costs $c_{z}$ are the sum of fuel cost $c^{b}_{z,t,i}$, emission cost $c^{e}_{z,t,i}$, operation and maintenance cost, capital costs of investment in conventional and intermittent generation ($c^{g}_{z}$, $c^{r}_{z}$), storage ($c^{s,v}_{z}$) and transmission ($c^{x}_{z}$) equipment, and the cost of non-served load ($c^{q}_{z}$) that accrues when demand is not met, e.g. when there is a power outage.
    \begin{align}
%\begin{split}
        c_{z} = \sum_{t,i}  c^{b}_{z,t,i} + \sum_{t,i} c^{e}_{z,t,i} + \sum_{i} c^{om}_{z,i} + \sum_{n} c^{om}_{z,n} + c^{g}_{z} +
        c^{r}_{z} + c^{s,v}_{z} + c^{x}_{z} + c^{q}_{z} \qquad \qquad \forall z
%\end{split}
    \end{align}
    The components of zonal system costs are calculated as given in equations (\ref{fuel_cost}) to (\ref{lost_load_cost}).
    Lower-case $c$ represent total cost, while upper-case $C$ denotes specific, annualized capital cost of technology investment.
    Prices for fuels and \COO are denoted by $P$.
    \begin{align}
        &c^{b}_{z,t,i}& =&\ \sum_{f} \left( P_{t,z,f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i \label{fuel_cost} \\
        &c^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: e_{z,t,i} \right) \qquad \qquad &\forall z,t,i\\
        &c^{om}_{z,i}& =&\ Q^{g}_{i} \left(\widetilde{G}_{z,i} - \widetilde{g}^{-}_{z,i} + \widetilde{g}^{+}_{z,i}\right) + \sum_{t} \sum_{m} \sum_{f} \left(O^{g}_{i} \: g_{z,t,i,m,f}\right) \qquad \qquad &\forall z,i \\
        &c^{om}_{z,n}& =&\ Q^{r}_{n} \left(\widetilde{R}_{z,n} - \widetilde{r}^{-}_{z,n} + \widetilde{r}^{+}_{z,n}\right) + \sum_{t} \left(O^{r}_{n} \: r_{z,t,n}\right) \qquad \qquad &\forall z,n \\
        &c^{g}_{z}& =&\ \sum_{i} \left( C^{g}_{z,i} \: \widetilde{g}^{+}_{z,i} \right) \qquad \qquad &\forall z\\
        &c^{r}_{z}& =&\ \sum_{n} \left( C^{r}_{z,n} \: \widetilde{r}^{+}_{z,n} \right) \qquad \qquad &\forall z\\
        &c^{s,v}_{z}& =&\ \sum_{k} \left( C^{s}_{z,k} \: \widetilde{s}^{+}_{z,k} + C^{v}_{z,k} \: v^{+}_{z,k} \right) \qquad \qquad &\forall z\\
        &c^{x}_{z}& =&\ \frac{1}{2} \: \sum_{zz} (C^{x} \: \delta_{z,zz} \: \widetilde{x}^{+}_{z,zz}) \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
        &c^{q}_{z}& =&\ \mu \sum_{t} \sum_{m} q^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
    \end{align}

    \paragraph{Market clearing}
    In each hour, the markets for electricity and heat have to clear.
    Equation (\ref{market_clearing_el}) ensures that the total supply from conventional and intermittent sources, and storages equals total electricity demand plus net exports, electricity stored and used for heat generation.
    Likewise, equation (\ref{market_clearing_ht}) clears the heat market by equating heat generation to heat demand.
    \begin{align}
        \begin{split}
            \sum_{i} \sum_{f} g_{z,t,i,\text{el},f} + \sum_{k} s^{out}_{z,t,k} + \sum_{n} r_{z,t,n} &=  \\ D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + & \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} - q^{-}_{z,t,\text{el}} + q^{+}_{z,t} \qquad \forall z,t
        \end{split}
        \label{market_clearing_el}
    \end{align}
    \begin{align}
        \sum_{i} \sum_{f} g_{z,t,i,\text{ht},f} = D_{z,t,\text{ht}} - q^{-}_{z,t,\text{ht}} \qquad \forall z,t \label{market_clearing_ht}
    \end{align}
    \emph{medea} can be thought of as representing energy-only electricity and heat markets without capacity payments. Then, the marginals of the market clearing equations (\ref{market_clearing_el}) and (\ref{market_clearing_ht}), $\partial C / \partial D_{z,t,m}$, can be interpreted as the zonal prices for electricity and heat, respectively.

    \paragraph{Energy generation}
    Energy generation $g_{z,t,i,m,f} \geq 0$ is constrained by available installed capacity, which can be adjusted through investment ($\widetilde{g}^{+}_{z,i} \geq 0$) and decommissioning $\widetilde{g}^{-}_{z,i} \geq 0$.
    \begin{align}
        \sum_{f} g_{z,t,i,m,f} \leq \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} - \widetilde{g}^{-}_{z,i} \qquad \qquad \forall z,t,i,m
    \end{align}
    Generator efficiency $\eta$ determines the amount of fuel $b_{z,t,i,f} \geq 0$ that needs to be spent in order to generate a given amount of energy.
    \begin{align}
        g_{z,t,i,m,f} = \sum_{f} \eta_{i,m,f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \notin J, f
    \end{align}

    \paragraph{Thermal co-generation}
    Co-generation units jointly generate heat and electricity. All feasible combinations of heat and electricity generation along with the corresponding fuel requirement are reflected in so-called `feasible operating regions'.
    The elements $l \in L$ span up a three-dimensional, convex feasible operating region for each co-generation technology.
    The weights $w_{z,t,i,l,f} \geq 0$ form a convex combination of the corners $l$, which are scaled to the available installed capacity of each co-generation technology.
    Defining weights over fuels allows co-generation units to switch fuels between multiple alternatives.
    Heat and electricity output along with the corresponding fuel requirement is then set according to the chosen weights.
    \begin{align}
        \sum_{l} \sum_{f} w_{z,t,i,l,f} = \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} - \widetilde{g}^{-}_{z,i} \qquad \qquad \forall z,t,i \in J \\
        g_{z,t,i,m,f} = \sum_{l} \sum_{f} \psi_{i,l,m} \: w_{z,t,i,l,f} \qquad \qquad \forall z,t,i \in J, m \\
        b_{z,t,i,f} = \sum_{l} \chi_{i,l,f} \: w_{z,t,i,l,f} \qquad \qquad \forall z,t,i \in J, f\\
        w(z,t,i,l,f) = 0 \qquad \qquad \forall z,t,i,k,f: \chi_{i,l,f} = 0
    \end{align}

    \paragraph{Intermittent electricity generation}
    Electricity generation from intermittent sources wind (on-shore and off-shore), solar irradiation, and river runoff follows generation profiles $\phi_{z,t,n} \in [0,1]$ and is scaled according to corresponding installed ($ \widetilde{R}_{z,n}$) and added ($\widetilde{r}^{+}_{z,n} \geq 0$) capacity.
    \begin{align}
        r_{z,t,n} = \phi_{z,t,n} \: \left( \widetilde{R}_{z,n} - \widetilde{r}^{-}_{z,n} + \widetilde{r}^{+}_{z,n} \right) \qquad \qquad \forall z,t,n
    \end{align}

    \paragraph{Electricity storages}
    Charging ($s^{in}_{z,t,k} \geq 0$) and discharging ($s^{out}_{z,t,k} \geq 0$) of storages is constrained by the storages' installed ($\widetilde{S}^{in}_{z,k}, \widetilde{S}^{out}_{z,k}$) and added ($\widetilde{s}^{+}_{z,k} \geq 0$) charging and discharging power, respectively. Similarly, the total energy that can be stored is constrained by the storage technology's initial ($\widetilde{V}_{z,k}$) and added ($\widetilde{v}^{+}_{z,k} \geq 0$) energy capacity.
    \begin{align}
        s^{out}_{z,t,k} \leq \widetilde{S}^{out}_{z,k} + \widetilde{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
        s^{in}_{z,t,k} \leq \widetilde{S}^{in}_{z,k} + \widetilde{s}^{+}_{z,k} \qquad \qquad \forall z,t,k \\
        v_{z,t,k} \leq \widetilde{V}_{z,k} + \widetilde{v}^{+}_{z,k} \qquad \qquad \forall z,t,k
    \end{align}
    Storage operation is subject to a storage balance, such that the current energy content must be equal to the previous period's energy content plus all energy flowing into the storage less all energy flowing out of the storage.
    \begin{align}
        v_{z,t,k} = \rho_{z,t,k} + \eta^{in}_{z,k} \: s^{in}_{z,t,k} - (\eta^{out}_{z,k})^{-1} \: s^{out}_{z,t,k} + v_{z,t-1,k} \qquad \qquad \forall z,t,k: t>1, \: \eta^{out}_{z,k} > 0
    \end{align}
    Since the model can add storage power capacity and energy capacity independently, we require a storage to hold at least as much energy as it could store in (or out) in one hour.
    \begin{align}
        \widetilde{v}^{+}_{z,k} \geq \widetilde{s}^{+}_{z,k} \qquad \qquad \forall z,k
    \end{align}

    \paragraph{Emission accounting}
    Burning fossil fuels for energy generation produces emissions of carbon dioxide (\COO). The amount of \COO emitted is
    tracked by the following equation
    \begin{align}
        e_{z,t,i} = \sum_{f} \left( \varepsilon_{f} \: b_{z,t,i,f} \right) \qquad \qquad \forall z,t,i
    \end{align}

    \paragraph{Electricity exchange}
    Implicitly, \emph{medea} assumes that there are no transmission constraints within market zones.
    However, electricity exchange between market zones is subject to several constraints.

    First, exchange between market zones is constrained by available transfer capacities. Transfer capacities can be expanded at constant, specific investment cost (see equation (\ref{transmission_expansion_cost})). This rules out economies of scale in transmission investment that might arise in interconnected, meshed grids.
    \begin{align}
        x_{z,zz,t} \leq \widetilde{X}_{z,zz} + \widetilde{x}^{+}_{z,zz} \qquad \qquad \forall z, zz, t \\
        x_{z,zz,t} \geq - \left( \widetilde{X}_{z,zz} + \widetilde{x}^{+}_{z,zz} \right) \qquad \qquad \forall z, zz, t
    \end{align}
    By definition, electricity net exports $x_{z,zz,t}$ from $z$ to $zz$ must equal electricity net imports of $zz$ from $z$.
    \begin{align}
        x_{z,zz,t} = -x_{zz,z,t} \qquad \qquad \forall z, zz, t
    \end{align}
    Added transmission capacities can be used in either direction.
    \begin{align}
        \widetilde{x}^{+}_{z,zz} = \widetilde{x}^{+}_{zz,z} \qquad \qquad \forall z, zz
    \end{align}
    Finally, electricity cannot flow between zones where there is no transmission infrastructure in place (including intra-zonal flows).
    \begin{align}
        x_{z,zz,t} = 0 \qquad \qquad \forall z, zz, t: \widetilde{X}_{z,zz} = 0 \\
        x_{zz,z,t} = 0 \qquad \qquad \forall z, zz, t: \widetilde{X}_{z,zz} = 0
    \end{align}

    \paragraph{Decommissioning}
    Keeping plant available for generation gives rise to quasi-fixed operation and maintenance costs.
    Such cost can be avoided by decommissioning an energy generator. This is modelled as a reduction in generation capacity, which cannot exceed installed capacity.
    \begin{align}
        \widetilde{g}^{-}_{z,i} &\leq \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i} \qquad \qquad \forall z,i \\
        \widetilde{r}^{-}_{z,n} &\leq \widetilde{R}_{z,n} + \widetilde{r}^{+}_{z,n} \qquad \qquad \forall z,n
    \end{align}

    \paragraph{Ancillary services}
    Power systems require various system services for secure and reliable operation, such as balancing services or voltage support through the provision of reactive power. Such system services can only be supplied by operational generators. Thus, we approximate system service provision by a requirement on the minimal amount of spinning reserves operating at each hour.
    We assume that ancillary services are supplied by conventional (thermal) power plants, hydro power plants, and storages.
    The requirement for spinning reserves is proportional to electricity peak load $\widehat{D}_{z,\text{el}} = \max_{t} D_{z,t,\text{el}}$ and peak generation from wind and solar resources, where $\widehat{\phi}_{z,n} = \max_{t} \phi_{z,t,n}$.
    \begin{align}
        \sum_{i} \sum_{f} \left( g_{z,t,i,\text{el},f} \right) + r_{z,t,\text{ror}} + \sum_{k} \left( s^{out}_{z,t,k} + s^{in}_{z,t,k} \right) \geq \lambda_{z} \widehat{D}_{z,\text{el}} + \sigma_{z} \sum_{n \setminus \{ \text{ror}\}} \widehat{\phi}_{z,n} (\widetilde{R}_{z,n} + \widetilde{r}^{+}_{z,n}) \qquad \forall z,t
    \end{align}

    \paragraph{Curtailment}
    Electricity generated from intermittent sources can be curtailed (disposed of) without any further cost (apart from implicit opportunity cost).
    \begin{align}
        q^{+}_{z,t} \leq \sum_{n} r_{z,t,n} \qquad \qquad \forall z, t
    \end{align}

\end{document}

### 2020-04-28 projects/asparagus/doc/paper_clean.tex
postprocessing of asparagus results

@@
 
-% graphics
+% graphics and figures
 \usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subcaption}
 
@@
     PV instead of wind power leads to an increase in dispatchable thermal generation.
-    As thermal generation in Austria is predominantly by fossil thermal, this induces an increase in \COO emissions due
-    to the increased use of PV.
+    As thermal generation in Austria is predominantly fossil thermal, this induces an increase in \COO emissions due
+    to the increased use of PV\@.
+
+    Curtailment in Austria is declining up to
 
@@
         \centering
-        \includegraphics[width=0.8 \textwidth]{./figures/undisturbed_base.pdf}
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{./figures/undisturbed_base_share.pdf}
+            \caption{Relative substitution}
+            \label{fig:rel_substitution}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{./figures/undisturbed_base.pdf}
+            \caption{Absolute substitution}
+            \label{fig::abs_substitution}
+        \end{subfigure}
         \caption{Opportunity cost of wind power assuming PV overnight cost of 630 EUR/kWp}

### 2020-04-28 projects/asparagus/doc/plots_asparagus.py

@@ -195,10 +195,38 @@ wind_add = results.loc[idx['base', :, :, 36715], idx['AT', 'add_wind_on']].unsta
 wind_add.index = wind_add.index.droplevel([0, 2])
 
 undisturbed_marginal = undisturbed_base.iloc[::-1].diff().divide(wind_add.iloc[::-1].diff().round(8), axis=0) * -1
+undisturbed_marginal_share = undisturbed_marginal.copy()
 undisturbed_marginal.columns = [f'CO2 Price: {p} €/t' for p in undisturbed_marginal.columns]
 
 plot_lines(undisturbed_marginal / 1000, os.path.join(FPATH, 'undisturbed_base.pdf'), xlim=[18, 0], ylim=[0, 75],
            xlabel='Added Capacity of Wind [GW]', ylabel='thousand € / MW', color=REFUEL_COLORS)
+
+# %% ------ ----- ----- plot opportunity cost of wind with SHARE of deployment
+uplim = results.loc[idx['base', PRICE_CO2, :, 36715], idx['AT', 'add_wind_on']].max()
+# divide index by uplim, delete indices larger than one except for the smallest and set the smallest to 1
+
+lgnd = []
+
+figure, axis = plt.subplots(1, 1, figsize=(8, 5))
+axis.set_ylabel('thousand € / MW')
+axis.set_xlabel(r'$\phi$')
+i = 0
+for p in undisturbed_marginal_share.columns:
+    oc_by_share = undisturbed_marginal_share.loc[:, p]
+    oc_by_share.index = undisturbed_marginal_share.index / wind_add.loc[:, p].max()
+
+    axis.plot(oc_by_share / 1000, color=REFUEL_COLORS[i])
+    i = i + 1
+    lgnd.append(rf'CO$_2$ Price: {p} €/t')
+
+axis.legend(lgnd)
+# axis.set_ylim(ylim)
+axis.set_xlim([1, 0])
+axis.grid()
+plt.rcParams.update({'font.size': 16})
+plt.tight_layout()
+plt.savefig(os.path.join(FPATH, 'undisturbed_base_share.pdf'))
+plt.close()
 
 # %% line plot of cost of undisturbed landscapes - low capital cost of pv
 undisturbed_cost_low = results.loc[idx['base', :, :, 16715], idx['AT', ['syscost', 'export_value', 'import_value',

### 2020-04-28 projects/asparagus/doc/slides.tex
updates asparagus slides

@@
             (1.0cm,2cm)
-            Consider a renewable energy technology $\omega$ that
+            Consider a renewable energy technology $T$ that
             \begin{itemize}
-                \item is \emph{system cost} minimizing, i.e.\ it saves cost relative to its best alternative $\mu$, but
+                \item is deployed at the \emph{system cost} minimizing level $T^\ast$, i.e.\ it saves system cost
+                relative to its best alternative $P$, but
                 \item comes at \emph{social cost}, e.g. due to unaccounted negative external effects
@@
 
-            A social planner should substitute $\omega$ with $\mu$ up to the point where the increase in system cost
-            equals the reduction in harm (from $\omega$'s negative externality unaccounted in system cost)
+            A social planner should substitute $T$ with $P$ up to the point where the increase in system cost
+            equals the reduction in harm (from $T$'s negative externality which is unaccounted for in system cost)
         \end{textblock*}
@@
                     ticks=none,
-                    xlabel={$\omega$ substituted with equivalent $\mu$},
+                    xlabel={$\Phi= \frac{T^\ast-T}{T^\ast}$},
                     ylabel={cost},
@@
                     \addplot[blue, thick] (x, 2*x^1.1-0.933);
-                    \node at (axis cs: 1.925,3) [blue, align=left] {additional\\ system\\ cost};
+                    \node at (axis cs: 2.05,2.95) [blue, align=left] {additional\\ system cost};
 
                     \addplot[bokugreen, thick] (x, -2*x^0.8+4);
-                    \node at (axis cs: 2.0,1) [bokugreen, align=left]{harm \\(from RET\\ externality)};
+                    \node at (axis cs: 1.9,0.925) [bokugreen, align=left]{harm};
+                    \node at (axis cs: 0.5,-0.25) []{$0$};
+                    \node at (axis cs: 1.925,-0.25) []{$1$};
                 \end{axis}
@@
             \centering
-            \begin{tabular}{| l | r | r | r | r | r |}
-                \hline
+            \begin{tabular}{lrrrrr}
+                %\hline
                 & 2018 & 2018\footnote[frame]{meteorological conditions as in 2016} & Policy & 2030 & 2030 \\
@@
                 Hydro (run-of-river) & 5.72 & 28.34 & +5 & 33.34 & 6.73 \\ \hline
-                Biomass & $\sim 1$ & 4.78 & +1 & 5.78 & $<2$ \\ \hline
+                Biomass & $\sim 1$ & 4.78 & +1 & 5.78 & $<2$ %\\ \hline
             \end{tabular}
@@
     \emph{medea} is available on \textcolor{blue}{\href{https://github.com/inwe-boku/medea}{github.com/inwe-boku/medea}}
-    under an open MIT license
+    (MIT license)
 \end{center}
@@
     \section{Results}
+    \begin{frame}
+        \frametitle{The cost of undisturbed landscapes}
+
+        \begin{textblock*}{10cm}
+            (1.25cm, 1.75cm) % {block width} (coords)
+            \centering
+            \begin{tikzpicture}
+                \draw (4.5,0.0) node[] {\small Opportunity Cost of Wind Power};
+            \end{tikzpicture}
+        \end{textblock*}
+
+        \begin{textblock*}{10cm}
+            (0.75cm, 2.25cm) % {block width} (coords)
+            \includegraphics[width=\textwidth, keepaspectratio]{./figures/undisturbed_base_share.pdf}
+        \end{textblock*}
+
+        \begin{textblock*}{4.5cm}
+            (10.9cm, 2.5cm)
+            \begin{small}
+                \begin{table}
+                    \begin{tabular}{p{1.8cm} | r l}
+                        Capital cost of solar PV & $630$ & \SI{}{\text{\euro}\per\kilo\watt\text{p}}
+                    \end{tabular}
+                \end{table}
+                \smallskip
+                \quad $\rightarrow$ about $\sfrac{2}{3}$ rooftop PV, \\
+                \quad $\sfrac{1}{3}$ open space PV
+
+                \vspace{3.75cm}
+                \quad wind power and solar PV \\ \quad are substitutes
+            \end{small}
+        \end{textblock*}
+    \end{frame}
 

### 2020-04-30 projects/asparagus/doc/paper_clean.tex
updates medea_asparagus.gms to restrict technology additions

@@
 %    •	LOOK AT WHAT HAPPENS WHEN GERMANY BUILDS MORE PV THAN ANTICIPATED (particularly in the low pv-cost scenario)
+    Effect of \COO prices on opportunity cost of wind power:
+    - lower \COO price means less 'voluntary' RET installed
+    - more PV means more ancillary capacities installed (?)
+    - more PV means more fuel used in dispatchable plant
+    - more PV means more curtailment (?)
+    - more PV means lower exports -- large dependency on \COO price. At price of 60-120, exports fall as less wind.
+    For low \COO price exports increase as less wind. -->> Need to look at capacities in Germany!
+    - more PV means
+    * PV capacity at full wind substitution is (almost) invariant to \COO price and comes in at 24.11-24.26 GW
+    * wind on capacity is sensitive to \COO price. At 120 \EUR \COO, 17.07 GW wind on added, at 0 \EUR \COO10.43 GW added
+    * austria builds coal plants -- this should be impossible! --> correct in re-run
+    * austria builds oil instead of nat gas plants. --> check and potentially correct cost versus HAt gas
+    * absolutely no heatpump added. why?
+
+    + Germany add lignite and coal. With rising \COO price more coal is added. (!)
+    + Germany adds oil. At lower \COO price more oil, less ng added
+    + No bio added in Germany (proly restriction)
+    + No heatpumps added --> Why???
+    + Germany adds PV and wind only with \COO price at 120
+
+    Need to re-run model with restricted additions
 

### 2020-04-30 projects/asparagus/doc/plots_asparagus.py
updated asparagus demand assumption;

@@ -109,23 +109,6 @@ def plot_lines(df, fname, xlim=None, ylim=None, xlabel=None, ylabel=None, color=
     plt.close()
 
 
-# %% ----- ----- ----- ----- settings ----- ----- ----- -----
-idx = pd.IndexSlice
-PRICE_CO2 = 90
-REFUEL_COLORS = ['#c72321', '#0d8085', '#f0c220', '#595959']
-RES_COLORS = ['#d69602', '#ffd53d', '#3758ba', '#7794dd']
-RES_COLORS3 = ['#d69602', '#e5b710', '#ffd53d', '#3758ba', '#3b68f9', '#7794dd']
-
-ANNUITY_FACTOR = 0.05827816
-FLH_PV = 857.4938
-FLH_WINDON = 2015.0359
-
-RPATH = os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', 'asparagus', 'results', 'results.csv')
-FPATH = os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', 'asparagus', 'doc', 'figures')
-
-if not os.path.exists(FPATH):
-    os.makedirs(FPATH)
-
 # %% ----- ----- ----- ----- read results ----- ----- ----- -----
 results = pd.read_csv(RPATH, decimal=',', delimiter=';', index_col=[0, 1, 2, 3, 4], header=[0])
 results = results.unstack(-1)

### 2020-04-30 projects/asparagus/settings_asparagus.py

@@ -18,19 +18,19 @@ accounts for Germany's nuclear exit and its proposed coal phase-out.
 As restrictions on land availability for wind turbines are not specified, we generate scenarios of deployable wind 
 turbine capacity in the range between 0 and unconstrained deployment. This is repeated for a range of CO2 price 
 assumptions.
-For sensitivity analysis, we also vary the capitalcost of solar PV. 
+For sensitivity analysis, we also vary the capital cost of solar PV. 
 """
 
 # campaigns
 dict_campaigns = {
     'base': {
-        'co2_price': range(120, 29, -30),
+        'co2_price': range(120, -1, -30),
         'wind_cap': range(18, -1, -2),
         'pv_cost': [36715, 16715]
     },
     'cheappv': {
-        'co2_price': [30, 60, 120],
-        'wind_cap': range(18, 17, -2),
+        'co2_price': range(120, -1, -30),
+        'wind_cap': [18],
         'pv_cost': range(36715, 16026, -1000)
     }
 }
@@ -65,7 +65,7 @@ scenario_2030 = {
         'wind_off': [0],
         'pv': [1.1],
         'ror': [6.7],
-        'd_power': [62038.515 + 17600]
+        'd_power': [62038.515 + 24000]
     },
     'DE': {
         'bio': [1.3],

### 2020-04-30 projects/asparagus/doc/slides.tex

@@
         \framesubtitle{Estimating the opportunity cost of wind power}
-        To determine the opportunity cost of substituting wind power with solar PV, we
+        To determine the opportunity cost of wind power versus solar PV, we
         \begin{enumerate}
@@
             \item restrict deployment of wind power by a small margin (so that the next best RET substitutes for wind power) and
-            observe net system cost $c_{net}$ (calculated as total system cost including air pollution cost net of trade balance) \label{step2}
-            \item repeat \ref{step2} till no wind power can be deployed
+            observe net system cost $c_{net}$ in the country of interest (calculated as total system cost including air
+            pollution cost net of trade balance) \label{step2}
+            \item repeat \ref{step2} till no wind power can be deployed anymore
             \item the opportunity cost of wind power is approximated by the change in $c_{net}$ in response to a change in wind
@@
         \frametitle{The cost of undisturbed landscapes}
-
+        \framesubtitle{Relative measure of substitution}
         \begin{textblock*}{10cm}
@@
 
-                \vspace{3.75cm}
-                \quad wind power and solar PV \\ \quad are substitutes
+                \medskip
+                At $\Phi=1$ RET deployment is as in the unconstrained system cost minimum.
+
+                \smallskip
+                At $\Phi=0$ the least cost technology is fully substituted by its best alternative.
             \end{small}
@@
         \frametitle{The cost of undisturbed landscapes}
-
+        \framesubtitle{Absolute measure of substitution}
         \begin{textblock*}{10cm}
@@
         \framesubtitle{Cost with restricted wind power}
-\begin{textblock*}{10cm}(0.75cm, 2.0cm) % {block width} (coords)
+        \begin{textblock*}{10cm}
+            (0.75cm, 2.0cm) % {block width} (coords)
             \includegraphics[width=\textwidth, keepaspectratio]{./figures/cost}
@@
 
-\begin{textblock*}{4cm}(11cm, 2.25cm)
+        \begin{textblock*}{4cm}
+            (11cm, 2.25cm)
             \begin{small}
@@
         \usebackgroundtemplate{
-        \begin{picture}(320,290)
+            \begin{picture}
+                (320,290)
                 \hspace{9.5cm}

### 2020-05-03 projects/asparagus/doc/paper_clean.tex
updates asparagus paper_clean.tex

--- 2020-04-28 projects/asparagus/doc/paper_clean.tex
@@
-\documentclass[final, 3p, times]{elsarticle} %review, twocolumn, 5p, 12pt, authoryear, twocolumn, 5p,
+\documentclass[review, 3p, times]{elsarticle} %final, twocolumn, 5p, 12pt, authoryear, twocolumn, 5p,
 
@@
     \section{Introduction} \label{sec:introduction}
-    Future, low carbon electricity systems should largely rely on renewable energy technologies (RET), such as solar
-    photovoltaics (PV) or wind turbines, according to a substantial body of literature~\citep{Rodriguez2015}.\todo{references}
-    For central Europe, it is typically found that system-cost\footnote{System cost are the sum of fuel and emission
-    cost,operation and maintenance cost, and investment cost} minimizing systems to rely on large shares of onshore
-    wind power~\citep{Rodriguez2015}\todo{references}.
-    Such undertakings frequently implement upper bounds on the deployment of RET, which are thought to reflect the
-    available potential for RET erection~\citep{references}\todo{references}.
-    These potentials, in turn, are usually derived from surveys of land available and suitable for the erection of RET\@.
-    The typical methodology is to exclude areas that are considered not suited because of insufficient resource
-    availability or because of technical or legal restrictions such as minimum distances to settlements.
-    Areas not excluded are then considered as feasible for RET deployment \citep[e.g.][]{McKenna2015, Bosch2017, Ryberg2020}.
-    \todo[inline]{refer to excluded areas as black, feasible areas as white}
-    However, researchers are increasingly aware that such a 'black-or-white'-approach fails to account for more nuanced
-    issues such as the social acceptance of interference with landscapes.
-    Indeed, several contributions sought to improve on such shortcomings, for example by participatory
-    modelling~\citep{Hoeltinger2016}, by xxx~\citep{Jaeger2016} or xxx~\citep{Harper2019}\todo[inline]{add mini review paragraph}
+    Future, low carbon energy systems are likely to largely rely on electricity from renewable energy technologies
+    (RET), such as solar photovoltaics (PV) or wind turbines, according to a substantial body of
+    literature~\citep[e.g.][]{Becker2014,Delucchi2011,Fernandes2014,Gils2017,Jacobson2011,Schmidt2016}.
+    \todo{references not good for statement}
+    A broad stream of papers~\citep[e.g.][]{Brown2018,Rodriguez2015,Schlachtberger2018,Scholz2017} investigates
+    conditions under which total system cost, conceived as the sum of fuel and emission cost, operation and maintenance
+    cost, and investment cost, can be minimized.
+    In this context, available resource potentials are an important boundary condition for the determination of such
+    least-cost system configurations.
+    Fortunately, such resource potentials are assessed by numerous scholars at various levels.
+    Frequently, these assessments combine indicators of the wind resource with technical or legal restrictions on land
+    available for RET deployment~\citep[e.g.][]{Bosch2017,McKenna2015,Ryberg2020}.
+    However, there is an increasing awareness that such approaches are not a proper reflection of deployable RET
+    capacites, as a more nuanced analysis would have to account for the social acceptance of RET.
+    Opposition to wind turbines, for example, is often based on perceived negative interference with
+    landscapes~\cite{Mattmann2016}, potential harm to wildlife~\cite{Loss2013, Voigt2015, Wang2015}, or
+    noise~\cite{Wang2015a}.
+    Zerrahn~\cite{Zerrahn2017} provides an excellent survey of the literature on wind power externalities.
 
-    In contrast, the work of~\cite{Drechsler2011} % Drechsler et al. 2011: Combining spatial modelling and choice experiments ...
-    \todo[inline]{mention later paper by Drechsler} \cite{Drechsler2017}
-    \todo[inline]{refer to negative externalities not included in system cost}
+    Based on this insight, several contributions sought to take social acceptance into account when assessing RET
+    potentials.
+    Höltinger et al.\ \cite{Hoeltinger2016}, for example, relied on participatory modeling to identify uncontested wind
+    power potentials in Austria.
+    Jäger et al.\ \cite{Jaeger2016} \todo[inline]{brief description of paper}
+    Harper et al.\ \cite{Harper2019} develop a spatially resolved multi-criteria analysis including technological, legal,
+    and social constraints to determine sites suitable for wind turbine erection.
+    % mention minimum distances to settlements
 
-    We improve on this work by xxx. \todo[inline]{explain what we do also in contrast to Drechsler et al.}
+    In contrast to the determination of such `augmented potentials' trying to identify sites that are certainly suited
+    for RET deployment, Drechsler et al.\ \cite{Drechsler2011} argue that a welfare-optimal spatial allocation of RET
+    needs to consider social costs and benefits.
+    Social conflict would then be mitigated by selecting sites with the lowest welfare loss.
+    To demonstrate their approach, Drechsler et al.\ conduct a case study for a region in Germany, where they survey
+    preferences of inhabitants to explicitly consider the trade-off between \emph{social cost} arising from negative
+    externalities of wind power and electricity generation cost of wind power due to spatially varying wind conditions.
+    Some of the authors studied implications of welfare optimal allocation of wind turbines also in more recent
+    contributions.
+    Salomon et al. (2020)~\citep{Salomon2020} study welfare-optimal rules on minimum distances between wind turbines and
+    human settlements.
+    Drechsler et al. (2017)~\citep{Drechsler2017} study welfare-optimal allocation of wind turbines and solar PV.\todo[inline]{expand on this}
 
-    \subsection{Framework}
+    We expand on this work by conceptualising welfare optimal RET deployment in more general settings, so that we can
+    account for the opportunity cost of RET substitution (arising from different system integration cost of RET), as
+    well as spatial differences in endowments with renewable resources and negative externalities from RET depending on
+    spatially diverse characteristics, as we lay out in section~\ref{subsec:framework} below.
+    \todo[inline]{how? accounting for everything could be done by spatially resolved model that has info on resource
+    availability and on negative impact for each location.
+    Alternatively, use model for opportunity cost only and calc deviation against mean resource?}
+
+    To undergird our proposed methodology, we use a numerical power system model (see
+    section~\ref{sec:data-and-methods}) to investigate the trade-off between deploying different renewable energy
+    technologies in terms of system cost, which constitute an important, though incomplete part of social cost.
+    We do this to underline the importance of RET substitution and the corresponding opportunity costs in the
+    identification of spatial allocations of RET as well as in power system planning in general.
+    In the latter context, we also contribute to the analysis of near-optimal power system
+    configurations~\citep[see e.g.][]{Neumann2019,Schlachtberger2017}.
+
+    % What do we want to show?
+    % We embed our demonstration in a wider framework suited to expand renewable energy technologies at minimal social cost,
+    % thereby mitigating potential social conflict over the transition towards low-carbon energy systems.
+    % The current contribution demonstrates that:
+    % - there is a tradeoff between different RET
+
+    \subsection{Framework} \label{subsec:framework}
     Similar to~\cite{Drechsler2011}, we regard the problem as one of welfare optimization, in which we seek to minimize
@@
     RET\@.
-    Such a social planner faces a trade-off between
+    The social planner faces a trade-off between
     (i) (potential) negative externalities of a RET at the local level (recall that externalities from emissions are
@@
     For illustration, consider figure~\ref{figure:social-optimum}, which depicts the effect of deviating from a
-    system-cost minimizing RET allocation by substituting some share of technology $A$ with its best alternative,
+    system-cost minimizing RET allocation by substituting some share $\alpha$ of technology $A$ with its best alternative,
     technology $B$.\footnote{Here, we consider the interesting case in which in system-cost minimum a RET causes higher
@@
     than the corresponding decline in harm.
-    \bigskip
-    \todo[inline]{the following subsection must be overhauled}
-    Within this framework, we quantify the opportunity cost of wind power ($s$ in Figure~\ref{figure:social-optimum})
-    for the case of Austria in the work at hand.
-    We chose Austria as our case study, as the government has comitted itself to meeting \SI{100}{\percent} of
+
+    \subsection{Case study}
+    Based on this framework, we use the case of Austria to quantify the opportunity cost of wind power (line $s$ in
+    figure~\ref{figure:social-optimum}).
+    We do not set out to quantify harm (i.e.\ negative external effects) induced by wind power.
+    While such an endeavor is beyond the scope of this paper, it is a frequent undertaking in the
+    literature~\citep[e.g.][]{Droees2016, Gibbons2015, Heintzelmann2017, Jensen2018, Kussel2019, Lang2014, Sims2008, Sunak2016, Vyn2014},
+    even though an assessment of negative wind turbine externalities has to the best of our knowledge not been conducted
+    for Austria yet.
+
+    We chose Austria as our case to study, as the government has committed itself to meeting \SI{100}{\percent} of
     electricity consumption from renewable sources (on annual balance, excluding industry own-consumption and system
@@
     electricity generation constant.
-    Moreover, the bulk of additional renewable electricity must come either from wind power or solar photovoltaics, as
-    the other renewables potentials are largely exploited already (see section~\ref{section:}), which simplifies our
-    analysis further.
-    In the context of Austria, as in many other European countries, least-cost renewable electricity systems are largely
-    based on wind power~\citep{Rodriguez2015}. \todo[inline]{references}
-    Harm is inflicted by local external effects of wind turbines, such as flickering, noise, negative impacts on
-    wildlife, or (perceived) disturbance of landscapes.
-    \cite{Mattmann2016} identifies the latter as a main trigger of opposition against wind power at the local level.
-    For a review of wind power externalities, see~\cite{Zerrahn2017}.
-    \todo[inline]{1) motivate Austrian case, 2) make clear that paper only looks at opportunity cost of wind power --
-    cost of undisturbed landscapes, NOT at harm, which is assessed in various ways by many papers, though not for
-    Austria}
+    Moreover, in Austria the bulk of additional renewable electricity must come either from wind power or solar
+    photovoltaics, as other renewables potentials are largely exploited already (see
+    section~\ref{subsubsec:assumptions-austria:}), which simplifies our analysis further.
+
+    In the context of Austria, we find a least-cost renewable electricity system to be largely based on wind power,
+    which allows us to quantify the opportunity cost of wind power in terms of its best alternative solar PV.
+    Corresponding results are given in section~\ref{sec:results-discussion}.
 
@@
     model \emph{medea}, summarized in section~\ref{subsec:medea}, which we instantiate with scenario assumptions
-    (see section~\ref{subsec:data}) and observed data of the year 2016 (see section~\ref{subsec:scenarios}).
-
+    and observed data of the year 2016 (see section~\ref{subsec:data}).
     For determining the opportunity cost of wind power, we proceed as follows:
@@
 % ***** medea *****
-
     \subsection{Power system model \emph{medea}} \label{subsec:medea}
@@
     prospective Austrian (and German) power system.
-
     The model is cast as a linear optimization seeking to minimize total system cost, which consist of fuel and emission
@@
     storage, and transmission assets, and (potential) cost of non-served load.
+    In terms of economics, the model reflects a perfectly competitive energy-only market with fully price-inelastic
+    demand and perfect foresight of all actors.
 
@@
     electrical efficiency, the electricity loss per unit of heat production, and the back pressure coefficient.
-    Electricity generation from intermittent sources (wind, water, solar) is subject to exogenous hourly generation
-    profiles, which are scaled according to total installed capacities.
+    Electricity generation from intermittent sources (wind, run-of-river hydro, solar) is subject to exogenous hourly
+    generation profiles, which are scaled according to total installed capacities.
     Electricity from these sources can be curtailed at no additional cost (free disposal).
 
-    \todo{Energy storage}
     Electricity can be stored in reservoir and pumped hydro storages or batteries.
-    Generation from hydro storage plants is constrained by turbine capacity and energy contained in reservoirs of
-    limited size.
-    Hydro reservoirs are filled by inflows or by pumping (pumped storages only).
-    To better capture operational differences, we model daily, weekly and seasonal reservoir and pumped storage plants
-    separately.
-
-    \todo{Electricity exchange}
-    Electricity can be exchanged between Austria and Germany.
-
-    \todo{Ancillary services}
-    To ensure a stable and secure operation of the electricity system, power plants must provide ancillary services
-    (e.g. frequency control, voltage support).
-    We assume that this requires generators with an installed capacity of at least 21 GW2 to be operational (either
-    generating or pumping in case of pumped hydro storages) at any point in time.
-
+    The capacity of hydro storages can not be expanded, as we assume existing potentials to be exhausted.
+    Battery capacity, on the other hand, can be added endogenously.
+    Generation from storages is constrained by installed capacity and energy contained in storage.
+    Hydro reservoirs are filled by exogenously given inflows.
+    Pumped hydro storages and batteries can actively store electricity for later use.
+    To better capture operational differences of hydro storage units, we model daily, weekly and seasonal reservoir and
+    pumped storage plants separately.
+    To ensure a stable and secure operation of the electricity system, ancillary services (e.g. frequency control,
+    voltage support) are required.
+    We model ancillary service needs as a minimum requirement on spinning reserves operating at any point in time.
+    Thus, we assume that ancillary services can be provided by thermal power plant, run-of-river hydro plant or any
+    type of storage technology.
+    We do not model transmission and distribution grids within Austria or Germany.
+    We do, however, account for the congestion management scheme introduced by German and Austrian authorities in 2018,
+    that limits cross-border electricity trade.
+    \todo[inline]{is this elegant? Should i mention that investment in x-border capacity is allowed but does not occur?}
     For a detailed, mathematical description please refer to \ref{sec:medea}.
 
-    Data processing is implemented in python, the optimization model is based on GAMS. Running the model on an Intel
-    i7-8700 machine with 16 GB RAM, using CPLEX 12.9 as a solver, takes 10--15 minutes, depending on the model
-    parameters.
-    \todo[inline]{perfect foresight, perfect competition, reflects energy-only market}
+    Data processing is implemented in python, while the optimization model is based on GAMS.
+    Running the model on an Intel i7-8700 machine with 16 GB RAM, using CPLEX 12.9 as a solver, takes 10--15 minutes,
+    depending on the model parameters.
 
@@
     \subsubsection{Scenario Assumptions for Austria}\label{subsubsec:assumptions-austria}
-    Austria has set itself the goal of generating \SI{100}{\percent} of electricity consumption from renewable sources
-    on annual balance by 2030.
-    However, industry own consumption and system services, which currently account for \SI{10}{\percent} of annual
+    The Austrian government has set itself the goal of generating \SI{100}{\percent} of electricity consumption from
+    renewable sources on annual balance by 2030~\citep{Regierungsprogramm2020}.
+    However, industry own consumption and system services, which currently account for about \SI{10}{\percent} of annual
     electricity consumption, are exempt.
-    Hence, in our analysis we set the policy goal to generating at least \SI{90}{\percent} of annual electricity
-    consumption from renewable sources.\footnote{Please note that this policy goal does not imply that the generated
-    electricity must actually be consumed.
-    Hence, we consider curtailed electricity as contributing to the policy goal.}
     The government plans to achieve this goal by generating an additional \SI{27}{\tera\watt\hour} of electricity
-    annually from renewable sources.
-    Consequently, we have set prospective electricity consumption accordingly.
-
+    annually from renewable sources\footnote{Please note that this policy goal does not imply that the generated
+    electricity must actually be consumed. Hence, we count curtailed electricity as contributing to the policy goal.}.
     New hydro power plants are thought to contribute \SI{5}{\tera\watt\hour} annually, while additional electricity
@@
     We add corresponding generation capacities to reach these targets.
-
-    The remainder is projected to come from solar PV (\SI{11}{\tera\watt\hour} annually) and onshore wind turbines
-    (\SI{10}{\tera\watt\hour} annually)~\cite{Regierungsprogramm2020}.
-
-    The latter two RET are the only low-carbon electricity generation technologies with significant expansion potential.
-    Austrians ruled out the use of nuclear power in a 1978 referendum~\citep{Pelinka1983}.
-    Biomass lacks ecological and economical sustainability~\citep{Erb2018}, while available hydro power potentials were
-    already largely exploited~\citep{Walder2014} even before adding another \SI{5}{\tera\watt\hour} of annual generation.
-
+    The government projects the remainder to come from solar PV (\SI{11}{\tera\watt\hour} annually) and onshore wind
+    turbines (\SI{10}{\tera\watt\hour} annually)~\cite{Regierungsprogramm2020}.
     As we are interested in determining the opportunity cost of wind power versus its best alternative, we allow for
     endogenous investment in wind and solar power, without enforcing announced targets.
+    Other low-carbon energy technologies are not feasible at large scale in the Austrian context.
+    The use of nuclear power was banned in a 1978 referendum~\citep{Pelinka1983}.
+    Biomass lacks ecological as well as economical sustainability~\citep{Erb2018}, and hydro power potentials were
+    already largely exploited~\citep{Walder2014} even before government plans of adding another \SI{5}{\tera\watt\hour}
+    in annual hydro power generation.
+    Initial generation capacities for Austria are given in table~\ref{tab:instcaps}.
 
@@
     Following these announcements, we anticipate an end to nuclear power generation, a (partial) coal exit according to
-    recommendations by~\cite{WSB2019}, and a further expansion of renewable electricity generation in line
-    with the German Renewable Energy Sources Act~\cite{noauthor_renewable_2017}.
-    We do, however, expect a lifting of the \SI{52}{\giga\watt} limit on solar PV\@.
+    recommendations by the `coal commission'~\cite{WSB2019}, and a further expansion of renewable electricity generation
+    in line with the German Renewable Energy Sources Act~\cite{noauthor_renewable_2017}.
+    We do, however, expect the \SI{52}{\giga\watt} solar PV limit to be lifted.
+    The corresponding capacity assumptions are displayed in table~\ref{tab:instcaps}.
 
@@
             Solar PV & $1.1$ & $73.0$ \\
-            Run-of-river Hydro & $6.7$ & $4.5$ \\
+            Run-of-river Hydro & $6.9$ & $4.5$ \\
             Biomass & $0.725$ & $8.4$ \\
@@
     We represent 27\todo{compare with table!} dispatchable energy conversion and storage technologies that are expected
-    to be operational by 2030.
-    In addition to exogenously set, initally installed capacities (see~\ref{tab:instcaps}), the model can endogenously
-    add further generation capacities that are compatible with stated policy objectives.
+    to be operated in 2030.
+    In addition to given, initally installed capacities (see~\ref{tab:instcaps}), the model can endogenously add further
+    generation capacities that are compatible with stated policy objectives.
     All technologies can also be decommissioned, so that we adopt a long-run perspective on the power system.
+    Parameters of admissible technologies are summarized in Table~\ref{itm:techcost}.
 
-    Assumed technology parameters are summarized in Table~\ref{itm:techcost}.
-
-    \begin{table*}[h]
+    \begin{table*}[h!t]
         \centering
@@
 
-    Hourly generation from non-dispatchable technologies solar photovoltaics, wind turbines and run-of-river hydro is
-    assumed to follow hourly profiles as observed in 2016.
-    \todo[inline]{more details on derivation of load profiles for Austria, source for Germany, scaling}
+    Generation from non-dispatchable technologies solar photovoltaics, wind turbines and run-of-river hydro is assumed
+    to follow hourly generation profiles as observed in 2016.
+    German solar and wind power profiles are sourced from Open Power System Data~\citep{opsd2019}.
+    For Austria, solar and wind power generation profiles were derived by dividing solar and wind power generation
+    timeseries reported by Open Power System Data~\cite{opsd2019} by installed capacities published by \cite{Biermayr}.
+    Similarly, generation profiles for run-of-river hydro power were derived based on generation data reported by
+    \cite{ENTSOE2020}.
 
-    Similarly, hourly inflows into hydro reservoirs (including pumped storage reservoirs) are inferred from weekly
-    reservoir levels in combination with hourly generation and pumping time series published by~\cite{ENTSO-E2020}.
-    However, all storage technologies (including hydro reservoir and pumped storage plant) are charged and dispatched
-    endogenously.
-    \todo[inline]{improve description of methodology for inflow approximation}
+    Data on inflows into Austrian hydro reservoirs is not publicly available.
+    However, ENTSO-E's transparency platform~\cite{ENTSOE2020} reports data on hydro storage filling levels, electricity
+    generation from hydro reservoirs, and pumping by pumped hydro storages.
+    Based on this data, we have approximated inflows as the part of change in (weekly) hydro storage fill levels not
+    explained by (upsampled hourly) pumping or generation.
+    To arrive at hourly inflows, we have interpolated our weekly estimates by piecewise cubic hermite interpolation.
+    Although inflows to hydro reservoirs are given exogenous, all storage technologies (including hydro reservoir and
+    pumped storage plant) are charged and dispatched endogenously.
 
     \subsection{Energy demand}\label{subsec:energy-demand}
-    According to projections of the Austrian government, additional renewable electricity generation of
-    \SI{27.0}{\tera\watt\hour} will suffice to meet policy goals by 2030 \citep{Regierungsprogramm2020}.
-    However, official documents do neither provide any information about expected electricity consumption in 2030,
-    nor about expected prices for \COO emissions.
-    Absent these information, we chose to set 2030 electricity demand to a level that results in \SI{27.0}{\tera\watt\hour}
-    additional electricity generation from renewable sources at \COO prices of 60 \EUR per tonne in a setting without
-    restrictions on renewables deployment.
-    The corresponding annual electricity consumption is \SI{86.0}{\tera\watt\hour}, \SI{24}{\tera\watt\hour} above
-    electricity end use observed in our base year 2016.
-    The difference in annual electricity generation stems from
-
-    Please note that the demand level consistent with generating \SI{27.0}{\tera\watt\hour} additional renewable
-    electricity depends on the price of \COO.
-    To keep scenarios comparable, however, we keep annual electricity consumption constant at \SI{86.0}{\tera\watt\hour}
-    throughout all scenarios.
-    In effect, added RET capacities fluctuate with the \COO price.
+    Regarding future electricity demand, government plans reveal little detail.
+    According to official documents, additional renewable electricity generation of \SI{27.0}{\tera\watt\hour} will be
+    required to meet electricity consummption entirely from renewable sources.\citep{Regierungsprogramm2020}.
+    However, industry self generation and consumption, amounting to \SI{4.75}{\tera\watt\hour} in the base
+    year~\citep{StatistikAustria2020}, is exempt~\citep{mission2030}.
+    Knowing that electricity generation from renewable sources amounted to
+    \SI{51.1}{\tera\watt\hour}~\citep{StatistikAustria2020}, we can infer expected total electricity consumption of
+    \SI{82.85}{\tera\watt\hour} by 2030.\footnote{Balancing energy is also exempt from the renewables obligation.
+    As we do not model balancing energy supply, we also do not account for this in the calculation of total demand.}
+    On a side-note, expected demand in 2030 is \SI{10.45}{\tera\watt\hour} above levels observed in our base year, even
+    though electrification of further sectors, most notably transport, is envisioned.
+    An attempt to judge the feasibility of these plans is beyond the scope of this paper.
 
@@
     In consequence, we scale reported hourly load to match annual electricity consumption published in national energy
-    balances \citep{AGEB}.
+    balances~\citep{AGEB}.
     \todo[inline]{cross-check code}
 
-    Annual heat consumption is derived from energy balances for Austria and Germany, respectively \citep{AGEB, OeSTAT2019}.
+    Annual heat consumption is derived from energy balances for Austria and Germany, respectively~\citep{AGEB, OeSTAT2019}.
     Subsequently, annual heat consumption is broken down to hourly heat consumption on the basis of standard natural gas
-    load profiles for space heating in the residential and commercial sectors \citep{Almbauer2008}.
+    load profiles for space heating in the residential and commercial sectors~\citep{Almbauer2008}.
     These load profiles make use of daily average temperatures to calculate daily heat demand.
-    We extract spatially resolved temperature data from ERA-5 climate data sets \citep{era-5} and compute a capacity-weighted
-    average of temperatures at locations of combined heat and power (CHP) generation units.
+    We extract spatially resolved temperature data from ERA-5 climate data sets~\citep{era-5} and compute a
+    capacity-weighted average of temperatures at locations of combined heat and power (CHP) generation units.
     Daily heat demand based on these weighted average temperatures is then broken down to hourly consumption on the
     basis of standardized factors accounting for weekday and time-of-day effects.
-    Descriptive statistics of electricity and heat consumption are provided in table \ref{tab:time-series}.
+    Descriptive statistics of electricity and heat consumption are provided in table~\ref{tab:time-series}.
 
-    \begin{table*}[ht]
+    \begin{table*}[h!t]
         \centering
@@
             \midrule
-            Electricity load & AT & GW & 7.15 & 7.14 & 10.5 & 4.21 & nose      \\
-            Electricity load & DE & GW & 60.2 & 59.6 & 82.8 & 34.6 & nose      \\
-            District heating load & AT & GW & 2.56 & 2.43 & 5.77 & 0.91 & bladder   \\
-            District heating load & DE & GW & 14.6 & 14.1 & 26.0 & 8.9 & bladder   \\
-            Wind onshore profile & AT & \%    & 0.229 & 0.142 & 0.929 & 0.000 &           \\
-            Wind onshore profile & DE & \%    & 0.176 & 0.133 & 0.764 & 0.003 &           \\
-            Wind offshore profile & DE & \%    & 0.371 & 0.327 & 0.900 & 0.000 &           \\
-            Solar PV profile & AT & \%    & 0.098 & 0.012 & 0.570 & 0.000 &           \\
-            Solar PV profile & DE & \%    & 0.099 & 0.003 & 0.661 & 0.000 &           \\
-            Run-of-river profile & AT & \%    & 0.564 & 0.538 & 0.971 & 0.202 &           \\
-            Run-of-river profile & DE & \%    & 0.437 & 0.420 & 0.637 & 0.208 &           \\
+            Electricity load & AT & GW & 7.15 & 7.14 & 10.5 & 4.21 & \cite{opsd2019}      \\
+            Electricity load & DE & GW & 60.2 & 59.6 & 82.8 & 34.6 & \cite{opsd2019}      \\
+            District heating load & AT & GW & 2.56 & 2.43 & 5.77 & 0.91 & own clculation based on \cite{era-5, OeSTAT}\\
+            District heating load & DE & GW & 14.6 & 14.1 & 26.0 & 8.9 & own calculation based on \cite{era-5, AGEB}  \\
+            Wind onshore profile & AT & \%    & 0.229 & 0.142 & 0.929 & 0.000 & \cite{opsd2019}          \\
+            Wind onshore profile & DE & \%    & 0.176 & 0.133 & 0.764 & 0.003 & \cite{opsd2019}          \\
+            Wind offshore profile & DE & \%    & 0.371 & 0.327 & 0.900 & 0.000 & \cite{opsd2019}          \\
+            Solar PV profile & AT & \%    & 0.098 & 0.012 & 0.570 & 0.000 & own calculation based on \cite{}          \\
+            Solar PV profile & DE & \%    & 0.099 & 0.003 & 0.661 & 0.000 & \cite{opsd2019}          \\
+            Run-of-river profile & AT & \%    & 0.564 & 0.538 & 0.971 & 0.202 & own calculation based on \cite{ENTSOE2020} \\
+            Run-of-river profile & DE & \%    & 0.437 & 0.420 & 0.637 & 0.208 & \cite{ENTSOE2020} \\
             \bottomrule
@@
 
-    Electricity transmission between the modelled market areas is constrained by net transfer capacities reported by
-    ENTSO-E for the year 20XX.
-    In addition, the model can expand transmission capacities at a cost of XXX EUR per MW and km.
+    Electricity transmission between the modelled market areas is limited to \SI{4.9}{\giga\watt}, in line with the
+    introduction of a congestion management scheme by German and Austrian authorities in 2018.
+    The congestion management scheme was put in place in response to so-called loop-flows from wind-rich Northern
+    Germany to demand centres in Southern Germany and Austria via Germany's eastern neighbors. \todo[inline]{we would
+    allow for an expansion of transmission capacities. This, however, does not happen. Shall we say we do not allow
+    transmission expansion (same outcome)? Or should we say we allow for expansio, but at highly unrealistic cost, as
+    the congestion scheme is artificial and removing it is a matter of a couple pieces of paper? Conduct a sensitivity run
+    with 10 GW transmission capacity (which is about physical capacity in place)?}
 
@@
     We convert these prices to \EUR/MWh based on the fuel's energy content and market exchange rates obtained from the
-    European Central Bank \citep{ECB}.
-    Finally, we resample prices to hourly frequency using piecewis cubic hermite interpolation.
-    The price of lignite is assumed at $1.50 EUR / MWh_{th}$, as reported by \cite{OekoInstitut2017}.
-    Biomass is assumed to be subsidized such that the cost of fuel is reduced to 6.5 $EUR / MWh$, which results in
-    continuous operation of biomass plant in our scenarios.\footnote{Subsidies are not included in system cost, as the
-    required level is uncertain. The total sum of subsidies is, however, constant across scenarios as biomass capacity
-    can not be expanded.}
+    European Central Bank~\citep{ECB}.
+    Finally, we resample prices to hourly frequency using piecewise cubic hermite interpolation.
+    Due to its low energy density lignite is not transported over large distances and consequently also not traded on
+    markets.
+    Instead, lignite-fired power plants are situated in proximity to lignite mines.
+    According to estimates from \cite{OekoInstitut2017}, the price of lignite in Germany is close to 1.50 \EUR per
+    MWh$_{th}$.
+    Biomass-fired power plant run on a wide variety of solid and gaseous fuels, some of which are marketed.
+    However, continued operation of biomass-fired plant typically relies on sufficient subsidies\todo{reference}.
+    As a first-order approximation to complex subsidy schemes, we assume subsidized fuel cost of 6.50 \EUR per
+    MWh$_{th}$.
+    In consequence, biomass-fired plant are operating at or close to capacity throughout the scenarios considered.
+    \footnote{Subsidies are not included in system cost, as the required level is uncertain.
+    The total sum of subsidies is, however, constant across scenarios as biomass capacity can not be expanded.}
 
-    \begin{table*}[h!]
+    \begin{table*}[h!t]
         \centering
@@
     \end{table*}
-    All price as in our baseline year 2016.
-    Transport costs apply to different regions.
 
-    The carbon intensity of fossil fuels is displayed in Table~\ref{table:carbon-intensity}, and is based on an
+    The assumed carbon intensity of fossil fuels is displayed in Table~\ref{table:carbon-intensity}, and is based on an
     analysis by the German environmental agency~\citep{Juhrich2016}.
 
-    \begin{table}[h!]
+    \begin{table}[h!t]
         \centering
@@
     wind turbines and solar PV is unconstrained.
+    Given the policy goal of generating \SI{21}{\tera\watt\hour} of electricity either from wind or solar power (with
+    the remainder coming from hydro power and biomass), required capacity expansion amounts to \SI{10.4}{\giga\watt}
+    wind power or \SI{24.1}{\giga\watt} of solar power if the policy is to be fulfilled with one technology only.
 
-    In our baseline scenario, we assume that about two thirds of solar PV installations are mounted on roof tops, while
-    the remainder is realized as open-space solar PV at utility scale.
-    This gives us capacity-weighted average capital cost for solar PV of $630$ \EUR per kW installed.
-    In section~\ref{subsec:sensitivity} we analyze the sensitivity of our results with respect to this assumption.
-    Moreover, we assume pricing of \COO emissions, i.e.\ all otherwise external cost of \COO emissions are
+    In the baseline scenario, we assume capacity-weighted average capital cost of solar PV of $630$ \EUR per kW
+    installed.
+    This corresponds to about 56\% of PV installations being mounted on roof tops, while the remainder (44\%) is
+    realized as open-space solar PV at utility scale.
+    In section~\ref{subsec:sensitivity} we analyze the sensitivity of our results with respect to capital cost of solar
+    PV\@.
+    Moreover, we assume efficient pricing of \COO emissions, i.e.\ all otherwise external cost of \COO emissions are
     internalized through \COO pricing.
-    As future \COO prices are uncertain but impact renewables deployment ~\citep{Brown2020, Kirchner2019}, we conduct
-    our analysis for \COO prices of 30, 60, 90, and 120 \EUR per tonne.
+    In the baseline scenario, we presume \COO prices as low as $30$ \EUR per tonne, as there is no incentive to add
+    carbon-free generation capacities above the policy target.
+    As future \COO prices are uncertain but do impact renewables deployment ~\citep{Brown2020, Kirchner2019}, we
+    extend the analysis to \COO prices of $60$, $90$, and $120$ \EUR per tonne.
 
-    Given our baseline assumptions, the cost-minimizing system set-up that is consistent with the stated policy
-    goal\footnote{Recall that the policy target is to generate at least 90\% of electricity demand from renewable sources, on annual balance}
-    adds \SI{16.5}{\giga\watt} wind power and \SI{0}{\giga\watt} solar PV to initially set capacities in Austria.
-    Overall, electricity generation from both sources totals \SI{38.5}{\tera\watt\hour} and
-    \SI{0.94}{\tera\watt\hour} annually, respectively.
-    For system services and heat generation, \SI{6.8}{\giga\watt} of fossil thermal co-generation capacities remain
-    active, giving rise to \SI{8.9}{} million tons of \COO emissions per year.
+    Given these baseline assumptions, the cost-minimizing system set-up that is consistent with policy objectives, adds
+    \SI{10.4}{\giga\watt} wind power and \SI{0}{\giga\watt} solar PV to initially available capacities in Austria.
+    Further, \SI{2.63}{\giga\watt} of natural gas fired capacity are added, while \SI{0.42}{\giga\watt} are decommissioned.
+    \SI{0.08}{\giga\watt} off oil-fired capacity are decommissioned.
+    In effect, \SI{6.23}{\giga\watt} of fossil thermal generation capacities remain active, giving rise to
+    \SI{10.5}{\mega\tonne} of \COO emissions per year.
     As Austria has ample hydro storage capacity in place, no further storage (e.g. from batteries) is added.
-    However, the addition of \SI{1.5}{\giga\watt} of compression heat pumps allows to use electricity for heat
-    generation and to limit curtailment to \SI{2.4}{\tera\watt\hour\per\year}.
-    System cost in Austria amount to $3.4$ billon \EUR in this scenario.\todo{add co2 emissions}
+    However, the addition of \SI{1.23}{\giga\watt} of compression heat pumps allows to use electricity for heat
+    generation and to limit curtailment to \SI{1.5}{\tera\watt\hour\per\year}.
+    Annual system cost in Austria amount to $2.9$ billon \EUR in this scenario.
+    Net earnings from exports total $0.7$ billion \EUR, so that net system cost amounts to $2.2$ billion \EUR.
 
     \subsection{The opportunity cost of wind power in Austria}\label{subsec:opportunity-cost-wind}
-    With a gradual restriction of wind power potentials, the system must increasingly rely on solar PV to generate the
-    amount of electricity from renewable sources that is required to meet policy goals.
+    %Effect of \COO prices on opportunity cost of wind power:
+    %- lower \COO price means less 'voluntary' RET installed
+    %- more PV means more ancillary capacities installed (?)
+    %- more PV means more fuel used in dispatchable plant
+    %- more PV means more curtailment (?)
+    %- more PV means lower exports -- large dependency on \COO price. At price of 60-120, exports fall as less wind.
+    %For low \COO price exports increase as less wind. -->> Need to look at capacities in Germany!
+    %- more PV means
+    %* PV capacity at full wind substitution is (almost) invariant to \COO price and comes in at 24.11-24.26 GW
+    %* wind onshore capacity is sensitive to \COO price. At 120 \EUR \COO, 17.07 GW wind on added, at 0 \EUR \COO10.43 GW added
+    %* absolutely no heatpump added. why?
+    %
+    %+ No bio added in Germany (proly restriction)
+    %+ Germany adds PV and wind only with \COO price at 120
+    With a gradual restriction of wind power potentials, the system must increasingly rely on solar PV to generate a
+    sufficient amount of electricity from renewable sources.
     Consequently, each \SI{}{\giga\watt} of wind power not installed needs to be replaced by \SI{2.35}{\giga\watt} of
     solar PV\@.
-    Due to the different generation profiles of these technologies, this induces changes in system operation and least
-    cost system configuration.
+    As generation profiles of wind and solar power differ, the substitution of wind power with solar PV brings about
+    changes in system operation, investment, and, consequently, system cost.
 
-    \todo[inline]{describe changes in system operation}
-    Substitution of wind power with solar PV leads to a reduction of electricity exports to Germany.
-    Due to the favorable complementarity of Austrian and German wind, Austria adds wind energy generation above the level
-    required for its own consumption.
+    At baseline \COO prices of $30$ \EUR per tonne,
+
+    At higher \COO prices, wind power installation in Austria increases, as low emission generation is valued higher
+    and because of the favorable complementarity of Austrian wind to (northern) German wind.
+
     As the admissible addition of wind power is reduced, electricity exports to Germany are being scaled back.
@@
     \todo[inline]{insert table with key scenario results}
+    \begin{table}
+        \caption{Key scenario results}
+        \begin{tabulary}{\textwidth}{LLC}
+            \toprule
+            column & column & column \\
+\midrule
+AAA & BBB & CCC \\
+\bottomrule
+\end{tabulary}
+\end{table}
 

### 2020-05-03 projects/asparagus/doc/plots_asparagus.py

@@ -196,7 +196,7 @@ wind_add.index = wind_add.index.droplevel([0, 2])
 
 undisturbed_marginal = undisturbed_base.iloc[::-1].diff().divide(wind_add.iloc[::-1].diff().round(8), axis=0) * -1
 undisturbed_marginal_share = undisturbed_marginal.copy()
-undisturbed_marginal.columns = [f'CO2 Price: {p} €/t' for p in undisturbed_marginal.columns]
+undisturbed_marginal.columns = [rf'CO$_2$ Price: {p} €/t' for p in undisturbed_marginal.columns]
 
 plot_lines(undisturbed_marginal / 1000, os.path.join(FPATH, 'undisturbed_base.pdf'), xlim=[18, 0], ylim=[0, 75],
            xlabel='Added Capacity of Wind [GW]', ylabel='thousand € / MW', color=REFUEL_COLORS)
@@ -297,3 +297,50 @@ axs[1].set_title('Winter week')
 plt.tight_layout()
 plt.savefig(os.path.join(FPATH, 'wind_not_pv.pdf'))
 plt.close()
+
+# %% data analysis
+
+# % response of capacity additions to \Phi
+capa = results.loc[idx['base', :, :, 36715], idx['AT', :]]
+capa.columns = capa.columns.droplevel(0)
+additions = pd.DataFrame()
+additions['lig'] = capa.loc[:, capa.columns.str.contains('add_lig')].sum(axis=1)
+additions['coal'] = capa.loc[:, capa.columns.str.contains('add_coal')].sum(axis=1)
+additions['oil'] = capa.loc[:, capa.columns.str.contains('add_oil')].sum(axis=1)
+additions['ng'] = capa.loc[:, capa.columns.str.contains('add_ng')].sum(
+    axis=1)  # - capa.loc[:, capa.columns.str.contains('add_ng_boiler')]
+additions['bio'] = capa.loc[:, capa.columns.str.contains('add_bio')].sum(axis=1)
+additions['heatpump'] = capa.loc[:, capa.columns.str.contains('add_heatpump')].sum(axis=1)
+additions['pv'] = capa.loc[:, capa.columns.str.contains('add_pv')].sum(axis=1)
+additions['wind'] = capa.loc[:, capa.columns.str.contains('add_wind_on')].sum(axis=1)
+
[…]

### 2020-05-03 projects/asparagus/settings_asparagus.py

@@ -24,7 +24,7 @@ For sensitivity analysis, we also vary the capitalcost of solar PV.
 # campaigns
 dict_campaigns = {
     'base': {
-        'co2_price': range(120, 29, -30),
+        'co2_price': range(120, -1, -30),
         'wind_cap': range(18, -1, -2),
         'pv_cost': [36715, 16715]
     },
@@ -64,8 +64,8 @@ scenario_2030 = {
         'wind_on': [2.6],
         'wind_off': [0],
         'pv': [1.1],
-        'ror': [6.7],
-        'd_power': [62038.515 + 17600]
+        'ror': [6.9],
+        'd_power': [62038.515 + 24000]  # 17600]
     },
     'DE': {
         'bio': [1.3],

### 2020-05-05 src/data/get_entsoe.py
improved get_entsoe.py * structured as function * checks and creates data directory if required

@@ -9,6 +9,7 @@ import config as cfg
 from logging_config import setup_logging
 
 # TODO: check file size and download larger files from ftp
+# TODO: download zipped csv and unpack after download
 
 setup_logging()
 
@@ -26,29 +27,47 @@ credentials = yaml.load(open(os.path.join(cfg.MEDEA_ROOT_DIR, 'credentials.yml')
 USER = credentials['entsoe']['user']
 PWD = credentials['entsoe']['pwd']
 
-# ======================================================================================================================
-# sFTP data download
-# ----------------------------------------------------------------------------------------------------------------------
 
-cnopts = pysftp.CnOpts()
-cnopts.hostkeys = None
-# connect to entsoe server via sFTP
-for cat in CATEGORIES:
-    ENTSOE_DIR = '/TP_export/' + cat
-    LOCAL_DIR = os.path.join(RAW_DATA_DIR, cat)
-    logging.info(f'connecting to {SERVER}')
-    with pysftp.Connection(SERVER, username=USER, password=PWD, cnopts=cnopts) as sftp:
-        sftp.chdir(ENTSOE_DIR)
+# ======================================================================================================================
+# %% sFTP data download
+# ----------------------------------------------------------------------------------------------------------------------
+def get_entsoe(connection_string, user, pwd, category, dir):
+    """
+    downloads dataset from ENTSO-E's transparency data sftp server.
+    contact ENTSO-E to receive login credentials.
+    :param connection_string: url of ENTSO-E transparency server, as of May 1, 2020: 'sftp-transparency.entsoe.eu'
+    :param user: user name required for connecting with sftp server
+    :param pwd: password required for connecting with sftp server
+    :param category: ENTSO-E data category to be downloaded
+    :param dir: directory where downloaded data is saved to. A separate subdirectory is created for each category
+    :return: downloaded dataset(s) in dir
+    """
+    # check if local_dir exists and create if it doesn't
+    local_dir = os.path.join(dir, category)
+    if not os.path.exists(local_dir):
+        os.makedirs(local_dir)
+    #
+    cnopts = pysftp.CnOpts()
+    cnopts.hostkeys = None
+    # connect to entsoe server via sFTP
+    entsoe_dir = f'/TP_export/{category}'
+    logging.info(f'connecting to {connection_string}')
[…]

### 2020-05-05 projects/asparagus/doc/paper_clean.tex
updated asparagus paper_clean.tex

@@
 \DeclareSIUnit{\year}{a}
+\DeclareSIUnit{\wattpeak}{Wp}
 
@@
     derive a measure of the (approximate) marginal impact of wind power expansion (or contraction).
-
-    \subsection{Baseline} \label{subsec:baseline}
-    As a point of reference for our analysis, we determine the least cost power system configuration when addition of
-    wind turbines and solar PV is unconstrained.
-    Given the policy goal of generating \SI{21}{\tera\watt\hour} of electricity either from wind or solar power (with
-    the remainder coming from hydro power and biomass), required capacity expansion amounts to \SI{10.4}{\giga\watt}
-    wind power or \SI{24.1}{\giga\watt} of solar power if the policy is to be fulfilled with one technology only.
-
-    In the baseline scenario, we assume capacity-weighted average capital cost of solar PV of $630$ \EUR per kW
-    installed.
-    This corresponds to about 56\% of PV installations being mounted on roof tops, while the remainder (44\%) is
-    realized as open-space solar PV at utility scale.
-    In section~\ref{subsec:sensitivity} we analyze the sensitivity of our results with respect to capital cost of solar
-    PV\@.
-    Moreover, we assume efficient pricing of \COO emissions, i.e.\ all otherwise external cost of \COO emissions are
-    internalized through \COO pricing.
-    In the baseline scenario, we presume \COO prices as low as $30$ \EUR per tonne, as there is no incentive to add
-    carbon-free generation capacities above the policy target.
-    As future \COO prices are uncertain but do impact renewables deployment ~\citep{Brown2020, Kirchner2019}, we
-    extend the analysis to \COO prices of $60$, $90$, and $120$ \EUR per tonne.
-
-    Given these baseline assumptions, the cost-minimizing system set-up that is consistent with policy objectives, adds
-    \SI{10.4}{\giga\watt} wind power and \SI{0}{\giga\watt} solar PV to initially available capacities in Austria.
-    Further, \SI{2.63}{\giga\watt} of natural gas fired capacity are added, while \SI{0.42}{\giga\watt} are decommissioned.
-    \SI{0.08}{\giga\watt} off oil-fired capacity are decommissioned.
-    In effect, \SI{6.23}{\giga\watt} of fossil thermal generation capacities remain active, giving rise to
-    \SI{10.5}{\mega\tonne} of \COO emissions per year.
-    As Austria has ample hydro storage capacity in place, no further storage (e.g. from batteries) is added.
-    However, the addition of \SI{1.23}{\giga\watt} of compression heat pumps allows to use electricity for heat
-    generation and to limit curtailment to \SI{1.5}{\tera\watt\hour\per\year}.
-    Annual system cost in Austria amount to $2.9$ billon \EUR in this scenario.
-    Net earnings from exports total $0.7$ billion \EUR, so that net system cost amounts to $2.2$ billion \EUR.
-
-    \subsection{The opportunity cost of wind power in Austria}\label{subsec:opportunity-cost-wind}
-    %Effect of \COO prices on opportunity cost of wind power:
-    %- lower \COO price means less 'voluntary' RET installed
-    %- more PV means more ancillary capacities installed (?)
-    %- more PV means more fuel used in dispatchable plant
-    %- more PV means more curtailment (?)
-    %- more PV means lower exports -- large dependency on \COO price. At price of 60-120, exports fall as less wind.
-    %For low \COO price exports increase as less wind. -->> Need to look at capacities in Germany!
-    %- more PV means
-    %* PV capacity at full wind substitution is (almost) invariant to \COO price and comes in at 24.11-24.26 GW
-    %* wind onshore capacity is sensitive to \COO price. At 120 \EUR \COO, 17.07 GW wind on added, at 0 \EUR \COO10.43 GW added
-    %* absolutely no heatpump added. why?
-    %
-    %+ No bio added in Germany (proly restriction)
-    %+ Germany adds PV and wind only with \COO price at 120
-    With a gradual restriction of wind power potentials, the system must increasingly rely on solar PV to generate a
-    sufficient amount of electricity from renewable sources.
-    Consequently, each \SI{}{\giga\watt} of wind power not installed needs to be replaced by \SI{2.35}{\giga\watt} of
-    solar PV\@.
-    As generation profiles of wind and solar power differ, the substitution of wind power with solar PV brings about
-    changes in system operation, investment, and, consequently, system cost.
-
-    At baseline \COO prices of $30$ \EUR per tonne,
-
-    At higher \COO prices, wind power installation in Austria increases, as low emission generation is valued higher
-    and because of the favorable complementarity of Austrian wind to (northern) German wind.
-
-    As the admissible addition of wind power is reduced, electricity exports to Germany are being scaled back.
-    This is reinforced by the similarities in generation profiles of solar PV across both countries.
-
-    Due to the concentration of solar PV generation in time (both on a daily and on a seasonal scale), installing solar
-    PV instead of wind power leads to an increase in dispatchable thermal generation.
-    As thermal generation in Austria is predominantly fossil thermal, this induces an increase in \COO emissions due
-    to the increased use of PV\@.
-
-    Curtailment in Austria is declining up to
-
-    \begin{figure*}
-        \centering
-        \includegraphics[width=0.8 \textwidth]{./figures/sysops.pdf}
-        \caption{Opportunity cost of wind power assuming PV overnight cost of 630 EUR/kWp}
-        \label{figure:system-operation-base}
-    \end{figure*}
-
-    \todo[inline]{describe changes in system configuration}
-
-    \todo[inline]{describe marginal change in system cost due to less wind deployed}
-
-    \todo[inline]{insert table with key scenario results}
-    \begin{table}
-        \caption{Key scenario results}
-        \begin{tabulary}{\textwidth}{LLC}
-            \toprule
-            column & column & column \\
-\midrule
-AAA & BBB & CCC \\
-\bottomrule
-\end{tabulary}
-\end{table}
-
-\begin{figure*}
+    A detailed discussion of model results is provided in \ref{sec:detailedresults}.
+
+    \begin{figure*}[h!t]
         \centering
@@
 
+    With a gradual restriction of deployable wind power capacities, we find an opportunity costs of wind power
+    to increase by about $3\,900$ \EUR for each \SI{}{\mega\watt} wind power substituted by solar PV across all scenarios
+    that we studied.
+    At PV overnight cost of $630$ \EUR per MWp (corresponding to a mix of 55\% rooftop and 45\% open space
+    installations), the opportunity cost of the last \SI{}{\mega\watt} wind power replaced reaches close to $62\,500$ \EUR.
+    At a wind power capacity consistent with the policy goal of generating \SI{21}{\tera\watt\hour} of electricity
+    from renewable sources, opportunity cost for wind power are in close to $20\,000$ \EUR per MW, with a notable
+    exception for \COO prices of $60$ \EUR per tonne.
+    In this case, opportunity cost of wind power are estimated at close to $10\,000$ \EUR per MW, about half the size of
+    opportunity cost estimated for all other \COO price scenarios.
+    \todo[inline]{investigate why this happens}
+    \todo[inline]{add brief explanation for why opportunity cost increase -- more thermal dispatch \& capacity etc.}
+
     \subsection{Sensitivity Analysis} \label{subsec:sensitivity}
-A major factor of influence for our results are the assumed capital cost of wind turbines and solar PV.
-Thus, we complement our baseline scenario with a sensitivity analysis on investment cost.
-For this purpose, we vary the investment cost of solar PV between our baseline assumption of 630 EUR/kW and 275
-EUR/kW that \cite{Vartiainen2019} project for utility-scale solar PV in the year 2030.
-
-\begin{figure*}
+    A major factor of influence for our results are the assumed capital cost of wind turbines and solar PV\@.
+    Therefore, we complement our baseline scenario with a sensitivity analysis on overnight investment cost of solar
+    photovoltaics.
+    We gradually lower the capital cost of solar PV from our baseline assumption of \SI{630}[\$]{\per\kilo\wattpeak}
+    (corresponding to \SI{55}{\percent} of rooftop and \SI{45}{\percent} open space capacity) to
+    \EUR~\SI{275}{\per\kilo\wattpeak}, which Vartiainen et al.~\cite{Vartiainen2019} project for open-space,
+    utility-scale solar PV in the year 2030.
+
+    \begin{figure*}[h!t]
         \centering
@@
 
-%    Further uncertainty surrounds
-%    •	Storage prices
-%    •	Weather / full-load hours
-%    •	LOOK AT WHAT HAPPENS WHEN GERMANY BUILDS MORE PV THAN ANTICIPATED (particularly in the low pv-cost scenario)
+    Under these assumptions, the impact of \COO prices on the opportunity cost of wind power increases.
+    If all additional PV is realized as open-space, utility-scale installation (i.e.\ PV capital cost of
+    \EUR~\SI{275}{\per\kilo\wattpeak}), the optimal deployment of wind power declines and approaches zero at \COO prices
+    of \EUR~\SI{30}{\per\tonne} or below.
+    In consequence, we can not determine opportunity cost of wind power given that both \COO prices and capital cost of
+    solar PV are very low.
+    For higher \COO prices, the optimal deployment of wind power is strictly positive, even at low capital cost of solar
+    PV and we find opportunity cost to increase by about $2\,550$ \EUR for each MW wind power foregone.
+    The last MW of wind power substituted with solar PV comes at opportunity cost between $20\,000$ and $56\,000$ \EUR,
+    depending on the \COO price prevailing.
 
@@
     Alas, uncertainty regarding future capital cost is high.
-\todo[inline]{WRITE SOMETHING ON THE IMPACT OF DIFFERENT COST ASSUMPTIONS ON TOTAL SYSTEM COST.}
+    \todo[inline]{write about impact of cost assumptions on total system cost}
 
@@
     Typically, it will be cheapest to add (at least some) fossil-fueled backup plants, even at \COO prices of EUR 70 per ton.
-\todo[inline]{HERE, WE COULD INVESTIGATE AT WHICH CO2-PRICE FOSSIL BACKUP VANISHES.}
-
-Moreover, the observed increase in producer surplus in line with higher CO2 penetration is robust to capital cost uncertainty.
-Higher levels of PV penetration require more efforts for system integration.
-\todo[inline]{THIS CAN'T BE THE FULL EXPLANATION AS PROD SURPLUS DECREASES AT VERY HIGH LEVELS OF PV}
+    \todo[inline]{do we know if there is a co2-price at which fossil backup vanishes? -> at 120 eur / t fossil backup is about halved.}
 
@@
 
+    \section{Detailed Description of scenario results} \label{sec:detailedresults}
+
+    \subsection{Baseline} \label{subsec:baseline}
+    As a point of reference for our analysis, we determine the least cost power system configuration when addition of
+    wind turbines and solar PV is unconstrained.
+    Given the policy goal of generating \SI{21}{\tera\watt\hour} of electricity either from wind or solar power (with
+    the remainder coming from hydro power and biomass), required capacity expansion amounts to \SI{10.4}{\giga\watt}
+    wind power or \SI{24.1}{\giga\watt} of solar power if the policy is to be fulfilled with one technology only.
+
+    In the baseline scenario, we assume capacity-weighted average capital cost of solar PV of $630$ \EUR per kW
+    installed.
+    This corresponds to about 56\% of PV installations being mounted on roof tops, while the remainder (44\%) is
+    realized as open-space solar PV at utility scale.
+    In section~\ref{subsec:sensitivity} we analyze the sensitivity of our results with respect to capital cost of solar
+    PV\@.
+    Moreover, we assume efficient pricing of \COO emissions, i.e.\ all otherwise external cost of \COO emissions are
+    internalized through \COO pricing.
+
+    Future \COO prices impact renewables deployment ~\citep{Brown2020, Kirchner2019} and affect the results of our
+    analysis.
+    As we can not foresee \COO prices in the year 2030, we have conducted our analysis for \COO prices in the range of
+    $0$ and $120$ \EUR per tonne.
+    In this context, it is worth noting that within the framework of our analysis a \COO price of $30$ \EUR per tonne
+    would be consistent with RET expansion goals stated by the Austrian government.
+    At \COO price above $30$ \EUR per tonne, there is incentive to invest in \COO emission-free generation beyond
+    government targets.
+
+    Given these baseline assumptions, the cost-minimizing system set-up that is consistent with policy objectives, adds
+    \SI{10.4}{\giga\watt} wind power and \SI{0}{\giga\watt} solar PV to initially available capacities in Austria.
+    Further, \SI{2.63}{\giga\watt} of natural gas fired capacity are added, while \SI{0.42}{\giga\watt} are decommissioned.
+    \SI{0.08}{\giga\watt} off oil-fired capacity are decommissioned.
+    In effect, \SI{6.23}{\giga\watt} of fossil thermal generation capacities remain active, giving rise to
+    \SI{10.5}{\mega\tonne} of \COO emissions per year.
+    As Austria has ample hydro storage capacity in place, no further storage (e.g. from batteries) is added.
+    However, the addition of \SI{1.23}{\giga\watt} of compression heat pumps allows to use electricity for heat
+    generation and to limit curtailment to \SI{1.5}{\tera\watt\hour\per\year}.
+    Annual system cost in Austria amount to $2.9$ billon \EUR in this scenario.
+    Net earnings from exports total $0.7$ billion \EUR, so that net system cost amounts to $2.2$ billion \EUR.
+
+    \subsection{The opportunity cost of wind power in Austria}\label{subsec:opportunity-cost-wind}
+    %Effect of \COO prices on opportunity cost of wind power:
+    %- lower \COO price means less 'voluntary' RET installed
+    %- more PV means more ancillary capacities installed (?)
+    %- more PV means more fuel used in dispatchable plant
+    %- more PV means more curtailment (?)
+    %- more PV means lower exports -- large dependency on \COO price. At price of 60-120, exports fall as less wind.
+    %For low \COO price exports increase as less wind. -->> Need to look at capacities in Germany!
+    %- more PV means
+    %* PV capacity at full wind substitution is (almost) invariant to \COO price and comes in at 24.11-24.26 GW
+    %* wind onshore capacity is sensitive to \COO price. At 120 \EUR \COO, 17.07 GW wind on added, at 0 \EUR \COO10.43 GW added
+    %* absolutely no heatpump added. why?
+    %
+    %+ No bio added in Germany (proly restriction)
+    %+ Germany adds PV and wind only with \COO price at 120
+    With a gradual restriction of wind power potentials, the system must increasingly rely on solar PV to generate a
+    sufficient amount of electricity from renewable sources.
+    Consequently, each \SI{}{\giga\watt} of wind power not installed needs to be replaced by \SI{2.35}{\giga\watt} of
+    solar PV\@.
+    As generation profiles of wind and solar power differ, the substitution of wind power with solar PV brings about
+    changes in system operation, investment, and, consequently, system cost.
+
+    \subsubsection{Changes in system configuration}
+    \todo[inline]{describe changes in system configuration}
+    Consistent across all \COO price scenarios, the few existing, small oil-fired units in Austria are shut down.
+    Conversely, natural gas-fired co-generation capacities are expanded.
+    Additions of natural-gas fired capacities are lower at high \COO prices and at high wind power installations.
+    However, there are huge differences in fossil capacity additions between \COO prices of $60$ \EUR per ton and $90$
+    \EUR per ton.
+    In the former case, there is very little difference to capacity addition at zero \COO price, while in the latter case
+    fossil capacity additions are more than halved at high wind shares.
+    Further heat generation is taking place in newly added compression heat pumps, of which between \SI{1.2}{\giga\watt}
+    and \SI{2.9}{\giga\watt} are active in Austria.
+    Generally, higher \COO prices are leading to higher installed capacities of heat pumps.
+    Similarly, optimally installed heatpump capacities tends to fall as wind power is substituted with solar PV.
+    This effect is more pronounced at high \COO prices and lower in magnitude than the level effect of \COO prices.
+
+    \subsubsection{Changes in system operation}
+    Curtailment of renewable electricity generation in Austria is the highest when \COO prices are high and the system
+    relies on large shares of wind power.
+    Curtailment reaches its minimum at intermediate \COO prices of $60$ \EUR per tonne and wind power installations of
+    \SI{8}{\giga\watt}.
+    At \COO prices of $30$ \EUR per tonne or lower curtailment tends to increase as wind power is substituted by
+    photovoltaics.
+
+    At baseline \COO prices of $30$ \EUR per tonne,
+
+    At higher \COO prices, wind power installation in Austria increases, as low emission generation is valued higher
+    and because of the favorable complementarity of Austrian wind to (northern) German wind.
+
+    As the admissible addition of wind power is reduced, electricity exports to Germany are being scaled back.
+    This is reinforced by the similarities in generation profiles of solar PV across both countries.
+
+    Due to the concentration of solar PV generation in time (both on a daily and on a seasonal scale), installing solar
+    PV instead of wind power leads to an increase in dispatchable thermal generation.
+    As thermal generation in Austria is predominantly fossil thermal, this induces an increase in \COO emissions due
+    to the increased use of PV\@.
+
+    Curtailment in Austria is declining up to
+
+    \begin{figure*}
+        \centering
+        \includegraphics[width=0.8 \textwidth]{./figures/sysops.pdf}
+        \caption{Opportunity cost of wind power assuming PV overnight cost of 630 EUR/kWp}
+        \label{figure:system-operation-base}
+    \end{figure*}
+
+    \todo[inline]{describe marginal change in system cost due to less wind deployed}
+    \todo[inline]{insert table with key scenario results}
+
+    \begin{table}
+        \centering
+        \caption{Key scenario results}
+        \begin{tabulary}{\textwidth}{LLC}
+            \toprule
+            column & column & column \\
+\midrule
+AAA & BBB & CCC \\
+\bottomrule
+\end{tabulary}
+\end{table}
+
 \end{document}

### 2020-05-05 projects/asparagus/doc/plots_asparagus.py

@@ -9,16 +9,17 @@ import config as cfg
 
 # %% ----- ----- ----- ----- settings ----- ----- ----- -----
 idx = pd.IndexSlice
-PRICE_CO2 = 30
+PRICE_CO2 = 60
 REFUEL_COLORS = ['#c72321', '#0d8085', '#f0c220', '#595959', '#3b68f9', '#7794dd']
 RES_COLORS = ['#d69602', '#ffd53d', '#3758ba', '#7794dd']
 RES_COLORS3 = ['#d69602', '#e5b710', '#ffd53d', '#3758ba', '#3b68f9', '#7794dd']
+RES_COLORS2 = ['#ffd53d', '#3758ba']
 
 ANNUITY_FACTOR = 0.05827816
 FLH_PV = 857.4938
 FLH_WINDON = 2015.0359
 
-RPATH = os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', 'asparagus', 'results', 'results.csv')
+RPATH = os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', 'asparagus', 'results', 'results_200503.csv')
 FPATH = os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', 'asparagus', 'doc', 'figures')
 
 if not os.path.exists(FPATH):
@@ -73,7 +74,7 @@ def plot_subn(df, fname, width=2, xlim=None, ylim=None, xlabel=None, ylabel=None
 
         axis[a, b].grid()
         axis[a, b].set_title(df.columns[col])
-    plt.rcParams.update({'font.size': 16})
+    # plt.rcParams.update({'font.size': 16})
     plt.tight_layout()
     plt.savefig(fname)
     plt.close()
@@ -120,7 +121,60 @@ def plot_lines(df, fname, xlim=None, ylim=None, xlabel=None, ylabel=None, color=
     axis.set_ylim(ylim)
     axis.set_xlim(xlim)
     axis.grid()
-    plt.rcParams.update({'font.size': 16})
+    # plt.rcParams.update({'font.size': 16})
+    plt.tight_layout()
+    plt.savefig(fname)
+    plt.close()
+
+
+def plot_sublines(df, fname, width=2, midx_level=0, subtitle=True, xlim=None, ylim=None, xlabel=None, ylabel=None,
+                  color=None):
+    """
+    plots multiindex pd.DataFrame to multiple line-subplots
+    :param df: Pandas MultiIndex DataFrame
[…]

### 2020-05-13 projects/asparagus/doc/paper_clean.tex
uodate of project asparagus

@@
     \section{Introduction} \label{sec:introduction}
+    Limiting global warming to `well below 2^{\circ}C compared to pre-industrial levels', as agreed on in the `Paris
+    agreement' requires rapid and substantial reductions in greenhouse gas emissions \citep{IPCC2019}.
+    The European Union has committed itself to xxx.
+    Within this framework, by 2030 Austria aims to generate 100\% of its electricity consumption from domestic renewable
+    sources on annual balance\footnote{Industry own use and ancillary service provision is exempt.
+    Together, these accounted for approximately 10\% of consumption in 2016.}.
+    According to government estimates, additional electricity generation of \SI{27}{\tera\watt\hour} from renewable
+    sources is required to meet policy goals.
+    Out of this total increase in renewable electricity generation, \SI{1}{\tera\watt\hour} is thought to be sourced
+    from biomass, and \SI{5}{\tera\watt\hour} from new hydropower generation, according to the government's climate and
+    energy strategy \citep{BMNT2018}.
+    The comparatively small projected increase in electricity generation from these sources results from the perception
+    of biomass being ecologically \citep{Erb2018} and economically unsustainable, while Austria's vast hydropower
+    potentials were already largely exploited \citep{Walder2014} even before the projected increase in hydro generation
+    by \SI{5}{\tera\watt\hour}.
+    As Austrians banned the use of nuclear power in a popular referendum in 1978 \citep{Pelinka1983}, wind and solar
+    power are the only remaining options for a large-scale expansion of renewable power generation.
+
+%    Erneuerbare sind notwendig
+%    -> Wind und PV die einzigen Technologien mit großem Potential
+%    -> Wind hat soziale Akzeptanzprobleme (bzw. negative welfare impacts)
+%    -> Wind kann aber möglicherweise auch positive Effekte auf die Systemkosten haben
+%    -> Keine Windkraft auszubauen und damit die Landschaft zu schützen hat also Opportunitätskosten
+%    -> Wie hoch sind diese Opportunitätskosten?
+%    -> Das ist interessant, weil man damit auch ausrechnen kann, wieviel man im Prinzip kompensieren könnte.
+%    Fertig.
+
     Future, low carbon energy systems are likely to largely rely on electricity from renewable energy technologies

### 2020-05-13 projects/asparagus/doc/plots_asparagus.py

@@ -1,15 +1,15 @@
-import itertools
+# %% imports
 import os
-
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
-
 import config as cfg
+from src.tools.visualize_data import plot_lines, plot_sublines, plot_subn
 
 # %% ----- ----- ----- ----- settings ----- ----- ----- -----
 idx = pd.IndexSlice
 PRICE_CO2 = 60
+# TODO: define color palette
 REFUEL_COLORS = ['#c72321', '#0d8085', '#f0c220', '#595959', '#3b68f9', '#7794dd']
 RES_COLORS = ['#d69602', '#ffd53d', '#3758ba', '#7794dd']
 RES_COLORS3 = ['#d69602', '#e5b710', '#ffd53d', '#3758ba', '#3b68f9', '#7794dd']
@@ -25,164 +25,43 @@ FPATH = os.path.join(cfg.MEDEA_ROOT_DIR, 'projects', 'asparagus', 'doc', 'figure
 if not os.path.exists(FPATH):
     os.makedirs(FPATH)
 
-
-# %% define plotting functions
-def plot_subn(df, fname, width=2, xlim=None, ylim=None, xlabel=None, ylabel=None, color=None):
-    """
-    plots each column of a pandas dataframe on a seperate subplot. Uses column names as subplot-titles.
-    :param df:
-    :param fname:
-    :param width:
-    :param xlim:
-    :param ylim:
-    :param xlabel:
-    :param ylabel:
-    :param color:
-    :return:
-    """
-    if isinstance(df, pd.DataFrame):
-        ncols = len(df.columns)
-    else:
-        raise TypeError
-
-    sub_length = np.ceil(ncols / width).astype(int)
-    sub_boxes = list(itertools.product(range(0, width), range(0, sub_length)))
-
-    figure, axis = plt.subplots(width, sub_length, figsize=(8, 5))
-    for col in range(0, ncols):
-        a = sub_boxes[col][0]
-        b = sub_boxes[col][1]
-
-        if color is not None:
-            axis[a, b].plot(df.iloc[:, col], linewidth=2, color=color[col])
-        else:
-            axis[a, b].plot(df.iloc[:, col], linewidth=2)
-
-        if xlabel is not None:
-            axis[a, b].set_xlabel(xlabel)
-
-        if (ylabel is not None) & (len(ylabel) == 1):
-            axis[a, b].set_ylabel(*ylabel)
-        elif (ylabel is not None) & (len(ylabel) == ncols):
[…]

### 2020-06-19 projects/asparagus/doc/asparagus.tex
uodate of project asparagus -- preprint status

\documentclass[review, 3p, times, 12pt]{elsarticle} %final, twocolumn, 5p, 12pt, authoryear, twocolumn, 5p,

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

% improved float environment
\usepackage{float}
% improved underlines (unterstreichungen)
\usepackage[normalem]{ulem}

% plotting
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.16}

% tables
\usepackage{threeparttable}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{longtable}
\usepackage{tabulary}
\usepackage{threeparttablex}
\usepackage{tabularx}
\usepackage{ltablex}
\usepackage{multirow}

\newcolumntype{M}{>{\raggedright\arraybackslash}X} % Defines new column type with variable length, similar to X in tabularx, but is left aligned rather than justified

%\usepackage{ctable}
\usepackage{makecell}

% SI units
\usepackage{siunitx}
\DeclareSIUnit{\thermal}{th}
\DeclareSIUnit{\electric}{el}
\DeclareSIUnit{\year}{a}
\DeclareSIUnit{\wattpeak}{Wp}
\DeclareSIUnit{\million}{million}

% Euro symbol
\usepackage[gen]{eurosym}

% citations
\usepackage{natbib}

% to-do notes
\setlength{\marginparwidth}{2cm}
\usepackage[draft]{todonotes}

% graphics and figures
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\COO}{\ensuremath{\mathrm{CO_2}} }

% neater date
\usepackage{etoolbox}
\patchcmd{\MaketitleBox}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}%
\patchcmd{\pprintMaketitle}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}%

\usepackage[colorlinks=true, citecolor=blue, linkcolor=blue, filecolor=blue,urlcolor=blue,pdfauthor=author]{hyperref}

% remove elsevier footer
\makeatletter
\def\ps@pprintTitle{%
\let\@oddhead\@empty
\let\@evenhead\@empty
\def\@oddfoot{}%
\let\@evenfoot\@oddfoot}
\makeatother

\begin{document}
\begin{frontmatter}
\title{The Cost of Undisturbed Landscapes}
\author[1]{Sebastian Wehrle\corref{cor1}}
\ead{sebastian.wehrle@boku.ac.at}
\author[1]{Johannes Schmidt}
\ead{johannes.schmidt@boku.ac.at}
\author[1]{Christian Mikovits}
\ead{christian.mikovits@boku.ac.at}
\cortext[cor1]{Corresponding author}
\address[1]{Institute for Sustainable Economic Development, University of Natural Resources and Life Sciences,
Feistmantelstrasse 4, 1180 Vienna, Austria}

\begin{abstract}
By 2030 Austria aims to meet \SI{100}{\percent} of its electricity demand from domestic renewable sources, predominantly from wind and solar energy.
While wind power reduces \COO emissions, it is also connected to negative impacts at the local level, such as interference with landscape aesthetics.
Yet, wind power comes at lower system integration cost than solar power, so that it effectively reduces system cost.
We quantify the opportunity cost of replacing wind turbines with solar power, using the power system model \emph{medea}.
Our findings suggest that these cost of undisturbed landscapes are considerable, particularly when PV is not entirely rolled out as utility-scale, open-space installations.
The opportunity cost is likely high enough to allow for significant compensation of the ones affected by local wind turbine externalities.
\end{abstract}

\begin{keyword}
externalities \sep wind power\sep system design\sep policy
\JEL L98 \sep Q42 \sep C61
\end{keyword}
\end{frontmatter}
\newpage

% %%%%% %%%%% %%%%% INTRODUCTION %%%%% %%%%% %%%%% %%%%%
\section{Introduction}\label{sec:introduction}
"Holding the increase in the global average temperature [\ldots] well below \SI{2}{\celsius} above pre-industrial levels and pursuing efforts to limit the temperature increase to \SI{1.5}{\celsius} above pre-industrial levels"~\citep{UNFCCC2015} requires rapid and substantial reductions in greenhouse gas emissions worldwide~\citep{IPCC2018}.
Increasing the share of electricity generated from renewable sources consequently is amongst the United Nations's sustainable development goals (indicator 7.\ 2.\ 1) and a prominent goal of many climate and energy strategies.
The European Union, for example, aims to meet at least \SI{32.5}{\percent} of its final energy consumption from renewable sources~\citep{EU2018}.
Against this background, Austria aims to generate \SI{100}{\percent} of its electricity consumption from domestic renewable sources on annual balance by 2030\footnote{Industry own use and ancillary service provision are exempt.
Together, these accounted for approximately \SI{10}{\percent} of consumption in 2016.}.
According to government estimates, additional electricity generation of \SI{27}{\tera\watt\hour} from renewable
sources is required to meet policy goals.
Out of this total increase in renewable electricity generation, \SI{1}{\tera\watt\hour} is thought to be sourced from biomass, and \SI{5}{\tera\watt\hour} from new hydropower generation, according to the government's climate and
energy strategy~\citep{BMNT2018}.
The comparatively small projected increase in electricity generation from these sources results from the perception
of biomass being ecologically~\citep{Erb2018} and economically unsustainable, while further development of Austria's
hydropower potentials was considered `challenging'~\citep{Wagner2015} even before the projected increase in hydro
generation by \SI{5}{\tera\watt\hour} was factored in.
As Austrians banned the use of nuclear power in a referendum in 1978~\citep{Pelinka1983}, wind and solar
power are the only remaining options for a large-scale expansion of low-carbon power generation in the country.

While wind power mitigates negative external effects at the global (\COO emissions) and the regional (air pollutants
from fossil thermal power generation) level, it also imposes negative external effects at the local level.
Wind turbines are found to negatively impact wildlife~\cite{Loss2013,Voigt2015,Wang2015}, to emit
noise~\cite{Wang2015a}, and to interfere negatively with landscape aesthetics~\citep{Jones2010, Meyerhoff2010}.
The latter is found to be the dominant negative external effect of wind turbines~\cite{Mattmann2016}.

Accordingly, several strands of literature seek to quantify the negative impact of wind turbines, for example
through their impact on property prices within a revealed preferences framework~\citep{Droees2016, Gibbons2015,
Heintzelmann2017, Jensen2018, Kussel2019, Lang2014, Sims2008, Sunak2016, Vyn2014}, or through choice
experiments and surveys~\citep{Drechsler2011, Mattmann2016, Meyerhoff2010} in a stated preferences-context.
Zerrahn~\cite{Zerrahn2017} provides a comprehensive survey of the literature on wind power externalities.

Apart from negative external effects at the local level, wind power is also found to have the potential to reduce
electricity system cost relative to solar PV.
For the case of Germany, Ueckerdt et al.~\cite{Ueckerdt2013} estimate solar PV integration cost being about four times higher than the integration cost of wind power at penetration levels of \SI{25}{\percent}.
Scholz et al. find \cite{Scholz2017} integration cost of \SI{40}[\euro]{\per \mega\watt\hour} for a mix of \SI{80}{\percent} solar PV and \SI{20}{\percent} wind power at \SI{100}{\percent} renewables penetration in Europe.
This reduces to below \SI{25}[\euro]{\per\mega\watt\hour} when the technology mix is changed to \SI{20}{\percent} solar PV and \SI{80}{\percent} wind power.
% Similarly, Gils et al. \cite{Gils2017}

These findings illustrate that abandoning wind power in favor of undisturbed landscapes potentially comes at significant opportunity cost.
We complement the literature on negative wind turbines impacts and set out to quantify this cost of undisturbed landscapes for the case of Austria.
First, Austrian policies targeting at a fixed share of renewable electricity generation are an ideal framework for
studying the effects of substituting renewable energy generation technologies (RET).
Moreover, Austria has large hydro storage capacities (including seasonal hydro storages) in place, which should limit PV integration cost.
Thus, our findings should be qualitatively generalizable to power systems with high renewable shares and lower storage capacities in similar climates.
Methodologically, we rely on the power system model \emph{medea}, which is described in section~\ref{sec:data-and-methods}.

Our analysis contributes to a more comprehensive understanding of the trade-offs faced in the design of highly renewable electricity systems, thereby helping to inform policies about the socially optimal expansion of RET.
Any such policy necessarily needs to account for the full social cost (including all externalities) of RET.
Moreover, we contribute to the analysis of near-optimal power system configurations~\citep[see e.g.][]{Neumann2019,
Schlachtberger2017}.
Corresponding results are presented in section~\ref{sec:results-discussion}.

Our results also provide insight into how potential compensation for negative external effects of wind turbines at the local level could be financed by system-wide benefits of wind power.
Moreover, based on our analysis we can approximate the valuation of undisturbed landscapes that is implicit in Austria's climate and energy strategy.
These policy implications and conclusions are summarized in section~\ref{sec:conclusions-policy-implication}.

% %%%%% %%%%% %%%%% DATA AND METHODS %%%%% %%%%% %%%%% %%%%%

\section{Data and Methods} \label{sec:data-and-methods}
To investigate the cost of undisturbed landscapes (i.e. the opportunity cost of wind power) we use the power system model \emph{medea}, as summarized in section~\ref{subsec:medea}.
We instantiate the model with scenario assumptions and data as observed in the year 2016, which was a low-wind year in Austria (see section~\ref{subsec:data}).

Based on this, we approximate the opportunity cost of wind power through the following procedure:
\begin{enumerate}
\item We derive the unrestricted system cost-minimizing deployment of wind and solar power, given the scenario
assumptions
\item We restrict deployment of wind power by a small margin (so that the next best RET substitutes for wind
power) and observe net system cost $c_{net}$ for Austria (calculated as total system cost including air
pollution cost net of trade).\label{enumerate:restrict}
\item We repeat step~\ref{enumerate:restrict} till no wind power can be deployed.
\item Finally, we approximate the opportunity cost of wind power (at given wind power capacity $w$) $OC_w$ by
the change in $c_{net}$ in response to a change in wind power capacity $w$ deployed, i.e.
\[OC_w = \frac{\Delta c_{net}}{\Delta w}\]
where $\Delta$ is the difference operator and opportunity cost $OC_w$ are expressed in \EUR per
\SI{}{\mega\watt} wind power foregone.
\end{enumerate}

% ***** medea *****
\subsection{Power system model \emph{medea}}\label{subsec:medea}
We make use of the power system model \emph{medea} to simulate (dis)investment in and hourly operation of the
prospective Austrian (and German) power system.
The model is cast as a linear optimization seeking to minimize total system cost, which consist of fuel and emission
cost, quasi-fixed and variable operation and maintenance (O\&M) cost, the costs of investment in energy generation,
storage, and transmission assets, and (potential) cost of non-served load.
From an economic perspective, the model reflects a perfectly competitive energy-only market with fully
price-inelastic demand and perfect foresight of all actors.

The system is required to meet exogenous and inelastic demand for electricity and heat in any hour of the modeled
year.
Energy supply, in turn, is constrained by available installed capacities of energy conversion, storage, and
transmission units.
Co-generation units convert fuel to heat and power subject to a feasible operating region defined by the unit's
electrical efficiency, the electricity loss per unit of heat production, and the back pressure coefficient.
Electricity generation from intermittent sources (wind, run-of-river hydro, solar) is subject to exogenous hourly
generation profiles, which are scaled according to total installed capacities.
Electricity from these sources can be curtailed at no additional cost (free disposal).

Electricity can be stored in reservoir and pumped hydro storages or batteries.
The capacity of hydro storages can not be expanded, as we assume existing potentials to be exhausted.
Battery capacity, on the other hand, can be added endogenously.
Generation from storages is constrained by installed capacity and energy contained in storage.
Hydro reservoirs are filled (exogenous) inflows.
Pumped hydro storages and batteries can actively store electricity for later use.
To better capture operational differences of hydro storage units, we model daily, weekly and seasonal reservoir and pumped storage plant separately.
To ensure a stable and secure operation of the electricity system, ancillary services (e.g. frequency control,
voltage support) are required.
We model ancillary service needs as a minimum requirement on spinning reserves operating at any point in time.
Thus, we assume that ancillary services can be provided by thermal power plant, run-of-river hydro plant or any
type of storage technology.
We do not model transmission and distribution grids within Austria or Germany.
We do, however, allow for cross-border electricity trade between both countries.
We also keep track of the cost of air pollution arising from burning fossil fuels.
As these cost are external, they do not enter cost minimization.
For a detailed, mathematical description of the model please refer to \ref{sec:medea-desc}.
Data processing is implemented in python, while the optimization model is based on GAMS.
Running the model in hourly resolution for one year takes 10--15 minutes, depending on the model parameters, on an Intel i7-8700 machine with 16 GB RAM, using CPLEX 12.9 as a solver.

% ***** Data *****
\subsection{Data}\label{subsec:data}
We set up our model with the goal to resemble Austria's prospective electricity and district heating systems in the year 2030.
We also include Austria's largest electricity trading partner, Germany, to account for potential effects from electricity trade.
Again, our scenario reflects Germany's currently announced electricity sector policies, so that we set generation capacities to levels consistent with policies effective by 2030.
These assumptions are laid out in sections~\ref{subsubsec:assumptions-austria} and \ref{subsubsec:assumptions-germany} and summarized in table~\ref{tab:instcaps}.

\subsubsection{Scenario Assumptions for Austria}\label{subsubsec:assumptions-austria}
The Austrian government has set itself the goal of generating \SI{100}{\percent} of electricity consumption from domestic renewable sources on annual balance by 2030~\citep{Regierungsprogramm2020}.
However, industry own consumption and system services, which currently account for about \SI{10}{\percent} of annual electricity consumption, are exempt.

The government plans to achieve this goal by generating an additional \SI{27}{\tera\watt\hour} of electricity annually from renewable sources\footnote{Please note that this policy goal does not imply that the generated
electricity must actually be consumed. Hence, we count curtailed electricity as contributing to the policy goal.}.
New hydro power plants are thought to contribute \SI{5}{\tera\watt\hour} annually, while additional electricity generation from biomass is envisaged to account for \SI{1}{\tera\watt\hour} annually.
In our scenarios, we add generation capacities sufficient for reaching these targets.

Further, the government projects the remainder to come from solar PV (\SI{11}{\tera\watt\hour} annually) and onshore wind turbines (\SI{10}{\tera\watt\hour} annually)~\cite{Regierungsprogramm2020}.
As we are interested in determining the opportunity cost of wind power versus its best alternative, we allow for endogenous investment in wind and solar power, without enforcing announced targets.
Other low-carbon energy technologies are not feasible at large scale in the Austrian context, as we have laid out in the introduction.
Initial generation capacities for Austria are summarized in table~\ref{tab:instcaps}.

\subsubsection{Scenario Assumptions for Germany}\label{subsubsec:assumptions-germany}
Germany has announced specific capacity targets for several power generation technologies.
Following these announcements, we anticipate an end to nuclear power generation, a (partial) coal exit according to
recommendations by the `coal commission'~\cite{WSB2019}, and a further expansion of renewable electricity generation
in line with the German Renewable Energy Sources Act~\cite{noauthor_renewable_2017}.
We do, however, expect the \SI{52}{\giga\watt} solar PV cap to be removed.
The corresponding capacity assumptions are displayed in table~\ref{tab:instcaps}.

\begin{table}[H]
\caption{Initial Generation Capacities} \label{tab:instcaps}
\centering
\begin{tabulary}{0.55\textwidth}{L C C}
\toprule
Technology & Austria & Germany\\
& \si{\giga\watt} & \si{\giga\watt}\\
\midrule
Wind Onshore & $2.6$ & $90.8$ \\
Wind Offshore  & $0.0$ & $15.0$ \\
Solar PV & $1.1$ & $73.0$ \\
Run-of-river Hydro  & $6.9$ & $4.5$ \\
Hydro storage & $7.7$ & $6.5$ \\
Biomass & $0.7$ & $8.4$ \\
Lignite & $0.0$ & $11.4$ \\
Coal & $0.0$ & $14.0$ \\
Natural Gas & $3.9$ & $24.2$ \\
Mineral Oil & $0.2$ & $3.5$ \\
Heat Pump & $1.0$ & $1.0$ \\
Gas Boiler & $3.9$ & $25.5$\\ \bottomrule
\end{tabulary}
\end{table}

\subsubsection{Energy supply}
We represent 20 dispatchable energy conversion and storage technologies that are expected to be operated in 2030.
In addition to given, initially installed capacities (see table~\ref{tab:instcaps}), the model can endogenously add further generation capacities that are compatible with stated policy objectives.
We calculated annualized investment cost of each technology on the basis of an assumed weighted average cost of capital of \SI{5}{\percent} over the plant's lifetime.
All technologies can also be decommissioned, so that we adopt a long-run perspective on the power system.
Parameters of admissible technologies are summarized in Table~\ref{itm:techcost}.

%\begin{table*}[h!t]
\begin{ThreePartTable}
\renewcommand\TPTminimum{\textwidth} %%% we want full width
\begin{TableNotes}
\small
\item SC -- subcritical, USC -- (ultra-)supercritical, ST -- steam turbine, GT -- gas turbine, CC -- combined cycle, PSP -- Pumped storage plant
\item[$\dagger$] own assumption
\item[$\ddagger$] \EUR / $(\text{MVA} \times \text{km})$
\end{TableNotes}

\begin{longtable}{l ccc ccc l}
\caption{Technology Input Parameters for 2030} \label{itm:techcost}\\
\toprule
&\multicolumn{2}{c}{Capital Cost} & \multicolumn{2}{c}{O\&M-Cost} & & & \\ \cmidrule(lr){2-3}\cmidrule(lr){4-5}
Technology & Power & Energy & Quasi- & Variable & Life- & Efficiency & Source   \\
& & & fixed & & time & &          \\
&\EUR/kW & \EUR/kWh & \EUR/MWa & \EUR/MWh & a & & \\
\midrule
\endhead
%\cmidrule{2-2}
\midrule
\multicolumn{8}{c}{\textit{continued on next page}} \endfoot
\bottomrule
\insertTableNotes
\endlastfoot
% the contents of the table
Wind Onshore         & $1040$ & NA    & $12600$ & $1.35$ & $30$ & NA & \cite{DEA2019}\\
Wind Offshore        & $1930$ & NA    & $36053$ & $2.70$ & $30$ & NA & \cite{DEA2019}\\
PV Rooftop           & $870$  & NA    & $10815$ & $0.00$ & $40$ & NA & \cite{DEA2019}\\
PV Open-space        & $380$  & NA    & $7250$  & $0.00$ & $40$ & NA & \cite{DEA2019}\\
Hydro run-of-river   & --     & NA    & $60000$ & $0.00$ & $60$ & NA & \cite{Schroeder2013, LacalArantegui2014} \\
\midrule
Lignite Adv          & --     & NA    & $40500$ & $0.85$ & $40$ & 0.439 & \cite{OekoInstitut2017}\\
Coal SC              & --     & NA    & $25000$ & $6.00$ & $25$ & 0.390 & \cite{Schroeder2013} \\
Coal USC             & --     & NA    & $31500$ & $3.00$ & $25$ & 0.460 & \cite{DEA2019}\\
Nat Gas ST           & $400$  & NA    & $15000$ & $3.00$ & $30$ & 0.413 & \cite{Schroeder2013}\\
Nat Gas GT           & $435$  & NA    & $7745$  & $4.50$ & $25$ & 0.410 & \cite{DEA2019}\\
Nat Gas CC           & $830$  & NA    & $27800$ & $4.20$ & $25$ & 0.580 & \cite{DEA2019}\\
Oil ST               & $400$  & NA    & $6000$  & $3.00$ & $30$ & 0.410 & \cite{Schroeder2013}\\
Oil GT               & $363$  & NA    & $7745$  & $4.50$ & $25$ & 0.400 & \cite{DEA2019}\\
%Oil CC             & $830$  & NA    & $27800$ & $4.20$ & $25$ & 0.580 & \cite{}\\
Biomass              & $3300$ & NA    & $96000$ & $4.60$ & $25$ & 0.270 & \cite{DEA2019}\\ \midrule
Hydro Reservoir      & --     & --    & $22000$ & $3.00$ & $60$ & $0.900$\tnote{$\dagger$} & \cite{LacalArantegui2014}\\
Hydro PSP            & --     & --    & $22000$ & $3.00$ & $60$ & $0.810$\tnote{$\dagger$} & \cite{LacalArantegui2014}\\
Battery Li-Ion       & $320$ & $302$ & $540$ & $1.80$ & $25$ & 0.920 & \cite{DEA2020}\\ \midrule
Electric Boiler      & $140$  & NA    & $1020$  & $0.50$ & $20$ & 0.990 & \cite{DEA2019}\\
Absorption heat pump & $510$  & NA    & $2000$  & $1.30$ & $25$ & 1.730 & \cite{DEA2019}\\
Natural gas boiler   & $50$   & NA    & $1900$  & $1.00$ & $25$ & 0.920\tnote{$\dagger$} & \cite{DEA2019} \\ \midrule
Transmission (AC)    & $455$\tnote{$\ddagger$} & NA    & $9$\tnote{$\ddagger$} & $0.00$ & $40$ & 1.000 & \cite{Brown2018, Hagspiel2014}\\
\end{longtable}
\end{ThreePartTable}
%\end{table*}

Power plants running on fossil fuels emit \COO into the atmosphere.
We approximate the amount of \COO released through the carbon intensity of fossil fuels, as displayed in Table~\ref{table:carbon-intensity}. These estimates are based on an analysis by the German environmental agency~\citep{Juhrich2016}.

\begin{table}[h!t]
\centering
\caption{\COO intensity of fossil fuels in tonnes \COO per MWh}
\label{table:carbon-intensity}
\begin{tabular}{c c c l}
\toprule
Lignite & Coal & Natural Gas & Mineral Oil  \\
\midrule
0.399 & 0.337 & 0.201 & 0.266        \\
\bottomrule
\end{tabular}
\end{table}

In the baseline scenario, we assume capital cost of solar PV of \SI{630}[\euro]{\per\kilo\watt} installed.
These cost are capacity-weighted average cost of rooftop and open-space PV installations, with roughly \SI{50}{\percent} of PV capacity being mounted on roof tops, while the remainder is realized as open-space solar PV at utility scale.
In section~\ref{subsec:sensitivity} we analyze the sensitivity of our results with respect to the capital cost of solar PV\@.

Generation from non-dispatchable technologies solar PV, wind turbines and run-of-river hydro is assumed to follow hourly generation profiles\footnote{A \emph{generation profile} tracks the share of installed intermittent capacity that is generating electricity over time.} as observed in 2016.
German solar and wind power profiles are sourced from Open Power System Data~\citep{opsd2019}.
For Austria, solar and wind power generation profiles were derived by dividing solar and wind power generation time series as reported by Open Power System Data~\cite{opsd2019} by installed capacities published by \cite{Biermayr2019}.
Similarly, generation profiles for run-of-river hydro power were derived based on generation data reported by \cite{ENTSOE2020b}.

Data on inflows into Austrian hydro reservoirs is not publicly available.
However, ENTSO-E's transparency platform reports weekly data on hydro storage filling levels~\cite{ENTSOE2020a} along with hourly electricity generation from hydro reservoirs, and pumping by pumped hydro storages~\cite{ENTSOE2020b}.
Based on this data, we have approximated inflows as the part of change in (weekly) hydro storage fill levels not explained by (upsampled hourly) pumping or generation.
To arrive at hourly inflows, we have interpolated our weekly estimates by piecewise cubic hermite interpolation.
Although inflows to hydro reservoirs are given exogenously, all storage technologies (including hydro reservoir and pumped storage plants) are charged and dispatched endogenously.

\subsubsection{Energy demand}\label{subsec:energy-demand}
By combining government policy targets with information about electricity generation in our base
year~\citep{StatistikAustria2020}, we can infer expected electricity demand of \SI{83}{\tera\watt\hour} in 2030.
To generate hourly load, we scale the 2016 load profile accordingly.
For Germany, we scale reported hourly load to match annual electricity consumption as reported in national energy balances~\citep{AGEB2018} for the year 2016.
For 2030 we assume electricity demand to remain unchanged from 2016 levels.

Annual heat consumption is derived from energy balances for Austria and Germany, respectively~\citep{AGEB2018,
StatistikAustria2020}.
Subsequently, annual heat consumption is broken down to hourly heat consumption on the basis of standard natural gas load profiles for space heating in the residential and commercial sectors~\citep{Almbauer2008}.
These load profiles make use of daily average temperatures to calculate daily heat demand.
We extract spatially resolved temperature data from ERA-5 climate data sets~\citep{CCCS2017} and compute a capacity-weighted average of temperatures at locations of combined heat and power (CHP) generation units.
Daily heat demand based on these weighted average temperatures is then broken down to hourly consumption on the basis of standardized factors accounting for weekday and time-of-day effects.
Descriptive statistics of electricity and heat consumption are provided in table~\ref{tab:time-series}.

\begin{table*}[h!t]
\centering
\begin{threeparttable}
\caption{Descriptive data of time series (2016)}\label{tab:time-series}
\begin{tabular}{l c c c c c c l}
\toprule
Name & Area & Unit & mean & median & max & min & Source   \\
\midrule
Electricity load & AT & GW & 7.15 & 7.14 & 10.5 & 4.21 &\cite{opsd2019}     \\
Electricity load & DE & GW & 60.2 & 59.6 & 82.8 & 34.6 &\cite{opsd2019}     \\
District heating load\tnote{$\dagger$} & AT & GW & 2.56 & 2.43 & 5.77 & 0.91 & \cite{CCCS2017, StatistikAustria2020}\\
District heating load\tnote{$\dagger$} & DE & GW & 14.6 & 14.1 & 26.0 & 8.9 & \cite{CCCS2017, AGEB2018} \\
Wind onshore profile & AT &\%   & 0.229 & 0.142 & 0.929 & 0.000 &\cite{opsd2019}         \\
Wind onshore profile & DE &\%   & 0.176 & 0.133 & 0.764 & 0.003 &\cite{opsd2019}         \\
Wind offshore profile & DE &\%   & 0.371 & 0.327 & 0.900 & 0.000 &\cite{opsd2019}         \\
Solar PV profile\tnote{$\dagger$ } & AT &\%   & 0.098 & 0.012 & 0.570 & 0.000 & \cite{opsd2019, Biermayr2019}\\
Solar PV profile & DE &\%   & 0.099 & 0.003 & 0.661 & 0.000 &\cite{opsd2019}         \\
Run-of-river profile & AT &\%    & 0.564 & 0.538 & 0.971 & 0.202 & \cite{ENTSOE2020b}$\dagger$\\
Run-of-river profile & DE &\%   & 0.437 & 0.420 & 0.637 & 0.208 &\cite{ENTSOE2020b}\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item[$\dagger$] own calculation based on referenced sources
\end{tablenotes}
\end{threeparttable}
\end{table*}

\subsubsection{Electricity Transmission}
Electricity transmission between the modelled market areas is limited to \SI{4.9}{\giga\watt}, in line with the introduction of a congestion management scheme by German and Austrian authorities in 2018~\citep{}.
%\ref{sec:transmission-sensitivity} explores the effects of doubling the transmission capacity between Germany and Austria.

\subsubsection{Prices}
Monthly prices for exchange-traded fuels (hard coal, crude oil (Brent), natural gas) are retrieved from the
International Monetary Fund's Commodity Data Portal.
We convert these prices to \SI[sticky-per]{}[\euro]{\per\mega\watt\hour} based on the fuel's energy content and market exchange rates obtained from the European Central Bank~\citep{ECB2020}.
Finally, we resample prices to hourly frequency using piecewise cubic hermite interpolation.
Due to its low energy density lignite is not transported over large distances and consequently also not traded on
markets.
Instead, lignite-fired power plants are situated in proximity to lignite mines.
According to estimates from~\cite{OekoInstitut2017}, the price of lignite in Germany is close to
\SI[sticky-per]{1.50}[\euro]{\per\mega\watt\hour}.
Biomass-fired power plant run on a wide variety of solid and gaseous fuels, some of which are marketed.
However, continued operation of biomass-fired plants typically relies on sufficient subsidies.
As a first-order approximation to more complex subsidy schemes, we assume subsidized fuel cost of
\SI[sticky-per]{6.50}[\euro]{\per\mega\watt\hour}.
In consequence, biomass-fired plants are operating at or close to capacity throughout the scenarios considered.
\footnote{Subsidies are not included in system cost, as the required level is uncertain.
The total sum of subsidies is, however, constant across scenarios as biomass capacity can not be expanded.}

\begin{table*}[h!t]
\centering
\begin{threeparttable}
\caption{Descriptive data of price time series}
\begin{tabular}{l c r r r r l}
\toprule
Name & Unit & mean & median & max & min & Source \\
\midrule
Lignite     &\EUR/MWh & 1.50  & 1.50  & 1.50  & 1.50  & \cite{OekoInstitut2017}\\
Coal        &\EUR/MWh & 8.58  & 8.13  & 11.90 & 6.69  & \cite{IMF2020}         \\
Natural gas &\EUR/MWh & 13.62 & 12.87 & 20.18 & 12.04 & \cite{IMF2020}         \\
Mineral oil &\EUR/MWh & 26.40 & 26.63 & 33.44 & 18.34 & \cite{IMF2020}         \\
Biomass (subsidized) &\EUR/MWh & 6.50 & 6.50 & 6.50 & 6.50 & $\dagger$         \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item[$\dagger$] own assumption
\end{tablenotes}
\end{threeparttable}
\end{table*}

Future \COO prices impact renewables deployment ~\citep{Brown2020, Kirchner2019} and affect the results of our analysis.
As the future efficient price of \COO is highly uncertain, we have conducted our analysis for \COO prices in the range of $0$ to $120$ Euro per tonne.
Implicitly, we assume efficient pricing of \COO emissions, i.e.\ all otherwise external cost of \COO emissions are internalized through the prevailing \COO price.

In this context, it is worth noting that within the framework of our analysis a \COO price of
\SI{30}[\euro]{\per\tonne} would be consistent with RET expansion goals stated by the Austrian government.
At higher \COO prices, there is incentive to invest in \COO emission-free generation beyond
government targets.

\subsubsection{The cost of air pollution}
In addition to \COO emissions, we also keep track of air pollutants such as nitrogen oxides, sulphur oxide, particulate matters, carbon monoxide, or heavy metals and the external cost imposed by their emission from power plants.
For the valuation of these external cost, we rely on assessments from the NEEDS project, as reported in \cite{Samadi2017}.
We have, however, converted these output-related estimates to input-related values.
The corresponding estimates for the external cost of air pollution from fossil fuel combustion are summarized in Table \ref{tab:air_pollution_cost}.

\begin{table}[ht]
\centering
\begin{tabular}{c c c c c}
\toprule
Lignite & Coal & Natural Gas & Mineral Oil & Biomass \\
\midrule
$4.06$ & $6.12$ & $2.36$ & $3.21$ & $4.04$ \\
\bottomrule
\end{tabular}
\caption{The external cost of air pollution in \SI{}[\euro]{\per\mega\watt\per\hour} of fuel burned}
\label{tab:air_pollution_cost}
\end{table}
Please note that these cost are considered external and consequently do not enter system cost minimization.

The model as well as all data retrieval and processing scripts are available at medea's github repository
\url{https://github.com/inwe-boku/medea} under the MIT license.

% %%%%% %%%%% %%%%% RESULTS AND DISCUSSION %%%%% %%%%% %%%%% %%%%%
\section{Results and Discussion}\label{sec:results-discussion}
To derive the social opportunity cost of not using wind power (but solar photovoltaics instead), we start by
determining the long-run power market equilibrium without constraints on renewable capacity addition.
This gives us our unconstrained baseline scenario with minimal system costs.

\subsection{Unconstrained Baseline}\label{subsec:baseline}
As a point of reference for our analysis, we determine the least cost power system configuration when addition of
wind turbines and solar PV is unconstrained.
Given the policy goal of generating \SI{21}{\tera\watt\hour} of electricity either from wind or solar power (with
the remainder coming from hydro power and biomass), required capacity expansion amounts to~\SI{10.4}{\giga\watt}
wind power or \SI{24.1}{\giga\watt} of solar power if the policy is to be fulfilled with a single technology.

Given our baseline assumptions, we find the cost-minimizing system set-up being consistent with policy objectives,
to rely entirely on wind power.
In the unconstrained baseline,  \SI{10.4}{\giga\watt} wind power and \SI{0}{\giga\watt} solar PV are added to
initially available capacities in Austria.
Wind power is complemented by \SI{2.63}{\giga\watt} of additional natural gas fired combined heat and power
generation capacity, while less efficient natural gas-fired units with a capacity of \SI{0.42}{\giga\watt} are
decommissioned along with \SI{0.08}{\giga\watt} oil-fired capacity.
In effect, \SI{6.23}{\giga\watt} of fossil thermal generation capacities are active, giving rise to \SI{10.5}{\mega\tonne} of \COO emissions per year.
This compares to current (2016) domestic \COO emissions of power generation from fossil fuels of \SI{8.2}{\mega\tonne}\footnote{For comparability, we calculated \COO emissions based on fossil fuel consumption reported in Austria's energy balance \citep{StatistikAustria2020} and applied the \COO intensity factors from table \ref{table:carbon-intensity}.}. The increase in \COO emissions occurs as domestic electricity consumption is increasing while electricity net imports, which accounted for \SI{9.9}{\percent} of total consumption in 2016, are declining.
As Austria has ample hydro storage capacity in place, no further storage (e.g. from batteries) is added.
However, the addition of \SI{1.23}{\giga\watt} of compression heat pumps allows to use electricity for heat
generation, keeping curtailment at \SI{1.5}{\tera\watt\hour\per\year}.
Annual system cost in Austria amount to $2.9$ billon euro in this scenario.
Net earnings from exports total $0.7$ billion euro, so that net system cost amounts to $2.2$ billion euro.

\subsection{Restricting wind power}\label{subsec:restricting-wind}
As we gradually restrict wind power potentials, the system is forced to increasingly rely on electricity generation
from solar photovoltaics.
To satisfy the politically targeted share of renewable energy generation, each \SI{}{\giga\watt} of wind power not
installed needs to be replaced by \SI{2.35}{\giga\watt} of solar PV\@.
As generation profiles of wind and solar power differ, the substitution of wind power with solar PV brings about
changes in the system's composition, in system operation, and, consequently, in system cost.

\subsubsection{Changes in system composition and system operation}
The possibility to exchange electricity with Germany strongly affects system operation in Austria.
As Austria's electricity and heat generation is less \COO intense than Germany's, variations in the price of \COO
have marked, potentially unexpected consequences for outcomes in Austria.
If \COO emissions are costless, annual net exports of electricity total around \SI{5.5}{\tera\watt\hour}, with
minor variation due to the composition of renewable electricity generation in Austria.
When \COO prices rise to \SI{30}[\euro]{\per\tonne}, electricity exports increase by around \SI{2}{\tera\watt\hour},
for any composition of renewable electricity generation in Austria.
This happens because the more emission intense dispatchable electricity generators in Germany become less competitive
versus their less emission intense counterparts in Austria.
In consequence, \COO emissions in Austria increase, while total \COO emissions decline.
The increase in fossil thermal generation in Austria reverts back to levels observed for no emission cost as the
price of \COO reaches \SI{60}[\euro]{\per\tonne}.
Higher \COO prices induce further reductions of \COO emissions in Austria and lead to a higher sensitivity of
mitigated \COO emissions to the composition of domestic renewable generation.

Moreover, an increase in the price of \COO incentivises renewables development, as electricity generation from
low-carbon sources becomes more competitive versus fossil generation.
Yet, our analysis reveals that \COO prices are not the only factor of importance.
Developed wind resources in Austria are favourably complementary to their German counterparts, with a Pearson
correlation coefficient of $r_{AT,DE} = 0.235$.
Solar resources, on the other hand, are highly correlated in Austria and Germany ($r_{AT,DE}=0.948$).
In effect, wind power has a higher export revenue potential than solar power, particularly at \COO prices of
\SI{60}[\euro]{\per\tonne} or higher.
However, exploiting wind power export potentials requires sufficient wind power capacities to be in place.

Neglecting the effects of electricity trade, gradual substitution of wind energy with solar PV in tendency leads to (i) higher capacities of fossil thermal
generators and (ii) lower capacities of heat pumps in the system, (iii) more \COO emissions, and (iv) more local
air pollution.
Consistent across all \COO price scenarios, (i) oil-fired units in Austria are largely shut down while natural gas-fired
co-generation capacities are expanded.
Additions of natural-gas fired capacities are lower at high \COO prices and tend to increase with the share of solar
PV in the system, particularly at high \COO prices.
However, there are huge differences in fossil capacity additions between \COO prices of \SI{60}[\euro]{\per\tonne}
and \SI{90}[\euro]{\per\tonne}.
In the former case, there is very little difference to capacity addition when \COO emissions come at no cost, while
in the latter case fossil capacity additions are more than halved at high wind shares.
The amount of wind power in the system (ii) also affects the deployment of compression heat pumps.
In any scenario considered, heat pumps with a capacity of at least \SI{1.2}{\giga\watt} are operational.
With an increase of the \COO price to \SI{120}[\euro]{\per\tonne} heat pumps reach a total capacity of
\SI{2.2}{\giga\watt} when all additional renewable electricity generation comes from solar PV\@.
Heat pump capacity increases further to a maximum of \SI{2.9}{\giga\watt} when additional renewable electricity is
sourced from wind power.
Overall, more heat tends to be generated from fossil sources if \COO prices are low and the system is dominated by
solar PV\@.
At high \COO prices and with more wind energy, heat generation is increasingly electrified.
In effect, a higher share of wind energy (iii) reduces \COO emissions of electricity generation as less energy is sourced from fossil thermal plant.
This (iv) also reduces emissions of other air pollutants associated with fossil thermal generation.

\begin{figure*}
\centering
\includegraphics[width=0.95\textwidth]{./figures/sysops.pdf}
\caption{Opportunity cost of wind power assuming PV overnight cost of 630 EUR/kWp}
\label{figure:system-operation-base}
\end{figure*}

\subsubsection{The opportunity cost of wind power in Austria}\label{subsec:opportunity-cost-wind}
By relating the change in installed wind power to the change in system cost (or other variables of interest), we
can derive a measure of the (approximate) marginal cost of wind power expansion (or contraction).
In consequence, the total cost of undisturbed landscape is equal to the area below the opportunity cost curve,
starting from the highest wind power capacity down to the level of wind power admitted.

\begin{figure*}[h!t]
\centering
\begin{subfigure}[b]{0.475\textwidth}
\centering
\includegraphics[width=\textwidth]{./figures/undisturbed_base_share.pdf}
\caption{Relative substitution}
\label{fig:rel_substitution}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.475\textwidth}
\centering
\includegraphics[width=\textwidth]{./figures/undisturbed_base.pdf}
\caption{Absolute substitution}
\label{fig:abs_substitution}
\end{subfigure}
\caption{Annual opportunity cost of wind power, assuming PV overnight cost of 630 EUR/kWp}
\label{figure:opportunity-cost-base}
\end{figure*}

With a gradual restriction of deployable wind power capacities, we find opportunity costs of wind power
to increase by about \SI{3900}[\euro]{} for each \SI{}{\mega\watt} wind power substituted by solar PV across all studied scenarios.
At PV overnight cost of \SI{630}[\euro]{\per\mega\watt} (corresponding to a mix of rouhgly \SI{50}{\percent} rooftop and
\SI{50}{\percent} open space installations), the annual opportunity cost of the last \SI{}{\mega\watt} wind power replaced reaches close to \SI{62500}[\euro]{}.

At a wind power capacity consistent with the policy goal of generating \SI{21}{\tera\watt\hour} of electricity
from renewable sources, opportunity cost for wind power are close to \SI{20000}[\euro]{\per\mega\watt}, with a
notable exception for \COO prices of \SI{60}[\euro]{\per\tonne}.
In this case, opportunity cost of wind power are estimated at close to \SI{10000}[\euro]{\per\mega\watt}, about half
the size of opportunity cost estimated for all other \COO price scenarios.

Based on these findings, we can also approximate the cost of undisturbed landscapes implied by government targets.
Given the goal of generating \SI{21}{\tera\watt\hour} electricity from renewable sources and the wind energy target
generation of \SI{10}{\tera\watt\hour}, we can infer $\Phi = 0.476$ according to government plans.
At this level, we estimate opportunity costs of wind power between \SI{25000}[\euro]{\per\mega\watt} and \SI{40000}[\euro]{\per\mega\watt} depending on the carbon price.
Over an assumed lifetime of $30$ years, this implies an annual gain from a contemporary \SI{3.5}{\mega\watt}
wind turbine between \SI{1.35}{} and \SI{2.15}{} million euro, if future gain is discounted at \SI{5}{\percent}.

\subsection{Sensitivity Analysis} \label{subsec:sensitivity}
A major factor of influence for our results are the assumed capital cost of wind turbines and solar PV\@.
Therefore, we complement our baseline scenario with a sensitivity analysis on overnight investment cost of solar
photovoltaics.
We gradually lower the capital cost of solar PV from our baseline assumption of \SI{630}[\$]{\per\kilo\wattpeak}
(corresponding to \SI{55}{\percent} of rooftop and \SI{45}{\percent} open space capacity) to
\SI{275}[\euro]{\per\kilo\wattpeak}, which Vartiainen et al.~\cite{Vartiainen2019} project for open-space,
utility-scale solar PV in the year 2030.

\begin{figure*}[h!t]
\centering
\includegraphics[width=0.8\textwidth]{./figures/undisturbed_low.pdf}
\caption{Opportunity cost of wind power}
\end{figure*}

Under these assumptions, the impact of \COO prices on the opportunity cost of wind power increases.
If all additional PV is realized as open-space, utility-scale installation (i.e.\ PV capital cost of
\SI{275}[\euro]{\per\kilo\wattpeak}), the optimal deployment of wind power declines and approaches zero at \COO prices
of \SI{30}[\euro]{\per\tonne} or below.
In consequence, we can not determine opportunity cost of wind power given that both \COO prices and capital cost of
solar PV are very low.
For higher \COO prices, the optimal deployment of wind power is strictly positive, even at low capital cost of solar
PV and we find opportunity cost to increase by about \SI{2550}[\euro]{} for each \SI{}{\mega\watt} wind power
foregone.
The last \SI{}{\mega\watt} of wind power substituted with solar PV comes at opportunity cost between
\SI{20000}[\euro]{} and \SI{56000}[\euro]{}, depending on the prevailing \COO price.

\subsection{Limitations of the Analysis}
It is worth noting that our analysis is not spatially resolved, which has important consequences.
Our analysis of the opportunity cost of wind power does not depend on a specific spatial allocation of RET or a specific grid topology. Hence, our results are generalizable to the extent that any assessment of specific RET expansion plans must consider the effects of spatially distributed RET on resource quality and grid cost.
However, there are good reasons to believe these factors will not change the results of our analysis substantially.
First, we implicitly assume that wind and solar resource quality at locations developed in 2016 is representative of resource quality at locations developed by 2030, i.e. after generation increased by \SI{21}{\tera\watt\hour}.
For Austria, Hoeltinger et al. \cite{Hoeltinger2016}  identified \SI{45}{\tera\watt\hour} technically feasible wind resource potential with better or similar quality than we assumed in our analysis.
Hence, an additional \SI{21}{\tera\watt\hour} of electricity generated from wind power is feasible under our assumptions.
Yet, technical potentials are currently restricted by law in four of Austria's nine federal states.
These regulations limit total wind energy potential in these states to approximately \SI{11}{\tera\watt\hour}, of which around \SI{6}{\tera\watt\hour} are already used.
Thus, it will be necessary to lift these restrictions and amend current regulations even under current policy objectives.

Second, we implicitly assume that the cost of grid expansion does not differ between scenarios relying entirely on wind power and scenarios relying on solar PV exclusively.
Here, it makes a difference whether solar PV is realized as rooftop systems in distribution grids or as utility-scale open-space systems which are connected to transmission grids.
In the latter case, there is little reason to believe grid cost would differ substantially from wind power expansion, which also happens in transmission grids.
Horowitz et al. \cite{Horowitz2018} survey the nascent literature on the cost of integrating distributed solar PV systems in distribution grids.
These cost are found to depend on numerous and uncertain factors, leading the authors to the conclusion that "no generalized `cost of grid integration' for PV can be obtained".
Nevertheless, a study for Agora Energiewende \cite{Fuerstenwerth2015} estimates grid integration cost of solar PV in Europe at \SI{6}[\euro]{\per\mega\watt\hour} in distribution grids, and at \SI{1.5}[\euro]{\per\mega\watt\hour} in transmission grids. Estimates for specific grids, however, vary widely.
Yet, the available evidence suggests that any grid cost advantage of PV over onshore wind is likely small.

% %%%%% %%%%% %%%%% CONCLUSIONS AND POLICY IMPLICATIONS %%%%% %%%%% %%%%% %%%%%
\section{Conclusions and Policy Implication} \label{sec:conclusions-policy-implication}
% * future research should consider the opportunity cost of RET in terms of grid cost (including distribution grids)
Not disturbing landscapes can come at considerable opportunity cost, as our analysis reveals.
How high this cost is, depends on the valuation of \COO and on the cost of the best alternative to wind power.
If the cost of \COO is low and the best alternative solar PV is realized at utility scale in open space, wind power's system cost advantage vanishes in Austria.
If, on the other hand, the cost of \COO is high, if open-space PV itself causes negative externalities at the local level, or if there is a preference for rooftop solar PV, there are considerable opportunity cost of wind power.
Given the current policy target of adding capacities for generating \SI{10}{\tera\watt\hour} ($\phi=0.48$) wind energy, under our baseline assumptions the present value of the cost of undisturbed landscapes amounts to at least \SI{1.35}[\euro]{\million} over the lifetime of each \SI{3.5}{\mega\watt} wind turbine not erected.\footnote{We assume a lifetime of 30 years and a discount factor of \SI{5}{\percent}.}
At solar PV overnight cost that correspond to \SI{100}{\percent} open-space PV, the opportunity cost of wind power is reduced to \SI{480000}[\euro]{} at \COO prices of \SI{60}[\euro]{\per\tonne}.
In Austria, only open-space, utility scale solar PV has the potential to substitute wind power at comparable system cost, and only if a tonne of \COO emitted is valued below \SI{60}{\euro}.
However, open-space PV itself interferes with landscapes.
Relying completely on open-space solar PV would require about \SI{722}{\square\kilo\meter} land\footnote{In line with \cite{Guennewig2007} we assume land usage of \SI{29.55}{\square\meter} per \SI{}{\kilo\wattpeak} of non-tracking silicon wafer technology.} for replacing the approximately 2\,970 wind turbines of the \SI{3.5}{\mega\watt} class that would be needed to meet policy goals by wind power alone.
In the light of our findings, policy makers need to consider whether a wind power-leaning expansion of RET together with a compensation of local residents for harm inflicted by wind turbines could be a socially acceptable and cost-saving alternative to the large-scale expansion of solar PV on roof-tops.
Such an undertaking would, however, require spatially explicit estimates of wind turbine harm.
Here, we see a fruitful field for further research.

Attention should also be paid to the distributive consequences of wind power expansion in Austria.
As we have shown, a large-scale expansion of wind turbines comes with an increase in electricity exports, i.e. at high penetration some wind power capacity is mostly used to generate electricity for exporting.
Revenues from these export are accruing to the owners of power generation assets, while the negative externalities of wind turbines have to be borne by local residents.
This imbalance could be narrowed down by policies requiring wind turbine owners to share wind turbine income with affected residents, either directly or through transfers.
Fair sharing of benefits and burdens has the potential to foster acceptance of wind power \citep{Scherhaufer2017}.

%\section{Data Availability}\label{sec:data-availability}
%Data and model is avaliable on github \url{}. Raw data is sourced from danish Energy Agency \url{}, ENTSO-E's transparency database \url{},

\newpage
\bibliography{asparagus}
\bibliographystyle{elsarticle-harv}

\newpage
\appendix

\section{Description of the power system model \emph{medea}}\label{sec:medea-desc}

\subsection{Sets}\label{sets}
Sets are denoted by upper-case latin letters, while set elements are denoted by lower-case latin letters.

\begin{longtable}{p{0.15\textwidth}p{0.15\textwidth}p{0.2\textwidth}p{0.45\textwidth}}%{LLLp{6cm}
\caption{Sets}\\
\toprule
mathematical symbol & programming symbol & description & elements\\
\midrule
\endhead
\bottomrule
\multicolumn{4}{c}{\textit{continued on next page}} \endfoot
\bottomrule
\endlastfoot
% the contents of the table
$f \in F$            &\texttt{f}      & fuels & \texttt{nuclear, lignite, coal, gas, oil, biomass, power} \\
$i \in I$            &\texttt{i}      & power generation technologies & \texttt{nuc, lig\_stm, lig\_stm\_chp, lig\_boa, lig\_boa\_chp, coal\_sub, coal\_sub\_chp, coal\_sc, coal\_sc\_chp, coal\_usc, coal\_usc\_chp, coal\_igcc, ng\_stm, ng\_stm\_chp, ng\_ctb\_lo, ng\_ctb\_lo\_chp, ng\_ctb\_hi, ng\_ctb\_hi\_chp, ng\_cc\_lo, ng\_cc\_lo\_chp, ng\_cc\_hi, ng\_cc\_hi\_chp, ng\_mtr, ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm, oil\_stm\_chp, oil\_ctb, oil\_ctb\_chp, oil\_cc, oil\_cc\_chp, bio, bio\_chp, heatpump\_pth}                                 \\
$h \in H \subset I$  &\texttt{h(i)}   & power to heat technologies & \texttt{heatpump\_pth} \\
$j \in J \subset I$  &\texttt{j(i)}   & CHP technologies & \texttt{lig\_stm\_chp, lig\_boa\_chp, coal\_sub\_chp, coal\_sc\_chp, coal\_usc\_chp, ng\_stm\_chp, ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp, ng\_cc\_lo\_chp, ng\_cc\_hi\_chp, ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm\_chp, oil\_ctb\_chp, oil\_cc\_chp, bio\_chp} \\
$k \in K$            &\texttt{k}      & storage technologies & \texttt{res\_day, res\_week, res\_season, psp\_day, psp\_week, psp\_season, battery} \\
$l \in L$            &\texttt{l}      & feasible operation region limits & \texttt{l1, l2, l3, l4}\\
$m \in M$            &\texttt{m}      & energy products & \texttt{el, ht} \\
$n \in N$            &\texttt{n}      & intermittent generators & \texttt{wind\_on, wind\_off, pv, ror} \\
$t \in T$            &\texttt{t}      & time periods (hours)               & \texttt{t1, t2, \ldots, t8760}\\
$z \in Z$            &\texttt{z}      & market zones & \texttt{AT, DE} \\
\end{longtable}

\newpage
\subsection{Parameters}\label{parameters}
Parameters are denoted either by lower-case greek letters or by upper-case latin letters.

\begin{longtable}{p{0.15\textwidth}p{0.3\textwidth}p{0.35\textwidth}p{0.15\textwidth}}
\caption{Parameters}\\
\toprule
mathematical symbol & programming symbol & description & unit\\
\midrule
\endhead
\bottomrule
\multicolumn{4}{c}{\textit{continued on next page}} \endfoot
\bottomrule
\endlastfoot
% the contents of the table
$\delta_{z,zz}$              &\texttt{DISTANCE(z,zz)}                           & distance between countries' center of gravity & km                        \\
$\varepsilon_{f}$            &\texttt{CO2\_INTENSITY(f)}                        & fuel emission intensity & $\text{t}_{\COO}$/ MWh  \\
$\eta_{i,m,f}$               &\texttt{EFFICIENCY\_G(i,m,f)}                     & power plant efficiency & MWh / MWh                 \\
$\eta^{out}_{z,k}$           &\makecell[l]{\texttt{EFFICIENCY\_S\_OUT(k)}}      & discharging efficiency &                           \\
$\eta^{in}_{z,k}$            &\makecell[l]{\texttt{EFFICIENCY\_S\_IN(k)}}       & charging efficiency &                           \\
$\lambda_{z}$                &\texttt{LAMBDA(z)}                                & scaling factor for peak load &                           \\
$\mu_{z}$                    &\texttt{VALUE\_NSE(z)}                            & value of lost load & \EUR/ MWh               \\
$\rho_{z,t,k}$               &\makecell[l]{\texttt{INFLOWS(z,t,k)}}             & inflows to storage reservoirs & MW                        \\
$\sigma_{z}$                 &\texttt{SIGMA(z)}                                 & scaling factor for peak intermittent generation &                           \\
$\phi_{z,t,n}$               &\texttt{GEN\_PROFILE(z,t,n)}                      & intermittent generation profile & $[0,1]$                  \\
$\widehat{\phi}_{z,n}$       &\texttt{PEAK\_PROFILE(z,n)}                       & peak intermittent generation profile & $[0,1]$                   \\
$\chi_{i,l,f}$               &\texttt{FEASIBLE\_INPUT(i,l,f)}                   & inputs of feasible operating region & $[0,1]$                  \\
$\psi_{i,l,m}$               &\texttt{FEASIBLE\_OUTPUT(i,l,m)}                  & output tuples of feasible operating region & $[0,1]$                  \\
% UPPER-CASE LATIN LETTERS
$C^{r}_{z,n}$                &\texttt{CAPITALCOST\_R(z,n)}                      & capital cost of intermittent generators (specific, annuity)     & \EUR/ MW                \\
$C^{g}_{z,i}$                &\texttt{CAPITALCOST\_G(z,i)}                      & capital cost of thermal generators (specific, annuity)          & \EUR/ MW                 \\
$C^{s}_{z,k}$                &\makecell[l]{\texttt{CAPITALCOST\_S(z,k)}}        & capital cost of storages - power (specific, annuity)            & \EUR/ MW                \\
$C^{v}_{z,k}$                &\makecell[l]{\texttt{CAPITALCOST\_V(z,k)}}        & capital cost of storages - energy (specific, annuity)           & \EUR/ MW                \\
$C^{x}$                      &\texttt{CAPITALCOST\_X}                           & capital cost of transmission capacity (specific, annuity)       & \EUR/ MW                \\
$D_{z,t,m}$                  &\texttt{DEMAND(z,t,m)}                            & energy demand & GW                        \\
$\widehat{D}_{z,m}$          &\texttt{PEAK\_LOAD(z,m)}                          & peak demand & GW                        \\
$\widetilde{G}_{z,i}$        &\texttt{INITIAL\_CAP\_G(z,tec)}                   & initial capacity of dispatchable generators & GW                        \\
$O^{g}_{i}$                  &\texttt{OM\_COST\_G\_VAR(i)}                      & variable O\&M cost of dispatchable generators & \EUR/ MWh               \\
$O^{r}_{z,n}$                &\texttt{OM\_COST\_R\_VAR(z,n)}                    & variable O\&M cost of intermittent generators & \EUR/ MWh               \\
$P^{e}_{t,z}$                &\texttt{PRICE\_CO2(t,z)}                          & \COO price &\EUR/$\text{t}_{\COO}$ \\
$P_{t,z,f}$                  &\texttt{PRICE\_FUEL(t,z,f)}                       & fuel price & \EUR/ MWh               \\
$Q^{g}_{i}$                  &\texttt{OM\_COST\_G\_QFIX(i)}                     & quasi-fixed O\&M cost of dispatchable generators & \EUR/ MW                \\
$Q^{r}_{z,n}$                &\texttt{OM\_COST\_R\_QFIX(z,n)}                   & quasi-fixed O\&M cost of intermittent generators & \EUR/ MW                \\
$\widetilde{R}_{z,n}$        &\texttt{INITIAL\_CAP\_R(z,n)}                     & initial capacity of intermittent generators & GW                        \\
$\widetilde{S}^{out}_{z,k}$  &\makecell[l]{\texttt{INITIAL\_CAP\_S\_OUT(z,k)}}  & initial discharging capacity of storages & GW                        \\
$\widetilde{S}^{in}_{z,k}$   &\makecell[l]{\texttt{INITIAL\_CAP\_S\_IN(z,k)}}   & initial charging capacity of storages & GW                        \\
$\widetilde{V}_{z,k}$        &\makecell[l]{\texttt{INITIAL\_CAP\_V(z,k)}}       & initial energy storage capacity &                           \\
$\widetilde{X}_{z,zz}$       &\texttt{INITIAL\_CAP\_X(z,zz)}                    & initial transmission capacity & GW                        \\
\end{longtable}

\newpage
\subsection{Variables}\label{variables}
Variables are denoted by lower-case latin letters.

\begin{longtable}{p{0.15\textwidth}p{0.3\textwidth}p{0.35\textwidth}p{0.15\textwidth}}
\caption{Variables}\\
\toprule
mathematical symbol & programming symbol & description & unit\\
\midrule
\endhead
\bottomrule
\multicolumn{4}{c}{\textit{continued on next page}} \endfoot
\bottomrule
\endlastfoot
% the contents of the table
$b_{z,t,i,f}$                &\texttt{b(z,t,i,f)}           & fuel burn for energy generation & GW        \\
$c$                          &\texttt{cost\_system}         & total system cost & k\EUR     \\
$c_{z}$                      &\texttt{cost\_zonal(z)}       & zonal system cost & k\EUR     \\
$c^{b}_{z,t,i}$              &\texttt{cost\_fuel(z,t,i)}    & fuel cost & k\EUR     \\
$c^{e}_{z,t,i}$              &\texttt{cost\_co2(z,t,i)}     & emission cost & k\EUR     \\
$c^{om}_{z,i}$               &\texttt{cost\_om\_g(z,i)}     & total o\&m cost of dispatchable generators & k\EUR     \\
$c^{om}_{z,n}$               &\texttt{cost\_om\_r(z,n)}     & total o\&m cost of intermittent generators & k\EUR     \\
$c^{g}_{z}$                  &\texttt{cost\_invest\_g(z)}   & capital cost of generators & k\EUR     \\
$c^{q}_{z}$                  &\texttt{cost\_nse(z)}         & total cost of non-served load & k\EUR     \\
$c^{r}_{z}$                  &\texttt{cost\_invest\_r(z)}   & capital cost of intermittent generators & k\EUR     \\
$c^{s,v}_{z}$                &\texttt{cost\_invest\_sv(z)}  & capital cost of storages & k\EUR     \\
$c^{x}_{z}$                  &\texttt{cost\_invest\_x(z)}   & capital cost of interconnectors & k\EUR     \\
$e_{z}$                      &\texttt{emission\_co2(z)}     & \COO emissions & t\COO    \\
$\widetilde{g}^{+}_{z,i}$    &\texttt{add\_g(z,i)}          & added capacity of dispatchables & GW        \\
$\widetilde{g}^{-}_{z,i}$    &\texttt{deco\_g(z,i)}         & decommissioned capacity of dispatchables & GW        \\
$g_{z,t,i,m,f}$              &\texttt{g(z,t,i,m,f)}         & energy generated by conventionals & GW        \\
$q^{+}_{z,t}$                &\texttt{q\_curtail(z,t)}      & curtailed energy & GW        \\
$q^{-}_{z,t,m}$              &\texttt{q\_nse(z,t,m)}        & non-served energy & GW        \\
$\widetilde{r}^{+}_{z,n}$    &\texttt{add\_r(z,n)}          & added capacity of intermittents & GW        \\
$\widetilde{r}^{-}_{z,n}$    &\texttt{deco\_r(z,n)}         & decommissioned capacity of intermittents & GW        \\
$r_{z,t,n}$                  &\texttt{r(z,t,n)}             & electricity generated by intermittents & GW        \\
$\widetilde{s}^{+}_{z,k}$    &\texttt{add\_s(z,k)}          & added storage capacity (power)              & GW        \\
$s^{in}_{z,t,k}$             &\texttt{s\_in(z,t,k)}         & energy stored in & GW        \\
$s^{out}_{z,t,k}$            &\texttt{s\_out(z,t,k)}        & energy stored out & GW        \\
$\widetilde{v}^{+}_{z,k}$    &\texttt{add\_v(z,k)}          & added storage capacity (energy)             & GWh       \\
$v_{z,t,k}$                  &\texttt{v(z,t,k)}             & storage energy content & GWh       \\
$w_{z,t,i,l,f}$              &\texttt{w(z,t,i,l,f)}         & operating region weight &           \\
$\widetilde{x}^{+}_{z,zz}$   &\texttt{add\_x(z,zz)}         & added transmission capacity & GW        \\
$x_{z,zz,t}$                 &\texttt{x(z,zz,t)}            & electricity net export & GW        \\
\end{longtable}

\newpage
\subsection{Naming system}
\begin{table*}
\centering
\begin{threeparttable}
\caption{Naming System}
\begin{tabulary}{\textwidth}{LLLLLL}
\toprule
& initial capacity\tnote{$\dagger$}    & added capacity\tnote{$\ddagger$}  & decommissioned capacity\tnote{$\ddagger$} & specific investment cost\tnote{$\dagger$} & dispatch\tnote{$\ddagger$}   \\
\midrule
thermal units &$\widetilde{G}_{z,i}$                &$\widetilde{g}^{+}_{z,i}$         &$\widetilde{g}^{-}_{z,i}$                &$C^{g}_{z,i}$                            &$g_{z,t,i,m,f}$              \\
intermittent units &$\widetilde{R}_{z,n}$                &$\widetilde{r}^{+}_{z,n}$        &$\widetilde{r}^{-}_{z,n}$                &$C^{r}_{z,n}$                            &$r_{z,t,n}$                  \\
storages (power)    &$\widetilde{S}_{z,k}$                &$\widetilde{s}^{+}_{z,k}$        & NA &$C^{s}_{z,k}$                             &$s_{z,t,k}$                  \\% $\widetilde{s}^{-}_{z,k}$
storages (energy)   &$\widetilde{V}_{z,k}$                &$\widetilde{v}^{+}_{z,k}$        & NA &$C^{v}_{z,k}$                            & NA                            \\% $\widetilde{v}^{-}_{z,t,k}$
transmission &$\widetilde{X}_{z,zz}$               &$\widetilde{x}^{+}_{z,zz}$       & NA &$C^{x}_{z,zz}$                           &$x_{z,zz,t}$                 \\% $\widetilde{x}^{-}_{z,zz}$
\bottomrule
\end{tabulary}

\begin{tablenotes}
\item[$\dagger$] parameter
\item[$\ddagger$] variable
\end{tablenotes}
\end{threeparttable}
\end{table*}

\newpage

\subsection{Mathematical description}\label{mathmodel}

\paragraph{Model objective}
\emph{medea} minimizes total system cost $c$, i.e. the total cost of generating electricity and heat from
technologies and capacities adequate to meet demand, over a large number of decision variables, essentially
representing investment and dispatch decisions in each market zone$z$of the modelled energy systems.
\begin{align}
\min c =\sum_{z} (c_{z})
\end{align}
Zonal system costs $c_{z}$ are the sum of fuel cost $c^{b}_{z,t,i}$, emission cost $c^{e}_{z,t,i}$, operation and
maintenance cost, capital costs of investment in conventional and intermittent generation ($c^{g}_{z}$,$c^{r}_{z}$),
storage ($c^{s,v}_{z}$) and transmission ($c^{x}_{z}$) equipment, and the cost of non-served load ($c^{q}_{z}$) that
accrues when demand is not met, e.g. when there is a power outage.
\begin{align}
%\begin{split}
c_{z} =\sum_{t,i}  c^{b}_{z,t,i} +\sum_{t,i} c^{e}_{z,t,i} +\sum_{i} c^{om}_{z,i} +\sum_{n} c^{om}_{z,n} + c^{g}_{z} +
c^{r}_{z} + c^{s,v}_{z} + c^{x}_{z} + c^{q}_{z} \qquad \qquad \forall z
%\end{split}
\end{align}
The components of zonal system costs are calculated as given in equations (\ref{fuel_cost}) to (\ref{lost_load_cost}).
Lower-case $c$ represent total cost, while upper-case$C$denotes specific, annualized capital cost of technology investment.
Prices for fuels and \COO are denoted by$P$.
\begin{align}
&c^{b}_{z,t,i}& =&\ \sum_{f}\left( P_{t,z,f}\:b_{t,z,i,f}\right)\qquad \qquad&\forall z,t,i\label{fuel_cost}\\
&c^{e}_{z,t,i}& =&\ \sum_{f}\left( P^{e}_{t,z}\:e_{z,t,i}\right)\qquad \qquad&\forall z,t,i\\
%&c^{om}_{z,i}& =&\Q^{g}_{i}\left( \widetilde{G}_{z,i} - \widetilde{g}^{-}_{z,i} + \widetilde{g}^{+}_{z,i}\right) + \sum_{t}\sum_{m}\sum_{f}\left( O^{g}_{i}\:g_{z,t,i,m,f}\right)\qquad \qquad&\forall z,i\\
%&c^{om}_{z,n}& =&\Q^{r}_{n}\left( \widetilde{R}_{z,n} - \widetilde{r}^{-}_{z,n} + \widetilde{r}^{+}_{z,n}\right) + \sum_{t}\left( O^{r}_{n}\:r_{z,t,n}\right)\qquad \qquad&\forall z,n\\
%&c^{g}_{z}& =&\ \sum_{i}\left( C^{g}_{z,i}\: \widetilde{g}^{+}_{z,i}\right)\qquad \qquad&\forall z\\
%&c^{r}_{z}& =&\ \sum_{n}\left( C^{r}_{z,n}\: \widetilde{r}^{+}_{z,n}\right)\qquad \qquad&\forall z\\
&c^{s,v}_{z}& =&\ \sum_{k}\left( C^{s}_{z,k}\: \widetilde{s}^{+}_{z,k} + C^{v}_{z,k} \:v^{+}_{z,k}\right)\qquad \qquad&\forall z\\
&c^{x}_{z}& =&\ \frac{1}{2}\: \sum_{zz} (C^{x}\: \delta_{z,zz}\: \widetilde{x}^{+}_{z,zz})\qquad \qquad&\forall z\label{transmission_expansion_cost}\\
&c^{q}_{z}& =&\ \mu \sum_{t}\sum_{m} q^{-}_{z,t,m}\qquad \qquad&\forall z\label{lost_load_cost}
\end{align}

\paragraph{Market clearing}
In each hour, the markets for electricity and heat have to clear.
Equation (\ref{market_clearing_el}) ensures that the total supply from conventional and intermittent sources, and
storages equals total electricity demand plus net exports, electricity stored and used for heat generation.
Likewise, equation (\ref{market_clearing_ht}) clears the heat market by equating heat generation to heat demand.
\begin{align}
\begin{split}
\sum_{i}\sum_{f} g_{z,t,i,\text{el},f} + \sum_{k} s^{out}_{z,t,k} + \sum_{n} r_{z,t,n} &= \\D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + & \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} - q^{-}_{z,t,\text{el}} + q^{+}_{z,t} \qquad \forall z,t
\end{split}
\label{market_clearing_el}
\end{align}
\begin{align}
\sum_{i}\sum_{f} g_{z,t,i,\text{ht},f} = D_{z,t,\text{ht}} - q^{-}_{z,t,\text{ht}}\qquad \forall z,t\label{market_clearing_ht}
\end{align}
\emph{medea} can be thought of as representing energy-only electricity and heat markets without capacity payments.
Then, the marginals of the market clearing equations (\ref{market_clearing_el}) and (\ref{market_clearing_ht}),
$\partial C / \partial D_{z,t,m}$, can be interpreted as the zonal prices for electricity and heat, respectively.

\paragraph{Energy generation}
Energy generation $g_{z,t,i,m,f} \geq 0$ is constrained by available installed capacity, which can be adjusted
through investment ($\widetilde{g}^{+}_{z,i} \geq 0$) and decommissioning $\widetilde{g}^{-}_{z,i} \geq 0$.
\begin{align}
\sum_{f} g_{z,t,i,m,f}\leq \widetilde{G}_{z,i} +\widetilde{g}^{+}_{z,i} -\widetilde{g}^{-}_{z,i}\qquad \qquad \forall z,t,i,m
\end{align}
Generator efficiency $\eta$ determines the amount of fuel$b_{z,t,i,f} \geq 0$that needs to be spent in order to
generate a given amount of energy.
\begin{align}
g_{z,t,i,m,f} = \sum_{f}\eta_{i,m,f}\:b_{z,t,i,f}\qquad \qquad \forall z,t,i \notin J, f
\end{align}

\paragraph{Thermal co-generation}
Co-generation units jointly generate heat and electricity. All feasible combinations of heat and electricity
generation along with the corresponding fuel requirement are reflected in so-called `feasible operating regions'.
The elements $l \in L$ span up a three-dimensional, convex feasible operating region for each co-generation technology.
The weights $w_{z,t,i,l,f} \geq 0$ form a convex combination of the corners $l$, which are scaled to the available
installed capacity of each co-generation technology.
Defining weights over fuels allows co-generation units to switch fuels between multiple alternatives.
Heat and electricity output along with the corresponding fuel requirement is then set according to the chosen weights.
\begin{align}
\sum_{l}\sum_{f} w_{z,t,i,l,f} =\widetilde{G}_{z,i} +\widetilde{g}^{+}_{z,i} -\widetilde{g}^{-}_{z,i}\qquad \qquad \forall z,t,i \in J\\
g_{z,t,i,m,f} =\sum_{l}\sum_{f}\psi_{i,l,m}\:w_{z,t,i,l,f}\qquad \qquad \forall z,t,i \in J, m\\
b_{z,t,i,f} =\sum_{l}\chi_{i,l,f}\:w_{z,t,i,l,f}\qquad \qquad \forall z,t,i \in J, f\\
w(z,t,i,l,f) = 0\qquad \qquad \forall z,t,i,k,f:\chi_{i,l,f} = 0
\end{align}

\paragraph{Intermittent electricity generation}
Electricity generation from intermittent sources wind (on-shore and off-shore), solar irradiation, and river runoff
follows generation profiles $\phi_{z,t,n} \in [0,1]$ and is scaled according to corresponding installed
($ \widetilde{R}_{z,n}$) and added ($\widetilde{r}^{+}_{z,n} \geq 0$) capacity.
\begin{align}
r_{z,t,n} = \phi_{z,t,n}\: \left( \widetilde{R}_{z,n} -\widetilde{r}^{-}_{z,n} +\widetilde{r}^{+}_{z,n}\right)\qquad \qquad \forall z,t,n
\end{align}

\paragraph{Electricity storages}
Charging ($s^{in}_{z,t,k} \geq 0$) and discharging ($s^{out}_{z,t,k} \geq 0$) of storages is constrained by the
storages' installed ($\widetilde{S}^{in}_{z,k}, \widetilde{S}^{out}_{z,k}$) and added
($\widetilde{s}^{+}_{z,k} \geq 0$) charging and discharging power, respectively.
Similarly, the total energy that can be stored is constrained by the storage technology's initial
($\widetilde{V}_{z,k}$) and added ($\widetilde{v}^{+}_{z,k} \geq 0$) energy capacity.
\begin{align}
s^{out}_{z,t,k} \leq \widetilde{S}^{out}_{z,k} + \widetilde{s}^{+}_{z,k}\qquad \qquad \forall z,t,k\\
s^{in}_{z,t,k}\leq \widetilde{S}^{in}_{z,k} + \widetilde{s}^{+}_{z,k}\qquad \qquad \forall z,t,k\\
v_{z,t,k}\leq \widetilde{V}_{z,k} +\widetilde{v}^{+}_{z,k}\qquad \qquad \forall z,t,k
\end{align}
Storage operation is subject to a storage balance, such that the current energy content must be equal to the
previous period's energy content plus all energy flowing into the storage less all energy flowing out of the storage.
\begin{align}
v_{z,t,k} = \rho_{z,t,k} + \eta^{in}_{z,k}\:s^{in}_{z,t,k} - (\eta^{out}_{z,k})^{-1}\:s^{out}_{z,t,k} + v_{z,t-1,k}\qquad \qquad \forall z,t,k: t>1,\: \eta^{out}_{z,k} > 0
\end{align}
Since the model can add storage power capacity and energy capacity independently, we require a storage to hold at
least as much energy as it could store in (or out) in one hour.
\begin{align}
\widetilde{v}^{+}_{z,k}\geq \widetilde{s}^{+}_{z,k}\qquad \qquad \forall z,k
\end{align}

\paragraph{Emission accounting}
Burning fossil fuels for energy generation produces emissions of carbon dioxide (\COO). The amount of \COO emitted
is tracked by the following equation
\begin{align}
e_{z,t,i} = \sum_{f}\left(\varepsilon_{f}\:b_{z,t,i,f}\right)\qquad \qquad \forall z,t,i
\end{align}

\paragraph{Electricity exchange}
Implicitly, \emph{medea} assumes that there are no transmission constraints within market zones.
However, electricity exchange between market zones is subject to several constraints.

First, exchange between market zones is constrained by available transfer capacities. Transfer capacities can be
expanded at constant, specific investment cost (see equation (\ref{transmission_expansion_cost})).
This rules out economies of scale in transmission investment that might arise in interconnected, meshed grids.
\begin{align}
x_{z,zz,t} \leq \widetilde{X}_{z,zz} + \widetilde{x}^{+}_{z,zz}\qquad \qquad \forall z, zz, t\\
x_{z,zz,t}\geq-\left( \widetilde{X}_{z,zz} +\widetilde{x}^{+}_{z,zz}\right)\qquad \qquad \forall z, zz, t
\end{align}
By definition, electricity net exports $x_{z,zz,t}$ from $z$ to $zz$ must equal electricity net imports of $zz$ from $z$.
\begin{align}
x_{z,zz,t} = -x_{zz,z,t} \qquad \qquad \forall z, zz, t
\end{align}
Added transmission capacities can be used in either direction.
\begin{align}
\widetilde{x}^{+}_{z,zz} = \widetilde{x}^{+}_{zz,z}\qquad \qquad \forall z, zz
\end{align}
Finally, electricity cannot flow between zones where there is no transmission infrastructure in place (including
intra-zonal flows).
\begin{align}
x_{z,zz,t} = 0 \qquad \qquad \forall z, zz, t:\widetilde{X}_{z,zz} = 0\\
x_{zz,z,t} = 0\qquad \qquad \forall z, zz, t:\widetilde{X}_{z,zz} = 0
\end{align}

\paragraph{Decommissioning}
Keeping plant available for generation gives rise to quasi-fixed operation and maintenance costs.
Such cost can be avoided by decommissioning an energy generator. This is modelled as a reduction in generation
capacity, which cannot exceed installed capacity.
\begin{align}
\widetilde{g}^{-}_{z,i} &\leq \widetilde{G}_{z,i} + \widetilde{g}^{+}_{z,i}\qquad \qquad \forall z,i\\
\widetilde{r}^{-}_{z,n} &\leq \widetilde{R}_{z,n} +\widetilde{r}^{+}_{z,n}\qquad \qquad \forall z,n
\end{align}

\paragraph{Ancillary services}
Power systems require various system services for secure and reliable operation, such as balancing services or
voltage support through the provision of reactive power. Such system services can only be supplied by operational
generators.
Thus, we approximate system service provision by a requirement on the minimal amount of spinning reserves operating
at each hour.
We assume that ancillary services are supplied by conventional (thermal) power plant, hydro power plant, and storage.
The requirement for spinning reserves is proportional to electricity peak load
$\widehat{D}_{z,\text{el}} = \max_{t} D_{z,t,\text{el}}$ and peak generation from wind and solar resources,
where $\widehat{\phi}_{z,n} = \max_{t} \phi_{z,t,n}$.
\begin{align}
\sum_{i}\sum_{f}\left( g_{z,t,i,\text{el},f}\right) + r_{z,t,\text{ror}}
+ \sum_{k}\left( s^{out}_{z,t,k} + s^{in}_{z,t,k}\right)\geq \lambda_{z}\widehat{D}_{z,\text{el}}
+ \sigma_{z}\sum_{n\setminus \{ \text{ror}\}}\widehat{\phi}_{z,n} (\widetilde{R}_{z,n}
+\widetilde{r}^{+}_{z,n})\qquad \forall z,t
\end{align}

\paragraph{Curtailment}
Electricity generated from intermittent sources can be curtailed (disposed of) without any further cost (apart from
implicit opportunity cost).
\begin{align}
q^{+}_{z,t} \leq \sum_{n} r_{z,t,n} \qquad \qquad \forall z, t
\end{align}

%\section{Effects of doubling transmission capacity} \label{sec:transmission-sensitivity}
\end{document}

### 2020-06-19 projects/asparagus/doc/plots_asparagus.py

@@ -56,12 +56,33 @@ capa = results.loc[idx['base', :, :, 36715], idx['AT', :]]
 capa.columns = capa.columns.droplevel(0)
 capa = capa.loc[:, (capa.columns.str.contains('add_g') | capa.columns.str.contains('deco'))]
 capa.columns = pd.MultiIndex.from_tuples(list(capa.columns.str.split('_g_')))
+capa.loc[:, idx['deco', :]] = - capa.loc[:, idx['deco', :]]
 capa = capa.groupby(level=1, axis=1).sum()
 capa.columns.name = 'variable'
 capa.index = capa.index.rename(['scenario', 'co2price', 'wind_lim', 'pv_cost'])
 capa = (capa + capi.T).dropna(axis=1, how='all')
+capa = capa.replace(0, np.nan).dropna(axis=1, how='all').replace(np.nan, 0)
 
-# plot_lines(capa, os.path.join(FPATH, 'capacity.pdf'),color=REFUEL_COLORS)
+plot_lines(capa, os.path.join(FPATH, 'capacity.pdf'), color=REFUEL_COLORS)
+
+# %% analyze system operation (net exports, curtailment, generation)
+var_of_interest = ['AnnCO2Emissions', 'AnnCurtail', 'AnnCurtailShare', 'AnnGBiomass', 'AnnGFossil_el', 'AnnG_el',
+                   'AnnR', 'AnnSIn', 'AnnSOut', 'AnnX']
+
+oper = results.loc[idx['base', :, :, 36715], idx['AT', :]]
+oper.loc[:, idx['AT', 'AnnX']] = results.loc[:, idx['DE', 'AnnX']]
+oper.columns = oper.columns.droplevel(0)
+oper['AnnGFossil_el'] = oper['AnnG_el'] - oper['AnnGBiomass']
+oper['AnnCurtailShare'] = oper['AnnCurtail'] / oper['AnnR']
+oper.index = oper.index.droplevel([0, 3])
+
+for var in var_of_interest:
+    plot_lines(oper.loc[:, var].unstack(0), os.path.join(FPATH, f'oper_{var}.pdf'), color=REFUEL_COLORS)
+
+# oper = oper.loc[:, oper.columns.str.contains('AnnGByTec')]
+# oper.columns = oper.columns.str.replace('AnnGByTec_', '')  #.str.replace('(', '').str.replace(')', '')
+# oper.columns = pd.MultiIndex.from_tuples(list(oper.columns.str.split('_g_')))
+
 
[…]

### 2020-07-07 src/tools/export2gdx.py
beginning to rewrite data export to gdx

@@ -1,0 +1,152 @@
+import logging
+import os
+
+import pandas as pd
+from gams import *
+
+import config as cfg
+from src.tools.gams_io import df2gdx
+from src.tools.preprocess_data import dict_sets, plant_data, ts_data
+
+# TODO: Add energy stored in hydro reservoirs - STORAGE_LEVEL
+
+idx = pd.IndexSlice
+# --------------------------------------------------------------------------- #
+# %% create workspace and database
+# --------------------------------------------------------------------------- #
+ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
+db = ws.add_database()
+
+# --------------------------------------------------------------------------- #
+# %% instantiate SETS
+# --------------------------------------------------------------------------- #
+f_set = df2gdx(db, dict_sets['f'], 'f', 'set', [])
+l_set = df2gdx(db, dict_sets['l'], 'l', 'set', [])
+m_set = df2gdx(db, dict_sets['m'], 'm', 'set', [])
+z_set = df2gdx(db, dict_sets['z'], 'z', 'set', [])
+k_set = df2gdx(db, dict_sets['k'], 'k', 'set', [])
+n_set = df2gdx(db, dict_sets['n'], 'n', 'set', [])
+t_set = df2gdx(db, dict_sets['t'], 't', 'set', [])
+i_set = df2gdx(db, dict_sets['i'], 'i', 'set', [])
+j_set = df2gdx(db, pd.DataFrame.from_dict({x: True for x in [y for y in dict_sets['i'].index if 'chp' in y]},
+                                          orient='index'), 'j', 'set', [])
+h_set = df2gdx(db, pd.DataFrame.from_dict({x: True for x in [y for y in dict_sets['i'].index if 'pth' in y]},
+                                          orient='index'), 'h', 'set', [])
+logging.info('medea sets instantiated')
+
+# --------------------------------------------------------------------------- #
+# %% selectors
+# --------------------------------------------------------------------------- #
+intermittent = plant_data['technology'].loc[plant_data['technology']['intermittent'] == 1].index
[…]

### 2020-07-07 src/tools/preprocess_data.py

@@ -1,0 +1,134 @@
+# %% imports
+import os
+
+import numpy as np
+import pandas as pd
+
+import config as cfg
+from src.tools.data_processing import hours_in_year
+
+# --------------------------------------------------------------------------- #
+# %% settings and initializing
+# --------------------------------------------------------------------------- #
+STATIC_FNAME = os.path.join(cfg.MEDEA_ROOT_DIR, 'data', 'processed', 'data_static.xlsx')
+idx = pd.IndexSlice
+
+### ### ### processes data and writes: dict_sets, dict_instantiate, static_data, plant_data, ts_data, invest_limits
+###
+### dict_sets includes: f, i, l, m, n, k, t, z
+### dict_instantiate includes:  'efficiency', 'DISTANCE', 'tec_props', 'ancil',
+#
+### static_data includes:       'potentials', 'tec', 'feasops', 'cost_transport',
+#
+### plant_data includes:        'hydro', 'conventional', 'active', 'hydro_clusters', 'storage_clusters'
+### ts_data includes:           'timeseries', 'zonal', 'price', 'inflows'
+### invest_limits includes:     'thermal', 'intermittent', 'storage', 'atc'
+
+# 'CAP_G', 'CAP_R', 'CAP_S_OUT', 'CAP_S_IN', 'CAP_V', 'CAP_X', 'CAPCOST_R', 'CAPCOST_K', 'CAPCOST_X',
+# 'EFFICIENCY_G', 'EFFICIENCY_S_OUT', 'EFFICIENCY_S_IN'
+#
+# 'FEASIBLE_INPUT', 'FEASIBLE_OUTPUT'
+#
+# 'OM_COST_G_QFIX', 'OM_COST_G_VAR', 'OM_COST_R_QFIX', 'OM_COST_R_VAR'
+#
+# 'AIR_POLLUTION', 'VALUE_NSE', 'LAMBDA', 'SIGMA', 'CO2_INTENSITY', 'PEAK_LOAD', 'PEAK_PROFILE', VALUE_NSE
+#
+# 'DEMAND', 'GEN_PROFILE', 'INFLOWS',
+
+# INITIAL_CAP_G, INITIAL_CAP_R, INITIAL_CAP_S_OUT, INITIAL_CAP_S_IN, INITIAL_CAP_V, INITIAL_CAP_X
+
+# OM_COST_G_QFIX, OM_COST_G_VAR, OM_COST_R_QFIX, OM_COST_R_VAR,
[…]

### 2021-01-08 projects/asparagus/doc/asparagus.tex
beginning to rewrite data export to gdx
updates asparagus.tex to version submitted on Dec 8, 2020

@@
-\documentclass[review, 3p, times, 12pt]{elsarticle} %final, twocolumn, 5p, 12pt, authoryear, twocolumn, 5p,
+\documentclass[review, 3p, times, 12pt, authoryear]{elsarticle}
 
@@
 
-% improved float environment
 \usepackage{float}
-% improved underlines (unterstreichungen)
 \usepackage[normalem]{ulem}
 
-% plotting
 \usepackage{tikz}
@@
 
-% tables
 \usepackage{threeparttable}
@@
 
-\newcolumntype{M}{>{\raggedright\arraybackslash}X} % Defines new column type with variable length, similar to X in tabularx, but is left aligned rather than justified
-
-%\usepackage{ctable}
+\newcolumntype{M}{>{\raggedright\arraybackslash}X}
 \usepackage{makecell}
 
-% SI units
 \usepackage{siunitx}
-\DeclareSIUnit{\thermal}{th}
-\DeclareSIUnit{\electric}{el}
+\DeclareSIUnit{\wattth}{Wth}
+\DeclareSIUnit{\wattel}{We}
 \DeclareSIUnit{\year}{a}
@@
 \DeclareSIUnit{\million}{million}
-
-% Euro symbol
+\DeclareSIUnit{\coo}{\ce{CO2}e}
+
 \usepackage[gen]{eurosym}
-
-% citations
-\usepackage{natbib}
-
-% to-do notes
-\setlength{\marginparwidth}{2cm}
-\usepackage[draft]{todonotes}
-
-% graphics and figures
+\usepackage[version=4]{mhchem}
+\defcitealias{BMNT2018}{BMNT and BMVIT, 2018}
+\defcitealias{Regierungsprogramm2020}{BKA, 2020}
+\defcitealias{WSB2019}{WSB, 2019}
+\defcitealias{ENTSOE2020a}{ENTSO-E, 2020a}
+\defcitealias{ENTSOE2020b}{ENTSO-E, 2020b}
+\defcitealias{AGEB2018}{AGEB, 2018}
+\defcitealias{CCCS2017}{C3S, 2017}
+\defcitealias{IMF2020}{IMF (2020)}
+\defcitealias{opsd2019}{OPSD (2019)}
+
 \usepackage{graphicx}
+\graphicspath{{./figures/}}
 \usepackage{caption}
@@
 \DeclareMathOperator*{\argmax}{arg\,max}
-\newcommand{\COO}{\ensuremath{\mathrm{CO_2}} }
-
-% neater date
+
 \usepackage{etoolbox}
-\patchcmd{\MaketitleBox}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}%
-\patchcmd{\pprintMaketitle}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}%
-
-\usepackage[colorlinks=true, citecolor=blue, linkcolor=blue, filecolor=blue,urlcolor=blue,pdfauthor=author]{hyperref}
+\patchcmd{\MaketitleBox}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}
+\patchcmd{\pprintMaketitle}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}
+
+\usepackage[hidelinks, colorlinks=false, citecolor=blue, linkcolor=blue, filecolor=blue,urlcolor=blue,pdfauthor=Sebastian Wehrle]{hyperref}
+
+\newcommand*{\figref}[2][]{%
+    \hyperref[{fig:#2}]{%
+        Figure~\ref*{fig:#2}%
+        \ifx\\#1\\%
+        \else
+        \,#1%
+        \fi
+    }%
+}
 
@@
         \ead{johannes.schmidt@boku.ac.at}
-\author[1]{Christian Mikovits}
-\ead{christian.mikovits@boku.ac.at}
         \cortext[cor1]{Corresponding author}
@@
         \begin{abstract}
-By 2030 Austria aims to meet \SI{100}{\percent} of its electricity demand from domestic renewable sources, predominantly from wind and solar energy.
-While wind power reduces \COO emissions, it is also connected to negative impacts at the local level, such as interference with landscape aesthetics.
-Yet, wind power comes at lower system integration cost than solar power, so that it effectively reduces system cost.
-We quantify the opportunity cost of replacing wind turbines with solar power, using the power system model \emph{medea}.
-Our findings suggest that these cost of undisturbed landscapes are considerable, particularly when PV is not entirely rolled out as utility-scale, open-space installations.
-The opportunity cost is likely high enough to allow for significant compensation of the ones affected by local wind turbine externalities.
+            By 2030, Austria aims to meet \SI{100}{\percent} of its electricity demand from domestic renewable sources, such as wind energy.
+            Apart from the benefit of reducing \ce{CO2} emissions and, potentially, system cost, wind power is also associated with negative impacts at the local level, particularly its interference with landscape aesthetics.
+            Some of these negative impacts might be avoided by using alternative renewable energy technologies.
+            To better understand the trade-offs involved, we quantify the opportunity cost of wind power versus its best feasible alternative, which is solar photovoltaics in Austria.
+Methodologically, our analysis relies on the power system model \emph{medea}, a technology-rich, partial-equilibrium model of the Austro-German electricity and district heating markets.
+            Our findings suggest that the cost of undisturbed landscapes is considerable, particularly when solar PV is mainly realized on rooftops.
+            Under a wide range of assumptions, the opportunity cost of wind power is high enough to allow for significant compensation of the ones affected by local, negative wind turbine externalities.
         \end{abstract}
@@
     \newpage
-
 % %%%%% %%%%% %%%%% INTRODUCTION %%%%% %%%%% %%%%% %%%%%
+
     \section{Introduction}\label{sec:introduction}
     "Holding the increase in the global average temperature [\ldots] well below \SI{2}{\celsius} above pre-industrial levels and pursuing efforts to limit the temperature increase to \SI{1.5}{\celsius} above pre-industrial levels"~\citep{UNFCCC2015} requires rapid and substantial reductions in greenhouse gas emissions worldwide~\citep{IPCC2018}.
-Increasing the share of electricity generated from renewable sources consequently is amongst the United Nations's sustainable development goals (indicator 7.\ 2.\ 1) and a prominent goal of many climate and energy strategies.
-The European Union, for example, aims to meet at least \SI{32.5}{\percent} of its final energy consumption from renewable sources~\citep{EU2018}.
-Against this background, Austria aims to generate \SI{100}{\percent} of its electricity consumption from domestic renewable sources on annual balance by 2030\footnote{Industry own use and ancillary service provision are exempt.
+    Increasing the share of electricity generated from renewable sources consequently is amongst the United Nations' sustainable development goals (indicator 7.2.1) and a prominent goal of many climate and energy strategies.
+    The European Union, for example, aims to meet at least \SI{32}{\percent} of its final energy consumption from renewable sources by 2030~\citep{EU2018}.\footnote{In its "European Green Deal" the EU Commission proposes to increase ambitions and reduce \ce{CO2} emissions by at least \SI{55}{\percent} by 2030.
+This compares to a \SI{40}{\percent} reduction target previously and is likely to have knock-on effects on the expansion of renewable energy generation.}
+    Against this background, Austria aims to generate \SI{100}{\percent} of its electricity consumption from domestic renewable sources on annual balance by 2030\footnote{Industry's own use and ancillary service provision are exempt.
 Together, these accounted for approximately \SI{10}{\percent} of consumption in 2016.}.
-According to government estimates, additional electricity generation of \SI{27}{\tera\watt\hour} from renewable
-sources is required to meet policy goals.
-Out of this total increase in renewable electricity generation, \SI{1}{\tera\watt\hour} is thought to be sourced from biomass, and \SI{5}{\tera\watt\hour} from new hydropower generation, according to the government's climate and
-energy strategy~\citep{BMNT2018}.
-The comparatively small projected increase in electricity generation from these sources results from the perception
-of biomass being ecologically~\citep{Erb2018} and economically unsustainable, while further development of Austria's
-hydropower potentials was considered `challenging'~\citep{Wagner2015} even before the projected increase in hydro
-generation by \SI{5}{\tera\watt\hour} was factored in.
-As Austrians banned the use of nuclear power in a referendum in 1978~\citep{Pelinka1983}, wind and solar
-power are the only remaining options for a large-scale expansion of low-carbon power generation in the country.
-
-While wind power mitigates negative external effects at the global (\COO emissions) and the regional (air pollutants
-from fossil thermal power generation) level, it also imposes negative external effects at the local level.
-Wind turbines are found to negatively impact wildlife~\cite{Loss2013,Voigt2015,Wang2015}, to emit
-noise~\cite{Wang2015a}, and to interfere negatively with landscape aesthetics~\citep{Jones2010, Meyerhoff2010}.
-The latter is found to be the dominant negative external effect of wind turbines~\cite{Mattmann2016}.
-
-Accordingly, several strands of literature seek to quantify the negative impact of wind turbines, for example
-through their impact on property prices within a revealed preferences framework~\citep{Droees2016, Gibbons2015,
-Heintzelmann2017, Jensen2018, Kussel2019, Lang2014, Sims2008, Sunak2016, Vyn2014}, or through choice
-experiments and surveys~\citep{Drechsler2011, Mattmann2016, Meyerhoff2010} in a stated preferences-context.
-Zerrahn~\cite{Zerrahn2017} provides a comprehensive survey of the literature on wind power externalities.
-
-Apart from negative external effects at the local level, wind power is also found to have the potential to reduce
-electricity system cost relative to solar PV.
-For the case of Germany, Ueckerdt et al.~\cite{Ueckerdt2013} estimate solar PV integration cost being about four times higher than the integration cost of wind power at penetration levels of \SI{25}{\percent}.
-Scholz et al.
-find \cite{Scholz2017} integration cost of \SI{40}[\euro]{\per \mega\watt\hour} for a mix of \SI{80}{\percent} solar PV and \SI{20}{\percent} wind power at \SI{100}{\percent} renewables penetration in Europe.
-This reduces to below \SI{25}[\euro]{\per\mega\watt\hour} when the technology mix is changed to \SI{20}{\percent} solar PV and \SI{80}{\percent} wind power.
-% Similarly, Gils et al. \cite{Gils2017}
-
-These findings illustrate that abandoning wind power in favor of undisturbed landscapes potentially comes at significant opportunity cost.
-We complement the literature on negative wind turbines impacts and set out to quantify this cost of undisturbed landscapes for the case of Austria.
-First, Austrian policies targeting at a fixed share of renewable electricity generation are an ideal framework for
-studying the effects of substituting renewable energy generation technologies (RET).
+    The required expansion of renewable energy generation shall be incentivized by technology-specific subsidies rather than \ce{CO2} pricing.
+
+    According to government estimates, an additional \SI{27}{\tera\watt\hour} of electricity needs to be generated from renewable sources annually to meet this policy goal.
+    Out of the targeted total increase in renewable electricity generation, \SI{1}{\tera\watt\hour} is thought to be sourced from biomass, and \SI{5}{\tera\watt\hour} from new hydropower generation, according to the government's climate and energy strategy~\citepalias{BMNT2018}.
+    The comparatively small projected increase in electricity generation from these sources results from the perception of biomass being ecologically\footnote{More specifically, increasing biomass harvests would decrease carbon stocks strongly, both through deforestation and management.
+Thus, a net positive greenhouse gas balance of bioenergy would only be possible after a rather long payback time.
+However, additions to the stock of atmospheric greenhouse gases need to be reduced rapidly.}
+    ~\citep{Erb2018} and economically unsustainable\footnote{The temporary expiration of Austria's subsidy scheme for biomass plants in 2019 resulted in a widespread shutdown of biomass-fired generators.}
+    , while further development of Austria's hydropower potential was considered `challenging'~\citep{Wagner2015} even before the projected \SI{5}{\tera\watt\hour} increase in hydro generation was factored in.
+    As Austrians banned the use of nuclear power in a referendum in 1978~\citep{Pelinka1983}, wind and solar power are the only remaining options for a large-scale expansion of low-carbon power generation in the country.
+
+    While wind power mitigates negative external effects at the global (\ce{CO2} emissions) and the regional (air pollutants from fossil thermal power generation) level, it also imposes negative external effects at the local level.
+    Wind turbines are found to impact wildlife negatively~\citep{Loss2013,Voigt2015,Wang2015} and to emit noise~\citep{Wang2015a}.
+    They are also associated with land-use and land-cover change~\citep{Turkovska2020} and with negative interference with landscape aesthetics in general~\citep{Jones2010, Meyerhoff2010}.
+    The latter is found to be the dominant, negative external effect of wind turbines~\citep{Mattmann2016}.
+    Accordingly, several strands of literature seek to quantify the negative impact of wind turbines, for example through their effect on property prices within a revealed preference framework~\citep{Droees2016, Gibbons2015,
+    Heintzelmann2017, Jensen2018, Kussel2019, Lang2014, Sims2008, Sunak2016, Vyn2014}, or through choice experiments and surveys~\citep{Drechsler2011, Mattmann2016, Meyerhoff2010} in a stated preferences-context.
+    \cite{Zerrahn2017} provides a comprehensive survey of the literature on wind power externalities.
+
+    Apart from negative external effects at the local level, wind power is also found to have the potential to reduce electricity system costs, in particular, relative to solar PV.
+    For the case of Germany, \cite{Ueckerdt2013} estimate solar PV integration cost to be about four times higher than the integration cost of wind power at penetration levels of \SI{25}{\percent}.
+    \cite{Scholz2017} find integration cost of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{40}[\euro]{\per\mega\watt\hour} for a mix of \SI{80}{\percent} solar PV and \SI{20}{\percent} wind power at \SI{100}{\percent} renewables penetration in Europe.
+    This reduces to below \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{25}[\euro]{\per\mega\watt\hour} when the technology mix consists of \SI{20}{\percent} solar PV and \SI{80}{\percent} wind power.
+    These findings illustrate that abandoning wind power for PV in favor of undisturbed landscapes potentially comes at significant opportunity cost.
+    We complement the literature on negative wind turbine impacts and set out to quantify this cost of undisturbed landscapes for the case of Austria.
+    First, Austrian policies targeting a fixed share of renewable electricity generation are an ideal framework for studying the effects of substituting renewable energy generation technologies (RET).
     Moreover, Austria has large hydro storage capacities (including seasonal hydro storages) in place, which should limit PV integration cost.
-Thus, our findings should be qualitatively generalizable to power systems with high renewable shares and lower storage capacities in similar climates.
-Methodologically, we rely on the power system model \emph{medea}, which is described in section~\ref{sec:data-and-methods}.
+    Thus, our findings should be qualitatively generalizable to power systems with high renewable shares but lower storage capacities with climates comparable to Austria.
+    Methodologically, we rely on the power system model \emph{medea}, which is described in \autoref{sec:data-and-methods}.
 
     Our analysis contributes to a more comprehensive understanding of the trade-offs faced in the design of highly renewable electricity systems, thereby helping to inform policies about the socially optimal expansion of RET.
-Any such policy necessarily needs to account for the full social cost (including all externalities) of RET.
-Moreover, we contribute to the analysis of near-optimal power system configurations~\citep[see e.g.][]{Neumann2019,
-Schlachtberger2017}.
-Corresponding results are presented in section~\ref{sec:results-discussion}.
-
-Our results also provide insight into how potential compensation for negative external effects of wind turbines at the local level could be financed by system-wide benefits of wind power.
-Moreover, based on our analysis we can approximate the valuation of undisturbed landscapes that is implicit in Austria's climate and energy strategy.
-These policy implications and conclusions are summarized in section~\ref{sec:conclusions-policy-implication}.
+    Any such policy necessarily needs to account for the full social cost, including all externalities, of RET.
+    Further, we contribute to the analysis of near-optimal power system configurations~\citep[see e.g.][]{Neumann2019, Schlachtberger2017}.
+    Corresponding results are presented in \autoref{sec:results-discussion}.
+    Finally, our investigation into the cost of undisturbed landscapes generates insight into the possibilities and limitations of funding potential compensating measures for the benefit of the ones negatively affected by wind turbines.
+    We illustrate these potentials by an approximation of the valuation of undisturbed landscapes that is implicit in Austria's climate and energy strategy.
+    \hyperref[sec:conclusions-policy-implication]{Section~\ref*{sec:conclusions-policy-implication}} summarizes policy implications and concludes.
 
@@
     \section{Data and Methods} \label{sec:data-and-methods}
-To investigate the cost of undisturbed landscapes (i.e. the opportunity cost of wind power) we use the power system model \emph{medea}, as summarized in section~\ref{subsec:medea}.
-We instantiate the model with scenario assumptions and data as observed in the year 2016, which was a low-wind year in Austria (see section~\ref{subsec:data}).
-
+    To investigate the cost of undisturbed landscapes (i.e. the opportunity cost of wind power) we use the power system model \emph{medea}, as summarized in \autoref{subsec:medea}.
+    We instantiate the model with scenario assumptions and data as observed in 2016, which was a low-wind year in Austria, while solar and hydro resources were close to long-term averages (see \autoref{subsec:data}).
     Based on this, we approximate the opportunity cost of wind power through the following procedure:
     \begin{enumerate}
-\item We derive the unrestricted system cost-minimizing deployment of wind and solar power, given the scenario
-assumptions
-\item We restrict deployment of wind power by a small margin (so that the next best RET substitutes for wind
-power) and observe net system cost $c_{net}$ for Austria (calculated as total system cost including air
-pollution cost net of trade).\label{enumerate:restrict}
+        \item We derive the unrestricted system cost-minimizing deployment of wind and solar power, given the scenario assumptions
+        \item We restrict deployment of wind power by a small margin (so that PV, the next best RET, substitutes for wind power) and observe net system cost $c_{net}$ for Austria, calculated as total system cost including air pollution cost net of the balance of trade.\label{enumerate:restrict}
         \item We repeat step~\ref{enumerate:restrict} till no wind power can be deployed.
-\item Finally, we approximate the opportunity cost of wind power (at given wind power capacity $w$) $OC_w$ by
-the change in $c_{net}$ in response to a change in wind power capacity $w$ deployed, i.e.
-\[OC_w = \frac{\Delta c_{net}}{\Delta w}\]
-where $\Delta$ is the difference operator and opportunity cost $OC_w$ are expressed in \EUR per
-\SI{}{\mega\watt} wind power foregone.
+        \item Finally, we approximate the opportunity cost of wind power (at given wind power capacity $w$) $OC_w$ by the change in $c_{net}$ in response to a change in wind power capacity $w$ deployed, i.e.\ \[OC_w = \frac{\Delta c_{net}}{\Delta w}\]
+        where $\Delta$ is the difference operator and opportunity cost $OC_w$ are expressed in \EUR per \SI{}{\mega\watt} wind power foregone.
     \end{enumerate}
@@
 % ***** medea *****
-\subsection{Power system model \emph{medea}}\label{subsec:medea}
-We make use of the power system model \emph{medea} to simulate (dis)investment in and hourly operation of the
-prospective Austrian (and German) power system.
-The model is cast as a linear optimization seeking to minimize total system cost, which consist of fuel and emission
-cost, quasi-fixed and variable operation and maintenance (O\&M) cost, the costs of investment in energy generation,
-storage, and transmission assets, and (potential) cost of non-served load.
-From an economic perspective, the model reflects a perfectly competitive energy-only market with fully
-price-inelastic demand and perfect foresight of all actors.
-
-The system is required to meet exogenous and inelastic demand for electricity and heat in any hour of the modeled
-year.
-Energy supply, in turn, is constrained by available installed capacities of energy conversion, storage, and
-transmission units.
-Co-generation units convert fuel to heat and power subject to a feasible operating region defined by the unit's
-electrical efficiency, the electricity loss per unit of heat production, and the back pressure coefficient.
-Electricity generation from intermittent sources (wind, run-of-river hydro, solar) is subject to exogenous hourly
-generation profiles, which are scaled according to total installed capacities.
+
+    \subsection{Power system model \emph{medea}} \label{subsec:medea}
+    We make use of the power system model \emph{medea} to simulate (dis)investment in, and hourly operation of, the prospective Austrian and German power systems in the year 2030.
+    The model is cast as a linear optimization seeking to minimize total system cost, which consists of fuel and emission costs, quasi-fixed and variable operation and maintenance (O\&M) costs, the costs of investment in energy generation, storage, and transmission assets, and the (potential) cost of non-served load.
+    From an economic perspective, the model reflects a perfectly competitive energy-only market with fully price-inelastic demand and perfect foresight of all actors.
+
+    The modeled system is required to meet exogenous and inelastic demand for electricity and heat at any hour of the year.
+    Energy supply, in turn, is constrained by available installed capacities of energy conversion, storage, and transmission units.
+    Co-generation units convert fuel to heat and power subject to a feasible operating region defined by the unit's electrical efficiency, the electricity loss per unit of heat production, and the backpressure coefficient.
+    Electricity generation from intermittent sources (wind, run-of-river hydro, solar) is subject to exogenous hourly generation profiles, which are scaled according to total installed capacities.
     Electricity from these sources can be curtailed at no additional cost (free disposal).
 
-Electricity can be stored in reservoir and pumped hydro storages or batteries.
+    Electricity can be stored in reservoir and pumped hydro storages, batteries, or hydrogen storage (more precisely: water electrolysis, hydrogen storage, and reconversion to electricity in fuel cells).
     The capacity of hydro storages can not be expanded, as we assume existing potentials to be exhausted.
-Battery capacity, on the other hand, can be added endogenously.
-Generation from storages is constrained by installed capacity and energy contained in storage.
-Hydro reservoirs are filled (exogenous) inflows.
-Pumped hydro storages and batteries can actively store electricity for later use.
-To better capture operational differences of hydro storage units, we model daily, weekly and seasonal reservoir and pumped storage plant separately.
-To ensure a stable and secure operation of the electricity system, ancillary services (e.g. frequency control,
-voltage support) are required.
+    Battery and hydrogen storage capacities, on the other hand, can be added endogenously.
+    Generation from storage is constrained by installed capacity and stored energy.
+    Inflows of water into reservoirs add to stored energy.
+    Pumped hydro storages, batteries, and hydrogen storage can actively store electricity for later use.
+    To better capture operational differences of hydro storage units, we model daily, weekly and seasonal reservoir and pumped storage plants separately.
+    To ensure the stable and secure operation of the electricity system, ancillary services (e. g. frequency control, voltage support) are required.
     We model ancillary service needs as a minimum requirement on spinning reserves operating at any point in time.
-Thus, we assume that ancillary services can be provided by thermal power plant, run-of-river hydro plant or any
-type of storage technology.
-We do not model transmission and distribution grids within Austria or Germany.
+    Thus, we assume that ancillary services can be provided by thermal power plants, run-of-river hydro plants, or any (active) storage technology.
+    We don't model transmission and distribution grids within Austria or Germany.
     We do, however, allow for cross-border electricity trade between both countries.
-We also keep track of the cost of air pollution arising from burning fossil fuels.
-As these cost are external, they do not enter cost minimization.
-For a detailed, mathematical description of the model please refer to \ref{sec:medea-desc}.
-Data processing is implemented in python, while the optimization model is based on GAMS.
-Running the model in hourly resolution for one year takes 10--15 minutes, depending on the model parameters, on an Intel i7-8700 machine with 16 GB RAM, using CPLEX 12.9 as a solver.
+    We also keep track of the cost of (non-\ce{CO2}) air pollution arising from burning fossil fuels.
+    As these costs are external, they do not enter cost minimization.
+    For a detailed mathematical description of the model, please refer to \ref{sec:medea-desc}.
+    Data processing is implemented in python, while the optimization model is based on GAMS. The model code is published at \url{https://github.com/inwe-boku/medea} under an open MIT license.
+    Running the model in hourly resolution for one year takes 10-15 minutes, depending on the model parameters, on an Intel i7-8700 machine with 16 GB RAM, using CPLEX 12.10 as a solver.
 
 % ***** Data *****
-\subsection{Data}\label{subsec:data}
+
+    \subsection{Data} \label{subsec:data}
     We set up our model with the goal to resemble Austria's prospective electricity and district heating systems in the year 2030.
     We also include Austria's largest electricity trading partner, Germany, to account for potential effects from electricity trade.
-Again, our scenario reflects Germany's currently announced electricity sector policies, so that we set generation capacities to levels consistent with policies effective by 2030.
-These assumptions are laid out in sections~\ref{subsubsec:assumptions-austria} and \ref{subsubsec:assumptions-germany} and summarized in table~\ref{tab:instcaps}.
-
-\subsubsection{Scenario Assumptions for Austria}\label{subsubsec:assumptions-austria}
-The Austrian government has set itself the goal of generating \SI{100}{\percent} of electricity consumption from domestic renewable sources on annual balance by 2030~\citep{Regierungsprogramm2020}.
+    For Germany, our scenario reflects current and announced future electricity sector policies up to 2030, so that we set generation capacities to levels consistent with these policies.
+    These assumptions are laid out in subsections \ref{subsubsec:assumptions-austria} and \ref{subsubsec:assumptions-germany}, and summarized in \autoref{tab:instcaps}.
+
+    \subsubsection{Scenario Assumptions for Austria} \label{subsubsec:assumptions-austria}
+    The Austrian government has set itself the goal of generating \SI{100}{\percent} of electricity consumption from domestic renewable sources on annual balance by 2030~\citepalias{Regierungsprogramm2020}.
     However, industry own consumption and system services, which currently account for about \SI{10}{\percent} of annual electricity consumption, are exempt.
-
-The government plans to achieve this goal by generating an additional \SI{27}{\tera\watt\hour} of electricity annually from renewable sources\footnote{Please note that this policy goal does not imply that the generated
-electricity must actually be consumed. Hence, we count curtailed electricity as contributing to the policy goal.}.
-New hydro power plants are thought to contribute \SI{5}{\tera\watt\hour} annually, while additional electricity generation from biomass is envisaged to account for \SI{1}{\tera\watt\hour} annually.
-In our scenarios, we add generation capacities sufficient for reaching these targets.
-
-Further, the government projects the remainder to come from solar PV (\SI{11}{\tera\watt\hour} annually) and onshore wind turbines (\SI{10}{\tera\watt\hour} annually)~\cite{Regierungsprogramm2020}.
-As we are interested in determining the opportunity cost of wind power versus its best alternative, we allow for endogenous investment in wind and solar power, without enforcing announced targets.
-Other low-carbon energy technologies are not feasible at large scale in the Austrian context, as we have laid out in the introduction.
-Initial generation capacities for Austria are summarized in table~\ref{tab:instcaps}.
-
-\subsubsection{Scenario Assumptions for Germany}\label{subsubsec:assumptions-germany}
+    The government plans to achieve this goal by generating an additional \SI{27}{\tera\watt\hour} of electricity annually from renewable sources\footnote{Please note that this policy goal does not imply that the generated electricity must actually be consumed.
+Hence, we count curtailed electricity as contributing to the policy goal.}
+    , for a total of \SI{78.1}{\tera\watt\hour} of renewable electricity generation in 2030.
+    New hydropower plants are thought to contribute \SI{5}{\tera\watt\hour} annually, while additional electricity generation from biomass is envisaged to account for \SI{1}{\tera\watt\hour} annually.
+    In our scenarios, we add generation capacities sufficient to reach these targets.
+    Further, the government projects the remainder to come from solar PV and onshore wind turbines with an annual contribution of \SI{11}{\tera\watt\hour} and \SI{10}{\tera\watt\hour} \citepalias{Regierungsprogramm2020}, respectively.
+
+    As we are interested in determining the opportunity cost of wind power versus its best alternative, we allow for endogenous investment in wind and solar power without enforcing announced technology-specific targets.
+    Other low-carbon energy technologies are not feasible at a large scale in the Austrian context, as we have laid out in the introduction.
+    Initial generation capacities for Austria are summarized in \autoref{tab:instcaps}.
+
+    \subsubsection{Scenario Assumptions for Germany} \label{subsubsec:assumptions-germany}
     Germany has announced specific capacity targets for several power generation technologies.
-Following these announcements, we anticipate an end to nuclear power generation, a (partial) coal exit according to
-recommendations by the `coal commission'~\cite{WSB2019}, and a further expansion of renewable electricity generation
-in line with the German Renewable Energy Sources Act~\cite{noauthor_renewable_2017}.
-We do, however, expect the \SI{52}{\giga\watt} solar PV cap to be removed.
-The corresponding capacity assumptions are displayed in table~\ref{tab:instcaps}.
-
-\begin{table}[H]
-\caption{Initial Generation Capacities} \label{tab:instcaps}
-\centering
-\begin{tabulary}{0.55\textwidth}{L C C}
+    Following these announcements, we anticipate an end to nuclear power generation, a (partial) coal exit according to recommendations by the `coal commission'~\citepalias{WSB2019}, and a further expansion of renewable electricity generation in line with the German Renewable Energy Sources Act~\citep{noauthor_renewable_2017}, except for the \SI{52}{\giga\watt} cap on solar PV, which was removed meanwhile.
+    Investment in additional capacity (as well as decommissioning of excess capacities) is incentivised through technology-specific government subsidies, so that we assume targeted capacities will be in place by 2030.
+    The corresponding capacity assumptions are displayed in \autoref{tab:instcaps}.
+    In addition, we allow for endogenous (dis-) investment in energy conversion capacities.
+    \begin{ThreePartTable}
+        \centering
+        \renewcommand\TPTminimum{0.8\textwidth}
+        \begin{TableNotes}
+            \footnotesize
+            \item[a] capacity scaled to match generation from energy balance \citep{StatistikAustria2020}
+        \end{TableNotes}
+        \begin{longtable}{l c c c c}
+            \caption{Initial Energy Conversion Capacities (2030)} \label{tab:instcaps} \\
             \toprule
-Technology & Austria & Germany\\
-& \si{\giga\watt} & \si{\giga\watt}\\
-\midrule
-Wind Onshore & $2.6$ & $90.8$ \\
-Wind Offshore  & $0.0$ & $15.0$ \\
-Solar PV & $1.1$ & $73.0$ \\
-Run-of-river Hydro  & $6.9$ & $4.5$ \\
-Hydro storage & $7.7$ & $6.5$ \\
-Biomass & $0.7$ & $8.4$ \\
-Lignite & $0.0$ & $11.4$ \\
-Coal & $0.0$ & $14.0$ \\
-Natural Gas & $3.9$ & $24.2$ \\
-Mineral Oil & $0.2$ & $3.5$ \\
-Heat Pump & $1.0$ & $1.0$ \\
-Gas Boiler & $3.9$ & $25.5$\\ \bottomrule
-\end{tabulary}
-\end{table}
+            Technology & \multicolumn{2}{c}{Austria} & \multicolumn{2}{c}{Germany} \\
+            \cmidrule(lr){2-3}\cmidrule(lr){4-5}
+            & \si{\giga\wattel} & \si{\giga\wattth} & \si{\giga\wattel} & \si{\giga\wattth} \\
+            \midrule
+            \endhead
+            \midrule
+            \multicolumn{5}{c}{\textit{continued on next page}}     \endfoot
+            \bottomrule
+            \insertTableNotes
+            \endlastfoot
+            Wind Onshore       & $2.6$             & --                & $90.8$            & --                \\
+            Wind Offshore      & --                & --                & $15.0$            & --                \\
+            Solar PV           & $1.1$             & --                & $73.0$            & --                \\
+            Run-of-river Hydro & $6.8$             & --                & $4.5$             & --                \\
+            Hydro storage      & $8.7$             & --                & $6.5$             & --                \\
+            Biomass            & $0.7$             & $0.4$             & $8.4$             & $4.9$             \\
+            Lignite            & --                & --                & $11.4$            & $8.8$             \\
+            Coal               & --                & --                & $14.0$            & $9.9$             \\
+            Natural Gas        & $3.9$             & $1.7$             & $24.2$            & $6.1$             \\
+            Mineral Oil        & $0.2$             & --                & $3.5$             & $1.7$             \\
+            Heat Pump          & --                & $1.0$             & --                & $1.0$             \\
+            Gas Boiler         & --                & $1.5$             & --                & $18.0$            \\
+            Biomass Boiler     & --                & $0.8$\tnote{a}    & --                & 2.6               \\ \bottomrule
+        \end{longtable}
+    \end{ThreePartTable}
 
     \subsubsection{Energy supply}
-We represent 20 dispatchable energy conversion and storage technologies that are expected to be operated in 2030.
-In addition to given, initially installed capacities (see table~\ref{tab:instcaps}), the model can endogenously add further generation capacities that are compatible with stated policy objectives.
-We calculated annualized investment cost of each technology on the basis of an assumed weighted average cost of capital of \SI{5}{\percent} over the plant's lifetime.
-All technologies can also be decommissioned, so that we adopt a long-run perspective on the power system.
-Parameters of admissible technologies are summarized in Table~\ref{itm:techcost}.
-
-%\begin{table*}[h!t]
+    We represent 21 energy conversion and storage technologies that are expected to be operated in 2030.
+    In addition to given, initially installed capacities (see \autoref{tab:instcaps}), the model can endogenously add further generation capacities that are compatible with stated policy objectives.
+    We calculate the annualized investment cost of each technology based on an assumed weighted average cost of capital of \SI{5}{\percent} over the plant's lifetime.
+    All technologies can also be decommissioned so that we adopt a long-run perspective on the power system.
+    \autoref{itm:techparms} summarizes the parameters of admissible technologies.
     \begin{ThreePartTable}
-\renewcommand\TPTminimum{\textwidth} %%% we want full width
+        \renewcommand\TPTminimum{\textwidth}
         \begin{TableNotes}
-\small
+            \footnotesize
             \item SC -- subcritical, USC -- (ultra-)supercritical, ST -- steam turbine, GT -- gas turbine, CC -- combined cycle, PSP -- Pumped storage plant
-\item[$\dagger$] own assumption
-\item[$\ddagger$] \EUR / $(\text{MVA} \times \text{km})$
+            \item[a] Alkaline electrolysis + cavern storage + PEM fuel cell
+            \item[b] own assumption
+            \item[c] \EUR / $(\text{MVA} \times \text{km})$
+            \item Sources:
+            [1] \cite{DEA2019},
+            [2] \cite{Schroeder2013},
+            [3] \cite{LacalArantegui2014},
+            [4] \cite{OekoInstitut2017},
+            [5] \cite{DEA2019},
+            [6] \cite{DEA2020a},
+            [7] \cite{Brown2018},
+            [8] \cite{Hagspiel2014}
         \end{TableNotes}
@@
         \begin{longtable}{l ccc ccc l}
-\caption{Technology Input Parameters for 2030} \label{itm:techcost}\\
+            \caption{Technology Input Parameters for 2030} \label{itm:techparms} \\
             \toprule
-&\multicolumn{2}{c}{Capital Cost} & \multicolumn{2}{c}{O\&M-Cost} & & & \\ \cmidrule(lr){2-3}\cmidrule(lr){4-5}
+            & \multicolumn{2}{c}{Capital Cost} & \multicolumn{2}{c}{O\& M-Cost} & & & \\ \cmidrule(lr){2-3}\cmidrule(lr){4-5}
             Technology                & Power          & Energy   & Quasi-       & Variable & Life- & Efficiency               & Source    \\
             &                &          & fixed        &          & time  &                          &           \\
-&\EUR/kW & \EUR/kWh & \EUR/MWa & \EUR/MWh & a & & \\
+            & \EUR/kW        & \EUR/kWh & \EUR/MWa     & \EUR/MWh & a     &                          &           \\
             \midrule
             \endhead
-%\cmidrule{2-2}
             \midrule
@@
             \endlastfoot
-% the contents of the table
-Wind Onshore         & $1040$ & NA    & $12600$ & $1.35$ & $30$ & NA & \cite{DEA2019}\\
-Wind Offshore        & $1930$ & NA    & $36053$ & $2.70$ & $30$ & NA & \cite{DEA2019}\\
-PV Rooftop           & $870$  & NA    & $10815$ & $0.00$ & $40$ & NA & \cite{DEA2019}\\
-PV Open-space        & $380$  & NA    & $7250$  & $0.00$ & $40$ & NA & \cite{DEA2019}\\
-Hydro run-of-river   & --     & NA    & $60000$ & $0.00$ & $60$ & NA & \cite{Schroeder2013, LacalArantegui2014} \\
-\midrule
-Lignite Adv          & --     & NA    & $40500$ & $0.85$ & $40$ & 0.439 & \cite{OekoInstitut2017}\\
-Coal SC              & --     & NA    & $25000$ & $6.00$ & $25$ & 0.390 & \cite{Schroeder2013} \\
-Coal USC             & --     & NA    & $31500$ & $3.00$ & $25$ & 0.460 & \cite{DEA2019}\\
-Nat Gas ST           & $400$  & NA    & $15000$ & $3.00$ & $30$ & 0.413 & \cite{Schroeder2013}\\
-Nat Gas GT           & $435$  & NA    & $7745$  & $4.50$ & $25$ & 0.410 & \cite{DEA2019}\\
-Nat Gas CC           & $830$  & NA    & $27800$ & $4.20$ & $25$ & 0.580 & \cite{DEA2019}\\
-Oil ST               & $400$  & NA    & $6000$  & $3.00$ & $30$ & 0.410 & \cite{Schroeder2013}\\
-Oil GT               & $363$  & NA    & $7745$  & $4.50$ & $25$ & 0.400 & \cite{DEA2019}\\
-%Oil CC             & $830$  & NA    & $27800$ & $4.20$ & $25$ & 0.580 & \cite{}\\
-Biomass              & $3300$ & NA    & $96000$ & $4.60$ & $25$ & 0.270 & \cite{DEA2019}\\ \midrule
-Hydro Reservoir      & --     & --    & $22000$ & $3.00$ & $60$ & $0.900$\tnote{$\dagger$} & \cite{LacalArantegui2014}\\
-Hydro PSP            & --     & --    & $22000$ & $3.00$ & $60$ & $0.810$\tnote{$\dagger$} & \cite{LacalArantegui2014}\\
-Battery Li-Ion       & $320$ & $302$ & $540$ & $1.80$ & $25$ & 0.920 & \cite{DEA2020}\\ \midrule
-Electric Boiler      & $140$  & NA    & $1020$  & $0.50$ & $20$ & 0.990 & \cite{DEA2019}\\
-Absorption heat pump & $510$  & NA    & $2000$  & $1.30$ & $25$ & 1.730 & \cite{DEA2019}\\
-Natural gas boiler   & $50$   & NA    & $1900$  & $1.00$ & $25$ & 0.920\tnote{$\dagger$} & \cite{DEA2019} \\ \midrule
-Transmission (AC)    & $455$\tnote{$\ddagger$} & NA    & $9$\tnote{$\ddagger$} & $0.00$ & $40$ & 1.000 & \cite{Brown2018, Hagspiel2014}\\
+            Wind Onshore              & $1040$         & NA       & $12600$      & $1.35$   & $30$  & NA                       & [1]       \\
+            Wind Offshore             & $1930$         & NA       & $36053$      & $2.70$   & $30$  & NA                       & [1]       \\
+            PV Rooftop                & $870$          & NA       & $10815$      & $0.00$   & $40$  & NA                       & [1]       \\
+            PV Open-space             & $380$          & NA       & $7250$       & $0.00$   & $40$  & NA                       & [1]       \\
+            Hydro run-of-river        & --             & NA       & $60000$      & $0.00$   & $60$  & NA                       & [2, 3]    \\
+            \midrule
+            Lignite Adv               & --             & NA       & $40500$      & $0.85$   & $40$  & 0.439                    & [4]       \\
+            Coal SC                   & --             & NA       & $25000$      & $6.00$   & $25$  & 0.390                    & [2]       \\
+            Coal USC                  & --             & NA       & $31500$      & $3.00$   & $25$  & 0.460                    & [1]       \\
+            Nat Gas ST                & $400$          & NA       & $15000$      & $3.00$   & $30$  & 0.413                    & [2]       \\
+            Nat Gas GT                & $435$          & NA       & $7745$       & $4.50$   & $25$  & 0.410                    & [1]       \\
+            Nat Gas CC                & $830$          & NA       & $27800$      & $4.20$   & $25$  & 0.580                    & [1]       \\
+            Oil ST                    & $400$          & NA       & $6000$       & $3.00$   & $30$  & 0.410                    & [2]       \\
+            Oil GT                    & $363$          & NA       & $7745$       & $4.50$   & $25$  & 0.400                    & [1]       \\
+            Biomass                   & $3300$         & NA       & $96000$      & $4.60$   & $25$  & 0.270                    & [1]       \\ \midrule
+            Hydro Reservoir           & --             & --       & $22000$      & $3.00$   & $60$  & $0.900$\tnote{$\dagger$} & [3]       \\
+            Hydro PSP                 & --             & --       & $22000$      & $3.00$   & $60$  & $0.810$\tnote{$\dagger$} & [3]       \\
+            Battery Li-Ion            & $320$          & $302$    & $540$        & $1.80$   & $25$  & 0.920                    & [1]       \\
+            Hydrogen Storage\tnote{a} & $2575$         & $2$      & 82500        & 0        & 25    & 0.333                    & [1, 5, 6] \\ \midrule
+            Electric Boiler           & $140$          & NA       & $1020$       & $0.50$   & $20$  & 0.990                    & [5]       \\
+            Absorption heat pump      & $510$          & NA       & $2000$       & $1.30$   & $25$  & 1.730                    & [5]       \\
+            Natural gas boiler        & $50$           & NA       & $1900$       & $1.00$   & $25$  & 0.920\tnote{b}           & [5]       \\ \midrule
+            Transmission (AC)         & $455$\tnote{c} & NA       & $9$\tnote{c} & $0.00$   & $40$  & 1.000                    & [7, 8]    \\
         \end{longtable}
     \end{ThreePartTable}
-%\end{table*}
-
-Power plants running on fossil fuels emit \COO into the atmosphere.
-We approximate the amount of \COO released through the carbon intensity of fossil fuels, as displayed in Table~\ref{table:carbon-intensity}.
+
+    Power plants running on fossil fuels emit \ce{CO2} into the atmosphere.
+    We approximate the amount of \ce{CO2} released through the carbon intensity of fossil fuels, as displayed in \autoref{table:carbon-intensity}.
     These estimates are based on an analysis by the German environmental agency~\citep{Juhrich2016}.
 
-\begin{table}[h!t]
-\centering
-\caption{\COO intensity of fossil fuels in tonnes \COO per MWh}
+    In the baseline scenario, we assume capital cost of solar PV of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{625}[\euro]{\per\kilo\wattpeak} installed.
+    These costs are capacity-weighted average cost of rooftop and open-space PV installations, with \SI{50}{\percent} of PV capacity being mounted on rooftops, while the remainder is realized as open-space solar PV at utility-scale.
+    In \autoref{subsec:sensitivity}, we analyze the sensitivity of our results to the capital cost of solar PV\@.
+
+    \begin{table}[t]
+        \centering
+        \begin{threeparttable}
+            \renewcommand\TPTminimum{0.8\textwidth}
+            \caption{\ce{CO2} intensity of fossil fuels in tonnes \ce{CO2} per MWh}
             \label{table:carbon-intensity}
-\begin{tabular}{c c c l}
+            \begin{tabular}{l c c c c c}
                 \toprule
-Lignite & Coal & Natural Gas & Mineral Oil  \\
-\midrule
-0.399 & 0.337 & 0.201 & 0.266        \\
+                & & Lignite & Coal & Natural Gas & Mineral Oil \\
+                \midrule
+                \ce{CO2} intensity & \si[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{\tonne\per\mega\watt\hour}
+                & 0.399
+                & 0.337
+                & 0.201
+                & 0.266
+                \\
                 \bottomrule
             \end{tabular}
+            \begin{tablenotes}
+                \footnotesize
+                \item Source: \cite{Juhrich2016}
+            \end{tablenotes}
+        \end{threeparttable}
     \end{table}
 
-In the baseline scenario, we assume capital cost of solar PV of \SI{630}[\euro]{\per\kilo\watt} installed.
-These cost are capacity-weighted average cost of rooftop and open-space PV installations, with roughly \SI{50}{\percent} of PV capacity being mounted on roof tops, while the remainder is realized as open-space solar PV at utility scale.
-In section~\ref{subsec:sensitivity} we analyze the sensitivity of our results with respect to the capital cost of solar PV\@.
-
-Generation from non-dispatchable technologies solar PV, wind turbines and run-of-river hydro is assumed to follow hourly generation profiles\footnote{A \emph{generation profile} tracks the share of installed intermittent capacity that is generating electricity over time.} as observed in 2016.
-German solar and wind power profiles are sourced from Open Power System Data~\citep{opsd2019}.
-For Austria, solar and wind power generation profiles were derived by dividing solar and wind power generation time series as reported by Open Power System Data~\cite{opsd2019} by installed capacities published by \cite{Biermayr2019}.
-Similarly, generation profiles for run-of-river hydro power were derived based on generation data reported by \cite{ENTSOE2020b}.
-
-Data on inflows into Austrian hydro reservoirs is not publicly available.
-However, ENTSO-E's transparency platform reports weekly data on hydro storage filling levels~\cite{ENTSOE2020a} along with hourly electricity generation from hydro reservoirs, and pumping by pumped hydro storages~\cite{ENTSOE2020b}.
-Based on this data, we have approximated inflows as the part of change in (weekly) hydro storage fill levels not explained by (upsampled hourly) pumping or generation.
-To arrive at hourly inflows, we have interpolated our weekly estimates by piecewise cubic hermite interpolation.
-Although inflows to hydro reservoirs are given exogenously, all storage technologies (including hydro reservoir and pumped storage plants) are charged and dispatched endogenously.
-
-\subsubsection{Energy demand}\label{subsec:energy-demand}
-By combining government policy targets with information about electricity generation in our base
-year~\citep{StatistikAustria2020}, we can infer expected electricity demand of \SI{83}{\tera\watt\hour} in 2030.
-To generate hourly load, we scale the 2016 load profile accordingly.
-For Germany, we scale reported hourly load to match annual electricity consumption as reported in national energy balances~\citep{AGEB2018} for the year 2016.
+    Generation from non-dispatchable technologies solar PV, wind turbines, and run-of-river hydro is assumed to follow hourly generation profiles\footnote{A \emph{generation profile} tracks the share of installed intermittent capacity that is generating electricity over time.}
+    as observed in 2016.
+    German solar and wind power profiles are sourced from \cite{opsd2019}.
+    For Austria, solar and wind power generation profiles are based on solar and wind power generation time series reported by \cite{opsd2019}, which we scaled to match annual generation as reported in the national energy balance~\citep{StatistikAustria2020}.
+    Subsequently, the scaled generation time series were divided by installed capacities as published in~\cite{Biermayr2019}.
+    Similarly, generation profiles for run-of-river hydropower were derived from generation data reported by \cite{ENTSOE2020b}, which we scaled to match annual run-of-river electricity generation as reported by Austria's regulatory body \cite{EControl2020}.
+
+    Data on inflows into Austrian hydro reservoirs are not publicly available.
+    However, the \cite{ENTSOE2020a} reports weekly data on hydro storage filling levels along with hourly electricity generation from hydro reservoirs and pumping by pumped hydro storages \citepalias{ENTSOE2020b}.
+    Based on this data, we have approximated inflows as the part of the change in (weekly) hydro storage fill levels that isn't explained by upsampled hourly pumping and generation.
+    To arrive at hourly inflows, we have interpolated our weekly estimates by piecewise cubic Hermite interpolation.
+    Although inflows to water reservoirs are given exogenously, all storage technologies (including hydro reservoir and pumped storage plants) are charged and dispatched endogenously.
+    \autoref{tab:time-series} summarizes the descriptive statistics of the abovementioned renewables time series.
+
+    \subsubsection{Energy demand} \label{subsec:energy-demand}
+    By combining government policy targets with information about electricity generation in our base year~\citep{StatistikAustria2020}, we can infer that an electricity demand of \SI{83}{\tera\watt\hour} is expected in 2030.
+    To generate a time series of hourly electricity demand, we scale Austria's electricity load as reported by \cite{opsd2019} accordingly.
+    For Germany, we scale reported hourly load to match annual electricity consumption as reported in national energy balances~\citepalias{AGEB2018} for the year 2016.
     For 2030 we assume electricity demand to remain unchanged from 2016 levels.
 
-Annual heat consumption is derived from energy balances for Austria and Germany, respectively~\citep{AGEB2018,
-StatistikAustria2020}.
-Subsequently, annual heat consumption is broken down to hourly heat consumption on the basis of standard natural gas load profiles for space heating in the residential and commercial sectors~\citep{Almbauer2008}.
+    Annual heat consumption is derived from energy balances for Austria and Germany, respectively~\citep{AGEB2018, StatistikAustria2020}.
+    Subsequently, we break annual heat consumption down into hourly heat consumption, based on standard natural gas load profiles for space heating in the residential and commercial sectors~\citep{Almbauer2008}.
     These load profiles make use of daily average temperatures to calculate daily heat demand.
-We extract spatially resolved temperature data from ERA-5 climate data sets~\citep{CCCS2017} and compute a capacity-weighted average of temperatures at locations of combined heat and power (CHP) generation units.
-Daily heat demand based on these weighted average temperatures is then broken down to hourly consumption on the basis of standardized factors accounting for weekday and time-of-day effects.
-Descriptive statistics of electricity and heat consumption are provided in table~\ref{tab:time-series}.
-
-\begin{table*}[h!t]
+    We extract spatially resolved temperature data from ERA-5 climate data sets~\citepalias{CCCS2017} and compute a capacity-weighted average of temperatures at locations of combined heat and power (CHP) generation units.
+    Daily heat demand based on these weighted average temperatures is then broken down to hourly consumption based on standardized factors accounting for weekday and time-of-day effects.
+    Descriptive statistics of electricity and heat consumption are provided in \autoref{tab:time-series}.
+    \begin{table*}[t]
         \centering
         \begin{threeparttable}
-\caption{Descriptive data of time series (2016)}\label{tab:time-series}
+            \caption{Descriptive statistics of time series (as observed in 2016)}\label{tab:time-series}
             \begin{tabular}{l c c c c c c l}
@@
                 \midrule
-Electricity load & AT & GW & 7.15 & 7.14 & 10.5 & 4.21 &\cite{opsd2019}     \\
-Electricity load & DE & GW & 60.2 & 59.6 & 82.8 & 34.6 &\cite{opsd2019}     \\
-District heating load\tnote{$\dagger$} & AT & GW & 2.56 & 2.43 & 5.77 & 0.91 & \cite{CCCS2017, StatistikAustria2020}\\
-District heating load\tnote{$\dagger$} & DE & GW & 14.6 & 14.1 & 26.0 & 8.9 & \cite{CCCS2017, AGEB2018} \\
-Wind onshore profile & AT &\%   & 0.229 & 0.142 & 0.929 & 0.000 &\cite{opsd2019}         \\
-Wind onshore profile & DE &\%   & 0.176 & 0.133 & 0.764 & 0.003 &\cite{opsd2019}         \\
-Wind offshore profile & DE &\%   & 0.371 & 0.327 & 0.900 & 0.000 &\cite{opsd2019}         \\
-Solar PV profile\tnote{$\dagger$ } & AT &\%   & 0.098 & 0.012 & 0.570 & 0.000 & \cite{opsd2019, Biermayr2019}\\
-Solar PV profile & DE &\%   & 0.099 & 0.003 & 0.661 & 0.000 &\cite{opsd2019}         \\
-Run-of-river profile & AT &\%    & 0.564 & 0.538 & 0.971 & 0.202 & \cite{ENTSOE2020b}$\dagger$\\
-Run-of-river profile & DE &\%   & 0.437 & 0.420 & 0.637 & 0.208 &\cite{ENTSOE2020b}\\
+                Electricity load               & AT   & GW   & 7.15  & 7.14   & 10.5  & 4.21  & [1]       \\
+                Electricity load               & DE   & GW   & 60.2  & 59.6   & 82.8  & 34.6  & [1]       \\
+                District heating load\tnote{a} & AT   & GW   & 2.56  & 2.43   & 5.77  & 0.91  & [2, 3]    \\
+                District heating load\tnote{a} & DE   & GW   & 14.6  & 14.1   & 26.0  & 8.9   & [2, 4]    \\
+                \midrule
+                Wind onshore profile           & AT   & \%   & 0.226 & 0.140  & 0.913 & 0.000 & [1, 3]    \\
+                Wind onshore profile           & DE   & \%   & 0.176 & 0.133  & 0.764 & 0.003 & [1]       \\
+                Wind offshore profile          & DE   & \%   & 0.371 & 0.327  & 0.900 & 0.000 & [1]       \\
+                Solar PV profile\tnote{a}      & AT   & \%   & 0.114 & 0.014  & 0.666 & 0.000 & [1, 3, 5] \\
+                Solar PV profile               & DE   & \%   & 0.099 & 0.003  & 0.661 & 0.000 & [1]       \\
+                Run-of-river profile\tnote{a}  & AT   & \%   & 0.585 & 0.559  & 0.971 & 0.202 & [6, 7]    \\
+                Run-of-river profile           & DE   & \%   & 0.437 & 0.420  & 0.637 & 0.208 & [7]       \\
                 \bottomrule
@@
             \begin{tablenotes}
-\small
-\item[$\dagger$] own calculation based on referenced sources
+                \footnotesize
+                \item[a] own calculation based on referenced sources
+                \item Sources:
+                [1] OPSD \citeyearpar{opsd2019},
+                [2] C3S \citeyearpar{CCCS2017},
+                [3] \cite{StatistikAustria2020},
+                [4] AGEB \citeyearpar{AGEB2018},
+                [5] \cite{Biermayr2019},
+                [6] ENTSO-E \citeyearpar{ENTSOE2020b},
+                [7] \cite{EControl2020}
             \end{tablenotes}
@@
     \subsubsection{Electricity Transmission}
-Electricity transmission between the modelled market areas is limited to \SI{4.9}{\giga\watt}, in line with the introduction of a congestion management scheme by German and Austrian authorities in 2018~\citep{}.
-%\ref{sec:transmission-sensitivity} explores the effects of doubling the transmission capacity between Germany and Austria.
+    Electricity transmission between the modeled market areas is limited to \SI{4.9}{\giga\watt}, in line with the introduction of a congestion management scheme by German and Austrian authorities in 2018~\citep{BNetzA2017}.
+    Section \ref{subsubsec:sens_transmission} explores the effects of increasing transmission capacity to \SI{10}{\giga\watt}.
 
     \subsubsection{Prices}
-Monthly prices for exchange-traded fuels (hard coal, crude oil (Brent), natural gas) are retrieved from the
-International Monetary Fund's Commodity Data Portal.
-We convert these prices to \SI[sticky-per]{}[\euro]{\per\mega\watt\hour} based on the fuel's energy content and market exchange rates obtained from the European Central Bank~\citep{ECB2020}.
-Finally, we resample prices to hourly frequency using piecewise cubic hermite interpolation.
-Due to its low energy density lignite is not transported over large distances and consequently also not traded on
-markets.
+    Monthly prices for exchange-traded fuels (hard coal, crude oil (Brent), natural gas) are retrieved from the International Monetary Fund's Commodity Data Portal.
+    We convert these prices to \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{}[\euro]{\per\mega\watt\hour} based on the fuel's energy content and market exchange rates obtained from the \cite{ECB2020}.
+    Finally, we resample prices to hourly frequency using piecewise cubic Hermite interpolation.
+    Due to its low energy density, lignite is not transported over large distances and consequently also not traded on markets.
     Instead, lignite-fired power plants are situated in proximity to lignite mines.
-According to estimates from~\cite{OekoInstitut2017}, the price of lignite in Germany is close to
-\SI[sticky-per]{1.50}[\euro]{\per\mega\watt\hour}.
-Biomass-fired power plant run on a wide variety of solid and gaseous fuels, some of which are marketed.
-However, continued operation of biomass-fired plants typically relies on sufficient subsidies.
-As a first-order approximation to more complex subsidy schemes, we assume subsidized fuel cost of
-\SI[sticky-per]{6.50}[\euro]{\per\mega\watt\hour}.
+    According to estimates from~\cite{OekoInstitut2017}, the price of lignite in Germany is close to \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{1.50}[\euro]{\per\mega\watt\hour}.
+    Biomass-fired power plants run on a wide variety of solid and gaseous fuels, some of which are marketed.
+    However, the continued operation of biomass-fired plants typically relies on sufficient subsidies.
+    As a first-order approximation to more complex subsidy schemes, we assume subsidized fuel cost of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{6.50}[\euro]{\per\mega\watt\hour}.
     In consequence, biomass-fired plants are operating at or close to capacity throughout the scenarios considered.
     \footnote{Subsidies are not included in system cost, as the required level is uncertain.
-The total sum of subsidies is, however, constant across scenarios as biomass capacity can not be expanded.}
-
-\begin{table*}[h!t]
+Biomass capacity can not be expanded and is constant across scenarios.
+Thus, electricity generation from biomass and the corresponding subsidies vary by no more than \SI{1.75}{\percent}.}
+    \begin{table*}[th]
         \centering
         \begin{threeparttable}
-\caption{Descriptive data of price time series}
-\begin{tabular}{l c r r r r l}
+            \caption{Descriptive data of price time series (as observed in 2016)}
+            \begin{tabular}{l c r r r r c}
                 \toprule
@@
                 \midrule
-Lignite     &\EUR/MWh & 1.50  & 1.50  & 1.50  & 1.50  & \cite{OekoInstitut2017}\\
-Coal        &\EUR/MWh & 8.58  & 8.13  & 11.90 & 6.69  & \cite{IMF2020}         \\
-Natural gas &\EUR/MWh & 13.62 & 12.87 & 20.18 & 12.04 & \cite{IMF2020}         \\
-Mineral oil &\EUR/MWh & 26.40 & 26.63 & 33.44 & 18.34 & \cite{IMF2020}         \\
-Biomass (subsidized) &\EUR/MWh & 6.50 & 6.50 & 6.50 & 6.50 & $\dagger$         \\
+                Lignite              & \EUR/MWh & 1.50  & 1.50   & 1.50  & 1.50  & [1]    \\
+                Coal                 & \EUR/MWh & 8.58  & 8.13   & 11.90 & 6.69  & [2]    \\
+                Natural gas          & \EUR/MWh & 13.62 & 12.87  & 20.18 & 12.04 & [2]    \\
+                Mineral oil          & \EUR/MWh & 26.40 & 26.63  & 33.44 & 18.34 & [2]    \\
+                Biomass (subsidized) & \EUR/MWh & 6.50  & 6.50   & 6.50  & 6.50  & [3]    \\
                 \bottomrule
@@
             \begin{tablenotes}
-\small
-\item[$\dagger$] own assumption
+                \footnotesize
+                \item Sources:
+                [1] \cite{OekoInstitut2017},
+                [2] \citetalias{IMF2020},
+                [3] own assumption
             \end{tablenotes}
@@
     \end{table*}
-
-Future \COO prices impact renewables deployment ~\citep{Brown2020, Kirchner2019} and affect the results of our analysis.
-As the future efficient price of \COO is highly uncertain, we have conducted our analysis for \COO prices in the range of $0$ to $120$ Euro per tonne.
-Implicitly, we assume efficient pricing of \COO emissions, i.e.\ all otherwise external cost of \COO emissions are internalized through the prevailing \COO price.
-
-In this context, it is worth noting that within the framework of our analysis a \COO price of
-\SI{30}[\euro]{\per\tonne} would be consistent with RET expansion goals stated by the Austrian government.
-At higher \COO prices, there is incentive to invest in \COO emission-free generation beyond
-government targets.
+    Future \ce{CO2} prices impact renewables deployment ~\citep{Brown2020, Kirchner2019} and affect the results of our analysis.
+    As the future efficient price of \ce{CO2} is highly uncertain, we have conducted our analysis for \ce{CO2} prices in the range of \euro$0$ to \euro$100$ per tonne.
+    Implicitly, we assume efficient pricing of \ce{CO2} emissions, i.e.\ all otherwise external cost of \ce{CO2} emissions are internalized through the prevailing \ce{CO2} price.
 
     \subsubsection{The cost of air pollution}
-In addition to \COO emissions, we also keep track of air pollutants such as nitrogen oxides, sulphur oxide, particulate matters, carbon monoxide, or heavy metals and the external cost imposed by their emission from power plants.
-For the valuation of these external cost, we rely on assessments from the NEEDS project, as reported in \cite{Samadi2017}.
-We have, however, converted these output-related estimates to input-related values.
-The corresponding estimates for the external cost of air pollution from fossil fuel combustion are summarized in Table \ref{tab:air_pollution_cost}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{c c c c c}
+    In addition to \ce{CO2} emissions, we also keep track of the external cost imposed by the emission of air pollutants such as nitrogen oxides, sulfur oxide, particulate matters, carbon monoxide, or heavy metals from fossil power plants.
+    For the valuation of these external costs, we rely on assessments from the NEEDS project, as reported in \cite{Samadi2017}.
+    We have, however, converted these output-related estimates to input-related values, so that variable cost refer to the energy content of fuel used to generate electricity.
+    The air pollution cost of RET mainly reflect the cost of air pollution during manufacturing of the plants.
+    To avoid distortions from different underlying assumptions on RET utilisation, we have converted these cost estimates to reflect the specific annual cost per unit of capacity installed.
+    The corresponding estimates for the external cost of air pollution from fossil fuel combustion are summarized in \autoref{tab:air_pollution_cost}.
+    \begin{table*}[ht]
+        \centering
+        \begin{threeparttable}
+            \caption{The external cost of air pollution}
+            \label{tab:air_pollution_cost}
+            \centering
+            \begin{tabular}{l c c}
                 \toprule
-Lignite & Coal & Natural Gas & Mineral Oil & Biomass \\
-\midrule
-$4.06$ & $6.12$ & $2.36$ & $3.21$ & $4.04$ \\
+                & Unit & Cost \\
+                \midrule
+                Lignite & \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{}[\euro]{\per\mega\watt\hour}
+                & $4.06$
+                \\
+                Coal & \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{}[\euro]{\per\mega\watt\hour}
+                & $6.12$
+                \\
+                Natural Gas & \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{}[\euro]{\per\mega\watt\hour}
+                & $2.36$
+                \\
+                Mineral Oil & \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{}[\euro]{\per\mega\watt\hour}
+                & $3.21$
+                \\
+                Biomass & \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{}[\euro]{\per\mega\watt\hour}
+                & $4.04$
+                \\
+                Wind power & \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{}[\euro]{\per\mega\watt}
+                & $2\,831$
+                \\
+                Solar Photovoltaics & \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{}[\euro]{\per\mega\wattpeak}
+                & $5\,028$
+                \\
                 \bottomrule
             \end{tabular}
-\caption{The external cost of air pollution in \SI{}[\euro]{\per\mega\watt\per\hour} of fuel burned}
-\label{tab:air_pollution_cost}
-\end{table}
+            \begin{tablenotes}
+                \footnotesize
+                \item Source: \cite{Samadi2017}, own calculations
+            \end{tablenotes}
+
+        \end{threeparttable}
+    \end{table*}
     Please note that these cost are considered external and consequently do not enter system cost minimization.
 
-The model as well as all data retrieval and processing scripts are available at medea's github repository
-\url{https://github.com/inwe-boku/medea} under the MIT license.
+    The data retrieval and processing scripts are published on GitHub \url{https://github.com/inwe-boku/medea} under the MIT license.
 
 % %%%%% %%%%% %%%%% RESULTS AND DISCUSSION %%%%% %%%%% %%%%% %%%%%
-\section{Results and Discussion}\label{sec:results-discussion}
-To derive the social opportunity cost of not using wind power (but solar photovoltaics instead), we start by
-determining the long-run power market equilibrium without constraints on renewable capacity addition.
+
+    \section{Results and Discussion} \label{sec:results-discussion}
+    To derive the social opportunity cost of not using wind power (but solar photovoltaics instead) in 2030, we start by determining the long-run power market equilibrium without constraints on renewable capacity addition.
     This gives us our unconstrained baseline scenario with minimal system costs.
-
-\subsection{Unconstrained Baseline}\label{subsec:baseline}
-As a point of reference for our analysis, we determine the least cost power system configuration when addition of
-wind turbines and solar PV is unconstrained.
-Given the policy goal of generating \SI{21}{\tera\watt\hour} of electricity either from wind or solar power (with
-the remainder coming from hydro power and biomass), required capacity expansion amounts to~\SI{10.4}{\giga\watt}
-wind power or \SI{24.1}{\giga\watt} of solar power if the policy is to be fulfilled with a single technology.
-
-Given our baseline assumptions, we find the cost-minimizing system set-up being consistent with policy objectives,
-to rely entirely on wind power.
-In the unconstrained baseline,  \SI{10.4}{\giga\watt} wind power and \SI{0}{\giga\watt} solar PV are added to
-initially available capacities in Austria.
-Wind power is complemented by \SI{2.63}{\giga\watt} of additional natural gas fired combined heat and power
-generation capacity, while less efficient natural gas-fired units with a capacity of \SI{0.42}{\giga\watt} are
-decommissioned along with \SI{0.08}{\giga\watt} oil-fired capacity.
-In effect, \SI{6.23}{\giga\watt} of fossil thermal generation capacities are active, giving rise to \SI{10.5}{\mega\tonne} of \COO emissions per year.
-This compares to current (2016) domestic \COO emissions of power generation from fossil fuels of \SI{8.2}{\mega\tonne}\footnote{For comparability, we calculated \COO emissions based on fossil fuel consumption reported in Austria's energy balance \citep{StatistikAustria2020} and applied the \COO intensity factors from table \ref{table:carbon-intensity}.}.
-The increase in \COO emissions occurs as domestic electricity consumption is increasing while electricity net imports, which accounted for \SI{9.9}{\percent} of total consumption in 2016, are declining.
-As Austria has ample hydro storage capacity in place, no further storage (e.g. from batteries) is added.
-However, the addition of \SI{1.23}{\giga\watt} of compression heat pumps allows to use electricity for heat
-generation, keeping curtailment at \SI{1.5}{\tera\watt\hour\per\year}.
-Annual system cost in Austria amount to $2.9$ billon euro in this scenario.
-Net earnings from exports total $0.7$ billion euro, so that net system cost amounts to $2.2$ billion euro.
-
-\subsection{Restricting wind power}\label{subsec:restricting-wind}
-As we gradually restrict wind power potentials, the system is forced to increasingly rely on electricity generation
-from solar photovoltaics.
-To satisfy the politically targeted share of renewable energy generation, each \SI{}{\giga\watt} of wind power not
-installed needs to be replaced by \SI{2.35}{\giga\watt} of solar PV\@.
-As generation profiles of wind and solar power differ, the substitution of wind power with solar PV brings about
-changes in the system's composition, in system operation, and, consequently, in system cost.
-
-\subsubsection{Changes in system composition and system operation}
-The possibility to exchange electricity with Germany strongly affects system operation in Austria.
-As Austria's electricity and heat generation is less \COO intense than Germany's, variations in the price of \COO
-have marked, potentially unexpected consequences for outcomes in Austria.
-If \COO emissions are costless, annual net exports of electricity total around \SI{5.5}{\tera\watt\hour}, with
-minor variation due to the composition of renewable electricity generation in Austria.
-When \COO prices rise to \SI{30}[\euro]{\per\tonne}, electricity exports increase by around \SI{2}{\tera\watt\hour},
-for any composition of renewable electricity generation in Austria.
-This happens because the more emission intense dispatchable electricity generators in Germany become less competitive
-versus their less emission intense counterparts in Austria.
-In consequence, \COO emissions in Austria increase, while total \COO emissions decline.
-The increase in fossil thermal generation in Austria reverts back to levels observed for no emission cost as the
-price of \COO reaches \SI{60}[\euro]{\per\tonne}.
-Higher \COO prices induce further reductions of \COO emissions in Austria and lead to a higher sensitivity of
-mitigated \COO emissions to the composition of domestic renewable generation.
-
-Moreover, an increase in the price of \COO incentivises renewables development, as electricity generation from
-low-carbon sources becomes more competitive versus fossil generation.
-Yet, our analysis reveals that \COO prices are not the only factor of importance.
-Developed wind resources in Austria are favourably complementary to their German counterparts, with a Pearson
-correlation coefficient of $r_{AT,DE} = 0.235$.
-Solar resources, on the other hand, are highly correlated in Austria and Germany ($r_{AT,DE}=0.948$).
-In effect, wind power has a higher export revenue potential than solar power, particularly at \COO prices of
-\SI{60}[\euro]{\per\tonne} or higher.
-However, exploiting wind power export potentials requires sufficient wind power capacities to be in place.
-
-Neglecting the effects of electricity trade, gradual substitution of wind energy with solar PV in tendency leads to (i) higher capacities of fossil thermal
-generators and (ii) lower capacities of heat pumps in the system, (iii) more \COO emissions, and (iv) more local
-air pollution.
-Consistent across all \COO price scenarios, (i) oil-fired units in Austria are largely shut down while natural gas-fired
-co-generation capacities are expanded.
-Additions of natural-gas fired capacities are lower at high \COO prices and tend to increase with the share of solar
-PV in the system, particularly at high \COO prices.
-However, there are huge differences in fossil capacity additions between \COO prices of \SI{60}[\euro]{\per\tonne}
-and \SI{90}[\euro]{\per\tonne}.
-In the former case, there is very little difference to capacity addition when \COO emissions come at no cost, while
-in the latter case fossil capacity additions are more than halved at high wind shares.
-The amount of wind power in the system (ii) also affects the deployment of compression heat pumps.
-In any scenario considered, heat pumps with a capacity of at least \SI{1.2}{\giga\watt} are operational.
-With an increase of the \COO price to \SI{120}[\euro]{\per\tonne} heat pumps reach a total capacity of
-\SI{2.2}{\giga\watt} when all additional renewable electricity generation comes from solar PV\@.
-Heat pump capacity increases further to a maximum of \SI{2.9}{\giga\watt} when additional renewable electricity is
-sourced from wind power.
-Overall, more heat tends to be generated from fossil sources if \COO prices are low and the system is dominated by
-solar PV\@.
-At high \COO prices and with more wind energy, heat generation is increasingly electrified.
-In effect, a higher share of wind energy (iii) reduces \COO emissions of electricity generation as less energy is sourced from fossil thermal plant.
-This (iv) also reduces emissions of other air pollutants associated with fossil thermal generation.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.95\textwidth]{./figures/sysops.pdf}
-\caption{Opportunity cost of wind power assuming PV overnight cost of 630 EUR/kWp}
-\label{figure:system-operation-base}
+    Subsequently, we report the system effects of a gradual restriction of wind power and the corresponding estimates of the cost of undisturbed landscapes (in terms of the opportunity cost of not installing wind turbines).
+
+    \subsection{Unconstrained Baseline} \label{subsec:baseline}
+    By 2030, Austria generates the bulk of its electricity from intermittent sources.
+    Given our baseline assumptions and carbon price of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{25}[\euro]{\per\tonne\coo}, this comes at a total system cost of  \SI{2.56}[\euro]{} billion.
+    \autoref{tab:cost_baseline} summarizes the breakdown of the total system cost.
+
+    Due to the large share of electricity generation from renewable sources, operation and maintenance cost of \SI{1}[\euro]{} billion exceeds spending on fuels and emission certificates required by thermal power plants.
+    Burning fossil fuels gives rise to \SI{7.1}{\mega\tonne} of \ce{CO2} emissions, which are priced at \SI{177}[\euro]{} million.
+
+    Thermal generation capacity, running on natural gas, waste, and biomass, is not expanded in the baseline scenario.
+    Hence, the annualized investment cost is entirely spent on the expansion of renewable electricity generation capacity.
+    \begin{table*}[th]
+        \centering
+        \begin{threeparttable}
+            \caption{System cost components and trade balance, baseline scenario (2030, \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{25}[\euro]{\per\tonne\coo})}
+            \label{tab:cost_baseline}
+            \begin{tabular}{l c c c c c}
+                \toprule
+                & Unit & Fuel \& \ce{CO2} & Investment & O\& M & Trade Balance \\
+                \midrule
+                Total Cost & million \euro & $824.2$ & $1019.3$ & $718.2$ & $-699.0$ \\
+                \bottomrule
+            \end{tabular}
+            \begin{tablenotes}
+                \footnotesize
+                \item Source: own calculations
+            \end{tablenotes}
+        \end{threeparttable}
+    \end{table*}
+    In our baseline scenario, \SI{10.6}{\giga\watt} of onshore wind turbines are added to the power plant stock, while no investment in solar PV or storage technologies, such as batteries and hydrogen storage, occurs.
+    In total, \SI{59.7}{\tera\watt\hour} electricity is generated from intermittent renewable sources, with another \SI{5.7}{\tera\watt\hour} generated by biomass-fired units and \SI{9.4}{\tera\watt\hour} from inflows into hydro storage reservoirs for a total of \SI{74.8}{\tera\watt\hour} of electricity generated from domestic renewable sources.
+    To reduce quasi-fixed and variable operation and maintenance cost, electricity generation units with a capacity of \SI{0.3}{\giga\watt} (natural gas-fired) and \SI{0.1}{\giga\watt} (oil-fired) are decommissioned in our baseline scenario.
+
+    Revenues from electricity net exports partly compensate for Austria's system costs.
+    On annual balance, Austria exports \SI{10.6}{\tera\watt\hour} of electricity to Germany.
+    This reversal in the current account\footnote{In 2016, Austria's net imports of electricity amounted to \SI{7.2}{\tera\watt\hour} or \SI{9.9}{\percent} of its total electricity consumption \cite{StatistikAustria2020}.}
+    is a necessary consequence of the government's policy target.
+    While (almost) as much electricity is generated from domestic renewable sources as is being consumed, fossil thermal generation units are still required to generate heat and to substitute for renewable generation in times of low electricity generation from intermittent renewable sources.
+
+    \begin{figure*}[h!t]
+        \centering
+        \includegraphics[width=\textwidth]{dispatch_AT}
+        \caption{Monthly electricity generation in Austria by source (2030) for (a) maximum wind power and (b) maximum solar power}
+        \label{fig:monthly-gen-el}
     \end{figure*}
 
-\subsubsection{The opportunity cost of wind power in Austria}\label{subsec:opportunity-cost-wind}
-By relating the change in installed wind power to the change in system cost (or other variables of interest), we
-can derive a measure of the (approximate) marginal cost of wind power expansion (or contraction).
-In consequence, the total cost of undisturbed landscape is equal to the area below the opportunity cost curve,
-starting from the highest wind power capacity down to the level of wind power admitted.
-
+    In the unconstrained baseline scenario with all additional renewable electricity generation coming from wind power, Austria is a net exporter of electricity throughout most of the year.
+    Minor net imports occur in January, as is shown in \figref[a]{monthly-gen-el}.
+    Net exports are relatively stable over the year, as the seasonality of onshore wind power generation is favorably complementary to generation from pre-existing run-of-river hydropower plants (see \autoref{fig:seasonal}), which historically dominate Austria's power generation mix.
+    Austria tends to export electricity to Germany when domestic wind power generation is high (Pearson correlation coefficient $r(X_{AT},wind_{AT})=0.46$, significant at the \SI{1}{\percent} level for hourly data), and to import electricity from Germany when German electricity generation from solar PV is high ($r(I_{AT},PV_{DE})=0.23$, significant at the \SI{1}{\percent} level).
+    Overall, Austria's system cost net of the electricity trade balance equals \SI{1.9}[\euro]{} billion, while air pollution costs (excluding the cost of \ce{CO2} emissions) amount to \SI{226}[\euro]{} million.
+
+    \begin{figure*}[ht]
+        \centering
+        \includegraphics{seasonal}
+        \caption{Seasonal pattern of electricity generation from intermittent sources and electricity consumption (monthly average relative to annual average) in Austria in 2016}
+        \label{fig:seasonal}
+    \end{figure*}
+
+    \subsection{Restricting wind power} \label{subsec:restricting-wind}
+    With the gradual restriction of wind turbine additions, electricity is increasingly sourced from solar PV.
+    This induces changes in the composition and the operation of the electricity system.
+    As long as the policy constraint of generating a given amount of electricity from renewable sources is binding, any decline in wind power generation must be offset by a corresponding increase in power generation from an alternative renewable power generation technology.
+    Thus, the installed capacity of solar PV increases by \SI{1.98}{\giga\watt} for each \si{\giga\watt} reduction in installed wind power capacity.
+    The corresponding change in investment cost is indicated in \autoref{fig:cost_components}.
+    (The corresponding changes in generation and trade are displayed in \autoref{fig:app_dispatch}.)
+    Adding some solar PV to the generation mix reduces the variance of total intermittent electricity generation somewhat, while the impact on minimal intermittent generation is minor.
+    In effect, resource adequacy can be achieved with slightly lower thermal capacity in our baseline scenario.
+    More thermal capacity can be decommissioned so that O\&M cost decreases slightly.
+    This reverses as wind power is phased out completely.
+    The variance of intermittent electricity generation is being reduced only marginally, but minimum intermittent generation declines stronger.
+    Thus, an electricity system dominated by solar PV requires slightly larger thermal generation capacities.
+    In effect, fewer thermal units can be decommissioned so that system-wide O\&M costs increase.
+
+    \begin{figure*}[ht]
+        \centering
+        \includegraphics{cost_components_base_25}
+        \caption{Change in Austria's annual net system cost by components induced by a gradual tightening of the upper limit on wind power installations in the baseline scenario with a \ce{CO2} price of \euro$25$ per tonne.
+Investment cost includes investment in RET.}
+        \label{fig:cost_components}
+    \end{figure*}
+
+    Changes in the cost of fuel and \ce{CO2} emissions are closely linked to trade flows.
+    As wind power deployment is gradually limited to \SI{6}{\giga\watt} from the initial unconstrained optimum, net exports increase by \SI{400}{\giga\watt\hour} in the baseline scenario.
+    The additional electricity exported is sourced from thermal units (between \SIrange{46}{73}{\percent} of the increase in exports) and lower curtailment of intermittent generation (between \SIrange{55}{28}{\percent}).
+    The increase in thermal generation comes in line with a \SI{94}{\kilo\tonne} increase in \ce{CO2} emissions.
+    A further limitation of wind power below \SI{6}{\giga\watt} leads to a \SI{30}{\percent} (\SI{+260}{\giga\watt\hour}) increase in curtailed electricity and a \SI{350}{\giga\watt\hour} decline in thermal electricity generation.
+    These changes come in line with a reduction of electricity net exports to Germany by \SI{810}{\giga\watt\hour}.
+    Moreover, the average price of imports increases by \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{4.69}[\euro]{\per\mega\watt\hour}, while the average price of exports declines by \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{1.76}[\euro]{\per\mega\watt\hour}.
+    Taken together, this weakens Austria's electricity trade balance with Germany by \SI{95}[\euro]{} million.
+    The move towards electricity generation from solar PV also leads to a shift in the seasonal pattern of electricity generation, imports, and exports.
+\autoref{fig:monthly-gen-el} shows the monthly electricity generation and electricity exchange for the two extreme scenarios in which additional adjustable renewable electricity generation is entirely sourced from wind power (\figref[a]{monthly-gen-el}) or solar power (\figref[b]{monthly-gen-el}).
+    Comparing these extremes, reduced thermal power generation in the wake of lower exports, particularly in winter months, results in a \SI{162}{\kilo\tonne} decline of \ce{CO2} emissions from the Austrian power sector.
+    However, total emissions in Austria and Germany are increasing as wind power is replaced by solar PV.
+    The reduction of Austria's \ce{CO2} emissions is possible only because thermal electricity generation, largely from lignite, expands in Germany.
+    Thus, total system emissions increase by \SI{2.3}{\mega\tonne} or approximately \SI{220}{\kilo\tonne} for each \si{\giga\watt} wind power avoided.
+
+    \subsection{The opportunity cost of wind power in Austria} \label{subsec:opportunity-cost-wind}
+    As laid out in \autoref{subsec:restricting-wind}, solar PV and wind power are not perfectly substitutable in Austria.
+    Thus, diverting from capacities deployed in optimum comes at an opportunity cost.
+    This opportunity cost is relatively small for the whole Austro-German market area.
+    With the complete replacement of wind power by solar PV in Austria, total system cost in Austria and Germany increases by \SI{0.5}{\percent}, while external air pollution cost increases by \SI{3.3}{\percent}.
+    However, the annual Austrian system cost net of trade increase by \SI{6.8}{\percent} or \SI{127}[\euro]{} million, while air pollution cost increases by \SI{75}[\euro]{} million or \SI{33}{\percent}.
+    Relating this cost increase to the change in installed wind power capacity allows us to estimate the marginal opportunity cost of wind power under the implicit assumption that solar PV does not give rise to external cost.
+    Accordingly, we estimate the marginal opportunity cost to increase by \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{2666}[\euro]{\per\mega\watt} wind power foregone averaged across the considered \ce{CO2} price scenarios and assuming PV overnight cost of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{625}[\euro]{\per\mega\wattpeak}.
+    The opportunity cost is small for the first MW of wind power displaced and, depending on the price of CO2, reaches between \SI{32400}[\euro]{} and \SI{41200}[\euro]{} for the last \si{\mega\watt} of wind power displaced.
+
+    Given these estimates, we can calculate the valuation of undisturbed landscapes implied by announced government policies.
+    The government intends to source an additional \SI{10}{\tera\watt\hour} annually from wind power, which corresponds to an additional \SI{5.05}{\giga\watt} wind turbines being installed.
+    At this level of wind power deployment, the opportunity cost of wind power averaged over all considered \ce{CO2} price scenarios is approximately \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{22500}[\euro]{\per\mega\watt} and year.
+    The present value of the cost of undisturbed landscapes implied by government policies amounts to approximately \SI{1.3}[\euro]{} million over the lifetime of a standard \SI{3.5}{\mega\watt} wind turbine at a \SI{5}{\percent} discount rate.
     \begin{figure*}[h!t]
         \centering
-\begin{subfigure}[b]{0.475\textwidth}
-\centering
-\includegraphics[width=\textwidth]{./figures/undisturbed_base_share.pdf}
-\caption{Relative substitution}
-\label{fig:rel_substitution}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{0.475\textwidth}
-\centering
-\includegraphics[width=\textwidth]{./figures/undisturbed_base.pdf}
-\caption{Absolute substitution}
-\label{fig:abs_substitution}
-\end{subfigure}
-\caption{Annual opportunity cost of wind power, assuming PV overnight cost of 630 EUR/kWp}
-\label{figure:opportunity-cost-base}
+        \includegraphics{oc_base}
+        \caption{Annual opportunity cost of wind power in Austria, assuming PV overnight cost of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{625}[\euro]{\per\kilo\wattpeak} (2030)}
+        \label{fig:opportunity-cost-base}
     \end{figure*}
 
-With a gradual restriction of deployable wind power capacities, we find opportunity costs of wind power
-to increase by about \SI{3900}[\euro]{} for each \SI{}{\mega\watt} wind power substituted by solar PV across all studied scenarios.
-At PV overnight cost of \SI{630}[\euro]{\per\mega\watt} (corresponding to a mix of rouhgly \SI{50}{\percent} rooftop and
-\SI{50}{\percent} open space installations), the annual opportunity cost of the last \SI{}{\mega\watt} wind power replaced reaches close to \SI{62500}[\euro]{}.
-
-At a wind power capacity consistent with the policy goal of generating \SI{21}{\tera\watt\hour} of electricity
-from renewable sources, opportunity cost for wind power are close to \SI{20000}[\euro]{\per\mega\watt}, with a
-notable exception for \COO prices of \SI{60}[\euro]{\per\tonne}.
-In this case, opportunity cost of wind power are estimated at close to \SI{10000}[\euro]{\per\mega\watt}, about half
-the size of opportunity cost estimated for all other \COO price scenarios.
-
-Based on these findings, we can also approximate the cost of undisturbed landscapes implied by government targets.
-Given the goal of generating \SI{21}{\tera\watt\hour} electricity from renewable sources and the wind energy target
-generation of \SI{10}{\tera\watt\hour}, we can infer $\Phi = 0.476$ according to government plans.
-At this level, we estimate opportunity costs of wind power between \SI{25000}[\euro]{\per\mega\watt} and \SI{40000}[\euro]{\per\mega\watt} depending on the carbon price.
-Over an assumed lifetime of $30$ years, this implies an annual gain from a contemporary \SI{3.5}{\mega\watt}
-wind turbine between \SI{1.35}{} and \SI{2.15}{} million euro, if future gain is discounted at \SI{5}{\percent}.
-
     \subsection{Sensitivity Analysis} \label{subsec:sensitivity}
-A major factor of influence for our results are the assumed capital cost of wind turbines and solar PV\@.
-Therefore, we complement our baseline scenario with a sensitivity analysis on overnight investment cost of solar
-photovoltaics.
-We gradually lower the capital cost of solar PV from our baseline assumption of \SI{630}[\$]{\per\kilo\wattpeak}
-(corresponding to \SI{55}{\percent} of rooftop and \SI{45}{\percent} open space capacity) to
-\SI{275}[\euro]{\per\kilo\wattpeak}, which Vartiainen et al.~\cite{Vartiainen2019} project for open-space,
-utility-scale solar PV in the year 2030.
-
-\begin{figure*}[h!t]
-\centering
-\includegraphics[width=0.8\textwidth]{./figures/undisturbed_low.pdf}
-\caption{Opportunity cost of wind power}
-\end{figure*}
-
-Under these assumptions, the impact of \COO prices on the opportunity cost of wind power increases.
-If all additional PV is realized as open-space, utility-scale installation (i.e.\ PV capital cost of
-\SI{275}[\euro]{\per\kilo\wattpeak}), the optimal deployment of wind power declines and approaches zero at \COO prices
-of \SI{30}[\euro]{\per\tonne} or below.
-In consequence, we can not determine opportunity cost of wind power given that both \COO prices and capital cost of
-solar PV are very low.
-For higher \COO prices, the optimal deployment of wind power is strictly positive, even at low capital cost of solar
-PV and we find opportunity cost to increase by about \SI{2550}[\euro]{} for each \SI{}{\mega\watt} wind power
-foregone.
-The last \SI{}{\mega\watt} of wind power substituted with solar PV comes at opportunity cost between
-\SI{20000}[\euro]{} and \SI{56000}[\euro]{}, depending on the prevailing \COO price.
-
-\subsection{Limitations of the Analysis}
+    To assess the robustness of our results, we analyze the sensitivity of model outcomes to changes in parameters that can be expected to have a particular influence.
+The effects of changes in capital cost and electricity transmission capacities are reported below.
+
+    \begin{figure}[h!t]
+        \centering
+        \includegraphics{Sensitivities}
+        \caption{Comparison of the annual opportunity cost of wind power in Austria derived from sensitivity analyses for low capital cost of PV (\SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{560}[\euro]{\per\kilo\wattpeak}) and high interconnector capacity (\SI{10}{\giga\watt}) with baseline.
+Each point represents a specific combination of a \ce{CO2} price assumption and an upper limit on deployable wind power.
+Lines represent cubic fit to scenario results.}
+        \label{fig:sensitivity}
+    \end{figure}
+
+    \subsubsection{Sensitivity to capital cost assumptions}
+    To assess the sensitivity of results to changes in the assumed capital cost of wind turbine and solar PV installations, we gradually lower the cost of solar PV from our baseline assumption of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{625}[\euro]{\per\kilo\wattpeak}, corresponding to a mix of \SI{50}{\percent} rooftop and \SI{50}{\percent} open space installations at 2030 prices, to \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{380}[\euro]{\per\kilo\wattpeak} which corresponds to \SI{100}{\percent} open-space utility-scale PV according to our baseline assumptions.
+    To account for uncertainty in price projections, we lower PV capital cost further down to \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{275}[\euro]{\per\kilo\wattpeak}, which Vartiainen et al.~\cite{Vartiainen2019} project for open-space utility-scale solar PV projects in 2030.
+
+    \begin{figure}[h!t]
+        \centering
+        \includegraphics{S_PVCost_P_WindOnshore}
+        \caption{Optimal deployment of wind power in Austria conditional on the capital cost of solar PV}
+        \label{fig:opt_wind_cap}
+    \end{figure}
+
+    As illustrated by \autoref{fig:opt_wind_cap}, we find the optimal level of wind power deployment declining in the capital cost of PV.
+    The marked difference in the effect of the capital cost of solar PV on the optimal level of wind power deployed in Austria is a result of the interactions between a renewables target and a carbon price.
+    \ref{subsec:interactions} provides further details about these interactions.
+    For emission prices of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{50}[\euro]{\per\tonne\coo} or lower, this decline is particularly pronounced, with a reduction of the optimally deployed wind power capacity by approximately \SI{0.6}{\giga\watt} for each \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{10}[\euro]{\per\kilo\wattpeak} decline in the capital cost of solar PV.
+    At capital costs of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{400}[\euro]{\per\kilo\wattpeak} or lower, solar PV covers all additionally required renewable energy generation if the price of \ce{CO2} is at \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{25}[\euro]{\per\tonne} or lower.
+    For a carbon price of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{50}[\euro]{\per\tonne\coo}, about \SI{2}{\giga\watt} of wind power is deployed in optimum at this capital cost.
+    As capital costs fall further, deployed wind power capacity descends slowly, approaching zero as PV capital cost reach \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{275}[\euro]{\per\kilo\wattpeak}.
+    With \ce{CO2} price at \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{75}[\euro]{\per\tonne} or higher, when government policy is not binding anymore, deployed wind power capacity remains sturdy.
+    Even at a capital cost of PV as low as \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{275}[\euro]{\per\kilo\wattpeak}, added wind power capacity tops \SI{6}{\giga\watt}.
+    Moreover, it is worth noting that the impact of the lower PV capital cost on \ce{CO2} emissions largely depends on the prevailing \ce{CO2} price level.
+    With emission prices of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{25}[\euro]{\per\tonne\coo} or lower, a halving of PV capital cost reduces total system emissions by only \SI{2}{\percent}.
+    In comparison, emission reductions of about \SI{15}{\percent} are realized when \ce{CO2} is priced at least at \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{50}[\euro]{\per\tonne}.
+
+    In line with less wind power capacities deployed, the opportunity cost of wind power declines.
+    Given PV's capital cost of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{560}[\euro]{\per\kilo\wattpeak}, a \SI{10}{\percent} reduction from our baseline, we estimate the opportunity cost of wind power at \SI{5.05}{\giga\watt} added at \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{17500}[\euro]{\per\mega\watt} wind power avoided, a \SI{21.9}{\percent} reduction compared to baseline.
+    However, the reduction in opportunity cost is smaller when restrictions on wind power deployment are tighter.
+    The last megawatt of wind power foregone comes at an opportunity cost of approximately \SI{27200}[\euro]{}, as long as emitting a tonne of \ce{CO2} costs \SI{75}[\euro]{} or less.
+
+    \subsubsection{Sensitivity to transmission constraints} \label{subsubsec:sens_transmission}
+    Following a decision by ACER, electricity transmission capacity between Austria and Germany was lowered from an initially (virtually) unlimited capacity to \SI{4.9}{\giga\watt} in 2018.
+    In consequence, the previously unified Austro-German electricity market zone was split up, allowing for diverging wholesale electricity prices.
+    The cap in transmission capacity was implemented to prevent so-called "loop flows" of electricity from Northern Germany through Eastern Europe to Austria and Southern Germany.
+    These loop flows arose from an imbalance of supply in the North and demand in the South, which could all too often not be resolved by the frequently congested inner-German transmission grid.
+
+    \begin{figure}[t]
+        \centering
+        \includegraphics{cost_components_no_bottleneck_25}
+        \caption{Change in Austria's annual net system cost by components with \SI{10}{\giga\watt} transmission capacity and a \ce{CO2} price of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{25}[\euro]{\per\tonne}}
+        \label{fig:sens_interconn_costco}
+    \end{figure}
+
+    Given that current energy policy goals turn Austria from a net importer into a net exporter of electricity, we have analyzed the effects of a potential lifting of ACER's decision to cap electricity transmission by setting transmission capacity to \SI{10}{\giga\watt}.
+    Under this assumption, the opportunity cost of wind power is higher compared to baseline at low penetrations of solar PV.
+    However, the increase in opportunity cost as the PV penetration increases is less pronounced compared to baseline.
+
+    Higher transmission capacity allows (fossil) power plants to generate and export electricity at times when interconnectors with lower transmission capacity would be congested.
+    In consequence, use of fossil fuels for thermal power generation and \ce{CO2} emissions are higher (by \SIrange{390}{670}{\kilo\tonne} at a carbon price of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{25}[\euro]{\per\tonne\coo}) compared to our baseline with lower interconnection capacity.
+    Moreover, thermal capacity is not being shut down (as in the baseline scenario), so that changes in O\&M cost are very small.
+
+    Higher interconnector capacity allows for higher quantities of imports and exports.
+    Indeed, Austrian imports are up by \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{3.1}{\tera\watt\hour\per\year}, while exports increase by about \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{2.6}{\tera\watt\hour\per\year}, compared to the unconstrained baseline scenario.
+
+    Even though net exports decline, Austria's (net) export surplus increases, mostly on the back of a lower average import price for electricity from Germany.
+    Moreover, increased interconnector capacity leads to a trade balance that is less responsive to changes in installed RET capacities.
+    Thus, the opportunity cost of wind power are higher at at high wind power penetration, but lower at low wind power penetration (see \autoref{fig:sens_interconn_costco}).
+
+    \subsection{Limitations of the Analysis} \label{subsec:limitations}
     It is worth noting that our analysis is not spatially resolved, which has important consequences.
@@
 after generation increased by \SI{21}{\tera\watt\hour}.
-For Austria, Hoeltinger et al. \cite{Hoeltinger2016}  identified \SI{45}{\tera\watt\hour} technically feasible wind resource potential with better or similar quality than we assumed in our analysis.
-Hence, an additional \SI{21}{\tera\watt\hour} of electricity generated from wind power is feasible under our assumptions.
+    However, Hoeltinger et al.
+\cite{Hoeltinger2016}, for example, identified technically feasible wind resource potential of \SI{45}{\tera\watt\hour} in Austria which is of better or similar quality than we assumed in our analysis.
+    Hence, \SI{21}{\tera\watt\hour} additional electricity generation from wind power is feasible under our assumptions.
     Yet, technical potentials are currently restricted by law in four of Austria's nine federal states.
@@
     Second, we implicitly assume that the cost of grid expansion does not differ between scenarios relying entirely on wind power and scenarios relying on solar PV exclusively.
-Here, it makes a difference whether solar PV is realized as rooftop systems in distribution grids or as utility-scale open-space systems which are connected to transmission grids.
+    Here, it makes a difference whether solar PV is realized as rooftop systems in distribution grids or as utility-scale open-space systems that are connected to transmission grids.
     In the latter case, there is little reason to believe grid cost would differ substantially from wind power expansion, which also happens in transmission grids.
-Horowitz et al. \cite{Horowitz2018} survey the nascent literature on the cost of integrating distributed solar PV systems in distribution grids.
-These cost are found to depend on numerous and uncertain factors, leading the authors to the conclusion that "no generalized `cost of grid integration' for PV can be obtained".
-Nevertheless, a study for Agora Energiewende \cite{Fuerstenwerth2015} estimates grid integration cost of solar PV in Europe at \SI{6}[\euro]{\per\mega\watt\hour} in distribution grids, and at \SI{1.5}[\euro]{\per\mega\watt\hour} in transmission grids.
+    \cite{Horowitz2018} survey the nascent literature on the cost of integrating distributed solar PV systems in distribution grids.
+    These costs are found to depend on numerous and uncertain factors, leading the authors to the conclusion that "no generalized `cost of grid integration' for PV can be obtained".
+    Nevertheless, \cite{Fuerstenwerth2015} estimate grid integration cost of solar PV in Europe at \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{6}[\euro]{\per\mega\watt\hour} in distribution grids, and at \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{1.5}[\euro]{\per\mega\watt\hour} in transmission grids.
     Estimates for specific grids, however, vary widely.
-Yet, the available evidence suggests that any grid cost advantage of PV over onshore wind is likely small.
-
-% %%%%% %%%%% %%%%% CONCLUSIONS AND POLICY IMPLICATIONS %%%%% %%%%% %%%%% %%%%%
+    Given the available evidence, we are inclined to assume that any grid cost advantage of PV over onshore wind is likely small.
+
+    Lastly, in our analysis we focus on the time-horizon up to 2030.
+    Thus, our findings are not necessarily transferable to the (almost) fully decarbonized electricity systems we might establish in the more distant future.
+    Yet, fundamental properties of renewable resources, such as seasonal and diurnal patterns in wind speeds or solar irradiation, will persist over these time frames.
+    Thus, we expect to find positive opportunity costs of wind power also in a decarbonized Austrian electricity system, particularly as future options to facilitate the system integration of variable renewables that we did not model, are unlikely to come at significantly lower cost than the included options.
+
     \section{Conclusions and Policy Implication} \label{sec:conclusions-policy-implication}
-% * future research should consider the opportunity cost of RET in terms of grid cost (including distribution grids)
     Not disturbing landscapes can come at considerable opportunity cost, as our analysis reveals.
-How high this cost is, depends on the valuation of \COO and on the cost of the best alternative to wind power.
-If the cost of \COO is low and the best alternative solar PV is realized at utility scale in open space, wind power's system cost advantage vanishes in Austria.
-If, on the other hand, the cost of \COO is high, if open-space PV itself causes negative externalities at the local level, or if there is a preference for rooftop solar PV, there are considerable opportunity cost of wind power.
-Given the current policy target of adding capacities for generating \SI{10}{\tera\watt\hour} ($\phi=0.48$) wind energy, under our baseline assumptions the present value of the cost of undisturbed landscapes amounts to at least \SI{1.35}[\euro]{\million} over the lifetime of each \SI{3.5}{\mega\watt} wind turbine not erected.\footnote{We assume a lifetime of 30 years and a discount factor of \SI{5}{\percent}.}
-At solar PV overnight cost that correspond to \SI{100}{\percent} open-space PV, the opportunity cost of wind power is reduced to \SI{480000}[\euro]{} at \COO prices of \SI{60}[\euro]{\per\tonne}.
-In Austria, only open-space, utility scale solar PV has the potential to substitute wind power at comparable system cost, and only if a tonne of \COO emitted is valued below \SI{60}{\euro}.
-However, open-space PV itself interferes with landscapes.
-Relying completely on open-space solar PV would require about \SI{722}{\square\kilo\meter} land\footnote{In line with \cite{Guennewig2007} we assume land usage of \SI{29.55}{\square\meter} per \SI{}{\kilo\wattpeak} of non-tracking silicon wafer technology.} for replacing the approximately 2\,970 wind turbines of the \SI{3.5}{\mega\watt} class that would be needed to meet policy goals by wind power alone.
-In the light of our findings, policy makers need to consider whether a wind power-leaning expansion of RET together with a compensation of local residents for harm inflicted by wind turbines could be a socially acceptable and cost-saving alternative to the large-scale expansion of solar PV on roof-tops.
-Such an undertaking would, however, require spatially explicit estimates of wind turbine harm.
+    How high this cost is, depends on the valuation of \ce{CO2} and the cost of the best alternative to wind power.
+    If the cost of \ce{CO2} is low and the best alternative solar PV is realized at utility-scale in open space, wind power's system cost advantage vanishes in Austria.
+    If, on the other hand, the cost of \ce{CO2} is high, if open-space PV itself causes negative external effects at the local level, or if there is a preference for rooftop solar PV, the opportunity cost of wind power is worthy of consideration.
+
+    Given the current policy target of adding capacities for generating \SI{10}{\tera\watt\hour} wind energy, the present value of the cost of undisturbed landscapes amounts to at least \SI{995000}[\euro]{} over the lifetime of each \SI{3.5}{\mega\watt} wind turbine not erected.\footnote{We assume a lifetime of 30 years and a discount factor of \SI{5}{\percent}.}
+
+    While these numbers indicate that in the absence of local wind turbine externalities, the government targets for RET deployment aren't economically efficient, a complete policy evaluation must weigh the cost of undisturbed landscapes against its benefits arising from the preservation of current landscapes aesthetics, for example.
+    Yet, any significant expansion of renewable electricity generation technologies will interfere with landscapes.
+    Relying completely on open-space solar PV would require about \SI{315}{\square\kilo\meter} land\footnote{We assume land usage of \SI{15}{\square\meter} per \si{\kilo\wattpeak} of non-tracking silicon wafer technology.}
+    for replacing the approximately 3\,030 wind turbines of the \SI{3.5}{\mega\watt} class that would be needed to meet policy goals by wind power alone.
+
+    For a comprehensive policy evaluation, our findings need to be complemented by a spatially highly resolved analysis of the local cost of disturbed landscapes.
+    Together, such findings would allow determining the socially optimal deployment of renewable energy generation technologies.
     Here, we see a fruitful field for further research.
@@
     As we have shown, a large-scale expansion of wind turbines comes with an increase in electricity exports, i.e.
-at high penetration some wind power capacity is mostly used to generate electricity for exporting.
-Revenues from these export are accruing to the owners of power generation assets, while the negative externalities of wind turbines have to be borne by local residents.
+at high penetration some wind power capacity is mostly used to generate electricity for export.
+    Revenues from these export are accruing to the owners of power generation assets, while negative external effects of wind turbines have to be borne by local residents.
     This imbalance could be narrowed down by policies requiring wind turbine owners to share wind turbine income with affected residents, either directly or through transfers.
@@
 
-%\section{Data Availability}\label{sec:data-availability}
-%Data and model is avaliable on github \url{}. Raw data is sourced from danish Energy Agency \url{}, ENTSO-E's transparency database \url{},
+    \section*{Acknowledgements}
+    Sebastian Wehrle would like to thank Katharina Gruber, Peter Regner and Claude Klöckl for helpful comments on early drafts of the manuscript.
+    We gratefully acknowledge support from the European Research Council (“reFUEL” ERC-2017-STG 758149).
 
@@
 
-\section{Description of the power system model \emph{medea}}\label{sec:medea-desc}
-
-\subsection{Sets}\label{sets}
+    \section{Change in dispatch and trade for a gradual tightening of the upper limit on wind power installations}
+    \begin{figure}[h!]
+        \centering
+        \begin{subfigure}[b]{0.49\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{gen_by_fuel_25}
+            \caption{Carbon price of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{25}[\euro]{\per\tonne\coo}}
+            \label{fig:app_dispatch_a}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.49\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{gen_by_fuel_50}
+            \caption{Carbon price of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{50}[\euro]{\per\tonne\coo}}
+            \label{fig:app_dispatch_b}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.49\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{gen_by_fuel_75}
+            \caption{Carbon price of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{75}[\euro]{\per\tonne\coo}}
+            \label{fig:app_dispatch_c}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.49\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{gen_by_fuel_100}
+            \caption{Carbon price of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{100}[\euro]{\per\tonne\coo}}
+            \label{fig:app_dispatch_c}
+        \end{subfigure}
+
+        \caption{Change in Austria's annual electricity generation by fuel and electricity exchange with Germany induced by a gradual tightening of the upper limit on wind power installations.}
+        \label{fig:app_dispatch}
+    \end{figure}
+
+    \section{The interaction of the policy target with \ce{CO2} prices} \label{subsec:interactions}
+    The Austrian government seeks to achieve its renewable electricity generation target through the provision of technology-specific subsidies.
+    The subsidization of renewable electricity generation induces an expansion of installed RET capacities.
+    Pricing carbon dioxide emissions would be an alternative policy to incentivize investment in renewable energy generation, as carbon pricing would improve the competitiveness of carbon-free generators versus their fossil counterparts.
+    The higher the price of \ce{CO2}, the stronger the incentive for investment in renewable energy technologies.
+    Hence, there is a \ce{CO2} price that is equivalent to reaching the same emission reduction.
+    In our baseline scenario, the policy reduces \ce{CO2} emissions by \SI{1.2}{\mega\tonne} versus the same scenario without the policy constraint.
+    An emission reduction of similar size could also be reached through raising the \ce{CO2} price to approximately \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{40}[\euro]{\per\tonne}.
+    However, the resulting system is not equivalent in terms of installed capacities of electricity generated from domestic renewable sources.
+    Reaching this policy objective would require a carbon price of approximately \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{63}[\euro]{\per\tonne\coo}.
+    For emission prices above this level, the government policy is not binding anymore.
+    A major difference between a binding government policy objective and a hypothetical \ce{CO2} price lies in the measures' effect on electricity trade.
+    Under the renewables target, Austria necessarily turns into a net exporter of electricity and benefits from the associated revenues.
+    Under \ce{CO2} pricing, Austria would remain a net importer of electricity and maintain a negative balance of payments in the electricity sector up to a \ce{CO2} price of approximately \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{45}[\euro]{\per\tonne}.
+
+    \section{Description of the power system model \emph{medea}} \label{sec:medea-desc}
+
+    \subsection{Sets} \label{sets}
     Sets are denoted by upper-case latin letters, while set elements are denoted by lower-case latin letters.
 
-\begin{longtable}{p{0.15\textwidth}p{0.15\textwidth}p{0.2\textwidth}p{0.45\textwidth}}%{LLLp{6cm}
-\caption{Sets}\\
+    \begin{longtable}{p{0.15\textwidth}p{0.15\textwidth}p{0.2\textwidth}p{0.45\textwidth}}
+        \caption{Sets} \\
         \toprule
-mathematical symbol & programming symbol & description & elements\\
+        mathematical symbol & programming symbol & description                      & elements                                                                                                                                                                                                                                                                                                                                                                                                                                                \\
         \midrule
@@
         \endlastfoot
-% the contents of the table
-$f \in F$            &\texttt{f}      & fuels & \texttt{nuclear, lignite, coal, gas, oil, biomass, power} \\
-$i \in I$            &\texttt{i}      & power generation technologies & \texttt{nuc, lig\_stm, lig\_stm\_chp, lig\_boa, lig\_boa\_chp, coal\_sub, coal\_sub\_chp, coal\_sc, coal\_sc\_chp, coal\_usc, coal\_usc\_chp, coal\_igcc, ng\_stm, ng\_stm\_chp, ng\_ctb\_lo, ng\_ctb\_lo\_chp, ng\_ctb\_hi, ng\_ctb\_hi\_chp, ng\_cc\_lo, ng\_cc\_lo\_chp, ng\_cc\_hi, ng\_cc\_hi\_chp, ng\_mtr, ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm, oil\_stm\_chp, oil\_ctb, oil\_ctb\_chp, oil\_cc, oil\_cc\_chp, bio, bio\_chp, heatpump\_pth}                                 \\
-$h \in H \subset I$  &\texttt{h(i)}   & power to heat technologies & \texttt{heatpump\_pth} \\
-$j \in J \subset I$  &\texttt{j(i)}   & CHP technologies & \texttt{lig\_stm\_chp, lig\_boa\_chp, coal\_sub\_chp, coal\_sc\_chp, coal\_usc\_chp, ng\_stm\_chp, ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp, ng\_cc\_lo\_chp, ng\_cc\_hi\_chp, ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm\_chp, oil\_ctb\_chp, oil\_cc\_chp, bio\_chp} \\
-$k \in K$            &\texttt{k}      & storage technologies & \texttt{res\_day, res\_week, res\_season, psp\_day, psp\_week, psp\_season, battery} \\
-$l \in L$            &\texttt{l}      & feasible operation region limits & \texttt{l1, l2, l3, l4}\\
-$m \in M$            &\texttt{m}      & energy products & \texttt{el, ht} \\
-$n \in N$            &\texttt{n}      & intermittent generators & \texttt{wind\_on, wind\_off, pv, ror} \\
-$t \in T$            &\texttt{t}      & time periods (hours)               & \texttt{t1, t2, \ldots, t8760}\\
-$z \in Z$            &\texttt{z}      & market zones & \texttt{AT, DE} \\
+        $f \in F$           & \texttt{f}         & fuels                            & \texttt{nuclear, lignite, coal, gas, oil, biomass, power}                                                                                                                                                                                                                                                                                                                                                                                               \\
+        $i \in I$           & \texttt{i}         & energy conversion technologies   & \texttt{nuc, lig\_stm, lig\_stm\_chp, lig\_boa, lig\_boa\_chp, coal\_sub, coal\_sub\_chp, coal\_sc, coal\_sc\_chp, coal\_usc, coal\_usc\_chp, coal\_igcc, ng\_stm, ng\_stm\_chp, ng\_ctb\_lo, ng\_ctb\_lo\_chp, ng\_ctb\_hi, ng\_ctb\_hi\_chp, ng\_cc\_lo, ng\_cc\_lo\_chp, ng\_cc\_hi, ng\_cc\_hi\_chp, ng\_mtr, ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm, oil\_stm\_chp, oil\_ctb, oil\_ctb\_chp, oil\_cc, oil\_cc\_chp, bio, bio\_chp, heatpump\_pth} \\
+        $h \in H \subset I$ & \texttt{h(i)}      & power to heat technologies       & \texttt{heatpump\_pth}                                                                                                                                                                                                                                                                                                                                                                                                                                  \\
+        $j \in J \subset I$ & \texttt{j(i)}      & CHP technologies                 & \texttt{lig\_stm\_chp, lig\_boa\_chp, coal\_sub\_chp, coal\_sc\_chp, coal\_usc\_chp, ng\_stm\_chp, ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp, ng\_cc\_lo\_chp, ng\_cc\_hi\_chp, ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm\_chp, oil\_ctb\_chp, oil\_cc\_chp, bio\_chp}                                                                                                                                                                                           \\
+        $k \in K$           & \texttt{k}         & storage technologies             & \texttt{res\_day, res\_week, res\_season, psp\_day, psp\_week, psp\_season, battery}                                                                                                                                                                                                                                                                                                                                                                    \\
+        $l \in L$           & \texttt{l}         & feasible operation region limits & \texttt{l1, l2, l3, l4}                                                                                                                                                                                                                                                                                                                                                                                                                                 \\
+        $m \in M$           & \texttt{m}         & energy products                  & \texttt{el, ht}                                                                                                                                                                                                                                                                                                                                                                                                                                         \\
+        $n \in N$           & \texttt{n}         & intermittent generators          & \texttt{wind\_on, wind\_off, pv, ror}                                                                                                                                                                                                                                                                                                                                                                                                                   \\
+        $t \in T$           & \texttt{t}         & time periods (hours)             & \texttt{t1, t2, \ldots, t8760}                                                                                                                                                                                                                                                                                                                                                                                                                          \\
+        $z \in Z$           & \texttt{z}         & market zones                     & \texttt{AT, DE}                                                                                                                                                                                                                                                                                                                                                                                                                                         \\
     \end{longtable}
@@
     \newpage
-\subsection{Parameters}\label{parameters}
+
+    \subsection{Parameters} \label{parameters}
     Parameters are denoted either by lower-case greek letters or by upper-case latin letters.
@@
     \begin{longtable}{p{0.15\textwidth}p{0.3\textwidth}p{0.35\textwidth}p{0.15\textwidth}}
-\caption{Parameters}\\
+        \caption{Parameters} \\
         \toprule
-mathematical symbol & programming symbol & description & unit\\
+        mathematical symbol & programming symbol & description & unit \\
         \midrule
@@
         \endlastfoot
-% the contents of the table
-$\delta_{z,zz}$              &\texttt{DISTANCE(z,zz)}                           & distance between countries' center of gravity & km                        \\
-$\varepsilon_{f}$            &\texttt{CO2\_INTENSITY(f)}                        & fuel emission intensity & $\text{t}_{\COO}$/ MWh  \\
-$\eta_{i,m,f}$               &\texttt{EFFICIENCY\_G(i,m,f)}                     & power plant efficiency & MWh / MWh                 \\
-$\eta^{out}_{z,k}$           &\makecell[l]{\texttt{EFFICIENCY\_S\_OUT(k)}}      & discharging efficiency &                           \\
-$\eta^{in}_{z,k}$            &\makecell[l]{\texttt{EFFICIENCY\_S\_IN(k)}}       & charging efficiency &                           \\
-$\lambda_{z}$                &\texttt{LAMBDA(z)}                                & scaling factor for peak load &                           \\
-$\mu_{z}$                    &\texttt{VALUE\_NSE(z)}                            & value of lost load & \EUR/ MWh               \\
-$\rho_{z,t,k}$               &\makecell[l]{\texttt{INFLOWS(z,t,k)}}             & inflows to storage reservoirs & MW                        \\
-$\sigma_{z}$                 &\texttt{SIGMA(z)}                                 & scaling factor for peak intermittent generation &                           \\
-$\phi_{z,t,n}$               &\texttt{GEN\_PROFILE(z,t,n)}                      & intermittent generation profile & $[0,1]$                  \\
-$\widehat{\phi}_{z,n}$       &\texttt{PEAK\_PROFILE(z,n)}                       & peak intermittent generation profile & $[0,1]$                   \\
-$\chi_{i,l,f}$               &\texttt{FEASIBLE\_INPUT(i,l,f)}                   & inputs of feasible operating region & $[0,1]$                  \\
-$\psi_{i,l,m}$               &\texttt{FEASIBLE\_OUTPUT(i,l,m)}                  & output tuples of feasible operating region & $[0,1]$                  \\
+        $\Bar{\alpha}_{f}$ & \texttt{AIR\_POL\_COST\_FIX(f)} & external cost of air pollution
+        & \SI{}[\euro]{\per\mega\watt}
+        \\
+        $\alpha_{f}$ & \texttt{AIR\_POL\_COST\_VAR(f)} & external cost of air pollution & \SI{}[\euro]{\per\mega\watt\hour} \\
+        $\delta_{z,zz}$ & \texttt{DISTANCE(z,zz)} & distance between countries' center of gravity
+        & km
+        \\
+        $\varepsilon_{f}$ & \texttt{CO2\_INTENSITY(f)} & fuel emission intensity & $\text{t}_{\ce{CO2}}$/ MWh \\
+        $\eta_{i,m,f}$ & \texttt{EFFICIENCY\_G(i,m,f)} & power plant efficiency & MWh / MWh \\
+        $\eta^{out}_{z,k}$ & \makecell[l]{\texttt{EFFICIENCY\_S\_OUT(k)}} & discharging efficiency
+        &
+        \\
+        $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{EFFICIENCY\_S\_IN(k)}} & charging efficiency
+        &
+        \\
+        $\lambda_{z}$ & \texttt{LAMBDA(z)} & scaling factor for peak load & \\
+        $\mu_{z}$ & \texttt{VALUE\_NSE(z)} & value of lost load & \EUR/ MWh \\
+        $\rho_{z,t,k}$ & \makecell[l]{\texttt{INFLOWS(z,t,k)}} & inflows to storage reservoirs
+        & MW
+        \\
+        $\sigma_{z}$ & \texttt{SIGMA(z)} & scaling factor for peak intermittent generation & \\
+        $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(z,t,n)} & intermittent generation profile & $[0,1]$ \\
+        $\widehat{\phi}_{z,n}$ & \texttt{PEAK\_PROFILE(z,n)} & peak intermittent generation profile
+        & $[0,1]$
+        \\
+        $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(i,l,f)} & inputs of feasible operating region
+        & $[0,1]$
+        \\
+        $\psi_{i,l,m}$ & \texttt{FEASIBLE\_OUTPUT(i,l,m)} & output tuples of feasible operating region
+        & $[0,1]$
+        \\
         % UPPER-CASE LATIN LETTERS
-$C^{r}_{z,n}$                &\texttt{CAPITALCOST\_R(z,n)}                      & capital cost of intermittent generators (specific, annuity)     & \EUR/ MW                \\
-$C^{g}_{z,i}$                &\texttt{CAPITALCOST\_G(z,i)}                      & capital cost of thermal generators (specific, annuity)          & \EUR/ MW                 \\
-$C^{s}_{z,k}$                &\makecell[l]{\texttt{CAPITALCOST\_S(z,k)}}        & capital cost of storages - power (specific, annuity)            & \EUR/ MW                \\
-$C^{v}_{z,k}$                &\makecell[l]{\texttt{CAPITALCOST\_V(z,k)}}        & capital cost of storages - energy (specific, annuity)           & \EUR/ MW                \\
-$C^{x}$                      &\texttt{CAPITALCOST\_X}                           & capital cost of transmission capacity (specific, annuity)       & \EUR/ MW                \\
-$D_{z,t,m}$                  &\texttt{DEMAND(z,t,m)}                            & energy demand & GW                        \\
-$\widehat{D}_{z,m}$          &\texttt{PEAK\_LOAD(z,m)}                          & peak demand & GW                        \\
-$\widetilde{G}_{z,i}$        &\texttt{INITIAL\_CAP\_G(z,tec)}                   & initial capacity of dispatchable generators & GW                        \\
-$O^{g}_{i}$                  &\texttt{OM\_COST\_G\_VAR(i)}                      & variable O\&M cost of dispatchable generators & \EUR/ MWh               \\
-$O^{r}_{z,n}$                &\texttt{OM\_COST\_R\_VAR(z,n)}                    & variable O\&M cost of intermittent generators & \EUR/ MWh               \\
-$P^{e}_{t,z}$                &\texttt{PRICE\_CO2(t,z)}                          & \COO price &\EUR/$\text{t}_{\COO}$ \\
-$P_{t,z,f}$                  &\texttt{PRICE\_FUEL(t,z,f)}                       & fuel price & \EUR/ MWh               \\
-$Q^{g}_{i}$                  &\texttt{OM\_COST\_G\_QFIX(i)}                     & quasi-fixed O\&M cost of dispatchable generators & \EUR/ MW                \\
-$Q^{r}_{z,n}$                &\texttt{OM\_COST\_R\_QFIX(z,n)}                   & quasi-fixed O\&M cost of intermittent generators & \EUR/ MW                \\
-$\widetilde{R}_{z,n}$        &\texttt{INITIAL\_CAP\_R(z,n)}                     & initial capacity of intermittent generators & GW                        \\
-$\widetilde{S}^{out}_{z,k}$  &\makecell[l]{\texttt{INITIAL\_CAP\_S\_OUT(z,k)}}  & initial discharging capacity of storages & GW                        \\
-$\widetilde{S}^{in}_{z,k}$   &\makecell[l]{\texttt{INITIAL\_CAP\_S\_IN(z,k)}}   & initial charging capacity of storages & GW                        \\
-$\widetilde{V}_{z,k}$        &\makecell[l]{\texttt{INITIAL\_CAP\_V(z,k)}}       & initial energy storage capacity &                           \\
-$\widetilde{X}_{z,zz}$       &\texttt{INITIAL\_CAP\_X(z,zz)}                    & initial transmission capacity & GW                        \\
+        $C^{r}_{z,n}$ & \texttt{CAPITALCOST\_R(z,n)} & capital cost of intermittent generators (specific, annuity)
+        & \EUR/ MW
+        \\
+        $C^{g}_{z,i}$ & \texttt{CAPITALCOST\_G(z,i)} & capital cost of thermal generators (specific, annuity)
+        & \EUR/ MW
+        \\
+        $C^{s}_{z,k}$ & \makecell[l]{\texttt{CAPITALCOST\_S(z,k)}} & capital cost of storages - power (specific, annuity)
+        & \EUR/ MW
+        \\
+        $C^{v}_{z,k}$ & \makecell[l]{\texttt{CAPITALCOST\_V(z,k)}} & capital cost of storages - energy (specific, annuity)
+        & \EUR/ MW
+        \\
+        $C^{x}$ & \texttt{CAPITALCOST\_X} & capital cost of transmission capacity (specific, annuity)
+        & \EUR/ MW
+        \\
+        $D_{z,t,m}$ & \texttt{DEMAND(z,t,m)} & energy demand & GW \\
+        $\widehat{D}_{z,m}$ & \texttt{PEAK\_LOAD(z,m)} & peak demand & GW \\
+        $\widetilde{G}_{z,i}$ & \texttt{INITIAL\_CAP\_G(z,tec)} & initial capacity of dispatchable generators
+        & GW
+        \\
+        $O^{g}_{i}$ & \texttt{OM\_COST\_G\_VAR(i)} & variable O\& M cost of dispatchable generators
+        & \EUR/ MWh
+        \\
+        $O^{r}_{z,n}$ & \texttt{OM\_COST\_R\_VAR(z,n)} & variable O\& M cost of intermittent generators
+        & \EUR/ MWh
+        \\
+        $\Bar{O}^{g}_{i}$ & \texttt{OM\_COST\_G\_QFIX(i)} & quasi-fixed O\& M cost of dispatchable generators
+        & \EUR/ MW
+        \\
+        $\Bar{O}^{r}_{z,n}$ & \texttt{OM\_COST\_R\_QFIX(z,n)} & quasi-fixed O\& M cost of intermittent generators
+        & \EUR/ MW
+        \\
+        $P^{e}_{t,z}$ & \texttt{PRICE\_CO2(t,z)} & \ce{CO2} price & \EUR/$\text{t}_{\ce{CO2}}$ \\
+        $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & fuel price & \EUR/ MWh \\
+        $\widetilde{R}_{z,n}$ & \texttt{INITIAL\_CAP\_R(z,n)} & initial capacity of intermittent generators
+        & GW
+        \\
+        $\widetilde{S}^{out}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_S\_OUT(z,k)}} & initial discharging capacity of storages
+        & GW
+        \\
+        $\widetilde{S}^{in}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_S\_IN(z,k)}} & initial charging capacity of storages
+        & GW
+        \\
+        $\widetilde{V}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_V(z,k)}} & initial energy storage capacity
+        &
+        \\
+        $\widetilde{X}_{z,zz}$ & \texttt{INITIAL\_CAP\_X(z,zz)} & initial transmission capacity
+        & GW
+        \\
     \end{longtable}
@@
     \newpage
-\subsection{Variables}\label{variables}
+
+    \subsection{Variables} \label{variables}
     Variables are denoted by lower-case latin letters.
@@
     \begin{longtable}{p{0.15\textwidth}p{0.3\textwidth}p{0.35\textwidth}p{0.15\textwidth}}
-\caption{Variables}\\
+        \caption{Variables} \\
         \toprule
-mathematical symbol & programming symbol & description & unit\\
+        mathematical symbol & programming symbol & description & unit \\
         \midrule
@@
         \endlastfoot
-% the contents of the table
-$b_{z,t,i,f}$                &\texttt{b(z,t,i,f)}           & fuel burn for energy generation & GW        \\
-$c$                          &\texttt{cost\_system}         & total system cost & k\EUR     \\
-$c_{z}$                      &\texttt{cost\_zonal(z)}       & zonal system cost & k\EUR     \\
-$c^{b}_{z,t,i}$              &\texttt{cost\_fuel(z,t,i)}    & fuel cost & k\EUR     \\
-$c^{e}_{z,t,i}$              &\texttt{cost\_co2(z,t,i)}     & emission cost & k\EUR     \\
-$c^{om}_{z,i}$               &\texttt{cost\_om\_g(z,i)}     & total o\&m cost of dispatchable generators & k\EUR     \\
-$c^{om}_{z,n}$               &\texttt{cost\_om\_r(z,n)}     & total o\&m cost of intermittent generators & k\EUR     \\
-$c^{g}_{z}$                  &\texttt{cost\_invest\_g(z)}   & capital cost of generators & k\EUR     \\
-$c^{q}_{z}$                  &\texttt{cost\_nse(z)}         & total cost of non-served load & k\EUR     \\
-$c^{r}_{z}$                  &\texttt{cost\_invest\_r(z)}   & capital cost of intermittent generators & k\EUR     \\
-$c^{s,v}_{z}$                &\texttt{cost\_invest\_sv(z)}  & capital cost of storages & k\EUR     \\
-$c^{x}_{z}$                  &\texttt{cost\_invest\_x(z)}   & capital cost of interconnectors & k\EUR     \\
-$e_{z}$                      &\texttt{emission\_co2(z)}     & \COO emissions & t\COO    \\
-$\widetilde{g}^{+}_{z,i}$    &\texttt{add\_g(z,i)}          & added capacity of dispatchables & GW        \\
-$\widetilde{g}^{-}_{z,i}$    &\texttt{deco\_g(z,i)}         & decommissioned capacity of dispatchables & GW        \\
-$g_{z,t,i,m,f}$              &\texttt{g(z,t,i,m,f)}         & energy generated by conventionals & GW        \\
-$q^{+}_{z,t}$                &\texttt{q\_curtail(z,t)}      & curtailed energy & GW        \\
-$q^{-}_{z,t,m}$              &\texttt{q\_nse(z,t,m)}        & non-served energy & GW        \\
-$\widetilde{r}^{+}_{z,n}$    &\texttt{add\_r(z,n)}          & added capacity of intermittents & GW        \\
-$\widetilde{r}^{-}_{z,n}$    &\texttt{deco\_r(z,n)}         & decommissioned capacity of intermittents & GW        \\
-$r_{z,t,n}$                  &\texttt{r(z,t,n)}             & electricity generated by intermittents & GW        \\
-$\widetilde{s}^{+}_{z,k}$    &\texttt{add\_s(z,k)}          & added storage capacity (power)              & GW        \\
-$s^{in}_{z,t,k}$             &\texttt{s\_in(z,t,k)}         & energy stored in & GW        \\
-$s^{out}_{z,t,k}$            &\texttt{s\_out(z,t,k)}        & energy stored out & GW        \\
-$\widetilde{v}^{+}_{z,k}$    &\texttt{add\_v(z,k)}          & added storage capacity (energy)             & GWh       \\
-$v_{z,t,k}$                  &\texttt{v(z,t,k)}             & storage energy content & GWh       \\
-$w_{z,t,i,l,f}$              &\texttt{w(z,t,i,l,f)}         & operating region weight &           \\
-$\widetilde{x}^{+}_{z,zz}$   &\texttt{add\_x(z,zz)}         & added transmission capacity & GW        \\
-$x_{z,zz,t}$                 &\texttt{x(z,zz,t)}            & electricity net export & GW        \\
+        $b_{z,t,i,f}$ & \texttt{b(z,t,i,f)} & fuel burn for energy generation & \si{\giga\watt} \\
+        $c$ & \texttt{cost\_system} & total system cost & k\EUR \\
+        $c_{z}$ & \texttt{cost\_zonal(z)} & zonal system cost & k\EUR \\
+        $c^{a}_{z,f}$ & \texttt{cost\_air\_pol(z,f)} & total external cost of air pollution & k\EUR \\
+        $c^{b}_{z,t,i}$ & \texttt{cost\_fuel(z,t,i)} & fuel cost & k\EUR \\
+        $c^{e}_{z,t,i}$ & \texttt{cost\_co2(z,t,i)} & emission cost & k\EUR \\
+        $c^{om}_{z,i}$ & \texttt{cost\_om\_g(z,i)} & total o\& m cost of dispatchable generators
+        & k\EUR
+        \\
+        $c^{om}_{z,n}$ & \texttt{cost\_om\_r(z,n)} & total o\& m cost of intermittent generators
+        & k\EUR
+        \\
+        $c^{g}_{z}$ & \texttt{cost\_invest\_g(z)} & capital cost of generators & k\EUR \\
+        $c^{q}_{z}$ & \texttt{cost\_nse(z)} & total cost of non-served load & k\EUR \\
+        $c^{r}_{z}$ & \texttt{cost\_invest\_r(z)} & capital cost of intermittent generators & k\EUR \\
+        $c^{s,v}_{z}$ & \texttt{cost\_invest\_sv(z)} & capital cost of storages & k\EUR \\
+        $c^{x}_{z}$ & \texttt{cost\_invest\_x(z)} & capital cost of interconnectors & k\EUR \\
+        $e_{z}$ & \texttt{emission\_co2(z)} & \ce{CO2} emissions & t\ce{CO2} \\
+        $\widetilde{g}^{+}_{z,i}$ & \texttt{add\_g(z,i)} & added capacity of dispatchables & GW \\
+        $\widetilde{g}^{-}_{z,i}$ & \texttt{deco\_g(z,i)} & decommissioned capacity of dispatchables
+        & GW
+        \\
+        $g_{z,t,i,m,f}$ & \texttt{g(z,t,i,m,f)} & energy generated by conventionals & GW \\
+        $q^{+}_{z,t}$ & \texttt{q\_curtail(z,t)} & curtailed energy & GW \\
+        $q^{-}_{z,t,m}$ & \texttt{q\_nse(z,t,m)} & non-served energy & GW \\
+        $\widetilde{r}^{+}_{z,n}$ & \texttt{add\_r(z,n)} & added capacity of intermittents & GW \\
+        $\widetilde{r}^{-}_{z,n}$ & \texttt{deco\_r(z,n)} & decommissioned capacity of intermittents
+        & GW
+        \\
+        $r_{z,t,n}$ & \texttt{r(z,t,n)} & electricity generated by intermittents & GW \\
+        $\widetilde{s}^{+}_{z,k}$ & \texttt{add\_s(z,k)} & added storage capacity (power) & GW \\
+        $s^{in}_{z,t,k}$ & \texttt{s\_in(z,t,k)} & energy stored in & GW \\
+        $s^{out}_{z,t,k}$ & \texttt{s\_out(z,t,k)} & energy stored out & GW \\
+        $\widetilde{v}^{+}_{z,k}$ & \texttt{add\_v(z,k)} & added storage capacity (energy) & GWh \\
+        $v_{z,t,k}$ & \texttt{v(z,t,k)} & storage energy content & GWh \\
+        $w_{z,t,i,l,f}$ & \texttt{w(z,t,i,l,f)} & operating region weight & \\
+        $\widetilde{x}^{+}_{z,zz}$ & \texttt{add\_x(z,zz)} & added transmission capacity & GW \\
+        $x_{z,zz,t}$ & \texttt{x(z,zz,t)} & electricity net export & GW \\
     \end{longtable}
@@
     \newpage
+
     \subsection{Naming system}
-\begin{table*}
+    \begin{table*}[h!]
         \centering
@@
                 intermittent units &$\widetilde{R}_{z,n}$                &$\widetilde{r}^{+}_{z,n}$        &$\widetilde{r}^{-}_{z,n}$                &$C^{r}_{z,n}$                            &$r_{z,t,n}$                  \\
-storages (power)    &$\widetilde{S}_{z,k}$                &$\widetilde{s}^{+}_{z,k}$        & NA &$C^{s}_{z,k}$                             &$s_{z,t,k}$                  \\% $\widetilde{s}^{-}_{z,k}$
-storages (energy)   &$\widetilde{V}_{z,k}$                &$\widetilde{v}^{+}_{z,k}$        & NA &$C^{v}_{z,k}$                            & NA                            \\% $\widetilde{v}^{-}_{z,t,k}$
-transmission &$\widetilde{X}_{z,zz}$               &$\widetilde{x}^{+}_{z,zz}$       & NA &$C^{x}_{z,zz}$                           &$x_{z,zz,t}$                 \\% $\widetilde{x}^{-}_{z,zz}$
+                storages (power)    &$\widetilde{S}_{z,k}$                &$\widetilde{s}^{+}_{z,k}$        & NA &$C^{s}_{z,k}$                             &$s_{z,t,k}$\\
+                storages (energy)   &$\widetilde{V}_{z,k}$                &$\widetilde{v}^{+}_{z,k}$        & NA &$C^{v}_{z,k}$                            & NA                            \\
+                transmission &$\widetilde{X}_{z,zz}$               &$\widetilde{x}^{+}_{z,zz}$       & NA &$C^{x}_{z,zz}$                           &$x_{z,zz,t}$\\
                 \bottomrule
@@
 
-\subsection{Mathematical description}\label{mathmodel}
+    \subsection{Mathematical description} \label{subsec:apdx_mathmodel}
 
@@
     accrues when demand is not met, e.g. when there is a power outage.
+    We set $c^{q}_{z} = 12500$.
     \begin{align}
@@
     \end{align}
-The components of zonal system costs are calculated as given in equations (\ref{fuel_cost}) to (\ref{lost_load_cost}).
-Lower-case $c$ represent total cost, while upper-case$C$denotes specific, annualized capital cost of technology investment.
-Prices for fuels and \COO are denoted by$P$.
-\begin{align}
-&c^{b}_{z,t,i}& =&\ \sum_{f}\left( P_{t,z,f}\:b_{t,z,i,f}\right)\qquad \qquad&\forall z,t,i\label{fuel_cost}\\
+    The components of zonal system costs are calculated according to \autoref{fuel_cost} to \autoref{lost_load_cost}.
+    Lower-case $c$ represent total cost, while upper-case $C$ denotes the specific, annualized capital cost of technology investment.
+    Prices for fuels and \ce{CO2} are denoted by $P$.
+    \begin{align}
+        &c^{b}_{z,t,i}& =&\ \sum_{f \in \Omega} \left( P_{t,z,f}\:b_{t,z,i,f}\right)\qquad \qquad&\forall z,t,i\label{fuel_cost}\\
         &c^{e}_{z,t,i}& =&\ \sum_{f}\left( P^{e}_{t,z}\:e_{z,t,i}\right)\qquad \qquad&\forall z,t,i\\
-%&c^{om}_{z,i}& =&\Q^{g}_{i}\left( \widetilde{G}_{z,i} - \widetilde{g}^{-}_{z,i} + \widetilde{g}^{+}_{z,i}\right) + \sum_{t}\sum_{m}\sum_{f}\left( O^{g}_{i}\:g_{z,t,i,m,f}\right)\qquad \qquad&\forall z,i\\
-%&c^{om}_{z,n}& =&\Q^{r}_{n}\left( \widetilde{R}_{z,n} - \widetilde{r}^{-}_{z,n} + \widetilde{r}^{+}_{z,n}\right) + \sum_{t}\left( O^{r}_{n}\:r_{z,t,n}\right)\qquad \qquad&\forall z,n\\
-%&c^{g}_{z}& =&\ \sum_{i}\left( C^{g}_{z,i}\: \widetilde{g}^{+}_{z,i}\right)\qquad \qquad&\forall z\\
-%&c^{r}_{z}& =&\ \sum_{n}\left( C^{r}_{z,n}\: \widetilde{r}^{+}_{z,n}\right)\qquad \qquad&\forall z\\
+        &c^{om}_{z,i}& =&\ \left( \widetilde{G}_{z,i} - \widetilde{g}^{-}_{z,i} + \widetilde{g}^{+}_{z,i}\right) \Bar{O}^{g}_{i} + \sum_{t}\sum_{m}\sum_{f \in \Omega}\left( O^{g}_{i}\:g_{z,t,i,m,f}\right)\qquad \qquad&\forall z,i\\
+        &c^{om}_{z,n}& =&\ \left( \widetilde{R}_{z,n} - \widetilde{r}^{-}_{z,n} + \widetilde{r}^{+}_{z,n}\right) \Bar{O}^{r}_{z,n} + \sum_{t}\left( O^{r}_{n}\:r_{z,t,n}\right)\qquad \qquad&\forall z,n\\
+        &c^{g}_{z}& =&\ \sum_{i}\left( C^{g}_{z,i}\: \widetilde{g}^{+}_{z,i}\right)\qquad \qquad&\forall z\\
+        &c^{r}_{z}& =&\ \sum_{n}\left( C^{r}_{z,n}\: \widetilde{r}^{+}_{z,n}\right)\qquad \qquad&\forall z\\
         &c^{s,v}_{z}& =&\ \sum_{k}\left( C^{s}_{z,k}\: \widetilde{s}^{+}_{z,k} + C^{v}_{z,k} \:v^{+}_{z,k}\right)\qquad \qquad&\forall z\\
@@
     In each hour, the markets for electricity and heat have to clear.
-Equation (\ref{market_clearing_el}) ensures that the total supply from conventional and intermittent sources, and
+    \autoref{market_clearing_el} ensures that the total supply from conventional and intermittent sources, and
     storages equals total electricity demand plus net exports, electricity stored and used for heat generation.
-Likewise, equation (\ref{market_clearing_ht}) clears the heat market by equating heat generation to heat demand.
+    Likewise, \autoref{market_clearing_ht} clears the heat market by equating heat generation to heat demand.
     \begin{align}
         \begin{split}
-\sum_{i}\sum_{f} g_{z,t,i,\text{el},f} + \sum_{k} s^{out}_{z,t,k} + \sum_{n} r_{z,t,n} &= \\D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + & \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} - q^{-}_{z,t,\text{el}} + q^{+}_{z,t} \qquad \forall z,t
+            \sum_{i}\sum_{f \in \Omega_{i,f}} g_{z,t,i,\text{el},f} + \sum_{n} r_{z,t,n} + \sum_{k} s^{out}_{z,t,k} &= \\D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + & \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} - q^{-}_{z,t,\text{el}} + q^{+}_{z,t} \qquad \forall z,t
         \end{split}
@@
     \begin{align}
-\sum_{i}\sum_{f} g_{z,t,i,\text{ht},f} = D_{z,t,\text{ht}} - q^{-}_{z,t,\text{ht}}\qquad \forall z,t\label{market_clearing_ht}
+        \sum_{i}\sum_{f \in \Omega_{i,f}} g_{z,t,i,\text{ht},f} = D_{z,t,\text{ht}} - q^{-}_{z,t,\text{ht}}\qquad \forall z,t\label{market_clearing_ht}
     \end{align}
@@
     \begin{align}
-\sum_{f} g_{z,t,i,m,f}\leq \widetilde{G}_{z,i} +\widetilde{g}^{+}_{z,i} -\widetilde{g}^{-}_{z,i}\qquad \qquad \forall z,t,i,m
-\end{align}
-Generator efficiency $\eta$ determines the amount of fuel$b_{z,t,i,f} \geq 0$that needs to be spent in order to
-generate a given amount of energy.
-\begin{align}
-g_{z,t,i,m,f} = \sum_{f}\eta_{i,m,f}\:b_{z,t,i,f}\qquad \qquad \forall z,t,i \notin J, f
+        \sum_{f \in  \Omega} g_{z,t,i,m,f}\leq \widetilde{G}_{z,i} +\widetilde{g}^{+}_{z,i} -\widetilde{g}^{-}_{z,i}\qquad \qquad \forall z,t,i,m
+    \end{align}
+    Generator efficiency $\eta$ determines the amount of fuel $b_{z,t,i,f} \geq 0$ that needs to be spent in order to generate a given amount of energy.
+    \begin{align}
+        g_{z,t,i,m,f} = \sum_{f}\eta_{i,m,f}\:&b_{z,t,i,f}\qquad \qquad \forall z,t,i \notin J, f \\
+        &b_{z,t,i,f \notin \Omega_{i,f}} = 0
     \end{align}
@@
     \begin{align}
-\sum_{l}\sum_{f} w_{z,t,i,l,f} =\widetilde{G}_{z,i} +\widetilde{g}^{+}_{z,i} -\widetilde{g}^{-}_{z,i}\qquad \qquad \forall z,t,i \in J\\
-g_{z,t,i,m,f} =\sum_{l}\sum_{f}\psi_{i,l,m}\:w_{z,t,i,l,f}\qquad \qquad \forall z,t,i \in J, m\\
+        \sum_{l}\sum_{f \in \Omega} w_{z,t,i,l,f} \leq \widetilde{G}_{z,i} +\widetilde{g}^{+}_{z,i} -\widetilde{g}^{-}_{z,i}\qquad \qquad \forall z,t,i \in J\\
+        g_{z,t,i,m,f} =\sum_{l}\psi_{i,l,m}\:w_{z,t,i,l,f}\qquad \qquad \forall z,t,i \in J, m, f\\
         b_{z,t,i,f} =\sum_{l}\chi_{i,l,f}\:w_{z,t,i,l,f}\qquad \qquad \forall z,t,i \in J, f\\
@@
     \end{align}
-Since the model can add storage power capacity and energy capacity independently, we require a storage to hold at
-least as much energy as it could store in (or out) in one hour.
+    Since the model can add storage power capacity and energy capacity independently, we require a storage to hold at least as much energy as it could store in (or out) in one hour.
     \begin{align}
@@
     \end{align}
+    To avoid last-round effects, i.e. a complete draw-down of the storage at year end, we require the storage to be filled to the same level at the beginning ($t=1$) and the end of the year ($t=T$).
+    \begin{align}
+        v_{z,1,k} \geq \Bar{v} \qquad \forall z, k\\
+        v_{z,T,k} \geq \Bar{v} \qquad \forall z, k
+    \end{align}
 
     \paragraph{Emission accounting}
-Burning fossil fuels for energy generation produces emissions of carbon dioxide (\COO). The amount of \COO emitted
+    Burning fossil fuels for energy generation produces emissions of carbon dioxide (\ce{CO2}). The amount of \ce{CO2} emitted
     is tracked by the following equation
@@
         e_{z,t,i} = \sum_{f}\left(\varepsilon_{f}\:b_{z,t,i,f}\right)\qquad \qquad \forall z,t,i
+    \end{align}
+
+    \paragraph{Air pollution}
+    Air pollutants are released during the manufacturing of plants and when fuel is burned.
+The following accounting equation keeps track of these external cost, which do not enter the objective function.
+    \begin{align}
+        \begin{split}
+            c^{a}_{z,f} =&
+            \sum_{i} \Bar{\alpha}_{f} \left(\widetilde{G}_{z,i}
+            +\widetilde{g}^{+}_{z,i} -\widetilde{g}^{-}_{z,i}\right) \Omega_{i,f}
+            + \sum_{n} \Bar{\alpha}_{f} \left( \widetilde{R}_{z,n} -\widetilde{r}^{-}_{z,n} +\widetilde{r}^{+}_{z,n}\right) \Omega_{n,f}
+            \\
+            + &\sum_{i} \alpha_{f} b_{z,t,i,f}
+            + \sum_{n} \alpha_{f} r_{z,t,n} \Omega_{n,f}
+        \end{split}
     \end{align}
@@
     First, exchange between market zones is constrained by available transfer capacities. Transfer capacities can be
-expanded at constant, specific investment cost (see equation (\ref{transmission_expansion_cost})).
+    expanded at constant, specific investment cost (see \autoref{transmission_expansion_cost}).
     This rules out economies of scale in transmission investment that might arise in interconnected, meshed grids.
@@
     We assume that ancillary services are supplied by conventional (thermal) power plant, hydro power plant, and storage.
-The requirement for spinning reserves is proportional to electricity peak load
+    The requirement for spinning reserves is assumed to be proportional to electricity peak load
     $\widehat{D}_{z,\text{el}} = \max_{t} D_{z,t,\text{el}}$ and peak generation from wind and solar resources,
@@
     \begin{align}
-\sum_{i}\sum_{f}\left( g_{z,t,i,\text{el},f}\right) + r_{z,t,\text{ror}}
+        \sum_{i}\sum_{f \in \Omega}\left( g_{z,t,i,\text{el},f}\right) + r_{z,t,\text{ror}}
         + \sum_{k}\left( s^{out}_{z,t,k} + s^{in}_{z,t,k}\right)\geq \lambda_{z}\widehat{D}_{z,\text{el}}
@@
     \begin{align}
-q^{+}_{z,t} \leq \sum_{n} r_{z,t,n} \qquad \qquad \forall z, t
-\end{align}
-
-%\section{Effects of doubling transmission capacity} \label{sec:transmission-sensitivity}
+        q^{+}_{z,t} \leq \sum_{n\setminus \{ \text{ror}\}} r_{z,t,n} \qquad \qquad \forall z, t
+    \end{align}
 \end{document}

### 2021-01-08 projects/asparagus/doc/figures_asparagus.py
remove blanks from file names

@@ -396,7 +396,7 @@ for v in var:
     # dfp = senspv[senspv.columns.intersection(mix)]
     dfp = senspv.loc[:, idx[v, :]]
     dfp.columns = dfp.columns.droplevel(0)
-    plot_lines(dfp, FIGPATH / f'S_PVCost_P_{v}.pdf',
+    plot_lines(dfp, FIGPATH / f'S_PVCost_P_{v}.pdf'.replace(' ', ''),
                xlabel='Capital Cost of PV [€/kWp]', xlim=[650, 250], ylim=[0, 16],
                ylabel='Added Capacity of Wind Turbines [GW]', color=REFUEL_COLORS)
 

### 2021-01-08 projects/asparagus/doc/asparagus_on_a_slide.tex
initial commit of slides for institute seminar on Jan 12, 2021

\documentclass[aspectratio=1610, xcolor=dvipsnames]{beamer}
\usetheme{Pittsburgh}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{siunitx}
\usepackage[version=4]{mhchem}

\usepackage[T1]{fontenc}

\usepackage{animate}
\usepackage{biblatex}

\usepackage{booktabs}
\usepackage{cmbright}
\usepackage[gen]{eurosym}
\usepackage{fancyvrb}
\usepackage{multicol}
\usepackage{setspace}
\usepackage[absolute,overlay]{textpos}

\usepackage{xcolor}
\usepackage{graphicx}

\newcommand{\COO}{\ensuremath{\mathrm{CO_2}} }
%\usepackage{natbib}
\usepackage{cellspace}
\setlength\cellspacetoplimit{4pt}
\setlength\cellspacebottomlimit{4pt}

% plotting with tikz and pgfplots
\usepackage{pgfplots}
\usepackage{tikz}
\usepackage{hyperref}
\usetikzlibrary{shapes.geometric, arrows, positioning, fit, shapes.arrows}
\usetikzlibrary{decorations.pathreplacing, decorations.markings}
\pgfplotsset{compat = 1.16}
% make tikz compatible with \againframe
\tikzset{onslide/.code args={<#1>#2}{%
    \only<#1>{\pgfkeysalso{#2}} % \pgfkeysalso doesn't change the path
}}
\tikzset{alt/.code args={<#1>#2#3}{%
    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}} % \pgfkeysalso doesn't change the path
}}
\tikzset{temporal/.code args={<#1>#2#3#4}{%
    \temporal<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}{\pgfkeysalso{#4}} % \pgfkeysalso doesn't change the path
}}

% horizontal alignment of \underbraces
\newcommand{\tempstrut}{\rule[-1.2\baselineskip]{0pt}{1.5\baselineskip}}

%gets rid of bottom navigation bars
\setbeamertemplate{footline}[frame number]{}

%gets rid of bottom navigation symbols
\setbeamertemplate{navigation symbols}{}

% colors
\definecolor{bokugreen}{rgb}{0.302, 0.675, 0.149}
\setbeamercolor{title}{fg=bokugreen, bg=white}
\setbeamercolor{frametitle}{fg=bokugreen}
\setbeamercolor{itemize item}{fg=bokugreen}
\setbeamercolor{enumerate item}{fg=bokugreen}
\setbeamercolor{local structure}{fg=bokugreen}

% wider slide
\newcommand\Wider[2][3em]{%
    \makebox[\linewidth][c]{%
        \begin{minipage}{\dimexpr\textwidth+#1\relax}
            \raggedright#2
        \end{minipage}%
    }%
}

% title page
\setbeamertemplate{title page}[default][colsep=-4bp,rounded=true]
\title[The Cost of Undisturbed Landscapes]{The Cost of Undisturbed Landscapes}
\author{Sebastian Wehrle}
\institute{Institute for Sustainable Economic Development \\University of Natural Resources and Life Sciences, Vienna}
\date{January 12, 2021}
% section page
\setbeamertemplate{section page}{
    \begin{textblock*}{4cm}
        (11.5cm, 4cm) % {block width} (coords)
            {\LARGE \insertsection}
    \end{textblock*}
}
% subsection page
\setbeamertemplate{subsection page}{
    \begin{textblock*}{4cm}
        (11.5cm, 4cm) % {block width} (coords)
            {\LARGE \insertsection} \\ \bigskip \insertsubsection
    \end{textblock*}
}

\begin{document}
{
    \begin{frame}[noframenumbering,plain]
        \maketitle
        \centering
        \includegraphics[height=1.7cm]{./figures/boku-logo}\\
    \end{frame}
}

    \begin{frame}
        \frametitle{Preface}
        \framesubtitle{Motivation for research}
        \structure{Wind power potentials -- state of the art in a nutshell} \\~\\
        %\medskip
        %\structure{How much wind power \emph{can} be installed?} \\~\\
        \begin{itemize}
            \item Wind power potentials typically derived by step-wise exclusion of areas
            \begin{itemize}
                \item technical restrictions (e.g.\ elevation)
                \item legal restrictions (e.g.\ protected areas)
                \item resource restrictions (i.e.\ sufficient wind at location)
            \end{itemize} %\\~\\
            \item Assumption that wind power is feasible in all areas not excluded
        \end{itemize}
        \medskip
        Yet, increasing understanding that "social acceptance" matters
        \medskip
        \begin{itemize}
            \item Exclude areas where erection of wind turbines is "socially unacceptable"
            \item In the literature, such areas are identified, for example, based on
            \begin{itemize}
                \item participatory processes
                \item approximation of public authorities' licensing decisions through "scenicness"
            \end{itemize} %\\~\\
            %\item Binary attributions
            %\item No guidance on choice of method
            %\item How is selection of areas justified?
        \end{itemize}
    \end{frame}

    \begin{frame}
        \frametitle{Preface}
        \framesubtitle{Motivation for research}
        \structure{Wind power potentials -- my take on advancing the state-of-the-art} \\~\\
        \begin{itemize}
            \item Abandon binary attributions which fall short of complexity involved
            \item Realize that there is no guaranteed way to come up with unanimous ("socially accepted") decisions in a society
            \begin{itemize}
                \item How much wind power \emph{can} be deployed ultimately is a matter of power
            \end{itemize}
        \end{itemize}
        \medskip
        Instead, consider how much wind power \emph{should} be installed
        \begin{itemize}
            \item Minimize damage caused
            \item Weigh cost and benefit we are willing to take (as a society)
        \end{itemize}
    \end{frame}

    \begin{frame}
        \frametitle{How much wind power \emph{should} be installed}
        \begin{columns}[T]
            \begin{column}{7cm}
                \begin{block}{The cost of undisturbed landscapes}
                    \begin{itemize}
                        \item change in electricity generation cost
                        \item change in GHG emissions
                        \item change in local air pollution
                        \item change in grid connection cost
                        \item change in capital construction cost
                        \item change in O\&M cost
                        \item change in cost of supply security
                    \end{itemize}
                \end{block}
            \end{column}

            \begin{column}{7cm}
                \begin{block}{The benefit of undisturbed landscapes}
                    \begin{itemize}
                        \item avoided impact on ecosystem services
                        \item avoided impact on amenity value of landscape
                        \item avoided impacts on wildlife
                        \item avoided loss of property value
                    \end{itemize}
                \end{block}
            \end{column}
        \end{columns}
    \end{frame}

    \begin{frame}
        \frametitle{The cost of undisturbed landscapes}
        \framesubtitle{Some background}
        \begin{itemize}
            \item By 2030, \SI{100}{\percent}\footnote[frame]{System services and industry own consumption,
            accounting for $\sim$\SI{10}{\percent} of total electricity demand and $\sim$\SI{40}{\percent} of \ce{CO2} emissions from power
            generation are exempt} of Austria's electricity demand to be sourced from renewables
            \item Sufficient potentials for large-scale expansion only for wind and solar power
        \end{itemize}
        \medskip
        {%
            \setbeamercolor{block body}{bg=bokugreen!30, fg=black}
            \setbeamertemplate{blocks}[rounded][shadow=false]
            \begin{block}{}
                \begin{itemize}
                    \item Not disturbing landscapes by wind turbines means substituting wind turbines with PV modules (potentially open-space)
                \end{itemize}
            \end{block}
        }%
        \medskip
        \begin{itemize}
            \item Substituting wind turbines with PV modules changes system operation and investment
            \item Which cost arises due to this substitution?
            \item In our case, this is the \emph{opportunity cost of wind power}, as the cost-minimizing system relies
            (almost) entirely on wind power
        \end{itemize}
    \end{frame}

    \begin{frame}
        \frametitle{The cost of undisturbed landscapes}
        \framesubtitle{Methodology}
        %, which includes all operation and
        %investment cost of the (domestic) electricity system, the cost of emissions (we also include air pollutants), plus
        %the cost (revenue) from electricity imports (exports).

        \medskip
        {%
            \setbeamercolor{block title}{bg=bokugreen, fg=white}
            \setbeamercolor{block body}{bg=bokugreen!20, fg=black}
            \setbeamertemplate{blocks}[rounded][shadow=false]
            \begin{block}{Approximation of the opportunity cost of wind turbines $OC_w$}
                %To approximate the opportunity cost of wind turbines $OC_w$, we proceed as follows:
                \begin{enumerate}
                    \item determine unconstrained level of RET that leads to minimal system cost $c_{net}$ \label{step1}
                    \item restrict the admissible expansion of the least-cost RET by a small margin, $\Delta w$, and observe
                    resulting change in system cost $\Delta c_{net}$ \label{step2}
                    \item approximate the opportunity cost of wind power by \label{step3}
                    \begin{align*}
                        OC_w = \frac{\Delta c_{net}}{\Delta w}
                    \end{align*}
                    \item repeat steps \ref{step2} and \ref{step3} till no more wind turbines can be erected
                \end{enumerate}
            \end{block}
        }%
    \end{frame}

    \begin{frame}
        \frametitle{The cost of undisturbed landscapes}
        \framesubtitle{Assumptions}
        We quantify the opportunity cost of wind power with the numerical power system model \emph{medea}, which we set up
        to resemble the Austrian and German electricity and district heating systems in 2030.
        \begin{columns}[T]
        {%
            \setbeamercolor{block title}{bg=bokugreen, fg=white}
            \setbeamercolor{block body}{bg=bokugreen!20, fg=black}
            \setbeamertemplate{blocks}[rounded][shadow=false]
            \begin{column}{7cm}
                \begin{block}{Model assumptions}
                    \begin{itemize}
                        \item perfect foresight, partial-equilibrium model of electricity and district heating sectors in
                        AT and DE
                        \item technology-rich representation of supply to meet price-inelastic demand
                        \item economic dispatch in hourly resolution plus investment
                    \end{itemize}
                \end{block}
            \end{column}

            \begin{column}{7cm}
                \begin{block}{Scenario assumptions (2030)}
                    \begin{itemize}
                        \item DE: nuclear and (partial) coal exit
                        \item DE: RET expansion as in EEG 2017
                        \item AT: must generate 79.3 TWh/a from RES
                        \begin{itemize}
                            \item + 5 TWh/a hydropower
                            \item + 1 TWh/a electricity from biomass
                        \end{itemize}
                        \item weather as in 2016 (a low wind year)
                    \end{itemize}
                \end{block}
            \end{column}
        }%
        \end{columns}
        \bigskip
        \url{https://github.com/inwe-boku/medea}
    \end{frame}

    \begin{frame}
        \frametitle{Change in annual net system cost -- Austria, 2030}
        \begin{figure}
            \includegraphics[width=0.95\textwidth]{./figures/cost_components_base_25.pdf} \label{fig:cost_breakdown}
        \end{figure}
    \end{frame}

    \begin{frame}
        \frametitle{Monthly electricity generation by source -- Austria, 2030}
        \framesubtitle{(a) maximum wind power, (b) maximum solar power}
        \begin{figure}
            \includegraphics[width=0.85\textwidth]{./figures/dispatch_AT.pdf} \label{fig:monthly_gen}
        \end{figure}
    \end{frame}

    \begin{frame}
        \frametitle{Annual opportunity cost of wind power -- Austria, 2030}
        \framesubtitle{PV capital cost: \euro 625/kWp}
        \begin{figure}
            \includegraphics[width=0.95\textwidth]{./figures/oc_base.pdf} \label{fig:opportunity_cost_base}
        \end{figure}
    \end{frame}

    \begin{frame}
        \frametitle{Sensitivity of annual opportunity cost of wind power -- Austria, 2030}
        \begin{figure}
            \includegraphics[width=0.95\textwidth]{./figures/Sensitivities.pdf} \label{fig:sensitivities}
        \end{figure}
    \end{frame}

\end{document}

### 2021-01-11 projects/asparagus/doc/asparagus_on_a_slide.tex
updates asparagus_on_a_slide.tex

@@
 \usepackage[version=4]{mhchem}
+\usepackage{qrcode}
 
@@
         {%
-            \setbeamercolor{block body}{bg=bokugreen!30, fg=black}
+            \setbeamercolor{block body}{bg=structure.fg!20, fg=black}
             \setbeamertemplate{blocks}[rounded][shadow=false]
@@
         {%
-            \setbeamercolor{block title}{bg=bokugreen, fg=white}
-            \setbeamercolor{block body}{bg=bokugreen!20, fg=black}
+            \setbeamercolor{block title}{bg=structure.fg!0, fg=structure.fg}
+            \setbeamercolor{block body}{bg=structure.fg!0, fg=black}%{bg=bokugreen!20, fg=black}
             \setbeamertemplate{blocks}[rounded][shadow=false]
@@
                 %To approximate the opportunity cost of wind turbines $OC_w$, we proceed as follows:
+                \medskip
                 \begin{enumerate}
@@
                     \end{align*}
-                    \item repeat steps \ref{step2} and \ref{step3} till no more wind turbines can be erected
+                    \item repeat steps~\ref{step2} and~\ref{step3} till no more wind turbines can be erected
                 \end{enumerate}
@@
         \framesubtitle{Assumptions}
+        {%
+            \setbeamercolor{block body}{bg=structure.fg!20, fg=black}
+            \setbeamertemplate{blocks}[rounded][shadow=false]
+            \begin{block}{}
                 We quantify the opportunity cost of wind power with the numerical power system model \emph{medea}, which we set up
-        to resemble the Austrian and German electricity and district heating systems in 2030.
+                to resemble the Austrian and German electricity and district heating systems in 2030
+            \end{block}
+        }%
         \begin{columns}[T]
-        {%
-            \setbeamercolor{block title}{bg=bokugreen, fg=white}
-            \setbeamercolor{block body}{bg=bokugreen!20, fg=black}
-            \setbeamertemplate{blocks}[rounded][shadow=false]
+            %{%
+            %\setbeamercolor{block title}{bg=bokugreen, fg=white}
+            %\setbeamercolor{block body}{bg=bokugreen!20, fg=black}
+            %\setbeamertemplate{blocks}[rounded][shadow=false]
             \begin{column}{7cm}
-                \begin{block}{Model assumptions}
+                %\begin{block}{Model assumptions}
+                \structure{Model assumptions}
                 \begin{itemize}
@@
                 \end{itemize}
-                \end{block}
+                %\end{block}
             \end{column}
@@
             \begin{column}{7cm}
-                \begin{block}{Scenario assumptions (2030)}
+                %\begin{block}{Scenario assumptions (2030)}
+                \structure{Scenario assumptions (2030)}
                 \begin{itemize}
@@
                     \item DE: RET expansion as in EEG 2017
-                        \item AT: must generate 79.3 TWh/a from RES
+                    \item AT: at least 74.8 TWh/a from RES
                     \begin{itemize}
@@
                 \end{itemize}
-                \end{block}
+                %\end{block}
             \end{column}
-        }%
+            %}%
         \end{columns}
         \bigskip
+        \begin{center}
             \url{https://github.com/inwe-boku/medea}
+        \end{center}
     \end{frame}
@@
     \begin{frame}
-        \frametitle{Change in annual net system cost -- Austria, 2030}
+        \frametitle{Results}
+        \begin{columns}[T]
+            \begin{column}{10.6 cm}
                 \begin{figure}
-            \includegraphics[width=0.95\textwidth]{./figures/cost_components_base_25.pdf} \label{fig:cost_breakdown}
+                    \caption{Change in annual net system cost -- Austria, 2030}
+                    \includegraphics[width=1.0\textwidth]{./figures/cost_components_base_25.pdf} \label{fig:cost_breakdown}
                 \end{figure}
+            \end{column}
+            \begin{column}{5.0 cm}
+                \bigskip
+                \begin{itemize}
+                    \item "100\%"-policy necessarily turns Austria from a net importer (2016: 7.2 TWh) into a
+                    net exporter of electricity as fossil generation needed (heat, system services)
+                    \item each avoided GW wind power replaced by 1.98 GWp PV
+                    \item resource adequacy drives O\&M cost
+                    \item lower exports also mean less fossil generation
+                \end{itemize}
+            \end{column}
+        \end{columns}
     \end{frame}
@@
     \begin{frame}
-        \frametitle{Monthly electricity generation by source -- Austria, 2030}
-        \framesubtitle{(a) maximum wind power, (b) maximum solar power}
+        \frametitle{Results}
+        \begin{columns}[T]
+            \begin{column}{10.6 cm}
                 \begin{figure}
-            \includegraphics[width=0.85\textwidth]{./figures/dispatch_AT.pdf} \label{fig:monthly_gen}
+                    \caption{\parbox[t]{8.6cm}{Monthly electricity generation by source -- Austria, 2030 (a) maximum wind power, (b) maximum solar power}}
+                    \includegraphics[width=0.875\textwidth]{./figures/dispatch_AT.pdf} \label{fig:monthly_gen}
                 \end{figure}
+            \end{column}
+            \begin{column}{5.0 cm}
+                \bigskip
+                \begin{itemize}
+                    \item Wind power is favourably complementary to run-of-river generation
+                    \item Solar PV positively correlated with run-of-river generation
+                    \item In PV-based system more imports needed in winter, while export price in summer lower
+                \end{itemize}
+            \end{column}
+        \end{columns}
     \end{frame}
@@
     \begin{frame}
-        \frametitle{Annual opportunity cost of wind power -- Austria, 2030}
-        \framesubtitle{PV capital cost: \euro 625/kWp}
+        \frametitle{Results}
+        %\framesubtitle{PV capital cost: \euro 625/kWp}
+        \begin{columns}[T]
+            \begin{column}{10.6 cm}
                 \begin{figure}
-            \includegraphics[width=0.95\textwidth]{./figures/oc_base.pdf} \label{fig:opportunity_cost_base}
+                    \caption{\parbox[t]{8.4cm}{Annual opportunity cost of wind power -- Austria, 2030 PV capital cost: \euro 625/kWp}}
+                    \includegraphics[width=0.975\textwidth]{./figures/oc_base.pdf} \label{fig:opportunity_cost_base}
                 \end{figure}
+            \end{column}
+            \begin{column}{5.0 cm}
+                \bigskip
+                \begin{itemize}
+                    \item Figure shows on marginal system cost of not erecting wind turbines
+                    \item Policy target implies marginal opportunity cost of 22\,500 \euro/MW per year
+                    \item Equals \euro 1.3 mn over lifetime of a single standard wind turbine
+                \end{itemize}
+            \end{column}
+        \end{columns}
     \end{frame}
@@
     \begin{frame}
-        \frametitle{Sensitivity of annual opportunity cost of wind power -- Austria, 2030}
+        \frametitle{Results}
+
+        \begin{columns}[T]
+            \begin{column}{10.6 cm}
                 \begin{figure}
-            \includegraphics[width=0.95\textwidth]{./figures/Sensitivities.pdf} \label{fig:sensitivities}
+                    \caption{\parbox[t]{8.0cm}{Sensitivity of annual opportunity cost of wind power PV capital cost: \euro 625/kWp}}
+                    \includegraphics[width=0.975\textwidth]{./figures/Sensitivities.pdf} \label{fig:sensitivities}
                 \end{figure}
+            \end{column}
+            \begin{column}{5.4 cm}
+                \bigskip
+                \begin{itemize}
+                    \item \emph{Low capital cost of PV} assumes overnight cost of PV of \euro 560/kWp
+                    \item \emph{No bottleneck} assumes 10 GW transmission capacity
+                \end{itemize}
+            \end{column}
+        \end{columns}
     \end{frame}
 
+%    \begin{frame}
+%        \frametitle{Summary}
+%        \begin{itemize}
+%            \item So-called 100\% goal necessarily turns Austria into a net exporter of electricity
+%            \item Choice of RET impacts installed capacities, system operation, and, most importantly, trade flows
+%            \item Austria's wind resources are favourable for exporting to Germany
+%            \item Export revenues accrue to electricity producers, not consumers
+%            \item Consumers benefit from low electricity prices
+%            \item Compensation for neighbors of wind parks makes sense
+%        \end{itemize}
+%    \end{frame}
+
+    \begin{frame}
+        \frametitle{Discussion}
+        \begin{itemize}
+            \item Renewable resource quality held constant as capacity is expanded
+            \item sub-national electricity grids neglected
+            \item external cost of open-space PV assumed to be zero
+        \end{itemize}
+    \end{frame}
+
+    \begin{frame}
+        \frametitle{Conclusions}
+        \begin{itemize}
+            \item If cost of $CO_2$ low and PV is realized as open-space, utility-scale installations (with zero external cost),
+            wind power does not come at a system cost advantage
+            \item With high $CO_2$ cost, a preference for rooftop PV, or in the presence of negative external effects of
+            open-space, utility-scale PV, wind power can reduce total system cost
+            \item Wind power induces a current account surplus, which accrues to exporters. Compensation of affected
+            residents could improve distributional fairness and, potentially, increase acceptance of wind power
+            \item In contrast to (relative) market value, opportunity cost of renewables consistent with fully renewable energy systems
+            \item Trade effects lead to much higher system cost impacts of capacity decisions than currently acknowledged in literature
+        \end{itemize}
+    \end{frame}
+
+    \begin{frame}
+        \frametitle{Thank you!}
+        %\vspace{1 cm}
+        \begin{center}
+            \structure{%
+                \href{https://arxiv.org/abs/2006.08009}{The Cost of Undisturbed Landscapes}
+            }\\%
+            \vspace{0.6cm}
+            \qrcode[height=2cm]{https://arxiv.org/abs/2006.08009}\\
+            \vspace{0.6 cm}
+            \href{https://arxiv.org/abs/2006.08009}{https://arxiv.org/abs/2006.08009}
+        \end{center}
+    \end{frame}
 \end{document}

### 2021-01-11 projects/asparagus/doc/figures_asparagus.py

@@ -12,6 +12,7 @@ import config as cfg
 from src.tools.data_processing import medea_path
 from src.tools.visualize_data import plot_lines
 
+# TODO: create /asparagus/doc/figures-subfolder before saving figures
 # TODO: Plots of first differrences (e.g. oc of wind) need to have their x-axis adjusted (shift to mean of adjacent categories)
 # TODO: plot sensitivity as scatters of baseline versus sensivitiy scenario plot
 

### 2021-02-09 src/utils/export2gdx.py
fixed bug in export2gdx.py that resulted in duplication of an empty line in data_static.xlsx

@@ -5,8 +5,8 @@ import pandas as pd
 from gams import *
 
 import config as cfg
-from src.tools.gams_io import df2gdx
-from src.tools.preprocess_data import dict_sets, plant_data, ts_data, estimates, invest_limits
+from src.utils.gams_io import df2gdx
+from src.utils.preprocess_data import dict_sets, plant_data, ts_data, estimates, invest_limits
 
 idx = pd.IndexSlice
 # --------------------------------------------------------------------------- #

### 2021-02-09 src/utils/preprocess_data.py

@@ -2,7 +2,7 @@
 import pandas as pd
 
 import config as cfg
-from src.tools.data_processing import hours_in_year, medea_path
+from src.utils.data_processing import hours_in_year, medea_path
 
 # --------------------------------------------------------------------------- #
 # %% settings and initializing
@@ -13,7 +13,7 @@ idx = pd.IndexSlice
 # --------------------------------------------------------------------------- #
 # %% Read Data
 plant_data = {
-    'technology': pd.read_excel(STATIC_FNAME, 'Technologies', header=[2], index_col=[2]),
+    'technology': pd.read_excel(STATIC_FNAME, 'Technologies', header=[2], index_col=[2]).dropna(axis=0, how='all'),
     'chp': pd.read_excel(STATIC_FNAME, 'FEASIBLE_INPUT-OUTPUT', header=[0], index_col=[0, 1, 2]),
     'installed': pd.read_excel(STATIC_FNAME, 'Capacities', header=[0], index_col=[0, 1, 2], skiprows=[0, 1, 2]),
     'CAP_X': pd.read_excel(STATIC_FNAME, 'ATC', index_col=[0]),

### 2021-02-09 projects/asparagus/doc/asparagus_on_a_slide.tex
renamed 'tools' to 'utils'

@@
             \begin{itemize}
-                \item How much wind power \emph{can} be deployed ultimately is a matter of power
+                \item Determining potentials means asking how much wind power \emph{can} be deployed
+                \item Yet, this ultimately is a matter of power
             \end{itemize}
@@
         \frametitle{The cost of undisturbed landscapes}
-        \framesubtitle{Some background}
+        \framesubtitle{Introduction}
         \begin{itemize}
@@
         \frametitle{The cost of undisturbed landscapes}
+        \framesubtitle{Policy consequences}
+        Immediate corollary of the "100\%"-policy:
+        \begin{itemize}
+            \item  Austria turns from a net importer of electricity (2016: 7.2 TWh, 9.9\% of total electricity
+            consumption) into a net exporter
+            \begin{itemize}
+                \item Heat generation requires running (fossil) co-generation units
+                \item (Fossil) generation also required to secure supply at all times
+            \end{itemize}
+            \item This is not the case under a (equivalent) carbon pricing regime
+        \end{itemize}
+    \end{frame}
+
+    \begin{frame}
+        \frametitle{The cost of undisturbed landscapes}
         \framesubtitle{Methodology}
@@
                     resulting change in system cost $\Delta c_{net}$ \label{step2}
-                    \item approximate the opportunity cost of wind power by \label{step3}
+                    \item approximate the opportunity cost of wind power $OC_w$ by \label{step3}
                     \begin{align*}
@@
                 \begin{itemize}
-                    \item "100\%"-policy necessarily turns Austria from a net importer (2016: 7.2 TWh) into a
-                    net exporter of electricity as fossil generation needed (heat, system services)
+                    %\item "100\%"-policy necessarily turns Austria from a net importer (2016: 7.2 TWh) into a
+                    %net exporter of electricity as fossil generation needed (heat, system services)
                     \item each avoided GW wind power replaced by 1.98 GWp PV
+                    \item mixture of RET slightly reduce variance and increases minimal renewable generation by a small margin
                     \item resource adequacy drives O\&M cost
@@
                     \item In PV-based system more imports needed in winter, while export price in summer lower
+                    \item PV pushes $CO_2$ emissions from Austria to Germany
                 \end{itemize}
@@
                 \begin{itemize}
-                    \item Figure shows on marginal system cost of not erecting wind turbines
-                    \item Policy target implies marginal opportunity cost of 22\,500 \euro/MW per year
-                    \item Equals \euro 1.3 mn over lifetime of a single standard wind turbine
-                \end{itemize}
-            \end{column}
-        \end{columns}
-    \end{frame}
-
-    \begin{frame}
-        \frametitle{Results}
-
-        \begin{columns}[T]
-            \begin{column}{10.6 cm}
-                \begin{figure}
-                    \caption{\parbox[t]{8.0cm}{Sensitivity of annual opportunity cost of wind power PV capital cost: \euro 625/kWp}}
-                    \includegraphics[width=0.975\textwidth]{./figures/Sensitivities.pdf} \label{fig:sensitivities}
-                \end{figure}
-            \end{column}
-            \begin{column}{5.4 cm}
-                \bigskip
-                \begin{itemize}
-                    \item \emph{Low capital cost of PV} assumes overnight cost of PV of \euro 560/kWp
-                    \item \emph{No bottleneck} assumes 10 GW transmission capacity
+                    \item Figure shows marginal system cost of not erecting wind turbines
+                    \item Policy target implies marginal opportunity cost of \euro 22\,500 per MW and year
+                    \item Equals \euro 1.3 mn over lifetime of a single 3.5 MW wind turbine (discounted at 5\%)
                 \end{itemize}
@@
             \item Renewable resource quality held constant as capacity is expanded
-            \item sub-national electricity grids neglected
-            \item external cost of open-space PV assumed to be zero
+            \begin{itemize}
+                \item Hölting et a. (2016) estimate 45 TWh of wind resources available in Austria at equal or lower cost
+                than we assumed
+                \item We consider additions of at most 32 TWh (at a carbon price of \euro 100 per t)
+            \end{itemize}
+            \item Sub-national electricity grids neglected
+            \begin{itemize}
+                \item grid cost of open-space PV and wind power should be similar
+                \item cost of rooftop PV in distribution grid underresearched
+                \item some estimates imply distribution cost not substantially different from wind power in transmission grids
+                \item however, huge variability of estimates
+            \end{itemize}
+            \item External cost of open-space PV assumed to be zero
+            \begin{itemize}
+                \item external cost of open-space PV underresearched
+                \item mixed findings from working papers ranging from "no impact" to "small impact with limited significance"
+            \end{itemize}
         \end{itemize}
@@
         \begin{itemize}
-            \item If cost of $CO_2$ low and PV is realized as open-space, utility-scale installations (with zero external cost),
+            \item If cost of $CO_2$ low and PV realized as open-space, utility-scale installations,
             wind power does not come at a system cost advantage
-            \item With high $CO_2$ cost, a preference for rooftop PV, or in the presence of negative external effects of
-            open-space, utility-scale PV, wind power can reduce total system cost
-            \item Wind power induces a current account surplus, which accrues to exporters. Compensation of affected
-            residents could improve distributional fairness and, potentially, increase acceptance of wind power
+            \item With high $CO_2$ cost, a preference for rooftop PV, or external costs of
+            open-space, utility-scale PV, wind power can reduce total system cost (including external cost)
+            \item Wind power increases electricity's current account surplus, which accrues to exporters.
+            Compensation of affected residents could improve distributional fairness and, potentially, increase
+            acceptance of wind power \\
+            \medskip
+            \item Trade effects lead to much higher system cost impacts of capacity decisions than currently acknowledged in literature
             \item In contrast to (relative) market value, opportunity cost of renewables consistent with fully renewable energy systems
-            \item Trade effects lead to much higher system cost impacts of capacity decisions than currently acknowledged in literature
         \end{itemize}
@@
     \end{frame}
+
+    \begin{frame}
+        \frametitle{Results}
+
+        \begin{columns}[T]
+            \begin{column}{10.6 cm}
+                \begin{figure}
+                    \caption{\parbox[t]{8.0cm}{Sensitivity of annual opportunity cost of wind power (Cubic fit across $CO_2$ price scenarios)}}
+                    \includegraphics[width=0.975\textwidth]{./figures/Sensitivities.pdf} \label{fig:sensitivities}
+                \end{figure}
+            \end{column}
+            \begin{column}{5.4 cm}
+                \bigskip
+                \begin{itemize}
+                    \item \emph{Low capital cost of PV} assumes overnight cost of PV of \euro 560/kWp
+                    \item \emph{No bottleneck} assumes 10 GW transmission capacity
+                \end{itemize}
+            \end{column}
+        \end{columns}
+    \end{frame}
+
+    \begin{frame}
+        \frametitle{Results}
+        \begin{columns}[T]
+            \begin{column}{10.6 cm}
+                \begin{figure}
+                    \caption{\parbox[t]{8.0cm}{Sensitivity of optimal wind power expansion}}
+                    \includegraphics[width=0.975\textwidth]{./figures/S_PVCost_P_WindOnshore.pdf} \label{fig:sens_wind_cap}
+                \end{figure}
+            \end{column}
+            \begin{column}{5.0 cm}
+                \bigskip
+                \begin{itemize}
+                    \item If RET expansion incentivised by carbon price (not "100\%"-policy), optimal wind power expansion is
+                    robust against capital cost of PV
+                \end{itemize}
+            \end{column}
+        \end{columns}
+    \end{frame}
+
 \end{document}

### 2021-02-09 logging_config.py

@@ -7,8 +7,8 @@ LOG_FILE = os.path.join(cfg.MEDEA_ROOT_DIR, 'data', 'logfile.log')
 
 
 def setup_logging(fname=LOG_FILE):
-    NO_COLOR = '\33[m'
-    RED, GREEN, ORANGE, BLUE, PURPLE, LBLUE, GREY = (
+    no_color = '\33[m'
+    red, green, orange, blue, purple, lblue, grey = (
         map('\33[%dm'.__mod__, range(31, 38)))
 
     logging_config = {
@@ -21,9 +21,7 @@ def setup_logging(fname=LOG_FILE):
             },
             'color': {
                 'format': '{}[%(asctime)s]{} {}%(levelname)-5s{} - '
-                          '{}%(name)-5s{}: %(message)s'.format(
-                    GREEN, NO_COLOR, PURPLE, NO_COLOR,
-                    ORANGE, NO_COLOR)
+                          '{}%(name)-5s{}: %(message)s'.format(green, no_color, purple, no_color, orange, no_color)
             }
         },
         'handlers': {

### 2021-02-09 projects/asparagus/doc/figures_asparagus.py

@@ -9,8 +9,8 @@ import pandas as pd
 import statsmodels.api as sm
 
 import config as cfg
-from src.tools.data_processing import medea_path
-from src.tools.visualize_data import plot_lines
+from src.utils.data_processing import medea_path
+from src.utils.visualize_data import plot_lines
 
 # TODO: create /asparagus/doc/figures-subfolder before saving figures
 # TODO: Plots of first differrences (e.g. oc of wind) need to have their x-axis adjusted (shift to mean of adjacent categories)

### 2021-02-09 projects/asparagus/postprocess/analysis_results.py

@@ -7,8 +7,8 @@ import pandas as pd
 from scipy.stats import pearsonr
 
 import config as cfg
-from src.tools.data_processing import medea_path
-from src.tools.visualize_data import plot_lines
+from src.utils.data_processing import medea_path
+from src.utils.visualize_data import plot_lines
 
 # %% settings
 APATH = medea_path('projects', 'asparagus')

### 2021-02-09 projects/asparagus/postprocess/monthly_dispatch.py

@@ -5,7 +5,7 @@ import matplotlib.dates as mdates
 import matplotlib.pyplot as plt
 import pandas as pd
 
-from src.tools.data_processing import medea_path
+from src.utils.data_processing import medea_path
 
 APATH = medea_path('projects', 'asparagus')
 RPATH = APATH / 'results'

### 2021-02-09 projects/asparagus/postprocess/plots_asparagus.py

@@ -6,7 +6,7 @@ import numpy as np
 import pandas as pd
 
 import config as cfg
-from src.tools.visualize_data import plot_lines, plot_sublines, plot_subn
+from src.utils.visualize_data import plot_lines, plot_sublines, plot_subn
 
 # %% ----- ----- ----- ----- settings ----- ----- ----- -----
 idx = pd.IndexSlice

### 2021-02-09 projects/asparagus/preprocess_asparagus.py

@@ -5,7 +5,7 @@ import pandas as pd
 from scipy.signal import savgol_filter
 
 import config as cfg
-from src.tools.data_processing import medea_path
+from src.utils.data_processing import medea_path
 
 # %% settings
 FNAME = medea_path('data', 'processed', 'medea_regional_timeseries.csv')

### 2021-02-09 projects/asparagus/results/analysis_results.py

@@ -5,7 +5,7 @@ from gams import *
 from scipy.stats import pearsonr
 
 import config as cfg
-from src.tools.gams_io import gdx2df
+from src.utils.gams_io import gdx2df
 
 idx = pd.IndexSlice
 

### 2021-02-09 projects/asparagus/results_asparagus.py

@@ -6,7 +6,7 @@ from gams import *
 
 import config as cfg
 from projects.asparagus.settings_asparagus import *
-from src.tools.gams_io import gdx2df
+from src.utils.gams_io import gdx2df
 
 # -------------------------------------------------------------------------------------------------------------------- #
 # %% prepare GAMS workspace

### 2021-02-09 projects/asparagus/run_asparagus.py

@@ -6,7 +6,7 @@ from gams import *
 
 import config as cfg
 from projects.asparagus.settings_asparagus import *
-from src.tools.gams_io import reset_symbol, gdx2df, df2gdx, run_medea_project
+from src.utils.gams_io import reset_symbol, gdx2df, df2gdx, run_medea_project
 
 idx = pd.IndexSlice
 # path to main medea model to use

### 2021-02-09 projects/opco/results_opco.py

@@ -1,47 +1,66 @@
-import numpy as np
+import os
+
 import pandas as pd
 from gams import *
 
 import config as cfg
-from src.templates.settings_template import *
-from src.tools.data_processing import medea_path
-from src.tools.gams_io import gdx2df
+from projects.opco.settings_opco import *
+from src.utils.gams_io import gdx2df
 
+idx = pd.IndexSlice
 # -------------------------------------------------------------------------------------------------------------------- #
-# prepare GAMS workspace
+# %% prepare GAMS workspace
 # -------------------------------------------------------------------------------------------------------------------- #
 ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
 
 # specify GAMS symbols to be read from output .gdx-file along with the dimensions of the corresponding output DataFrame
 # example: generate pandas DataFrame holding system cost and co2 emissions for each CO2 price scenario
 symbols_to_read = {
-    'cost_system': ([], ['z']),
-    'emission_co2': ([], ['z'])
+    'cost_zonal': ([], ['z']),
+    'emission_co2': ([], ['z']),
+    'add_r': (['n'], ['z'])
 }
 
-# iterate over all output .gdx-files
-df_results = pd.DataFrame()
-for price_co2 in range_co2price:
-    # generate name of .gdx-file to read
-    filename = f'medea_out_{output_naming}.gdx'.format(price_co2)
-    # create database of .gdx-data
-    db_output = ws.add_database_from_gdx(medea_path('projects', project_name, 'opt', filename))
+# %% iterate over all output .gdx-files
+df_result = pd.DataFrame(columns=cfg.zones)
+for campaign in dict_campaigns.keys():
+    # update campaign dictionary
+    dict_camp = dict_base.copy()
+    dict_camp.update(dict_campaigns[campaign])
 
-    # read symbols from database into DataFrames with symbol name as index and CO2-price/zone as multiindex-columns
-    df_i = pd.DataFrame(columns=cfg.zones)
-    for symbol, sets in symbols_to_read.items():
-        df = gdx2df(db_output, symbol, sets[0], sets[1])
-        # set index to symbol name
-        if df.index.any() == 'Value':
-            df.index = [symbol]
-        df_i = df_i.append(df)
-        # set column names to multiindex of CO2-price and market zone
-        df_i.columns = pd.MultiIndex.from_product([[f'PCO2_{price_co2}'], df_i.columns])
-    df_results = pd.concat([df_results, df_i], axis=1)
[…]

### 2021-02-09 projects/opco/run_opco.py

@@ -5,8 +5,8 @@ from gams import *
 
 import config as cfg
 from projects.opco.settings_opco import *
-from src.tools.data_processing import medea_path
-from src.tools.gams_io import reset_symbol, df2gdx, run_medea_project
+from src.utils.data_processing import medea_path
+from src.utils.gams_io import reset_symbol, df2gdx, run_medea_project
 
 idx = pd.IndexSlice
 # -------------------------------------------------------------------------------------------------------------------- #
@@ -42,6 +42,8 @@ os.chdir(medea_path('projects', PROJECT_NAME, 'opt'))
 # create empty scenario parameter in GAMS database so that it can be modified subsequently
 RE_SHARE = df2gdx(db_input, pd.DataFrame(data=[0]), 'RE_SHARE', 'par', 0, 'Minimum share of RE generation')
 WIND_ON_LIMIT = df2gdx(db_input, pd.DataFrame(data=[0]), 'WIND_ON_LIMIT', 'par', 0, 'max wind_on capacity addition')
+CO2_BUDGET = df2gdx(db_input, pd.DataFrame(data=[0]), 'CO2_BUDGET', 'par', 0, 'max amount of co2 emitted')
+CO2_SCENARIO = df2gdx(db_input, pd.DataFrame(data=[0]), 'CO2_SCENARIO', 'par', 0, 'CO2 price scenario')
 SWITCH_ANCILLARY = df2gdx(db_input, pd.DataFrame(data=[0]), 'SWITCH_ANCILLARY', 'par', 0,
                           'switch to activate ancillary service demand')
 
@@ -54,21 +56,26 @@ for campaign in dict_campaigns.keys():
 
     # (de)activate must-run
     reset_symbol(db_input, 'SWITCH_ANCILLARY', pd.DataFrame(data=dict_camp['must_run']))
-
     # set campaign-parameters
     for re_share in dict_camp['re_share']:
         reset_symbol(db_input, 'RE_SHARE', pd.DataFrame(data=[re_share]))
+
         for wind_limit in dict_camp['wind_on_cap']:
-            # modify upper limit for wind installation
             reset_symbol(db_input, 'WIND_ON_LIMIT', pd.DataFrame(data=[wind_limit]))
 
-            # generate identifier / name of scenario
-            identifier = f'{PROJECT_NAME}_{campaign}_{re_share}_{wind_limit}'
+            for price_co2 in dict_camp['carbon_price']:
[…]

### 2021-02-09 projects/opco/settings_opco.py

@@ -6,12 +6,20 @@ PROJECT_NAME = 'opco'
 dict_base = {
     're_share': range(100, 24, -25),
     'wind_on_cap': range(120, -1, -20),
+    'carbon_price': [25],
+    'carbon_limit': [float('inf')],
     'must_run': [1]
 }
 
 # campaigns
 dict_campaigns = {
-    'base': {
+    'carb_pricing': {
+        're_share': [0],
+        'carbon_price': range(900, 1, -400)
+    },
+    'carb_budget': {
+        're_share': [0],
+        'carbon_limit': [1000, 10000]
     }
 }
 

### 2021-02-09 src/data/compile_CommercialFlows.py

@@ -3,7 +3,7 @@ import os
 import pandas as pd
 
 import config as cfg
-from src.tools.data_processing import medea_path
+from src.utils.data_processing import medea_path
 
 directory = medea_path('data', 'raw', 'ScheduledCommercialExchanges')
 

### 2021-02-09 src/data/compile_HydroGeneration.py

@@ -3,7 +3,7 @@ import os
 import pandas as pd
 
 import config as cfg
-from src.tools.data_processing import medea_path
+from src.utils.data_processing import medea_path
 
 directory = medea_path('data', 'raw', 'AggregatedGenerationPerType')
 df_ror = pd.DataFrame()

### 2021-02-09 src/data/compile_ReservoirFilling.py

@@ -3,7 +3,7 @@ import os
 import pandas as pd
 
 import config as cfg
-from src.tools.data_processing import medea_path
+from src.utils.data_processing import medea_path
 
 directory = medea_path('data', 'raw', 'AggregateFillingRateWaterReservoirs')
 

### 2021-02-09 src/data/compile_capacities.py

@@ -2,7 +2,7 @@
 import numpy as np
 import pandas as pd
 
-from src.tools.data_processing import medea_path
+from src.utils.data_processing import medea_path
 
 # --------------------------------------------------------------------------- #
 # %% settings and initializing

### 2021-02-09 src/data/compile_timeseries.py

@@ -5,7 +5,7 @@ import numpy as np
 import pandas as pd
 
 import config as cfg
-from src.tools.data_processing import download_file, medea_path, download_energy_balance, process_energy_balance
+from src.utils.data_processing import download_file, medea_path, download_energy_balance, process_energy_balance
 
 idx = pd.IndexSlice
 
@@ -235,8 +235,8 @@ eca_at_rsvr = pd.read_excel(medea_path('data', 'raw', 'BStGes-JR1_Bilanz.xlsx'),
                             index_col=[0], nrows=37)
 eca_at_rsvr.drop('Einheit', inplace=True)
 eca_at_rsvr.replace(to_replace='- ', value=np.nan, inplace=True)
-eca_at_nflw = eca_at_hydro.loc[:, 'Speicher-\nkraftwerke'] / eta_hydro_storage - eca_at_rsvr.loc[:,
-                                                                                 'Verbrauch\nfür Pump-\nspeicher'] * eta_hydro_storage
+eca_at_nflw = eca_at_hydro.loc[:, 'Speicher-\nkraftwerke'] / eta_hydro_storage - \
+              eca_at_rsvr.loc[:, 'Verbrauch\nfür Pump-\nspeicher'] * eta_hydro_storage
 
 # 4) inflows in line with energy balance
 nbal_at_nflw = eca_at_nflw * scaling_factor_nflw

### 2021-02-09 src/data/get_entsoe.py

@@ -6,7 +6,7 @@ import pysftp
 import yaml
 
 from logging_config import setup_logging
-from src.tools.data_processing import medea_path
+from src.utils.data_processing import medea_path
 
 # TODO: check file size and download larger files from ftp
 # TODO: download zipped csv and unpack after download

### 2021-02-09 src/data/get_heatload.py

@@ -6,7 +6,7 @@ import pandas as pd
 
 import config as cfg
 from logging_config import setup_logging
-from src.tools.data_processing import medea_path, download_energy_balance, resample_index, heat_yr2day, heat_day2hr
+from src.utils.data_processing import medea_path, download_energy_balance, resample_index, heat_yr2day, heat_day2hr
 
 # ----------------------------------------------------------------------------
 # %% settings

### 2021-02-09 src/data/get_pricedata.py

@@ -4,7 +4,7 @@ import pandas as pd
 import yaml
 
 from logging_config import setup_logging
-from src.tools.data_processing import download_file, medea_path
+from src.utils.data_processing import download_file, medea_path
 
 setup_logging()
 

### 2021-02-09 src/data/get_temperatures.py

@@ -8,7 +8,7 @@ from scipy import interpolate
 
 import config as cfg
 from logging_config import setup_logging
-from src.tools.data_processing import download_era_temp, days_in_year, medea_path
+from src.utils.data_processing import download_era_temp, days_in_year, medea_path
 
 setup_logging()
 

### 2021-02-09 src/templates/results_template.py

@@ -4,8 +4,8 @@ from gams import *
 
 import config as cfg
 from src.templates.settings_template import *
-from src.tools.data_processing import medea_path
-from src.tools.gams_io import gdx2df
+from src.utils.data_processing import medea_path
+from src.utils.gams_io import gdx2df
 
 # -------------------------------------------------------------------------------------------------------------------- #
 # prepare GAMS workspace

### 2021-02-09 src/templates/run_template.py

@@ -6,8 +6,8 @@ from gams import *
 
 import config as cfg
 from src.templates.settings_template import *
-from src.tools.data_processing import medea_path
-from src.tools.gams_io import reset_symbol, gdx2df, df2gdx
+from src.utils.data_processing import medea_path
+from src.utils.gams_io import reset_symbol, gdx2df, df2gdx
 
 # -------------------------------------------------------------------------------------------------------------------- #
 # %% initialize GAMS, GAMS workspace and load model data
@@ -15,7 +15,6 @@ from src.tools.gams_io import reset_symbol, gdx2df, df2gdx
 # import base data from gdx
 ws = GamsWorkspace(system_directory=cfg.GMS_SYS_DIR)
 db_input = ws.add_database_from_gdx(medea_path('projects', project_name, 'opt', 'medea_main_data.gdx'))
-
 
 # %% read parameters that change in scenarios (and corresponding sets)
 # -------------------------------------------------------------------------------------------------------------------- #

### 2021-02-09 src/utils/create_project.py

@@ -3,9 +3,10 @@ import os
 import sys
 from shutil import copyfile
 
+from logging_config import setup_logging
+
 dirname = os.path.dirname
 sys.path.append(os.path.join(dirname(dirname(dirname(__file__)))))
-from logging_config import setup_logging
 
 
 def create_project(project_name, medea_dir):

### 2021-02-09 src/utils/data_processing.py

@@ -224,7 +224,6 @@ def download_energy_balance(country, years=range(2012, 2019)):
 
     if country == 'DE':
         # German energy balance as provided by AGEB
-        ht_enduse_de = pd.DataFrame()
         url_extension_bal = {12: 'xlsx', 13: 'xls', 14: 'xls', 15: 'xlsx', 16: 'xls', 17: 'xlsx', 18: 'xls'}
         url_extension_sat = {12: 'xlsx', 13: 'xls', 14: 'xls', 15: 'xlsx', 16: 'xls', 17: 'xlsx', 18: 'xlsx'}
         for yr in [x - 2000 for x in years]:
@@ -272,7 +271,7 @@ def scale_timeseries(time_series, target_sum):
     scales a time series such that its sum is equal to 'target_sum'.
     Typical use is to scale generation time series to match data reported in annual energy balances
     :param time_series: a pd.Series with high-frequency data
-    :param annual_sum: a float
+    :param target_sum: annual sum which the time series should be scaled to
     :return: time series scaled to target sum
     """
     scaled_generation = time_series * target_sum / time_series.sum()

### 2021-02-09 src/utils/gams_io.py

@@ -15,10 +15,9 @@ def timesort(df, index_sets='t', timeindex_name='t', timeindex_string='t'):
     :param df: A dataframe indexed by a float-string combination
     :param index_sets: column name(s) to be used as index
     :param timeindex_name: name of df's index. Defaults to 't'
-    :param timeindex_pattern: string part of the index to be ignored in sorting
+    :param timeindex_string:
     :return:
     """
-
     df.reset_index(inplace=True)
     df['tix'] = pd.to_numeric(df[timeindex_name].str.split(pat=timeindex_string).str.get(1))
     df.sort_values(by=['tix'], inplace=True)
@@ -130,6 +129,7 @@ def gdx2plot(db_gams, symbol, index_list, column_list, base_year, slicer=None, s
     :param column_list: set(s) to be used as columns
     :param base_year: year of model simulation
     :param slicer: slices the column-data
+    :param stacked:
     :return:
     """
     idx = pd.IndexSlice
@@ -207,8 +207,6 @@ def run_medea_project(project_name, scenario_id):
 def run_medea_test(test_data_name):
     """
     runs / solves a project of power system model medea with strict project directory conventions
-    :param test_name: string of medea-project name
-    :param scenario_id: string of project-scenario (typically one iteration)
     :return:
     """
     # generate file names

### 2021-02-09 src/utils/instantiate_gdx.py

@@ -5,8 +5,8 @@ import pandas as pd
 from gams import *
 
 import config as cfg
-from src.tools.gams_io import df2gdx
-from src.tools.prepare_data import dict_sets, dict_instantiate, static_data, plant_data, ts_data, invest_limits
+from src.utils.gams_io import df2gdx
+from src.utils.prepare_data import dict_sets, dict_instantiate, static_data, plant_data, ts_data, invest_limits
 
 # TODO: Add energy stored in hydro reservoirs - STORAGE_LEVEL
 

### 2021-02-09 src/utils/prepare_data.py

@@ -4,7 +4,7 @@ import numpy as np
 import pandas as pd
 
 import config as cfg
-from src.tools.data_processing import hours_in_year, medea_path
+from src.utils.data_processing import hours_in_year, medea_path
 
 # --------------------------------------------------------------------------- #
 # %% settings and initializing
